---
title: "RTSL Cardiovascular Health Program Impact Assessment: Projected Health Outcomes 2017–2050"
date: "`r Sys.Date()`"
author: "University of Washington"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE, echo=FALSE}
rm(list = ls())
knitr::opts_chunk$set(
  echo    = TRUE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 5,
  fig.width  = 9,
  fig.align  = "center"
)
```

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(DT)
library(scales)
library(openxlsx)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

```{r paths, message=FALSE, warning=FALSE, echo=FALSE}
wd       <- "C:/Users/wrgar/OneDrive - UW/02Work/ResolveToSaveLives/Assessment/UWRTSL-sodium-policy/"
wd_data  <- paste0(wd, "data/processed/")
wd_outp  <- paste0(wd, "output/")
wd_scen  <- paste0(wd, "scenarios/")
wd_raw   <- paste0(wd, "data/raw/")
```

```{r helpers, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Formatters --------------------------------------------------------------
fmt_ci <- function(est, lo, hi, digits = 1) {
  f <- function(x) formatC(round(x / 1e6, digits), format = "f",
                            digits = digits, big.mark = ",")
  paste0(f(est), " (", f(lo), "\u2013", f(hi), ")")
}
fmt_m <- function(x, digits = 1)
  formatC(round(x / 1e6, digits), format = "f", digits = digits, big.mark = ",")
fmt_k <- function(x, digits = 0)
  formatC(round(x / 1e3, digits), format = "f", digits = digits, big.mark = ",")
fmt_M <- function(x, digits = 1)
  formatC(round(x / 1e6, digits), format = "f", digits = digits, big.mark = ",")

# ---- Cause labels ------------------------------------------------------------
cause_map <- c(
  ihd     = "Ischemic heart disease",
  istroke = "Ischemic stroke",
  hstroke = "Intracerebral hemorrhage",
  hhd     = "Hypertensive heart disease"
)

# ---- Intervention display labels for RTSL Assessment -------------------------
# Model output scenarios (from 06_run_scenarios_assessment.R):
#   baseline  -> Counterfactual (BAU)
#   bp_only   -> HTN Control (Program)
#   tfa_only  -> TFA Elimination
#   bp_tfa    -> HTN Control + TFA Elimination
#
# Each scenario is run twice via rate_scenario:
#   "program"   -> Observed RTSL program control rates (2017-2024, maintained to 2050)
#   "bau_upper" -> Counterfactual (zero incremental program effect)
#
# The program-attributable impact = deaths under bau_upper - deaths under program

int_labels <- c(
  "Counterfactual"                = "Counterfactual",
  "HTN Control (Program)"        = "HTN Control (Program)",
  "TFA Elimination"              = "TFA Elimination",
  "HTN Control + TFA Elimination" = "HTN Control + TFA Elimination"
)

int_order <- c(
  "HTN Control (Program)",
  "TFA Elimination",
  "HTN Control + TFA Elimination"
)

# ---- Colour palettes ---------------------------------------------------------
int_cols <- c(
  "HTN Control (Program)"         = "#1f78b4",
  "TFA Elimination"               = "#e31a1c",
  "HTN Control + TFA Elimination" = "#6a3d9a",
  "Counterfactual"                = "#999999"
)

cause_cols <- c(
  ihd     = "#56B4E9",
  istroke = "#CC79A7",
  hstroke = "#F0E442",
  hhd     = "#E69F00"
)

cause_labels_plot <- c(
  ihd     = "Ischemic heart disease",
  istroke = "Ischemic stroke",
  hstroke = "Intracerebral hemorrhage",
  hhd     = "Hypertensive heart disease"
)

# ---- WHO standard population (ages 20-95+) for ASMR -------------------------
full_who_std <- data.table(
  age    = 0:100,
  weight = c(
    rep(0.013, 10), rep(0.014, 10), rep(0.013, 10), rep(0.012, 10),
    rep(0.011, 10), rep(0.009, 10), rep(0.007, 10), rep(0.005, 10),
    rep(0.003, 10), rep(0.001, 10), 0.001
  )
)
who_std <- full_who_std[age >= 20 & age <= 94]
who_std <- rbind(
  who_std,
  data.table(age = 95, weight = full_who_std[age >= 95, sum(weight)])
)
who_std[, weight := weight / sum(weight)]

calculate_asmr <- function(dt, std_pop = who_std,
                            deaths_col = "deaths_dx",
                            population_col = "Nx",
                            disagg_var = "intervention") {
  dt <- copy(dt)
  dt[, age := as.integer(gsub("\\+", "", age))]
  dt <- merge(dt, std_pop, by = "age", all.x = TRUE)
  dt[, rate          := get(deaths_col) / get(population_col)]
  dt[, weighted_rate := rate * weight]
  group_vars <- c("year", disagg_var)
  dt[, .(ASMR = sum(weighted_rate, na.rm = TRUE) * 1e5), by = group_vars]
}

# ---- Deaths-delayed calculator -----------------------------------------------
deaths_summary <- function(dt,
                           baseline_label = "Counterfactual",
                           by_cols = c("age_group", "sex", "year",
                                       "location", "cause", "age"),
                           value_col = "dead") {
  stopifnot(is.data.table(dt))
  n_interventions <- uniqueN(dt[intervention != baseline_label, intervention])
  baseline_cols   <- setdiff(by_cols, "intervention")
  baseline   <- dt[intervention == baseline_label,
                   .(baseline_deaths = sum(get(value_col), na.rm = TRUE)),
                   by = baseline_cols]
  nonbaseline <- dt[intervention != baseline_label,
                    .(deaths_projected = sum(get(value_col), na.rm = TRUE)),
                    by = by_cols]
  if (!("intervention" %in% by_cols))
    nonbaseline[, deaths_projected := deaths_projected / n_interventions]
  merged <- merge(nonbaseline, baseline, by = baseline_cols,
                  all.x = TRUE, allow.cartesian = TRUE)
  merged[, deaths_delayed     := baseline_deaths - deaths_projected]
  merged[, deaths_delayed_ave := deaths_delayed / uniqueN(dt$year)]
  merged[, deaths_delayed_pct := deaths_delayed / baseline_deaths]
  merged[]
}
```

```{r country-metadata, message=FALSE, warning=FALSE, echo=FALSE}
countries_dt <- fread(paste0(wd_data, "Country_groupings_extended.csv"))

# ---- Standardise country names if needed ------------------------------------
countries_dt[location == "United States of America", location := "United States"]
countries_dt[location == "Congo, Dem. Rep.", location := "Democratic Republic of the Congo"]
countries_dt[location == "Vietnam", location := "Viet Nam"]
countries_dt[location == "Bolivia (Plurinational State of)", location := "Bolivia"]

# ---- Add missing regions for specific ISO3s ---------------------------------
new_countries <- data.table(
  iso3     = c("LCA", "VCT"),
  location = c("Saint Lucia", "Saint Vincent and the Grenadines"),
  region   = c("Latin America and Caribbean", "Latin America and Caribbean")
)
countries_dt <- rbind(countries_dt, new_countries[!iso3 %in% countries_dt$iso3],
                      use.names = TRUE, fill = TRUE)

# Assign Argentina and Chile to Latin America and Caribbean (not high income)
countries_dt[location %in% c("Argentina", "Chile"),
             region := "Latin America and Caribbean"]
```

```{r load-model-output, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Load all country-level model output RDS files --------------------------
# Generated by 06_run_scenarios_assessment.R (program-impact mode)
out_model_path <- paste0(wd_outp, "out_model/")
files <- list.files(path = out_model_path, pattern = "\\.rds$", full.names = TRUE)

if (length(files) == 0)
  stop(paste0("No .rds files found in: ", out_model_path,
              "\nRun 06_run_scenarios_assessment.R first."))

data.out <- rbindlist(
  lapply(files, function(f) { dt <- readRDS(f); setDT(dt); dt }),
  use.names = TRUE, fill = TRUE
)
data.out <- data.out[!is.na(location)]

# Drop effect-ratio tracking columns if present
for (col in c("eff_ir", "eff_cf"))
  if (col %in% names(data.out)) data.out[, (col) := NULL]

# ---- Standardise scenario -> intervention labels ----------------------------
# The output has a 'scenario' column (baseline, bp_only, tfa_only, bp_tfa)
# and a 'rate_scenario' column (program, bau_upper)
if ("scenario" %in% names(data.out) && !"intervention" %in% names(data.out)) {
  setnames(data.out, "scenario", "intervention")
} else if ("scenario" %in% names(data.out)) {
  data.out[, intervention := scenario]
  data.out[, scenario := NULL]
}

# Ensure rate_scenario column exists
if (!"rate_scenario" %in% names(data.out))
  data.out[, rate_scenario := "program"]

# ---- Remap intervention labels -----------------------------------------------
data.out[intervention %in% c("baseline", "BAU", "b.a.u"),
         intervention := "Counterfactual"]
data.out[intervention %in% c("bp_only", "antihypertensive", "BP",
                              "Antihypertensive therapy"),
         intervention := "HTN Control (Program)"]
data.out[intervention %in% c("tfa_only", "tfa", "TFA"),
         intervention := "TFA Elimination"]
data.out[intervention %in% c("bp_tfa"),
         intervention := "HTN Control + TFA Elimination"]

# Keep only recognised scenarios
data.out <- data.out[intervention %in% c("Counterfactual", int_order)]

# HTN control countries
dt_htn_control_scenarios <- fread(paste0(wd_data, "htn_control_scenarios_by_loc_year.csv"))

htn_countries_include <- unique(dt_htn_control_scenarios[, location])

# Remove countries not included in RTSL TFA calculations (if any slipped in)
dt_tfa_scenarios <- readRDS(file = paste0(wd_data,"tfa_policy_scenarios_assessment.rds"))

tfa_countries_include <- unique(dt_tfa_scenarios[tfa_rtsl_include == "Yes", location])

countries_include <- unique(c(htn_countries_include, tfa_countries_include))

## Filter model output to included countries only
data.out <- data.out[location %in% countries_include]
# ---- Separate program and counterfactual runs --------------------------------
# Program-attributable impact:
#   deaths_averted = deaths(bau_upper) - deaths(program)
# For the report, we use:
#   - rate_scenario == "bau_upper" as the Counterfactual
#   - rate_scenario == "program"   as the Intervention
data_program <- data.out[rate_scenario == "program"]
data_bau     <- data.out[rate_scenario == "bau_upper"]

# ---- Age groups --------------------------------------------------------------
data_program[, age_group := cut(age, breaks = c(0, 70, 80, 90, Inf),
                                labels = c("<70", "70-79", "80-89", "90+"),
                                right = FALSE, include.lowest = TRUE)]
data_bau[, age_group := cut(age, breaks = c(0, 70, 80, 90, Inf),
                            labels = c("<70", "70-79", "80-89", "90+"),
                            right = FALSE, include.lowest = TRUE)]

# ---- Merge region metadata ---------------------------------------------------
if (nrow(countries_dt) > 0) {
  region_lookup <- countries_dt[, .(location, region)]
  region_lookup <- unique(region_lookup)
  data_program <- merge(data_program, region_lookup, by = "location", all.x = TRUE)
  data_bau     <- merge(data_bau,     region_lookup, by = "location", all.x = TRUE)
}
if (!"region" %in% names(data_program)) data_program[, region := NA_character_]
if (!"region" %in% names(data_bau))     data_bau[, region := NA_character_]

# ---- Key summary counts ------------------------------------------------------
n_countries <- uniqueN(data_program$location)
year_min    <- min(data_program$year, na.rm = TRUE)
year_max    <- max(data_program$year, na.rm = TRUE)
study_years <- paste0(year_min, "\u2013", year_max)
```

---

# Background

This report presents modelled projections of cardiovascular disease (CVD) mortality
attributable to **Resolve to Save Lives (RTSL) cardiovascular health programs**
operating in `r n_countries` countries from 2017 to 2024. The central research question is:

> *What is the expected health impact --- in terms of deaths averted ---
> of RTSL cardiovascular health programs from 2017--2024 in targeted populations,
> projected to 2050?*

Two interventions are assessed:

1. **Improved hypertension control** --- observed program-level blood pressure control
   rates (2017--2024) are maintained at 2024 levels through 2050, compared to a
   counterfactual scenario with no incremental program effect.
2. **Industrial trans-fat elimination (TFA)** --- policy-driven elimination of
   industrial trans-fatty acids in countries with enacted or effective TFA policies
   by end of 2024.

Four causes of CVD death are modelled jointly: ischemic heart disease (IHD), ischemic
stroke, intracerebral hemorrhage (hemorrhagic stroke), and hypertensive heart disease (HHD).
A discrete-time multi-state transition model (Well $\rightarrow$ Sick $\rightarrow$ Dead) is calibrated to GBD 2023
data and projected forward to 2050 using UNWPP 2024 population estimates.

---

# Study Design

## Countries Included

The `r n_countries` countries with RTSL program data are:

```{r country-list, message=FALSE, warning=FALSE, echo=FALSE}
dt_countries <- unique(data_program[, .(location, region)])
setorder(dt_countries, region, location)
countries_by_region <- dt_countries[,
  .(Countries = paste0(sort(location), collapse = ", ")),
  by = .(Region = region)
]
setorder(countries_by_region, Region)

kable(countries_by_region,
      caption = "Countries included in the assessment, by region.",
      align = "ll") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE)
```

## Intervention Definitions

| Intervention | Description |
|---|---|
| **HTN Control (Program)** | Antihypertensive therapy scale-up using observed RTSL program control rates. Baseline BP distributions are shifted based on Ettehad et al. cumulative effect sizes for each 10 mmHg BP bin. Program enrollment data (2017--2024) define the treated population; control rates are maintained at 2024 levels through 2050. |
| **TFA Elimination** | Complete elimination of industrial trans-fatty acids (target: 0% energy from TFA). Applied to countries with enacted/effective TFA best-practice policies by end of 2024, with a 2-year lag from policy passage to population-level effect. |
| **HTN Control + TFA Elimination** | Both interventions applied simultaneously with multiplicative effect combination. |

## Scenario Structure

The model is run in **program-rate mode**: each country is simulated twice:

1. **Program run** (`rate_scenario = "program"`): observed RTSL control rates applied
2. **Counterfactual run** (`rate_scenario = "bau_upper"`): zero incremental program effect

The **program-attributable impact** is the difference: deaths under counterfactual minus deaths under program.

## Projection Horizon

- **Program enrollment data**: 2017--2024 (from HTN global tracker)
- **Model projection**: `r study_years`
- **Impact assessment window**: 2017--2050
- **Age range**: 20--95 years

---

# Hypertension Control Program Data

```{r htn-program-data, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
htn_scenarios <- tryCatch(
  fread(paste0(wd_data, "htn_control_scenarios_by_loc_year.csv")),
  error = function(e) NULL
)

if (!is.null(htn_scenarios)) {
  # Plot observed + projected program control rates
  dt_plot <- htn_scenarios[year <= 2050, .(location, year,
                                            control_rate = control_rate_program,
                                            observed = control_rate_observed)]

  g <- ggplot(dt_plot, aes(x = year, y = control_rate * 100, group = location)) +
    geom_line(color = "grey60", linewidth = 0.6, alpha = 0.7) +
    # Highlight observed data points
    geom_point(data = dt_plot[observed == 1],
               aes(x = year, y = control_rate * 100),
               color = "#1f78b4", size = 0.8, alpha = 0.5) +
    scale_x_continuous(breaks = seq(2017, 2050, 5),
                       limits = c(2017, 2051), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 60), expand = c(0, 0)) +
    labs(
      title    = "RTSL Hypertension Control Rates by Country",
      subtitle = "Observed (dots) and projected (maintained at 2024 level) program control rates",
      x = "Year", y = "Hypertension Control Rate (%)"
    ) +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
  print(g)
  ggsave(paste0(wd_outp, "assessment_policy_htn_program.png"), plot = g,
         width = 11, height = 5, dpi = 300)
} else {
  message("htn_control_scenarios_by_loc_year.csv not found - skipping HTN program plot.")
}
```

```{r htn-enrollment, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
# 
# # Program enrollment trajectory (TOTAL only)
# if (!is.null(htn_scenarios)) {
# 
#   dt_enroll <- htn_scenarios[
#     !is.na(enrolled) & enrolled > 0,
#     .(location, year, enrolled)
#   ]
# 
#   dt_enroll_total <- dt_enroll[
#     , .(total_enrolled = sum(enrolled, na.rm = TRUE)),
#     by = year
#   ]
# 
#   g <- ggplot(dt_enroll_total, aes(x = year, y = total_enrolled / 1e6)) +
#     geom_line(color = "#1f78b4", linewidth = 1.3) +
#     scale_x_continuous(breaks = seq(2015, 2050, 5)) +
#     scale_y_continuous(labels = comma) +
#     labs(
#       title    = "RTSL Hypertension Program Enrollment Over Time",
#       subtitle = "Total enrollment across all countries (millions)",
#       x = "Year",
#       y = "Enrolled (millions)"
#     ) +
#     theme_minimal(base_size = 13) +
#     theme(
#       plot.title = element_text(face = "bold"),
#       panel.grid.minor = element_blank()
#     )
# 
#   print(g)
# 
#   ggsave(
#     paste0(wd_outp, "assessment_enrollment_trend.png"),
#     plot = g,
#     width = 11, height = 5, dpi = 300
#   )
# }

if (!is.null(htn_scenarios)) {

  dt_enroll <- htn_scenarios[
    !is.na(enrolled) & enrolled > 0,
    .(
      location,
      year,
      enrolled,
      control_rate_imputed
    )
  ][
    ,
    controlled_pop := enrolled * control_rate_imputed
  ]

  dt_enroll_total <- dt_enroll[
    ,
    .(
      total_enrolled   = sum(enrolled, na.rm = TRUE),
      total_controlled = sum(controlled_pop, na.rm = TRUE)
    ),
    by = year
  ]

  g <- ggplot(dt_enroll_total, aes(x = year)) +
    geom_line(
      aes(y = total_enrolled / 1e6, color = "Enrolled population"),
      linewidth = 1.3
    ) +
    geom_line(
      aes(y = total_controlled / 1e6, color = "Controlled population"),
      linewidth = 1.3,
      linetype = "solid"
    ) +
    scale_x_continuous(breaks = seq(2015, 2050, 5)) +
    scale_y_continuous(labels = comma) +
    scale_color_manual(
      values = c(
        "Enrolled population"   = "#1f78b4",
        "Controlled population" = "#33a02c"
      ),
      name = NULL
    ) +
    labs(
      title    = "RTSL Hypertension Program Enrollment and Controlled Population Over Time",
      subtitle = "Total enrolled and controlled population across all countries (millions)",
      x = "Year",
      y = "Population (millions)"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      legend.position = "top"
    )

  print(g)

  ggsave(
    paste0(wd_outp, "assessment_enrollment_trend.png"),
    plot = g,
    width = 11, height = 5, dpi = 300
  )
}
```

---

# Disease Trends (Baseline)

## Incidence and Prevalence Over Time

```{r disease-trends, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
# Use counterfactual (bau_upper) as the baseline trend
dt_trend <- data_bau[intervention == "Counterfactual" & year >= 2017,
                     .(new_cases = sum(newcases, na.rm = TRUE),
                       sick      = sum(sick,     na.rm = TRUE),
                       pop       = sum(pop,      na.rm = TRUE)),
                     by = year]

dt_melt <- melt(dt_trend, id.vars = "year",
                measure.vars  = c("new_cases", "sick"),
                variable.name = "metric", value.name = "count")
dt_melt[, metric_lbl := fifelse(metric == "new_cases", "Incident cases", "Prevalent cases")]

g <- ggplot(dt_melt, aes(x = year, y = count / 1e6,
                          color = metric_lbl, linetype = year > 2024)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Incident cases"  = "steelblue",
                                "Prevalent cases" = "tomato"),
                     name = NULL) +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Incident and Prevalent CVD Cases \u2014 Counterfactual Scenario",
    subtitle = paste0("All four causes combined across ", n_countries,
                      " RTSL countries; solid = historical, dashed = projected"),
    x = "Year", y = "Persons (millions)",
    caption = "Source: Model output; ages 20\u201395+"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_disease_trend.png"), plot = g,
       width = 11, height = 5, dpi = 300)
```

## Age-Standardised Mortality Rate by Cause

```{r asmr-cause, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=6}
dt_asmr_in <- data_bau[intervention == "Counterfactual" & year >= 2017,
                        .(deaths_dx = sum(dead, na.rm = TRUE),
                          well      = sum(well, na.rm = TRUE),
                          sick      = sum(sick, na.rm = TRUE)),
                        by = .(cause, age, year)]
dt_asmr_in[, Nx := well + sick]

asmr_cause <- calculate_asmr(dt_asmr_in, deaths_col = "deaths_dx",
                              population_col = "Nx", disagg_var = "cause")
asmr_base  <- asmr_cause[year == 2017, .(cause, ASMR_base = ASMR)]
asmr_cause <- merge(asmr_cause, asmr_base, by = "cause")
asmr_cause[, index := ASMR / ASMR_base * 100]
asmr_cause[, cause_lbl := fcase(
  cause == "ihd",     "Ischemic heart disease",
  cause == "istroke", "Ischemic stroke",
  cause == "hstroke", "Intracerebral hemorrhage",
  cause == "hhd",     "Hypertensive heart disease",
  default = cause
)]

g <- ggplot(asmr_cause, aes(x = year, y = index,
                             color = cause_lbl, linetype = year > 2024)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 100, linetype = "dotted", color = "grey50") +
  scale_color_manual(values = setNames(cause_cols, cause_labels_plot),
                     name = "Cause of death") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(
    title    = "Age-Standardised CVD Mortality Rate Index \u2014 Counterfactual",
    subtitle = paste0("Index: 2017 = 100; ", n_countries, " RTSL countries"),
    x = "Year", y = "ASMR Index (2017 = 100)",
    caption = "WHO World Standard Population (2000\u20132025), rescaled to ages 20\u201395+"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_asmr_cause.png"), plot = g,
       width = 11, height = 5, dpi = 300)
```

---

# Program-Attributable Deaths Averted (2017--2050)

```{r compute-deaths-averted, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Core: program-attributable deaths averted --------------------------------
# deaths_averted = deaths(Counterfactual, bau_upper) - deaths(intervention, program)
#
# The counterfactual is ALWAYS "Counterfactual" in rate_scenario == "bau_upper".
# Merging on intervention was wrong for TFA: bau_upper has no "TFA Elimination" rows.

grp_keys <- c("age_group", "sex", "year", "location", "cause", "age")

# Counterfactual deaths: Counterfactual intervention, bau_upper rate scenario
dt_cf <- data_bau[intervention == "Counterfactual" & year >= 2017,
                  .(cf_deaths = sum(dead, na.rm = TRUE)),
                  by = grp_keys]

# Program deaths: all non-Counterfactual interventions, program rate scenario
dt_pgm_agg <- data_program[intervention %in% int_order & year >= 2017,
                            .(pgm_deaths = sum(dead, na.rm = TRUE)),
                            by = c(grp_keys, "intervention")]

# Merge: each intervention row gets the single counterfactual baseline
dt_merged <- merge(dt_pgm_agg, dt_cf, by = grp_keys, all.x = TRUE)
dt_merged[is.na(cf_deaths),  cf_deaths  := 0]
dt_merged[is.na(pgm_deaths), pgm_deaths := 0]
dt_merged[, deaths_averted := cf_deaths - pgm_deaths]

# ---- Cumulative totals per intervention --------------------------------------
dt_cumul <- dt_merged[, .(deaths_averted = sum(deaths_averted, na.rm = TRUE)),
                       by = intervention]

# ---- Annual time-series ------------------------------------------------------
dt_annual <- dt_merged[, .(deaths_averted = sum(deaths_averted, na.rm = TRUE)),
                        by = .(intervention, year)]
dt_annual[, intervention := factor(intervention, levels = int_order)]
setorder(dt_annual, intervention, year)
dt_annual[, cum_deaths_averted := cumsum(deaths_averted), by = intervention]

# ---- Per-scenario totals for inline text ------------------------------------
total_combined <- dt_cumul[intervention == "HTN Control + TFA Elimination", deaths_averted]
total_bp       <- dt_cumul[intervention == "HTN Control (Program)",         deaths_averted]
total_tfa      <- dt_cumul[intervention == "TFA Elimination",               deaths_averted]

n_years <- uniqueN(dt_merged$year)
```

```{r per1m-scalars, message=FALSE, warning=FALSE, echo=FALSE}
# Per-1M scalars for inline narrative (average per-country cumulative per 1M)
lives_per_1M_htn      <- round(total_bp       / n_countries)
lives_per_1M_tfa      <- round(total_tfa      / n_countries)
lives_per_1M_combined <- round(total_combined / n_countries)
```

Between 2017 and 2050, the combined RTSL hypertension control and trans-fat elimination
programs are projected to avert approximately **`r fmt_k(total_combined)` thousand deaths**
(`r fmt_k(lives_per_1M_combined)` thousand per million treated)
across `r n_countries` countries.

- **HTN Control (Program)** alone would avert `r fmt_k(total_bp)` thousand deaths
  (`r fmt_k(lives_per_1M_htn)` thousand per million treated).
- **TFA Elimination** alone would avert `r fmt_k(total_tfa)` thousand deaths
  (`r fmt_k(lives_per_1M_tfa)` thousand per million population).

## Table 1 --- Cumulative Deaths Averted by Intervention (2017--2050)

```{r table1, message=FALSE, warning=FALSE, echo=FALSE}
tab1 <- copy(dt_cumul)
tab1[, intervention := factor(intervention, levels = int_order)]
setorder(tab1, intervention)

tab1[, deaths_K      := round(deaths_averted / 1e3, 1)]
tab1[, deaths_annual := round(deaths_averted / n_years / 1e3, 1)]

tab1[, description := fcase(
  intervention == "HTN Control (Program)",
    "Observed RTSL program HTN control rates (2017\u20132024) maintained to 2050",
  intervention == "TFA Elimination",
    "Industrial TFA elimination via best-practice policies enacted by 2024",
  intervention == "HTN Control + TFA Elimination",
    "Both interventions combined (multiplicative effects)",
  default = ""
)]

tab1_fmt <- tab1[, .(
  Intervention       = intervention,
  Description        = description,
  Deaths_averted_K   = comma(deaths_K,      accuracy = 0.1),
  Avg_per_year       = comma(deaths_annual,  accuracy = 0.1)
)]
setnames(tab1_fmt,
         old = c("Description", "Deaths_averted_K", "Avg_per_year"),
         new = c("Intervention Description",
                 "Deaths averted (thousands, 2017\u20132050)",
                 "Average per year (thousands)"))

kable(tab1_fmt,
      caption = "Table 1. Cumulative program-attributable deaths averted (2017\u20132050) by intervention.",
      align   = "llrr") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  column_spec(1, width = "6cm") |>
  column_spec(2, width = "10cm") |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "Deaths averted = deaths under counterfactual (no program effect) minus deaths ",
    "under program scenario, summed over 2017\u20132050. ",
    "Causes: IHD, ischemic stroke, intracerebral hemorrhage, hypertensive heart disease."
  ))

dt_excel_tab1 <- copy(tab1_fmt)
```

## Figure 1 --- Deaths Averted by Intervention (Bar Chart)

```{r figure1, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
dt_fig1 <- tab1[, .(
  intervention   = factor(intervention, levels = rev(int_order)),
  deaths_averted = deaths_averted
)]

g <- ggplot(dt_fig1, aes(x = deaths_averted / 1e3, y = intervention,
                          fill = as.character(intervention))) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = int_cols, guide = "none") +
  scale_x_continuous(expand = c(0, 0), labels = comma) +
  labs(
    title   = "Program-Attributable Deaths Averted by Intervention, 2017\u20132050",
    x       = "Deaths averted (thousands)",
    y       = NULL,
    caption = paste0("Deaths averted vs. counterfactual (no program effect); ",
                     "all CVD causes combined; ", n_countries, " RTSL countries.")
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank())
print(g)
ggsave(paste0(wd_outp, "assessment_fig1_deaths_intervention.png"), plot = g,
       width = 9, height = 5, dpi = 300)
```

## Table 2 --- Cumulative Deaths Averted by Intervention and Cause

```{r table2, message=FALSE, warning=FALSE, echo=FALSE}
dt_tab2 <- dt_merged[
  intervention %in% int_order,
  .(deaths_averted = sum(deaths_averted, na.rm = TRUE)),
  by = .(intervention, cause)
]

# TFA affects only IHD (via CHD risk pathway)
dt_tab2 <- dt_tab2[!(cause %in% c("istroke", "hstroke", "hhd") &
                        intervention == "TFA Elimination")]

dt_tab2[, cause_lbl := fcase(
  cause == "ihd",     "Ischemic heart disease",
  cause == "istroke", "Ischemic stroke",
  cause == "hstroke", "Intracerebral hemorrhage",
  cause == "hhd",     "Hypertensive heart disease",
  default = cause
)]

dt_tab2_wide <- dcast(dt_tab2, intervention ~ cause_lbl,
                      value.var = "deaths_averted", fill = NA)
dt_tab2_wide[, intervention := factor(intervention, levels = int_order)]
setorder(dt_tab2_wide, intervention)

for (col in names(dt_tab2_wide)[names(dt_tab2_wide) != "intervention"]) {
  dt_tab2_wide[, (col) := fifelse(
    is.na(get(col)), "\u2013",
    comma(round(get(col) / 1e3, 1), accuracy = 0.1)
  )]
}

kable(dt_tab2_wide,
      caption = "Table 2. Cumulative deaths averted (thousands, 2017\u20132050) by intervention and cause of death.",
      align   = c("l", rep("r", ncol(dt_tab2_wide) - 1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  add_header_above(setNames(c(1L, ncol(dt_tab2_wide) - 1L),
                            c(" ", "Deaths averted, thousands (2017\u20132050)"))) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "\u2013 = intervention does not affect this cause. ",
    "TFA elimination affects IHD only. ",
    "Antihypertensive interventions affect all four CVD causes."
  ))

dt_excel_tab2 <- copy(dt_tab2_wide)
```

## Figure 2 --- Deaths Averted by Cause (Stacked Bar)

```{r figure2, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
dt_fig2 <- dt_merged[
  intervention %in% int_order & cause %in% names(cause_map),
  .(deaths_averted = sum(deaths_averted, na.rm = TRUE)),
  by = .(intervention, cause)
]
dt_fig2[, intervention := factor(intervention, levels = int_order)]
dt_fig2[, cause_lbl := cause_labels_plot[cause]]
dt_fig2[, cause_lbl := factor(cause_lbl, levels = unname(cause_labels_plot))]

g <- ggplot(dt_fig2, aes(x = deaths_averted / 1e3, y = intervention, fill = cause_lbl)) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = setNames(cause_cols, cause_labels_plot),
                    name = "Cause of death") +
  scale_x_continuous(expand = c(0, 0), labels = comma) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(
    title = "Deaths Averted by Cause of Death and Intervention, 2017\u20132050",
    x     = "Deaths averted (thousands)",
    y     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position    = "bottom",
        panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title         = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_fig2_deaths_cause.png"), plot = g,
       width = 9, height = 5, dpi = 300)
```

## Figure 3 --- Time Series: Annual Deaths Averted by Intervention

```{r figure3-annual, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
g <- ggplot(dt_annual[!intervention %in% c("HTN Control + TFA Elimination")],
            aes(x = year, y = deaths_averted / 1e3,
                color = intervention, fill = intervention)) +
  geom_area(position = "stack", alpha = 0.7, color = "grey20", linewidth = 0.15) +
  scale_color_manual(values = int_cols, name = NULL) +
  scale_fill_manual(values  = int_cols, name = NULL) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Annual Program-Attributable Deaths Averted by Intervention, 2017\u20132050",
    x     = "Year",
    y     = "Deaths averted (thousands)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_fig3_annual_averted.png"), plot = g,
       width = 11, height = 5, dpi = 300)
```

## Figure 4 --- Cumulative Deaths Averted Over Time

```{r figure4-cumulative, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
g <- ggplot(dt_annual,
            aes(x = year, y = cum_deaths_averted / 1e3, color = intervention)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = int_cols, name = NULL) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Cumulative Deaths Averted by Intervention, 2017\u20132050",
    x     = "Year",
    y     = "Cumulative deaths averted (thousands)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_fig4_cum_averted.png"), plot = g,
       width = 11, height = 5, dpi = 300)
```

---

# Outcomes per 1 Million

```{r per1m-compute, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Reconstruct actual population denominators from raw inputs ----
# Replicating the logic from 05_build_baseline.R (not from model output)

# PROGRAM POPULATION (HTN denominator): actual enrolled per location-year
dt_program_pop <- tryCatch({
  # Step a: Load htn_control_scenarios, keep location, year, enrolled
  htn_csv <- fread(paste0(wd_data, "htn_control_scenarios_by_loc_year.csv"))
  htn_csv <- htn_csv[, .(location, year, enrolled)]

  # Step b: Compute age/sex population proportions from model output
  #         (data_bau, Counterfactual, cause == "ihd", year >= 2017)
  dt_pop_dist <- data_bau[
    intervention == "Counterfactual" & cause == "ihd" & year >= 2017,
    .(population = sum(pop, na.rm = TRUE)),
    by = .(location, year, age, sex)
  ]
  dt_pop_dist[, population_proportion := population / sum(population),
              by = .(location, year)]

  # Step c: Merge enrolled x population_proportion to get Nx_program
  dt_enr <- merge(htn_csv, dt_pop_dist,
                  by = c("location", "year"), all.y = TRUE)
  dt_enr[is.na(enrolled), enrolled := 0]
  dt_enr[, Nx_program := enrolled * population_proportion]

  # Step d: Sum by location-year to get total program_population
  dt_enr[, .(program_population = sum(Nx_program, na.rm = TRUE)),
         by = .(location, year)]
}, error = function(e) {
  message("htn_control_scenarios_by_loc_year.csv not found ",
          "\u2014 program_population unavailable.")
  NULL
})

# TOTAL COUNTRY POPULATION age 20+ per location-year
# (from UNWPP, same source as 05_build_baseline.R)
dt_total_pop <- tryCatch({
  # save for reporting estimates in the report
  pop_raw <- readRDS(file = paste0(wd_data,"dt_pop_estimates_output_report.rds"))
  
  # pop_raw <- fread(paste0(wd_data, "PopulationsAge20_2050.csv"))
  if ("year_id" %in% names(pop_raw)) setnames(pop_raw, "year_id", "year")
  pop_raw[, .(total_population = sum(population, na.rm = TRUE)),
          by = .(location, year)]
}, error = function(e) {
  message("PopulationsAge20_2050.csv not found ",
          "\u2014 total_population unavailable.")
  NULL
})

# ---- Per-1M summary by intervention x cause ----
# Model outputs are already per 1M treated (upstream normalisation).
# HTN scenarios: per 1M treated patients
# TFA scenarios: per 1M population (approximately equal per-person rate)

dt_per1m_int_cause <- dt_merged[
  year >= 2017,
  .(deaths_per_1M        = round(sum(pgm_deaths,     na.rm = TRUE), 1),
    deaths_averted_per_1M = round(sum(deaths_averted, na.rm = TRUE), 1)),
  by = .(intervention, cause)
]
dt_per1m_int_cause[, cause_lbl := cause_map[cause]]
dt_per1m_int_cause[, denominator := fifelse(
  intervention == "TFA Elimination",
  "per 1M population", "per 1M treated"
)]

# Totals by intervention (for bar chart)
dt_per1m_bar <- dt_merged[
  year >= 2017,
  .(deaths_averted_per_1M = sum(deaths_averted, na.rm = TRUE)),
  by = intervention
]
dt_per1m_bar[, intervention := factor(intervention, levels = int_order)]

# ---- Rescale to actual absolute deaths averted ----
dt_rescale_detail <- dt_merged[
  year >= 2017,
  .(per_1M_averted = sum(deaths_averted, na.rm = TRUE)),
  by = .(intervention, location, year)
]

if (!is.null(dt_program_pop)) {
  dt_rescale_detail <- merge(dt_rescale_detail, dt_program_pop,
                             by = c("location", "year"), all.x = TRUE)
} else {
  dt_rescale_detail[, program_population := NA_real_]
}
if (!is.null(dt_total_pop)) {
  dt_rescale_detail <- merge(dt_rescale_detail, dt_total_pop,
                             by = c("location", "year"), all.x = TRUE)
} else {
  dt_rescale_detail[, total_population := NA_real_]
}

# HTN: multiply per-1M by (actual program_population / 1e6)
dt_rescale_detail[
  intervention %in% c("HTN Control (Program)",
                       "HTN Control + TFA Elimination"),
  abs_averted := per_1M_averted * (program_population / 1e6)
]
# TFA: multiply per-1M by (total_population / 1e6)
dt_rescale_detail[
  intervention == "TFA Elimination",
  abs_averted := per_1M_averted * (total_population / 1e6)
]

# Aggregate for comparison table
dt_rescale_agg <- dt_rescale_detail[,
  .(rescaled_abs = sum(abs_averted, na.rm = TRUE),
    model_per_1M = sum(per_1M_averted, na.rm = TRUE)),
  by = intervention
]
dt_rescale_agg[, model_per_1M_K  := round(model_per_1M  / 1e3, 1)]
dt_rescale_agg[, rescaled_abs_K  := round(rescaled_abs  / 1e3, 1)]

# Flag large discrepancies (> 10%) between rescaled and model per-1M
# (expected to differ because of population scaling; flag only unexpected mismatches)
dt_rescale_agg[, ratio := rescaled_abs / pmax(model_per_1M, 1)]

# ---- Additional per-1M scalars for narrative ----
# (lives_per_1M_htn/tfa/combined already computed in per1m-scalars chunk)

# 5-year window (2020-2024) average per country
lives_per_1M_htn_5yr <- round(
  dt_merged[intervention == "HTN Control (Program)" &
              year %in% 2020:2024,
            sum(deaths_averted, na.rm = TRUE)] / n_countries
)

# Rescaled absolute totals
total_bp_rescaled <- sum(
  dt_rescale_detail[intervention == "HTN Control (Program)",
                    abs_averted], na.rm = TRUE)
total_tfa_rescaled <- sum(
  dt_rescale_detail[intervention == "TFA Elimination",
                    abs_averted], na.rm = TRUE)
total_combined_rescaled <- sum(
  dt_rescale_detail[intervention == "HTN Control + TFA Elimination",
                    abs_averted], na.rm = TRUE)
```

## Per 1M Treated (Normalised Model Output)

```{r per1m-table, message=FALSE, warning=FALSE, echo=FALSE}
dt_per1m_display <- dt_per1m_int_cause[, .(
  Intervention              = intervention,
  Cause                     = cause_lbl,
  `Deaths per 1M`           = comma(deaths_per_1M,        accuracy = 0.1),
  `Deaths averted per 1M`   = comma(deaths_averted_per_1M, accuracy = 0.1),
  Denominator               = denominator
)]
setorder(dt_per1m_display, Intervention, Cause)

kable(dt_per1m_display,
      caption = paste0(
        "Deaths and deaths averted per 1 million, by intervention and ",
        "cause (", study_years, ")."),
      align = c("l", "l", "r", "r", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "HTN scenarios: per 1M patients treated (normalised model output). ",
    "TFA scenarios: per 1M population age 20+. ",
    "Figures are cumulative sums across all ", n_countries,
    " program countries."))
```

```{r per1m-barchart, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
g <- ggplot(dt_per1m_bar,
            aes(x = deaths_averted_per_1M / 1e3,
                y = intervention,
                fill = as.character(intervention))) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = int_cols, guide = "none") +
  scale_x_continuous(expand = c(0, 0), labels = comma) +
  labs(
    title = paste0("Deaths Averted per 1 Million by Intervention, ",
                   study_years),
    x     = "Deaths averted per 1M (thousands, cumulative across countries)",
    y     = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title         = element_text(face = "bold"),
        panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank())
print(g)
```

## Rescaled to Actual Population

```{r per1m-rescaled, message=FALSE, warning=FALSE, echo=FALSE}
dt_rescale_display <- dt_rescale_agg[, .(
  Intervention                     = intervention,
  `Model per-1M (thousands)`       = comma(model_per_1M_K,  accuracy = 0.1),
  `Rescaled absolute (thousands)`  = comma(rescaled_abs_K,   accuracy = 0.1)
)]
dt_rescale_display[, Intervention := factor(Intervention, levels = int_order)]
setorder(dt_rescale_display, Intervention)

kable(dt_rescale_display,
      caption = paste0(
        "Rescaled absolute deaths averted vs. normalised per-1M model ",
        "output (", study_years, ")."),
      align = c("l", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "HTN rescaled by actual program enrollment; ",
    "TFA rescaled by total population age 20+. ",
    "Per-1M figures sum normalised model output across all countries."))
```

## Narrative Summary

The RTSL HTN program is projected to save approximately
`r fmt_k(lives_per_1M_htn)` thousand lives per million patients treated
(cumulative `r study_years`), equivalent to
`r fmt_k(total_bp_rescaled)` thousand deaths averted across all
`r n_countries` program countries. Over a 5-year window (2020--2024),
the average program country averts `r fmt_k(lives_per_1M_htn_5yr)` thousand
deaths per million treated.

TFA elimination policies are projected to avert
`r fmt_k(lives_per_1M_tfa)` thousand deaths per million population
(cumulative), rescaling to `r fmt_k(total_tfa_rescaled)` thousand actual
deaths averted across program countries.

Combined, `r fmt_k(lives_per_1M_combined)` thousand deaths per million
treated are averted, equivalent to `r fmt_k(total_combined_rescaled)` thousand
deaths averted in absolute terms.

---

# Outputs by Demographics

## By Sex and Age Group

```{r demo-age-sex, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
dt_age_sex <- dt_merged[
  intervention %in% int_order,
  .(deaths_averted = sum(deaths_averted, na.rm = TRUE)),
  by = .(intervention, age_group, sex)
]

order_lvls <- dt_age_sex[, .(total = sum(deaths_averted, na.rm = TRUE)), by = intervention
                         ][order(total), intervention]
dt_age_sex[, intervention := factor(intervention, levels = order_lvls)]

g <- ggplot(dt_age_sex,
            aes(x = deaths_averted / 1e3, y = intervention, fill = age_group)) +
  geom_col(width = 0.7) +
  facet_wrap(~sex, nrow = 1) +
  scale_x_continuous(expand = c(0, 0), labels = comma) +
  labs(
    title = "Deaths Averted by Age Group, Sex, and Intervention (2017\u20132050)",
    x     = "Deaths averted (thousands)",
    y     = NULL,
    fill  = "Age group"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.5),
        strip.text       = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_fig5_demo_age.png"), plot = g,
       width = 10, height = 5, dpi = 300)
```

## Table 3 --- Deaths Averted by Age Group and Intervention

```{r table3-age, message=FALSE, warning=FALSE, echo=FALSE}
# Build age-group summary using program vs bau comparison
dt_tab3_base <- data_bau[year >= 2017 & intervention %in% int_order,
                          .(baseline_deaths = sum(dead, na.rm = TRUE)),
                          by = .(age_group, intervention)]
dt_tab3_pgm  <- data_program[year >= 2017 & intervention %in% int_order,
                              .(pgm_deaths = sum(dead, na.rm = TRUE)),
                              by = .(age_group, intervention)]

dt_tab3 <- merge(dt_tab3_pgm, dt_tab3_base,
                 by = c("age_group", "intervention"), all = TRUE)
dt_tab3[, deaths_averted    := baseline_deaths - pgm_deaths]
dt_tab3[, deaths_averted_yr := deaths_averted / n_years]
dt_tab3[, pct_averted       := deaths_averted / baseline_deaths]

dt_tab3[, intervention := factor(intervention, levels = int_order)]
setorder(dt_tab3, age_group, intervention)

dt_tab3_fmt <- dt_tab3[, .(
  Age_group          = age_group,
  Intervention       = intervention,
  Counterfactual     = comma(baseline_deaths,   accuracy = 1),
  Program            = comma(pgm_deaths,        accuracy = 1),
  Deaths_averted     = comma(deaths_averted,    accuracy = 1),
  Deaths_averted_yr  = comma(deaths_averted_yr, accuracy = 1),
  Pct_averted        = percent(pct_averted, accuracy = 0.1)
)]
setnames(dt_tab3_fmt,
         old = c("Age_group", "Counterfactual", "Program",
                 "Deaths_averted", "Deaths_averted_yr", "Pct_averted"),
         new = c("Age group", "Counterfactual deaths", "Program deaths",
                 "Deaths averted", "Deaths averted/year", "% averted"))

kable(dt_tab3_fmt,
      caption = "Table 3. Deaths averted (2017\u20132050) by age group and intervention.",
      align   = c("l","l","r","r","r","r","r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12) |>
  row_spec(0, bold = TRUE)

dt_excel_tab3 <- copy(dt_tab3_fmt)
```

---

# Outputs by Geography

## Table 4 --- Deaths Averted by Country (HTN + TFA Combined)

```{r table4-country, message=FALSE, warning=FALSE, echo=FALSE}
dt_tab4_base <- data_bau[year >= 2017 &
                           intervention == "HTN Control + TFA Elimination",
                          .(baseline_deaths = sum(dead, na.rm = TRUE)),
                          by = .(region, location)]
dt_tab4_pgm  <- data_program[year >= 2017 &
                               intervention == "HTN Control + TFA Elimination",
                              .(pgm_deaths = sum(dead, na.rm = TRUE)),
                              by = .(region, location)]

dt_tab4 <- merge(dt_tab4_pgm, dt_tab4_base, by = c("region", "location"), all = TRUE)
dt_tab4[, deaths_averted    := baseline_deaths - pgm_deaths]
dt_tab4[, deaths_averted_yr := deaths_averted / n_years]
dt_tab4[, pct_averted       := deaths_averted / baseline_deaths]
setorder(dt_tab4, region, -deaths_averted)

dt_tab4_fmt <- dt_tab4[, .(
  Region             = region,
  Country            = location,
  Counterfactual     = comma(baseline_deaths,    accuracy = 1),
  Program            = comma(pgm_deaths,         accuracy = 1),
  Deaths_averted     = comma(deaths_averted,     accuracy = 1),
  Deaths_averted_yr  = comma(deaths_averted_yr,  accuracy = 1),
  Pct_averted        = percent(pct_averted, accuracy = 0.1)
)]
setnames(dt_tab4_fmt,
         old = c("Counterfactual", "Program",
                 "Deaths_averted", "Deaths_averted_yr", "Pct_averted"),
         new = c("Counterfactual deaths", "Program deaths",
                 "Deaths averted", "Deaths averted/year", "% averted"))

datatable(
  dt_tab4_fmt,
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left;",
    "Table 4. Country-level program-attributable deaths averted (HTN + TFA combined), 2017\u20132050."
  ),
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 15, scrollX = TRUE, dom = "lftipr")
)

dt_excel_tab4 <- copy(dt_tab4_fmt)
```

## Figure 5 --- Top Countries by Deaths Averted

```{r figure5-top-countries, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=7}
n_show <- min(n_countries, 20)
dt_top <- dt_tab4[order(-deaths_averted)][1:n_show]
dt_top[, location := factor(location, levels = rev(location))]

g <- ggplot(dt_top, aes(x = deaths_averted / 1e3, y = location, fill = region)) +
  geom_col(width = 0.7) +
  scale_x_continuous(expand = c(0, 0), labels = comma) +
  labs(
    title = paste0("Top ", n_show, " Countries by Program-Attributable Deaths Averted\n",
                   "(HTN + TFA, 2017\u20132050)"),
    x     = "Deaths averted (thousands)",
    y     = NULL,
    fill  = "Region"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position    = "right",
        panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title         = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_fig5_top_countries.png"), plot = g,
       width = 10, height = 7, dpi = 300)
```

## Figure 6 --- World Map: Deaths Averted per 100,000 Population

```{r figure6-map, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
wpp <- tryCatch(
  as.data.table(readRDS(paste0(wd_data, "PopulationsSingleAge0050.rds"))),
  error = function(e) {
    tryCatch(
      as.data.table(readRDS(paste0(wd_data, "UN2024/PopulationsSingleAge0050.rds"))),
      error = function(e2) NULL
    )
  }
)

if (!is.null(wpp)) {
  pop_2019 <- wpp[
    year_id == 2019 & age >= 20,
    .(pop = sum(Nx, na.rm = TRUE)),
    by = location
  ]
  
  dt_map <- merge(dt_tab4, pop_2019, by = "location", all.x = TRUE)
  dt_map[, da_per1m := deaths_averted / pop * 1e6]

  if (nrow(countries_dt) > 0) {
    world <- ne_countries(scale = "medium", returnclass = "sf")
    
    world_dt <- merge(
      world,
      merge(dt_map, countries_dt[, .(location, iso3)],
            by = "location", all.x = TRUE),
      by.x = "adm0_a3", by.y = "iso3", all.x = TRUE
    )

    g <- ggplot(world_dt) +
      geom_sf(aes(fill = da_per1m), color = "grey80", linewidth = 0.1) +
      scale_fill_viridis_c(
        name = "per 1,000,000",
        na.value = "grey90",
        option = "C",
        direction = -1
      ) +
      labs(
        title = "Program-Attributable Deaths Averted per 1,000,000 Population, 2017–2050",
        subtitle = "HTN control + TFA elimination combined",
        caption = "Source: Model output; denominator = UNWPP 2024 age 20+ population"
      ) +
      theme_void(base_size = 12) +
      theme(
        legend.position = "right",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40")
      )

    print(g)

    ggsave(
      paste0(wd_outp, "assessment_fig6_map.png"),
      plot = g,
      width = 11, height = 5, dpi = 300
    )
  }
} else {
  message("Population file not found - skipping map.")
}
```

## Table 5 --- Deaths Averted by Region

```{r table5-region, message=FALSE, warning=FALSE, echo=FALSE}
dt_tab5_base <- data_bau[year >= 2017 &
                           intervention == "HTN Control + TFA Elimination",
                          .(baseline_deaths = sum(dead, na.rm = TRUE)), by = region]
dt_tab5_pgm  <- data_program[year >= 2017 &
                               intervention == "HTN Control + TFA Elimination",
                              .(pgm_deaths = sum(dead, na.rm = TRUE)), by = region]

dt_tab5 <- merge(dt_tab5_pgm, dt_tab5_base, by = "region", all = TRUE)
dt_tab5[, deaths_averted    := baseline_deaths - pgm_deaths]
dt_tab5[, deaths_averted_yr := deaths_averted / n_years]
dt_tab5[, pct_averted       := deaths_averted / baseline_deaths]
setorder(dt_tab5, -deaths_averted)

dt_tab5_fmt <- dt_tab5[, .(
  Region             = region,
  Counterfactual     = comma(baseline_deaths,    accuracy = 1),
  Program            = comma(pgm_deaths,         accuracy = 1),
  Deaths_averted     = comma(deaths_averted,     accuracy = 1),
  Deaths_averted_yr  = comma(deaths_averted_yr,  accuracy = 1),
  Pct_averted        = percent(pct_averted, accuracy = 0.1)
)]
setnames(dt_tab5_fmt,
         old = c("Counterfactual", "Program",
                 "Deaths_averted", "Deaths_averted_yr", "Pct_averted"),
         new = c("Counterfactual deaths", "Program deaths",
                 "Deaths averted", "Deaths averted/year", "% averted"))

kable(dt_tab5_fmt,
      caption = "Table 5. Deaths averted by region (HTN + TFA combined), 2017\u20132050.",
      align   = c("l","r","r","r","r","r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE)

dt_excel_tab5 <- copy(dt_tab5_fmt)
```

---

# ASMR Under Program vs Counterfactual

```{r asmr-interventions, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=7}
# Compute ASMR for program and counterfactual runs (combined intervention = HTN+TFA)
dt_asmr_pgm <- data_program[
  intervention == "HTN Control + TFA Elimination" & year >= 2017,
  .(dead = sum(dead, na.rm = TRUE),
    well = sum(well, na.rm = TRUE),
    sick = sum(sick, na.rm = TRUE)),
  by = .(age, year)]
dt_asmr_pgm[, Nx := well + sick]
dt_asmr_pgm[, scenario := "Program (HTN + TFA)"]

dt_asmr_bau <- data_bau[
  intervention == "HTN Control + TFA Elimination" & year >= 2017,
  .(dead = sum(dead, na.rm = TRUE),
    well = sum(well, na.rm = TRUE),
    sick = sum(sick, na.rm = TRUE)),
  by = .(age, year)]
dt_asmr_bau[, Nx := well + sick]
dt_asmr_bau[, scenario := "Counterfactual"]

dt_asmr_both <- rbind(dt_asmr_pgm, dt_asmr_bau)
asmr_int <- calculate_asmr(dt_asmr_both, deaths_col = "dead",
                            population_col = "Nx", disagg_var = "scenario")

g <- ggplot(asmr_int,
            aes(x = year, y = ASMR, color = scenario, linetype = year > 2024)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Counterfactual"       = "#999999",
                                "Program (HTN + TFA)"  = "#6a3d9a"),
                     name = NULL) +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(
    title    = "Age-Standardised CVD Mortality Rate: Program vs Counterfactual",
    subtitle = paste0("All causes combined; solid = historical, dashed = projected; ",
                      n_countries, " RTSL countries"),
    x        = "Year",
    y        = "ASMR (per 100,000 population)",
    caption  = "WHO World Standard Population (2000\u20132025), rescaled to ages 20\u201395+"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
print(g)
ggsave(paste0(wd_outp, "assessment_asmr_interventions.png"), plot = g,
       width = 12, height = 6, dpi = 300)
```

---

# Premature Mortality (Deaths < 70 years)

```{r premature-mortality, message=FALSE, warning=FALSE, echo=FALSE}
dt_u70 <- dt_merged[intervention %in% int_order & age_group == "<70",
                     .(deaths_averted = sum(deaths_averted, na.rm = TRUE)),
                     by = .(intervention, sex)]

dt_u70_total <- dt_merged[intervention == "HTN Control + TFA Elimination" &
                            age_group == "<70",
                          .(deaths_averted_u70 = sum(deaths_averted, na.rm = TRUE)),
                          by = location]

dt_u70_all <- dt_merged[intervention == "HTN Control + TFA Elimination",
                         .(deaths_averted_total = sum(deaths_averted, na.rm = TRUE)),
                         by = location]

dt_u70_loc <- merge(dt_u70_total, dt_u70_all, by = "location")
dt_u70_loc[, prop_u70 := fifelse(deaths_averted_total == 0, NA_real_,
                                  deaths_averted_u70 / deaths_averted_total)]
setorder(dt_u70_loc, -deaths_averted_u70)

kable(dt_u70[, .(Intervention = intervention, Sex = sex,
                  `Deaths averted <70` = comma(deaths_averted, accuracy = 1))],
      caption = "Deaths averted among people under 70, by intervention and sex (2017\u20132050).",
      align = "llr") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE)
```

```{r premature-country-table, message=FALSE, warning=FALSE, echo=FALSE}
dt_u70_fmt <- dt_u70_loc[, .(
  Country = location,
  `Deaths averted <70` = comma(deaths_averted_u70, accuracy = 1),
  `Share of total averted` = percent(prop_u70, accuracy = 0.1)
)]

datatable(
  dt_u70_fmt,
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left;",
    "Deaths averted under age 70 by country (HTN + TFA combined, 2017\u20132050)."
  ),
  filter   = "top",
  rownames = FALSE,
  options  = list(pageLength = 15, scrollX = TRUE, dom = "lftipr")
)
```

---

# Excel Export

```{r excel-export, message=FALSE, warning=FALSE, echo=FALSE}
wb <- createWorkbook()

add_sheet <- function(wb, sheet_name, dat, title_str) {
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, title_str, startRow = 1, startCol = 1)
  addStyle(wb, sheet_name,
           style = createStyle(fontSize = 12, textDecoration = "bold"),
           rows = 1, cols = 1)
  writeData(wb, sheet_name, dat, startRow = 3, startCol = 1,
            headerStyle = createStyle(textDecoration = "bold",
                                      fgFill = "#2E75B6", fontColour = "white",
                                      border = "Bottom"))
  setColWidths(wb, sheet_name, cols = seq_len(ncol(dat)), widths = "auto")
}

add_sheet(wb, "T1_Deaths_by_Intervention", as.data.frame(dt_excel_tab1),
          "Table 1. Program-attributable deaths averted (2017-2050) by intervention.")
add_sheet(wb, "T2_Deaths_by_Cause", as.data.frame(dt_excel_tab2),
          "Table 2. Deaths averted (thousands) by intervention and cause of death.")
add_sheet(wb, "T3_Deaths_by_Age", as.data.frame(dt_excel_tab3),
          "Table 3. Deaths averted by age group and intervention.")
add_sheet(wb, "T4_Deaths_by_Country", as.data.frame(dt_excel_tab4),
          "Table 4. Country-level deaths averted (HTN + TFA, 2017-2050).")

if (exists("dt_excel_tab5") && nrow(dt_excel_tab5) > 0)
  add_sheet(wb, "T5_Deaths_by_Region", as.data.frame(dt_excel_tab5),
            "Table 5. Deaths averted by region (HTN + TFA, 2017-2050).")

add_sheet(wb, "Countries", as.data.frame(countries_by_region),
          "Countries included in the assessment, by region.")

# Premature mortality
if (exists("dt_u70_loc") && nrow(dt_u70_loc) > 0)
  add_sheet(wb, "U70_by_Country", as.data.frame(dt_u70_loc),
            "Deaths averted under age 70 by country (HTN + TFA, 2017-2050).")

# ---- Per1M_Summary sheet ----
# One row per location x year x scenario, all scenarios stacked
dt_cf_ly <- data_bau[
  intervention == "Counterfactual" & year >= 2017,
  .(deaths = sum(dead, na.rm = TRUE)),
  by = .(location, year)
]
dt_cf_ly[, scenario := "Counterfactual"]

dt_int_ly <- dt_merged[
  year >= 2017,
  .(deaths         = sum(pgm_deaths,     na.rm = TRUE),
    deaths_averted = sum(deaths_averted, na.rm = TRUE)),
  by = .(location, year, intervention)
]
setnames(dt_int_ly, "intervention", "scenario")

# Counterfactual deaths reference (for pct_averted)
dt_cf_ref_ly <- dt_cf_ly[, .(location, year, cf_deaths_ref = deaths)]

dt_excel_per1m <- rbind(
  dt_cf_ly[, .(location, year, scenario, deaths, deaths_averted = 0)],
  dt_int_ly[, .(location, year, scenario, deaths, deaths_averted)],
  fill = TRUE
)
dt_excel_per1m <- merge(dt_excel_per1m, dt_cf_ref_ly,
                        by = c("location", "year"), all.x = TRUE)

# Add population denominators
if (!is.null(dt_total_pop)) {
  dt_excel_per1m <- merge(dt_excel_per1m, dt_total_pop,
                          by = c("location", "year"), all.x = TRUE)
} else {
  dt_excel_per1m[, total_population := NA_real_]
}

if (!is.null(dt_program_pop)) {
  dt_excel_per1m <- merge(dt_excel_per1m, dt_program_pop,
                          by = c("location", "year"), all.x = TRUE)
} else {
  dt_excel_per1m[, program_population := NA_real_]
}

# For TFA and Counterfactual: program_population = total_population
dt_excel_per1m[
  scenario %in% c("TFA Elimination", "Counterfactual"),
  program_population := total_population
]

# Rename total_population -> population
setnames(dt_excel_per1m, "total_population", "population")

# Compute per-1M columns (model output is already per 1M; values are direct)
dt_excel_per1m[, deaths_per_1M         := round(deaths, 1)]
dt_excel_per1m[, deaths_averted_per_1M := round(deaths_averted, 1)]
dt_excel_per1m[scenario == "Counterfactual", deaths_averted_per_1M := 0]

# Absolute deaths averted (rescaled)
dt_excel_per1m[
  scenario %in% c("HTN Control (Program)",
                   "HTN Control + TFA Elimination"),
  n_deaths_averted := deaths_averted * (program_population / 1e6)
]
dt_excel_per1m[
  scenario == "TFA Elimination",
  n_deaths_averted := deaths_averted * (population / 1e6)
]
dt_excel_per1m[scenario == "Counterfactual", n_deaths_averted := 0]

# pct_averted
dt_excel_per1m[, pct_averted := round(
  fifelse(cf_deaths_ref > 0 & scenario != "Counterfactual",
          deaths_averted / cf_deaths_ref * 100, 0), 2)]
dt_excel_per1m[scenario == "Counterfactual", pct_averted := 0]

# Assigne deaths averted per 1M and pct averted to 0 for any rows where population denominator is 0
dt_excel_per1m[program_population == 0, `:=`(
  deaths_averted_per_1M = 0,
  pct_averted = 0
)]


# Select final columns and sort
dt_excel_per1m_final <- dt_excel_per1m[, .(
  location, year, scenario, population, program_population,
  deaths_per_1M, deaths_averted_per_1M, n_deaths_averted, pct_averted
)]
setorder(dt_excel_per1m_final, location, scenario, year)

add_sheet(wb, "Per1M_Summary", as.data.frame(dt_excel_per1m_final),
          "Per 1M Summary: deaths and deaths averted by location, year, and scenario.")

out_xlsx <- paste0(wd_scen, "RTSL_Assessment_Results.xlsx")
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
message("Tables exported to: ", out_xlsx)
```

# Slide Deck Outputs

```{r slides-setup, message=FALSE, warning=FALSE, echo=FALSE}
slides_dir <- paste0(wd_outp, "slides/")
if (!dir.exists(slides_dir)) dir.create(slides_dir, recursive = TRUE)
message("Slide artefacts will be written to: ", slides_dir)
```


```{r pareto-charts, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
# # ---- Pareto Charts: Top 20 Countries by Deaths Averted ----------------------
# 
# # Helper: build a top-20 + "Rest of countries" summary
# pareto_top20 <- function(dt, scenario_filter) {
#   dt_agg <- dt[scenario == scenario_filter,
#                .(deaths_averted = sum(n_deaths_averted, na.rm = TRUE)),
#                by = location]
#   setorder(dt_agg, -deaths_averted)
#   top20 <- head(dt_agg, 20)
#   rest  <- dt_agg[!location %in% top20$location,
#                   .(location = "Rest of countries",
#                     deaths_averted = sum(deaths_averted, na.rm = TRUE))]
#   out <- rbind(top20, rest)
#   out[, location := factor(location, levels = out$location)]
#   out
# }
# 
# # --- 1. HTN Control Pareto ---
# dt_pareto_htn <- pareto_top20(dt_excel_per1m_final, "HTN Control (Program)")
# 
# p_pareto_htn <- ggplot(dt_pareto_htn,
#                         aes(x = location, y = deaths_averted)) +
#   geom_col(fill = int_cols["HTN Control (Program)"]) +
#   scale_y_continuous(labels = scales::comma) +
#   labs(x = "Country", y = "Deaths Averted (2017\u20132050)",
#        title = "HTN Control: Top 20 Countries by Deaths Averted") +
#   theme_minimal(base_size = 12) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 
# # --- 2. TFA Elimination Pareto ---
# dt_pareto_tfa <- pareto_top20(dt_excel_per1m_final, "TFA Elimination")
# 
# p_pareto_tfa <- ggplot(dt_pareto_tfa,
#                         aes(x = location, y = deaths_averted)) +
#   geom_col(fill = int_cols["TFA Elimination"]) +
#   scale_y_continuous(labels = scales::comma) +
#   labs(x = "Country", y = "Deaths Averted (2017\u20132050)",
#        title = "TFA Elimination: Top 20 Countries by Deaths Averted") +
#   theme_minimal(base_size = 12) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 
# # --- 3. Combined Stacked Pareto ---
# # Sum HTN and TFA separately per country (do NOT use the combined scenario row)
# dt_htn_by_loc <- dt_excel_per1m_final[
#   scenario == "HTN Control (Program)",
#   .(deaths_averted = sum(n_deaths_averted, na.rm = TRUE)),
#   by = location
# ]
# dt_tfa_by_loc <- dt_excel_per1m_final[
#   scenario == "TFA Elimination",
#   .(deaths_averted = sum(n_deaths_averted, na.rm = TRUE)),
#   by = location
# ]
# 
# dt_combined <- merge(
#   dt_htn_by_loc[, .(location, htn = deaths_averted)],
#   dt_tfa_by_loc[, .(location, tfa = deaths_averted)],
#   by = "location", all = TRUE
# )
# dt_combined[is.na(htn), htn := 0]
# dt_combined[is.na(tfa), tfa := 0]
# dt_combined[, total := htn + tfa]
# setorder(dt_combined, -total)
# 
# top20_locs <- head(dt_combined$location, 20)
# rest_row <- dt_combined[!location %in% top20_locs,
#                         .(location = "Rest of countries",
#                           htn = sum(htn), tfa = sum(tfa),
#                           total = sum(total))]
# dt_combined_top <- rbind(dt_combined[location %in% top20_locs], rest_row)
# dt_combined_top[, location := factor(location, levels = dt_combined_top$location)]
# 
# # Melt for stacked bar
# dt_stacked <- melt(dt_combined_top, id.vars = "location",
#                    measure.vars = c("htn", "tfa"),
#                    variable.name = "intervention",
#                    value.name = "deaths_averted")
# dt_stacked[, intervention := fifelse(intervention == "htn",
#                                       "HTN Control (Program)",
#                                       "TFA Elimination")]
# 
# p_pareto_combined <- ggplot(dt_stacked,
#                              aes(x = location, y = deaths_averted,
#                                  fill = intervention)) +
#   geom_col() +
#   scale_fill_manual(values = int_cols[c("HTN Control (Program)",
#                                          "TFA Elimination")]) +
#   scale_y_continuous(labels = scales::comma) +
#   labs(x = "Country", y = "Deaths Averted (2017\u20132050)",
#        title = "Combined Interventions: Top 20 Countries by Deaths Averted",
#        fill = "Intervention") +
#   theme_minimal(base_size = 12) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 
# # Save plot objects for presentation_donors.Rmd
# saveRDS(p_pareto_htn,      paste0(slides_dir, "sl_pareto_htn.rds"))
# saveRDS(p_pareto_tfa,      paste0(slides_dir, "sl_pareto_tfa.rds"))
# saveRDS(p_pareto_combined, paste0(slides_dir, "sl_pareto_combined.rds"))
# message("Pareto chart objects saved to slides directory.")
# ---- Pareto Charts: Top 20 Countries by Deaths Averted ----------------------

# Helper: build a top-20 + "Rest of countries" summary with cumulative %
pareto_top20 <- function(dt, scenario_filter) {
  dt_agg <- dt[scenario == scenario_filter,
               .(deaths_averted = sum(n_deaths_averted, na.rm = TRUE)),
               by = location]
  setorder(dt_agg, -deaths_averted)

  top20 <- head(dt_agg, 20)
  rest  <- dt_agg[!location %in% top20$location,
                  .(location = "Rest of countries",
                    deaths_averted = sum(deaths_averted, na.rm = TRUE))]
  
  out <- rbind(top20, rest, fill = TRUE)
  setorder(out, -deaths_averted)

  out[, cum_deaths := cumsum(deaths_averted)]
  out[, cum_pct := cum_deaths / sum(deaths_averted)]
  out[, location := factor(location, levels = location)]
  
  out
}

# Helper: make a Pareto plot
make_pareto_plot <- function(dt_pareto, fill_color, title_text) {
  y_max <- max(dt_pareto$deaths_averted, na.rm = TRUE)

  ggplot(dt_pareto, aes(x = location)) +
    geom_col(aes(y = deaths_averted), fill = fill_color) +
    geom_line(aes(y = cum_pct * y_max, group = 1), linewidth = 1) +
    geom_point(aes(y = cum_pct * y_max), size = 2) +
    scale_y_continuous(
      labels = scales::comma,
      name   = "Deaths Averted (2017–2050)",
      sec.axis = sec_axis(
        ~ . / y_max,
        name   = "Cumulative percentage",
        labels = scales::percent_format(accuracy = 1)
      )
    ) +
    labs(x = "Country", title = title_text) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

# --- 1. HTN Control Pareto ---
dt_pareto_htn <- pareto_top20(dt_excel_per1m_final, "HTN Control (Program)")

p_pareto_htn <- make_pareto_plot(
  dt_pareto = dt_pareto_htn,
  fill_color = int_cols["HTN Control (Program)"],
  title_text = "HTN Control: Top 20 Countries by Deaths Averted"
)

# --- 2. TFA Elimination Pareto ---
dt_pareto_tfa <- pareto_top20(dt_excel_per1m_final, "TFA Elimination")

p_pareto_tfa <- make_pareto_plot(
  dt_pareto = dt_pareto_tfa,
  fill_color = int_cols["TFA Elimination"],
  title_text = "TFA Elimination: Top 20 Countries by Deaths Averted"
)

# --- 3. Combined Pareto (HTN + TFA summed by country) ---
# Sum HTN and TFA separately per country (do NOT use the combined scenario row)
dt_htn_by_loc <- dt_excel_per1m_final[
  scenario == "HTN Control (Program)",
  .(htn = sum(n_deaths_averted, na.rm = TRUE)),
  by = location
]

dt_tfa_by_loc <- dt_excel_per1m_final[
  scenario == "TFA Elimination",
  .(tfa = sum(n_deaths_averted, na.rm = TRUE)),
  by = location
]

dt_combined <- merge(dt_htn_by_loc, dt_tfa_by_loc, by = "location", all = TRUE)
dt_combined[is.na(htn), htn := 0]
dt_combined[is.na(tfa), tfa := 0]
dt_combined[, deaths_averted := htn + tfa]
setorder(dt_combined, -deaths_averted)

top20 <- head(dt_combined, 20)
rest  <- dt_combined[!location %in% top20$location,
                     .(location = "Rest of countries",
                       htn = sum(htn),
                       tfa = sum(tfa),
                       deaths_averted = sum(deaths_averted))]
dt_pareto_combined <- rbind(top20, rest, fill = TRUE)
setorder(dt_pareto_combined, -deaths_averted)

dt_pareto_combined[, cum_deaths := cumsum(deaths_averted)]
dt_pareto_combined[, cum_pct := cum_deaths / sum(deaths_averted)]
dt_pareto_combined[, location := factor(location, levels = location)]

p_pareto_combined <- make_pareto_plot(
  dt_pareto = dt_pareto_combined,
  fill_color = "grey60",
  title_text = "Combined Interventions: Top 20 Countries by Deaths Averted"
)

# Save plot objects for presentation_donors.Rmd
saveRDS(p_pareto_htn,      paste0(slides_dir, "sl_pareto_htn.rds"))
saveRDS(p_pareto_tfa,      paste0(slides_dir, "sl_pareto_tfa.rds"))
saveRDS(p_pareto_combined, paste0(slides_dir, "sl_pareto_combined.rds"))
message("Pareto chart objects saved to slides directory.")
```

---

```{r slides-scalars, message=FALSE, warning=FALSE, echo=FALSE}
bau_total_deaths <- data_bau[intervention == "HTN Control + TFA Elimination" &
                               year >= 2017,
                              sum(dead, na.rm = TRUE)]
n_htn_countries <- length(htn_countries_include)
n_tfa_countries <- length(tfa_countries_include)

sl_scalars <- list(
  n_countries      = n_countries,
  n_htn_countries  = n_htn_countries,
  n_tfa_countries  = n_tfa_countries,
  study_years      = study_years,
  year_min         = year_min,
  year_max         = year_max,
  n_years          = n_years,
  total_combined   = total_combined,
  total_bp         = total_bp,
  total_tfa        = total_tfa,
  bau_total_deaths = bau_total_deaths,
  lives_per_1M_htn      = lives_per_1M_htn,
  lives_per_1M_tfa      = lives_per_1M_tfa,
  lives_per_1M_combined = lives_per_1M_combined
)
saveRDS(sl_scalars, paste0(slides_dir, "sl_scalars.rds"))
message("SL-01 saved: sl_scalars.rds")
```

```{r slides-tab-summary, message=FALSE, warning=FALSE, echo=FALSE}
sl_tab_summary <- copy(dt_cumul)
sl_tab_summary[, pct    := deaths_averted / bau_total_deaths * 100]
sl_tab_summary[, cum_K  := paste0(fmt_k(deaths_averted), " K")]
sl_tab_summary[, avg_yr := paste0(fmt_k(deaths_averted / n_years), " K")]
sl_tab_summary[, pct_s  := paste0(round(pct, 1), "%")]
sl_tab_summary[, intervention := factor(intervention, levels = int_order)]
setorder(sl_tab_summary, intervention)
saveRDS(sl_tab_summary, paste0(slides_dir, "sl_tab_summary.rds"))
message("SL-02 saved: sl_tab_summary.rds")
```

```{r slides-tab-region, message=FALSE, warning=FALSE, echo=FALSE}
sl_tab_region <- merge(
  data_program[intervention == "HTN Control + TFA Elimination" & year >= 2017,
               .(pgm = sum(dead, na.rm = TRUE)), by = region],
  data_bau[intervention == "HTN Control + TFA Elimination" & year >= 2017,
           .(bau = sum(dead, na.rm = TRUE)), by = region],
  by = "region"
)
sl_tab_region[, averted := bau - pgm]
sl_tab_region[, pct     := averted / bau * 100]
setorder(sl_tab_region, -averted)
saveRDS(sl_tab_region, paste0(slides_dir, "sl_tab_region.rds"))
message("SL-03 saved: sl_tab_region.rds")
```

```{r slides-tab-top15, message=FALSE, warning=FALSE, echo=FALSE}
sl_tab_top15 <- merge(
  data_program[intervention == "HTN Control + TFA Elimination" & year >= 2017,
               .(pgm_dead = sum(dead, na.rm = TRUE),
                 region   = region[1]), by = location],
  data_bau[intervention == "HTN Control + TFA Elimination" & year >= 2017,
           .(bau = sum(dead, na.rm = TRUE)), by = location],
  by = "location"
)
sl_tab_top15[, averted := bau - pgm_dead]
setorder(sl_tab_top15, -averted)
sl_tab_top15 <- sl_tab_top15[1:min(15, .N)]
saveRDS(sl_tab_top15, paste0(slides_dir, "sl_tab_top15.rds"))
message("SL-04 saved: sl_tab_top15.rds")
```

```{r slides-dt-cause, message=FALSE, warning=FALSE, echo=FALSE}
sl_dt_cause <- dt_merged[intervention %in% int_order & cause %in% names(cause_map),
                          .(averted = sum(deaths_averted, na.rm = TRUE)),
                          by = .(cause, intervention)]
sl_dt_cause[, cause_lbl := fcase(
  cause == "ihd",     "IHD",
  cause == "istroke", "Ischemic stroke",
  cause == "hstroke", "Hemorrhagic stroke",
  cause == "hhd",     "Hypertensive HD",
  default = cause
)]
sl_dt_cause[, intervention := factor(intervention, levels = int_order)]
saveRDS(sl_dt_cause, paste0(slides_dir, "sl_dt_cause.rds"))
message("SL-05 saved: sl_dt_cause.rds")
```

```{r slides-dt-annual, message=FALSE, warning=FALSE, echo=FALSE}
saveRDS(copy(dt_annual)[intervention %in% int_order],
        paste0(slides_dir, "sl_dt_annual.rds"))
message("SL-06 saved: sl_dt_annual.rds")
```

```{r slides-fig-deaths-averted, message=FALSE, warning=FALSE, echo=FALSE}
g_ann_sl <- ggplot(dt_annual[!intervention %in% c("HTN Control + TFA Elimination")],
                    aes(x = year, y = deaths_averted / 1e3, fill = intervention)) +
  geom_area(alpha = 0.85, position = "stack") +
  scale_fill_manual(values = int_cols[int_order], name = NULL) +
  scale_y_continuous(labels = comma) +
  labs(x = "Year", y = "Deaths averted (thousands)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7),
        panel.grid.minor = element_blank()) +
  guides(fill = guide_legend(nrow = 2))
ggsave(paste0(slides_dir, "sl_fig_deaths_averted.png"), plot = g_ann_sl,
       width = 8, height = 4, dpi = 200)
message("SL-07 saved: sl_fig_deaths_averted.png")
```

```{r slides-fig-top15, message=FALSE, warning=FALSE, echo=FALSE}
sl_t15_plot <- copy(sl_tab_top15)
sl_t15_plot[, location := factor(location, levels = rev(location))]
g_top15_sl <- ggplot(sl_t15_plot,
                      aes(x = averted / 1e3, y = location, fill = region)) +
  geom_col(width = 0.7) +
  scale_x_continuous(labels = comma, expand = c(0, 0)) +
  labs(x = "Deaths averted (thousands)", y = NULL, fill = "Region") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8))
ggsave(paste0(slides_dir, "sl_fig_top15_countries.png"), plot = g_top15_sl,
       width = 8, height = 5, dpi = 200)
message("SL-08 saved: sl_fig_top15_countries.png")
```

```{r slides-fig-cumulative, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
g <- ggplot(dt_annual,
            aes(x = year, y = cum_deaths_averted / 1e3, color = intervention)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = int_cols, name = NULL) +
  scale_y_continuous(labels = comma) +
  labs(x = "Year", y = "Cumulative deaths averted (thousands)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"))
print(g)
ggsave(paste0(slides_dir, "sl_fig_cum_averted.png"), plot = g,
       width = 11, height = 5, dpi = 300)
message("SL-09 saved: sl_fig_cum_averted.png")
```

```{r slides-bangladesh, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Bangladesh country case artefact for presentation ---------------------
bgd_rds <- paste0(wd_outp, "out_model/model_output_Bangladesh_program_impact.rds")
if (file.exists(bgd_rds)) {
  dt_bgd <- setDT(readRDS(bgd_rds))

  # Program-attributable deaths by scenario
  dt_bgd_pgm <- dt_bgd[rate_scenario == "program" & year >= 2017,
                         .(pgm_deaths = sum(dead, na.rm = TRUE)),
                         by = .(scenario, year)]
  dt_bgd_cf  <- dt_bgd[rate_scenario == "bau_upper" &
                         scenario == "baseline" & year >= 2017,
                         .(cf_deaths = sum(dead, na.rm = TRUE)),
                         by = year]

  dt_bgd_merge <- merge(dt_bgd_pgm, dt_bgd_cf, by = "year")
  dt_bgd_merge[, averted := cf_deaths - pgm_deaths]

  # Summary table: total averted by scenario
  sl_bgd_summary <- dt_bgd_merge[, .(total_averted = sum(averted)), by = scenario]

  # Annual time-series for Bangladesh
  sl_bgd_annual <- dt_bgd_merge[, .(year, scenario, averted)]
  sl_bgd_annual[, cum_averted := cumsum(averted), by = scenario]

  # By cause
  dt_bgd_cause_pgm <- dt_bgd[rate_scenario == "program" & year >= 2017,
                               .(pgm_deaths = sum(dead, na.rm = TRUE)),
                               by = .(scenario, cause)]
  dt_bgd_cause_cf  <- dt_bgd[rate_scenario == "bau_upper" &
                               scenario == "baseline" & year >= 2017,
                               .(cf_deaths = sum(dead, na.rm = TRUE)),
                               by = cause]
  sl_bgd_cause <- merge(dt_bgd_cause_pgm, dt_bgd_cause_cf, by = "cause")
  sl_bgd_cause[, averted := cf_deaths - pgm_deaths]

  sl_bangladesh <- list(
    summary = sl_bgd_summary,
    annual  = sl_bgd_annual,
    cause   = sl_bgd_cause
  )
  saveRDS(sl_bangladesh, paste0(slides_dir, "sl_bangladesh.rds"))
  message("SL-10 saved: sl_bangladesh.rds")
} else {
  message("Bangladesh model output not found — skipping sl_bangladesh.rds")
}
```

```{r slides-tab-scenario, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Scenario comparison table for presentation ----------------------------
sl_tab_scenario <- data.table(
  Intervention = c("HTN Control (Program)", "TFA Elimination",
                    "HTN Control + TFA Elimination"),
  Mechanism = c(
    "Antihypertensive therapy using observed RTSL program control rates",
    "Elimination of industrial TFA via enacted best-practice policies",
    "Both interventions combined (multiplicative effect ratios)"
  ),
  Key_Param = c(
    "Observed control rates (2017-2024); maintained at 2024 level to 2050",
    "Target: 0% energy from TFA; 2-year lag from policy passage to effect",
    "epsilon_IR_total = epsilon_IR_BP x epsilon_IR_TFA"
  ),
  Deaths_averted_K = c(
    round(total_bp / 1e3, 1),
    round(total_tfa / 1e3, 1),
    round(total_combined / 1e3, 1)
  )
)
saveRDS(sl_tab_scenario, paste0(slides_dir, "sl_tab_scenario.rds"))
message("SL-11 saved: sl_tab_scenario.rds")
```

## Objects for Presentation — Data Sources & Key Variables Section

```{r sl-fig-htn-control-rates, message=FALSE, warning=FALSE, echo=FALSE}
# ---- A1: fig_htn_control_rates — Observed (2017–2025) vs Projected (2026–2050) ----
# Source: htn_scenarios (loaded earlier from htn_control_scenarios_by_loc_year.csv)
# Highlights: China, Ecuador, Bangladesh
library(ggrepel)

if (!is.null(htn_scenarios)) {
  dt_htn_cr <- htn_scenarios[year <= 2050,
    .(location, year,
      control_rate = control_rate_program,
      observed     = control_rate_observed)]

  highlight_countries <- c("China", "Ecuador", "Bangladesh")

  # Build a period column: Historical (2017–2025) vs Projected (2026–2050)
  # For the line we need continuous connection, so we duplicate 2025 into both periods
  dt_hist <- dt_htn_cr[year <= 2025]
  dt_hist[, period := "Observed (2017-2025)"]
  dt_proj <- dt_htn_cr[year >= 2025]
  dt_proj[, period := "Projected (2026-2050)"]
  dt_all <- rbind(dt_hist, dt_proj)

  # Highlight vs background

  dt_all[, highlighted := location %in% highlight_countries]
  dt_all[, alpha_val := fifelse(highlighted, 1.0, 0.3)]
  dt_all[, size_val  := fifelse(highlighted, 0.9, 0.4)]

  # Labels at rightmost point for highlighted countries
  dt_labels <- dt_all[highlighted == TRUE & year == max(year),
                       .(location, year, control_rate, period)]
  dt_labels <- unique(dt_labels, by = "location")

  fig_htn_control_rates <- ggplot(dt_all,
         aes(x = year, y = control_rate * 100, group = interaction(location, period))) +
    # Background countries
    geom_line(data = dt_all[highlighted == FALSE],
              aes(color = period, linetype = period),
              linewidth = 0.35, alpha = 0.3) +
    geom_point(data = dt_all[highlighted == FALSE & observed == 1],
               aes(x = year, y = control_rate * 100),
               color = "grey60", size = 0.5, alpha = 0.3) +
    # Highlighted countries
    geom_line(data = dt_all[highlighted == TRUE],
              aes(color = period, linetype = period),
              linewidth = 0.9, alpha = 1.0) +
    geom_point(data = dt_all[highlighted == TRUE & observed == 1],
               aes(x = year, y = control_rate * 100),
               color = "#2171B5", size = 1.5, alpha = 0.9) +
    # Country labels for highlighted
    geom_label_repel(data = dt_labels,
                     aes(x = year, y = control_rate * 100, label = location),
                     size = 2.5, nudge_x = 1, segment.size = 0.3,
                     fontface = "bold", fill = "white", label.size = 0.2,
                     inherit.aes = FALSE) +
    scale_color_manual(
      name = NULL,
      values = c("Observed (2017-2025)" = "#2171B5",
                 "Projected (2026-2050)" = "#D4880F"),
      guide = guide_legend(override.aes = list(linewidth = 1, alpha = 1))
    ) +
    scale_linetype_manual(
      name = NULL,
      values = c("Observed (2017-2025)" = "solid",
                 "Projected (2026-2050)" = "dashed")
    ) +
    scale_x_continuous(breaks = seq(2017, 2050, 5),
                       limits = c(2017, 2054), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
    labs(
      title = "Observed and Projected Hypertension Program Control Rates",
      x = "Year", y = "Control Rate (%)",
      caption = "Dots = observed; lines = projected at 2024 level (2026-2050). China, Ecuador, and Bangladesh highlighted."
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          plot.caption = element_text(size = 7, hjust = 0))

  saveRDS(fig_htn_control_rates, paste0(slides_dir, "sl_fig_htn_control_rates.rds"))
  message("SL-A1 saved: sl_fig_htn_control_rates.rds")
} else {
  fig_htn_control_rates <- NULL
  message("htn_scenarios not available — skipping fig_htn_control_rates")
}
```

```{r sl-tbl-tfa-countries, message=FALSE, warning=FALSE, echo=FALSE}
# ---- A2: tbl_tfa_countries — TFA policy table, RTSL-engaged countries only ----
# Source: tfa_policy_scenarios_assessment.rds
# Engagement column: tfa_rtsl_engagement (values "Yes"/"No", from 04_define_interventions.R)
dt_tfa_scenarios <- tryCatch(
  as.data.table(readRDS(paste0(wd_data, "tfa_policy_scenarios_assessment.rds"))),
  error = function(e) NULL
)

if (!is.null(dt_tfa_scenarios)) {
  # Inspect actual column names (logged for verification)
  message("TFA scenario columns: ", paste(names(dt_tfa_scenarios), collapse = ", "))

  # CRITICAL FILTER: keep only RTSL-engaged countries (case-insensitive)
  dt_tfa_pol <- dt_tfa_scenarios[
    !is.na(tfa_bpp_date_passed) &
    tolower(tfa_rtsl_engagement) == "yes"
  ]

  # One row per country: earliest record (policy establishment)
  dt_tfa_pol <- dt_tfa_pol[, .(
    `Year Established` = as.integer(min(tfa_bpp_date_passed, na.rm = TRUE)),
    `Year in Effect`   = as.integer(min(tfa_bpp_date_in_effect, na.rm = TRUE)),
    `RTSL Engaged`     = "Yes"
  ), by = .(Country = location)]

  # Sort by year established ascending
  setorder(dt_tfa_pol, `Year Established`)

  tbl_tfa_countries <- dt_tfa_pol
  saveRDS(tbl_tfa_countries, paste0(slides_dir, "sl_tbl_tfa_countries.rds"))
  message("SL-A2 saved: sl_tbl_tfa_countries.rds")
} else {
  tbl_tfa_countries <- NULL
  message("tfa_policy_scenarios_assessment.rds not found — skipping tbl_tfa_countries")
}
```

```{r sl-tbl-htn-program-vs-country, message=FALSE, warning=FALSE, echo=FALSE}
# ---- A3: tbl_htn_program_vs_country — Program vs National Control Rate ----
# Source: htn_control_scenarios_by_loc_year.csv
# Columns used: location, year, control_rate_program, htncov2
if (!is.null(htn_scenarios)) {
  message("HTN scenario columns: ", paste(names(htn_scenarios), collapse = ", "))

  # Keep rows where both control_rate_program and htncov2 are available
  dt_htn_compare <- htn_scenarios[
    !is.na(control_rate_program) & !is.na(htncov2),
    .(location, year, htncov2, control_rate_program)
  ]

  # Use most recent available year per country
  dt_htn_compare[, max_year := max(year), by = location]
  dt_htn_compare <- dt_htn_compare[year == max_year]
  dt_htn_compare[, max_year := NULL]

  # Compute difference: positive = program outperforms, negative = underperforms
  dt_htn_compare[, program_vs_country_diff := control_rate_program - htncov2]

  # Sort by difference ascending (worst performers first)
  setorder(dt_htn_compare, program_vs_country_diff)

  # Format for display
  dt_htn_compare[, `:=`(
    `National Control Rate` = paste0(round(htncov2 * 100, 1), "%"),
    `Program Control Rate`  = paste0(round(control_rate_program * 100, 1), "%"),
    `Difference (pp)`       = round(program_vs_country_diff * 100, 1)
  )]

  # Identify underperforming rows for highlighting
  underperforming_rows <- which(dt_htn_compare$program_vs_country_diff < 0)

  tbl_htn_program_vs_country <- dt_htn_compare[, .(
    Country = location,
    `National Control Rate`,
    `Program Control Rate`,
    `Difference (pp)`
  )]

  # Store the row indices for highlighting in presentation
  attr(tbl_htn_program_vs_country, "underperforming_rows") <- underperforming_rows

  saveRDS(tbl_htn_program_vs_country, paste0(slides_dir, "sl_tbl_htn_program_vs_country.rds"))
  message("SL-A3 saved: sl_tbl_htn_program_vs_country.rds")
} else {
  tbl_htn_program_vs_country <- NULL
  message("htn_scenarios not available — skipping tbl_htn_program_vs_country")
}
```

```{r sl-tbl-countries-by-region, message=FALSE, warning=FALSE, echo=FALSE}
# ---- A5: tbl_countries_by_region — Formatted table for presentation ----
# Source: countries_by_region (computed earlier in country-list chunk)
# Columns: Region, Countries (comma-separated string)
if (exists("countries_by_region") && nrow(countries_by_region) > 0) {
  message("countries_by_region columns: ", paste(names(countries_by_region), collapse = ", "))

  tbl_countries_by_region <- countries_by_region
  saveRDS(tbl_countries_by_region, paste0(slides_dir, "sl_tbl_countries_by_region.rds"))
  message("SL-A5 saved: sl_tbl_countries_by_region.rds")
} else {
  tbl_countries_by_region <- NULL
  message("countries_by_region not available — skipping tbl_countries_by_region")
}
```

## Objects for Presentation — Results Section

```{r sl-tbl-total-lives-saved, message=FALSE, warning=FALSE, echo=FALSE}
# ---- B1: tbl_total_lives_saved — Absolute total deaths averted by country ----
# Source: dt_excel_per1m (computed above)
# IMPORTANT: HTN uses program_population denominator; TFA uses total country population
# n_deaths_averted is already computed as absolute totals in dt_excel_per1m
if (exists("dt_excel_per1m") && nrow(dt_excel_per1m) > 0) {
  message("dt_excel_per1m columns: ", paste(names(dt_excel_per1m), collapse = ", "))

  # Summarize cumulative absolute deaths averted per country x scenario
  dt_lives <- dt_excel_per1m[
    scenario %in% c("HTN Control (Program)", "TFA Elimination") & year >= 2017,
    .(total_averted = sum(n_deaths_averted, na.rm = TRUE)),
    by = .(location, scenario)
  ]

  dt_lives_wide <- dcast(dt_lives, location ~ scenario,
                          value.var = "total_averted", fill = 0)
  # Rename columns for display
  old_names <- names(dt_lives_wide)
  setnames(dt_lives_wide,
           old = old_names,
           new = c("Country", "HTN Deaths Averted", "TFA Deaths Averted"))

  # Do NOT show combined column — populations differ
  # Sort by HTN deaths averted descending
  setorder(dt_lives_wide, -`HTN Deaths Averted`)

  # Format with comma separators
  dt_lives_wide[, `HTN Deaths Averted` := formatC(round(`HTN Deaths Averted`),
                                                    format = "d", big.mark = ",")]
  dt_lives_wide[, `TFA Deaths Averted` := formatC(round(`TFA Deaths Averted`),
                                                    format = "d", big.mark = ",")]

  tbl_total_lives_saved <- dt_lives_wide
  saveRDS(tbl_total_lives_saved, paste0(slides_dir, "sl_tbl_total_lives_saved.rds"))
  message("SL-B1 saved: sl_tbl_total_lives_saved.rds")
} else {
  tbl_total_lives_saved <- NULL
  message("dt_excel_per1m not available — skipping tbl_total_lives_saved")
}
```

```{r sl-tbl-grand-total-lives-saved, message=FALSE, warning=FALSE, echo=FALSE}
# ---- B4: tbl_grand_total_lives_saved — Grand total across all countries ----
# Reuses dt_lives (intermediate from B1 above) — do not re-derive
if (exists("dt_lives") && nrow(dt_lives) > 0) {
  # Sum across all countries per intervention
  dt_grand <- dt_lives[, .(total_averted = sum(total_averted, na.rm = TRUE)),
                        by = scenario]

  # Build 3-row summary: HTN, TFA, Combined
  grand_htn <- dt_grand[scenario == "HTN Control (Program)", total_averted]
  grand_tfa <- dt_grand[scenario == "TFA Elimination", total_averted]
  grand_combined <- grand_htn + grand_tfa

  tbl_grand_total_lives_saved <- data.table(
    Intervention = c("Hypertension Control", "TFA Elimination",
                     "Combined (All Interventions)"),
    `Deaths Averted` = c(
      formatC(round(grand_htn), format = "d", big.mark = ","),
      formatC(round(grand_tfa), format = "d", big.mark = ","),
      formatC(round(grand_combined), format = "d", big.mark = ",")
    ),
    `Time Horizon` = rep(paste0(year_min, "-", year_max), 3)
  )

  # Store raw values for inline R expressions in presentation
  attr(tbl_grand_total_lives_saved, "grand_htn") <- grand_htn
  attr(tbl_grand_total_lives_saved, "grand_tfa") <- grand_tfa
  attr(tbl_grand_total_lives_saved, "grand_combined") <- grand_combined

  saveRDS(tbl_grand_total_lives_saved, paste0(slides_dir, "sl_tbl_grand_total_lives_saved.rds"))
  message("SL-B4 saved: sl_tbl_grand_total_lives_saved.rds")
} else {
  tbl_grand_total_lives_saved <- NULL
  message("dt_lives not available — skipping tbl_grand_total_lives_saved")
}
```

```{r sl-fig-global-cumulative, message=FALSE, warning=FALSE, echo=FALSE}
# ---- B5: fig_global_cumulative — Stacked area: cumulative deaths averted, all countries ----
# Source: dt_excel_per1m, aggregated across ALL countries, time preserved
# HTN uses n_deaths_averted (program pop); TFA uses n_deaths_averted (total pop)
# n_deaths_averted already applies the correct population denominator per intervention
if (exists("dt_excel_per1m") && nrow(dt_excel_per1m) > 0) {
  # Aggregate absolute deaths averted across all countries by year and scenario
  dt_global_ts <- dt_excel_per1m[
    scenario %in% c("HTN Control (Program)", "TFA Elimination") & year >= 2017,
    .(n_deaths_averted = sum(n_deaths_averted, na.rm = TRUE)),
    by = .(year, scenario)
  ]

  # Compute cumulative over time by scenario
  setorder(dt_global_ts, scenario, year)
  dt_global_ts[, cum_averted := cumsum(n_deaths_averted), by = scenario]

  # Display labels
  dt_global_ts[, intervention := fifelse(
    scenario == "HTN Control (Program)",
    "Hypertension Control", "TFA Elimination"
  )]

  # Grand total at final year for annotation
  max_year_global <- max(dt_global_ts$year)
  grand_total_global <- dt_global_ts[year == max_year_global, sum(cum_averted, na.rm = TRUE)]
  # Per-intervention totals at final year for inline use
  global_cum_htn <- dt_global_ts[year == max_year_global & scenario == "HTN Control (Program)", cum_averted]
  global_cum_tfa <- dt_global_ts[year == max_year_global & scenario == "TFA Elimination", cum_averted]

  fig_global_cumulative <- ggplot(dt_global_ts,
         aes(x = year, y = cum_averted, fill = intervention)) +
    geom_area(alpha = 0.8, position = "stack") +
    scale_fill_manual(
      values = c("Hypertension Control" = "#2171B5",
                 "TFA Elimination"      = "#E31A1C"),
      name = "Intervention"
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_x_continuous(breaks = seq(2017, 2050, 5)) +
    annotate("text", x = max_year_global, y = grand_total_global,
             label = paste0("Total: ", formatC(round(grand_total_global),
                            format = "d", big.mark = ",")),
             hjust = 1, vjust = 0.5, size = 3.5, fontface = "bold") +
    labs(
      title = "Cumulative Deaths Averted Over Time — All Countries",
      x = "Year", y = "Cumulative Deaths Averted",
      caption = "Cumulative deaths averted by intervention, all RTSL-engaged countries, 2024-2050.\nHTN reflects program population; TFA reflects national population."
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          plot.caption = element_text(size = 7, hjust = 0))

  # Store per-intervention totals for inline use in presentation
  attr(fig_global_cumulative, "global_cum_htn") <- global_cum_htn
  attr(fig_global_cumulative, "global_cum_tfa") <- global_cum_tfa
  attr(fig_global_cumulative, "grand_total") <- grand_total_global

  saveRDS(fig_global_cumulative, paste0(slides_dir, "sl_fig_global_cumulative.rds"))
  message("SL-B5 saved: sl_fig_global_cumulative.rds")
} else {
  fig_global_cumulative <- NULL
  message("dt_excel_per1m not available — skipping fig_global_cumulative")
}
```

```{r sl-tbl-bangladesh-summary, message=FALSE, warning=FALSE, echo=FALSE}
# ---- B2: tbl_bangladesh_summary — Bangladesh cumulative impact summary ----
# Source: dt_excel_per1m, filtered to Bangladesh
# One row per intervention (HTN, TFA)
if (exists("dt_excel_per1m") && nrow(dt_excel_per1m) > 0) {
  dt_bd <- dt_excel_per1m[
    location == "Bangladesh" &
    scenario %in% c("HTN Control (Program)", "TFA Elimination") &
    year >= 2017
  ]

  dt_bd_summary <- dt_bd[, .(
    deaths_averted_per_1M = sum(deaths_averted_per_1M, na.rm = TRUE),
    total_deaths_averted  = sum(n_deaths_averted, na.rm = TRUE),
    year_min = min(year),
    year_max = max(year)
  ), by = scenario]

  dt_bd_summary[, Intervention := fifelse(
    scenario == "HTN Control (Program)",
    "Hypertension Control", "TFA Elimination"
  )]
  dt_bd_summary[, `Deaths Averted per 1M` := formatC(
    round(deaths_averted_per_1M), format = "d", big.mark = ",")]
  dt_bd_summary[, `Total Deaths Averted` := formatC(
    round(total_deaths_averted), format = "d", big.mark = ",")]
  dt_bd_summary[, `Time Horizon` := paste0(year_min, "-", year_max)]

  tbl_bangladesh_summary <- dt_bd_summary[, .(
    Intervention, `Deaths Averted per 1M`, `Total Deaths Averted`, `Time Horizon`
  )]

  # Store raw total for inline use in presentation
  total_bd_averted <- sum(dt_bd_summary$total_deaths_averted, na.rm = TRUE)
  attr(tbl_bangladesh_summary, "total_bd_averted") <- total_bd_averted

  saveRDS(tbl_bangladesh_summary, paste0(slides_dir, "sl_tbl_bangladesh_summary.rds"))
  message("SL-B2 saved: sl_tbl_bangladesh_summary.rds")
} else {
  tbl_bangladesh_summary <- NULL
  message("dt_excel_per1m not available — skipping tbl_bangladesh_summary")
}
```

```{r sl-fig-bangladesh-cumulative, message=FALSE, warning=FALSE, echo=FALSE}
# ---- B3: fig_bangladesh_cumulative — Stacked area: cumulative deaths averted ----
# Source: dt_excel_per1m, Bangladesh, time dimension preserved
# y-axis: ABSOLUTE cumulative deaths averted (not per 1M)
if (exists("dt_excel_per1m") && nrow(dt_excel_per1m) > 0) {
  dt_bd_ts <- dt_excel_per1m[
    location == "Bangladesh" &
    scenario %in% c("HTN Control (Program)", "TFA Elimination") &
    year >= 2017,
    .(year, scenario, n_deaths_averted)
  ]

  # Compute cumulative over time by scenario
  setorder(dt_bd_ts, scenario, year)
  dt_bd_ts[, cum_averted := cumsum(n_deaths_averted), by = scenario]

  # Intervention labels for display
  dt_bd_ts[, intervention := fifelse(
    scenario == "HTN Control (Program)",
    "Hypertension Control", "TFA Elimination"
  )]

  # Total at final year for annotation
  total_bd <- dt_bd_ts[year == max(year), sum(cum_averted, na.rm = TRUE)]
  max_year <- max(dt_bd_ts$year)

  fig_bangladesh_cumulative <- ggplot(dt_bd_ts,
         aes(x = year, y = cum_averted, fill = intervention)) +
    geom_area(alpha = 0.8, position = "stack") +
    scale_fill_manual(
      values = c("Hypertension Control" = "#2171B5",
                 "TFA Elimination"      = "#E31A1C"),
      name = "Intervention"
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_x_continuous(breaks = seq(2017, 2050, 5)) +
    annotate("text", x = max_year, y = total_bd * 0.95,
             label = paste0("Total: ", formatC(round(total_bd),
                            format = "d", big.mark = ",")),
             hjust = 1, vjust = 0, size = 3.5, fontface = "bold") +
    labs(
      title = "Bangladesh: Cumulative Deaths Averted Over Time",
      x = "Year", y = "Cumulative Deaths Averted",
      caption = "Cumulative deaths averted by intervention type, Bangladesh 2024-2050.\nHTN figures reflect program population; TFA figures reflect national population."
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          plot.caption = element_text(size = 7, hjust = 0))

  # Store total for inline use in presentation
  attr(fig_bangladesh_cumulative, "total_bd") <- total_bd

  saveRDS(fig_bangladesh_cumulative, paste0(slides_dir, "sl_fig_bangladesh_cumulative.rds"))
  message("SL-B3 saved: sl_fig_bangladesh_cumulative.rds")
} else {
  fig_bangladesh_cumulative <- NULL
  message("dt_excel_per1m not available — skipping fig_bangladesh_cumulative")
}
```

```{r sl-per1m-result-figures, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Separate HTN and TFA result figures from dt_excel_per1m ----
# dt_excel_per1m columns (verified above):
#   location, year, scenario, deaths, deaths_averted,
#   cf_deaths_ref, population, program_population,
#   deaths_per_1M, deaths_averted_per_1M, n_deaths_averted, pct_averted

# Cumulative deaths averted per 1M by country and scenario
dt_per1m_cum <- dt_excel_per1m[
  scenario != "Counterfactual" & year >= 2017,
  .(deaths_averted_per_1M = sum(deaths_averted_per_1M, na.rm = TRUE)),
  by = .(location, scenario)
]

# --- fig_htn_per1m: HTN results only ---
dt_htn_per1m <- dt_per1m_cum[scenario == "HTN Control (Program)"]
setorder(dt_htn_per1m, -deaths_averted_per_1M)
dt_htn_per1m[, location := factor(location, levels = rev(location))]

fig_htn_per1m <- ggplot(dt_htn_per1m,
       aes(x = deaths_averted_per_1M, y = location)) +
  geom_col(fill = "#2171B5", width = 0.7) +
  scale_x_continuous(labels = comma, expand = c(0, 0)) +
  labs(x = "Cumulative Deaths Averted per 1M (2017-2050)", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 6))
saveRDS(fig_htn_per1m, paste0(slides_dir, "sl_fig_htn_per1m.rds"))

# --- fig_tfa_per1m: TFA results only ---
dt_tfa_per1m <- dt_per1m_cum[scenario == "TFA Elimination"]
setorder(dt_tfa_per1m, -deaths_averted_per_1M)
dt_tfa_per1m[, location := factor(location, levels = rev(location))]

fig_tfa_per1m <- ggplot(dt_tfa_per1m,
       aes(x = deaths_averted_per_1M, y = location)) +
  geom_col(fill = "#E31A1C", width = 0.7) +
  scale_x_continuous(labels = comma, expand = c(0, 0)) +
  labs(x = "Cumulative Deaths Averted per 1M (2017-2050)", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 6))
saveRDS(fig_tfa_per1m, paste0(slides_dir, "sl_fig_tfa_per1m.rds"))

message("SL-NEW saved: sl_fig_htn_per1m.rds, sl_fig_tfa_per1m.rds")
```

```{r sl-fig-combined-per1m, message=FALSE, warning=FALSE, echo=FALSE}
# ---- fig_combined_per1m: Both interventions side by side per 1M ----
dt_combined <- dt_per1m_cum[scenario %in% c("HTN Control (Program)", "TFA Elimination")]
dt_combined[, intervention := scenario]

# Order countries by total combined averted (descending)
dt_country_order <- dt_combined[, .(total = sum(deaths_averted_per_1M)), by = location]
setorder(dt_country_order, total)
dt_combined[, location := factor(location, levels = dt_country_order$location)]

fig_combined_per1m <- ggplot(dt_combined,
       aes(x = deaths_averted_per_1M, y = location, fill = intervention)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(
    values = c("HTN Control (Program)" = "#2171B5",
               "TFA Elimination"       = "#E31A1C"),
    name = "Intervention"
  ) +
  scale_x_continuous(labels = comma, expand = c(0, 0)) +
  labs(x = "Cumulative Deaths Averted per 1 Million (2017-2050)", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 6))
saveRDS(fig_combined_per1m, paste0(slides_dir, "sl_fig_combined_per1m.rds"))
message("SL-NEW saved: sl_fig_combined_per1m.rds")
```

```{r sl-fig-tfa-timeline, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
# ---- TFA Ban Implementation Timeline (bar + line, dual y-axes) ----
# Uses dt_tfa_scenarios loaded earlier in sl-tbl-tfa-countries chunk
# and population data from PopulationsSingleAge0050.rds

sl_fig_tfa_timeline <- NULL

if (exists("dt_tfa_scenarios") && !is.null(dt_tfa_scenarios)) {

  # One row per country: earliest date_passed, plus tfa_rtsl_include flag
  dt_tfa_timeline <- dt_tfa_scenarios[
    !is.na(tfa_bpp_date_passed),
    .(year_passed    = as.integer(min(tfa_bpp_date_passed, na.rm = TRUE)),
      tfa_rtsl_include = first(tfa_rtsl_include)),
    by = .(location)
  ]

  # Load population for cumulative coverage line
  wpp_tfa <- tryCatch(
    as.data.table(readRDS(paste0(wd_data, "PopulationsSingleAge0050.rds"))),
    error = function(e) NULL
  )

  if (!is.null(wpp_tfa)) {
    pop_total <- wpp_tfa[year_id == 2024, .(pop = sum(Nx, na.rm = TRUE)), by = location]
    dt_tfa_timeline <- merge(dt_tfa_timeline, pop_total, by = "location", all.x = TRUE)
    dt_tfa_timeline[is.na(pop), pop := 0]
  } else {
    dt_tfa_timeline[, pop := 0]
  }

  # Build cumulative counts by year
  year_range <- seq(min(dt_tfa_timeline$year_passed), max(dt_tfa_timeline$year_passed))

  dt_cumul_tfa <- rbindlist(lapply(year_range, function(yr) {
    passed <- dt_tfa_timeline[year_passed <= yr]
    data.table(
      year = yr,
      n_rtsl_include_yes = nrow(passed[tolower(tfa_rtsl_include) == "yes"]),
      n_rtsl_include_no  = nrow(passed[tolower(tfa_rtsl_include) == "no"]),
      cum_pop            = sum(passed$pop, na.rm = TRUE)
    )
  }))

  dt_cumul_tfa[, n_total := n_rtsl_include_yes + n_rtsl_include_no]
  dt_cumul_tfa[, cum_pop_b := cum_pop / 1e9]

  # Melt for stacked bars
  dt_bar <- melt(dt_cumul_tfa, id.vars = c("year", "n_total", "cum_pop_b"),
                 measure.vars = c("n_rtsl_include_yes", "n_rtsl_include_no"),
                 variable.name = "category", value.name = "n_countries")
  dt_bar[, category := fifelse(category == "n_rtsl_include_yes",
                                "RTSL Assessment",
                                "Other Countries")]
  dt_bar[, category := factor(category, levels = c("Other Countries", "RTSL Assessment"))]

  # Scaling factor for dual y-axis: map population to country-count scale
  max_countries <- max(dt_cumul_tfa$n_total)
  max_pop_b     <- max(dt_cumul_tfa$cum_pop_b)
  scale_factor  <- max_countries / max_pop_b

  # After line 2632, add:
  dt_cumul_tfa[, cum_pop_scaled := cum_pop_b * scale_factor]

  scale_factor <- max_countries / max_pop_b

# Build sec_axis using a formula with the literal value substituted in
sec_axis_formula <- substitute(~ . / sf, list(sf = scale_factor))

  sl_fig_tfa_timeline <- ggplot() +
    # Stacked bars: cumulative countries
    geom_col(data = dt_bar,
             aes(x = year, y = n_countries, fill = category),
             width = 0.7) +
    # Line: cumulative population (rescaled to country-count axis)
    geom_line(data = dt_cumul_tfa,
              aes(x = year, y = cum_pop_scaled),
              color = "#2c3e50", linewidth = 1.2) +
    geom_point(data = dt_cumul_tfa,
               aes(x = year, y = cum_pop_scaled),
               color = "#2c3e50", size = 2.5) +
    # Data labels: cumulative country count on top of bars
    geom_text(data = dt_cumul_tfa,
              aes(x = year, y = n_total, label = n_total),
              vjust = -0.5, size = 3.5, fontface = "bold") +
    # WHO REPLACE annotation
    geom_vline(xintercept = 2018, linetype = "dashed", color = "#c0392b",
               linewidth = 0.7) +
    annotate("text", x = 2018.15, y = max_countries * 0.92,
             label = "WHO REPLACE\nInitiative (2018)",
             hjust = 0, size = 3.2, color = "#c0392b", fontface = "italic") +
    # Scales
    scale_fill_manual(
      values = c("RTSL Assessment"  = "#1f78b4",
                 "Other Countries"   = "#a6cee3"),
      name = NULL
    ) +
    scale_x_continuous(breaks = year_range,
                       expand = expansion(mult = c(0.02, 0.05))) +
    scale_y_continuous(
      name   = "Cumulative Number of Countries",
      expand = expansion(mult = c(0, 0.15)),
      sec.axis = sec_axis(
  trans = as.formula(sec_axis_formula),
  name  = "Cumulative Population Covered (billions)"
)
    ) +
    labs(
      title   = "Trans Fat Elimination: Global Best-Practice Policy Adoption",
      subtitle = "Cumulative countries with TFA bans and population covered over time",
      x       = "Year of Policy Passage",
      caption = "Population: UNWPP 2024 total population. Line = cumulative population covered."
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey40", size = 11),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position  = "top",
      legend.justification = "left",
      axis.title.y.right = element_text(color = "#2c3e50"),
      axis.text.y.right  = element_text(color = "#2c3e50"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

  print(sl_fig_tfa_timeline)
  saveRDS(sl_fig_tfa_timeline, paste0(slides_dir, "sl_fig_tfa_timeline.rds"))
  ggsave(paste0(wd_outp, "assessment_tfa_timeline.png"), plot = sl_fig_tfa_timeline,
         width = 12, height = 6, dpi = 300)
  message("SL-NEW saved: sl_fig_tfa_timeline.rds")
} else {
  message("dt_tfa_scenarios not available — skipping TFA timeline plot")
}
```

```{r sl-fig-country-map, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
# ---- World Map: Country Coverage (HTN Program, TFA Ban, Both) ----
# HTN program countries: from data_program (unique locations)
# TFA ban countries: from dt_tfa_scenarios (countries with tfa_bpp_date_passed)
# Engagement: tfa_rtsl_engagement

sl_fig_country_map <- NULL

program_countries <- unique(data_program[intervention %in% int_order, location])

dt_htn_control_scenarios <- fread(paste0(wd_data, "htn_control_scenarios_by_loc_year.csv"))
htn_countries <- unique(dt_htn_control_scenarios$location)


if (exists("dt_tfa_scenarios") && !is.null(dt_tfa_scenarios)) {

  # TFA countries with a ban (date_passed is not NA) that are in RTSL assessment
  dt_tfa_map <- dt_tfa_scenarios[
    !is.na(tfa_bpp_date_passed),
    .(tfa_rtsl_engagement = first(tfa_rtsl_engagement)),
    by = .(location)
  ]
  tfa_countries <- dt_tfa_map$location

  # Build classification
  dt_class <- data.table(
    location = union(htn_countries, tfa_countries)
  )
  dt_class[, has_htn := location %in% htn_countries]
  dt_class[, has_tfa := location %in% tfa_countries]

  # Merge engagement info
  dt_class <- merge(dt_class, dt_tfa_map, by = "location", all.x = TRUE)

  dt_class[, coverage := fcase(
    has_htn & has_tfa, "HTN Program + TFA Ban",
    has_htn & !has_tfa, "HTN Program Only",
    !has_htn & has_tfa & tolower(tfa_rtsl_engagement) == "yes",
      "TFA Ban Only",
    default = NA_character_
  )]

  dt_class[is.na(coverage) & has_tfa & !is.na(tfa_rtsl_engagement), coverage := "TFA Ban Only"]
  
  # Remove countries that don't match any display category
  dt_class <- dt_class[!is.na(coverage)]
  
  # remove countries not counted in TF not HTN program, eg USA
  dt_class <- dt_class[location %in% program_countries,]

  # Merge with ISO3 codes for map join
  dt_class_iso <- merge(dt_class, countries_dt[, .(location, iso3)],
                         by = "location", all.x = TRUE)

  # Load world map
  world <- ne_countries(scale = "medium", returnclass = "sf")

  world_map <- merge(world, dt_class_iso[, .(iso3, coverage)],
                     by.x = "adm0_a3", by.y = "iso3", all.x = TRUE)

  # Color palette
  coverage_cols <- c(
    "HTN Program + TFA Ban"         = "#6a3d9a",
    "HTN Program Only"              = "#1f78b4",
    "TFA Ban Only"   = "#e31a1c"
  )

  sl_fig_country_map <- ggplot(world_map) +
    geom_sf(aes(fill = coverage), color = "grey70", linewidth = 0.15) +
    scale_fill_manual(
      values   = coverage_cols,
      na.value = "grey90",
      name     = "Intervention Coverage",
      na.translate = FALSE
    ) +
    coord_sf(crs = "+proj=robin") +
    labs(
      title    = "Global Coverage: HTN and TFA Interventions",
      subtitle = "Countries with RTSL-associated hypertension programs and/or TFA best-practice policies",
      caption  = "Source: RTSL program data and WHO TFA policy tracker"
    ) +
    theme_void(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey40", size = 11),
      legend.position = "bottom",
      legend.title  = element_text(face = "bold"),
      legend.key.size = unit(0.5, "cm"),
      plot.margin   = margin(10, 10, 10, 10)
    ) +
    guides(fill = guide_legend(nrow = 1))

  print(sl_fig_country_map)
  saveRDS(sl_fig_country_map, paste0(slides_dir, "sl_fig_country_map.rds"))
  ggsave(paste0(wd_outp, "assessment_country_map.png"), plot = sl_fig_country_map,
         width = 12, height = 6, dpi = 300)
  message("SL-NEW saved: sl_fig_country_map.rds")
} else {
  message("dt_tfa_scenarios not available — skipping country map")
}
```

<!-- ======================================================================
     PARTNERS & DONORS PRESENTATION ARTEFACTS
     All computations for presentation_cvh_partners_donors.Rmd live here.
     - Combined (HTN + TFA) results only (no disaggregation in outputs)
     - HTN and TFA have different target populations so they are filtered
       separately then summed (do NOT use a pre-combined scenario row).
     - PAHO merge: all Latin America & Caribbean countries combined into a
       single "PAHO (Latin America & Caribbean)" bar on the Pareto chart.
     - Problem-framing scalars for the new opening slides (sourced from GBD
       2023, WHO Hypertension 2025 report, WHO REPLACE, Lancet 2025,
       and resolvetosavelives.org program pages).
     - Separate Excel export: presentation_cvh_partners_donors.xlsx
     ====================================================================== -->

```{r pd-combined-country, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Per-country combined deaths averted (HTN + TFA summed by country) ------
# HTN and TFA are filtered separately (different target populations) then
# summed at the country level to produce the combined total used throughout
# the donors/partners presentation.
dt_pd_htn <- dt_excel_per1m[
  scenario == "HTN Control (Program)" & year >= 2017,
  .(htn_averted = sum(n_deaths_averted, na.rm = TRUE)),
  by = location
]
dt_pd_tfa <- dt_excel_per1m[
  scenario == "TFA Elimination" & year >= 2017,
  .(tfa_averted = sum(n_deaths_averted, na.rm = TRUE)),
  by = location
]
dt_pd_country <- merge(dt_pd_htn, dt_pd_tfa, by = "location", all = TRUE)
dt_pd_country[is.na(htn_averted), htn_averted := 0]
dt_pd_country[is.na(tfa_averted), tfa_averted := 0]
dt_pd_country[, combined_averted := htn_averted + tfa_averted]

# Attach region metadata
dt_pd_country <- merge(dt_pd_country, countries_dt[, .(location, region)],
                       by = "location", all.x = TRUE)
setorder(dt_pd_country, -combined_averted)

saveRDS(dt_pd_country, paste0(slides_dir, "sl_pd_country_combined.rds"))
message("PD saved: sl_pd_country_combined.rds")
```

```{r pd-annual-combined, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Global annual combined deaths averted (HTN + TFA, sum over countries) --
dt_pd_annual <- dt_excel_per1m[
  scenario %in% c("HTN Control (Program)", "TFA Elimination") & year >= 2017,
  .(deaths_averted = sum(n_deaths_averted, na.rm = TRUE)),
  by = year
]
setorder(dt_pd_annual, year)
dt_pd_annual[, cum_averted := cumsum(deaths_averted)]

# Cumulative-deaths-averted chart used in place of the HTN-vs-TFA stacked area
fig_pd_global_cumulative <- ggplot(dt_pd_annual,
       aes(x = year, y = cum_averted)) +
  geom_area(alpha = 0.85, fill = "#4B2E83") +
  geom_line(color = "#4B2E83", linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2017, 2050, 5)) +
  annotate("text",
           x = max(dt_pd_annual$year),
           y = max(dt_pd_annual$cum_averted),
           label = paste0("Total: ",
                          formatC(round(max(dt_pd_annual$cum_averted)),
                                  format = "d", big.mark = ",")),
           hjust = 1, vjust = 1.3, size = 4.5, fontface = "bold") +
  labs(
    title = "Cumulative Deaths Averted: RTSL CVH Programs",
    x = "Year",
    y = "Cumulative deaths averted",
    caption = "Combined hypertension control and trans-fat elimination across RTSL-engaged countries (2017-2050)."
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.caption = element_text(size = 8, hjust = 0))

saveRDS(fig_pd_global_cumulative,
        paste0(slides_dir, "sl_pd_fig_global_cumulative.rds"))
saveRDS(dt_pd_annual, paste0(slides_dir, "sl_pd_annual_combined.rds"))
message("PD saved: sl_pd_fig_global_cumulative.rds, sl_pd_annual_combined.rds")
```

```{r pd-pareto-combined-paho, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
# ---- Combined Pareto chart with Latin America & Caribbean merged to "PAHO" --
# Take per-country combined totals, fold all LAC countries into a single
# pseudo-country, then rank and keep Top 20 + "Rest of countries".
dt_pareto_paho <- copy(dt_pd_country)

# Merge all Latin America & Caribbean countries into a single PAHO row
lac_mask <- dt_pareto_paho$region == "Latin America and Caribbean" &
              !is.na(dt_pareto_paho$region)
paho_total <- dt_pareto_paho[lac_mask,
  .(location = "PAHO (Latin America & Caribbean)",
    htn_averted      = sum(htn_averted,      na.rm = TRUE),
    tfa_averted      = sum(tfa_averted,      na.rm = TRUE),
    combined_averted = sum(combined_averted, na.rm = TRUE),
    region           = "Latin America and Caribbean")]
dt_pareto_paho <- rbind(dt_pareto_paho[!lac_mask], paho_total, fill = TRUE)

# Rank by combined deaths averted
setorder(dt_pareto_paho, -combined_averted)

top20 <- head(dt_pareto_paho, 20)
rest <- dt_pareto_paho[!location %in% top20$location,
  .(location = "Rest of countries",
    htn_averted      = sum(htn_averted,      na.rm = TRUE),
    tfa_averted      = sum(tfa_averted,      na.rm = TRUE),
    combined_averted = sum(combined_averted, na.rm = TRUE),
    region           = NA_character_)]

dt_pareto_combined_paho <- rbind(top20, rest, fill = TRUE)
setorder(dt_pareto_combined_paho, -combined_averted)
dt_pareto_combined_paho[, deaths_averted := combined_averted]
dt_pareto_combined_paho[, cum_deaths := cumsum(deaths_averted)]
dt_pareto_combined_paho[, cum_pct := cum_deaths / sum(deaths_averted)]
dt_pareto_combined_paho[, location := factor(location, levels = location)]

y_max_p <- max(dt_pareto_combined_paho$deaths_averted, na.rm = TRUE)
dt_pareto_combined_paho[, y_max_p := y_max_p]
dt_pareto_combined_paho[, cum_scaled := cum_pct * y_max_p]

# Build sec_axis formula with literal value so the plot is self-contained when saved as RDS
sec_axis_pareto <- substitute(~ . / ymp, list(ymp = y_max_p))

p_pd_pareto_combined <- ggplot(dt_pareto_combined_paho, aes(x = location)) +
  geom_col(aes(y = deaths_averted), fill = "#4B2E83") +
  geom_line(aes(y = cum_scaled, group = 1),
            color = "#85754D", linewidth = 1) +
  geom_point(aes(y = cum_scaled),
             color = "#85754D", size = 2) +
  scale_y_continuous(
    labels = scales::comma,
    name   = "Deaths Averted (2017-2050)",
    sec.axis = sec_axis(
      trans = as.formula(sec_axis_pareto),
      name = "Cumulative percentage",
      labels = scales::percent_format(accuracy = 1)
    )
  ) +
  labs(
    x = NULL,
    title = "Combined Interventions: Top Countries by Deaths Averted",
    subtitle = "Latin America & Caribbean countries aggregated as PAHO"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))

saveRDS(p_pd_pareto_combined,
        paste0(slides_dir, "sl_pd_pareto_combined.rds"))
saveRDS(dt_pareto_combined_paho,
        paste0(slides_dir, "sl_pd_pareto_combined_data.rds"))
message("PD saved: sl_pd_pareto_combined.rds (PAHO-merged Pareto)")
```

```{r pd-tbl-region-combined, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Regional combined deaths averted for the donors/partners presentation --
sl_pd_tab_region <- dt_pd_country[!is.na(region),
  .(deaths_averted = sum(combined_averted, na.rm = TRUE),
    n_countries    = .N),
  by = region]
setorder(sl_pd_tab_region, -deaths_averted)
sl_pd_tab_region[, `Deaths Averted` := formatC(round(deaths_averted),
                                                format = "d", big.mark = ",")]
sl_pd_tab_region[, Countries := n_countries]
setnames(sl_pd_tab_region, "region", "Region")

saveRDS(sl_pd_tab_region[, .(Region, Countries, `Deaths Averted`)],
        paste0(slides_dir, "sl_pd_tab_region.rds"))
message("PD saved: sl_pd_tab_region.rds")
```

```{r pd-tbl-top-countries, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Top 10 countries (combined, PAHO-merged) -------------------------------
sl_pd_tab_top <- head(dt_pareto_combined_paho[location != "Rest of countries"], 10)
sl_pd_tab_top[, `Deaths Averted` := formatC(round(deaths_averted),
                                              format = "d", big.mark = ",")]
sl_pd_tab_top[, `Share of Total` := scales::percent(cum_pct - shift(cum_pct, fill = 0),
                                                     accuracy = 0.1)]
saveRDS(sl_pd_tab_top[, .(Country = as.character(location), `Deaths Averted`,
                          `Share of Total`)],
        paste0(slides_dir, "sl_pd_tab_top_countries.rds"))
message("PD saved: sl_pd_tab_top_countries.rds")
```

```{r pd-combined-scalars, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Combined totals and framing scalars for inline R refs ------------------
grand_combined_pd <- sum(dt_pd_country$combined_averted, na.rm = TRUE)

sl_pd_scalars <- list(
  # Model outputs - combined totals
  n_countries           = n_countries,
  n_htn_countries       = n_htn_countries,
  n_tfa_countries       = n_tfa_countries,
  study_years           = study_years,
  year_min              = year_min,
  year_max              = year_max,
  grand_combined        = grand_combined_pd,
  grand_combined_fmt    = formatC(round(grand_combined_pd), format = "d",
                                   big.mark = ","),
  grand_combined_m      = formatC(round(grand_combined_pd / 1e6, 1),
                                   format = "f", digits = 1, big.mark = ","),
  daily_averted_fmt     = formatC(round(grand_combined_pd / ((year_max - 2017 + 1) * 365)),
                                   format = "d", big.mark = ","),
  n_years               = year_max - year_min + 1,

  # --- Problem-framing scalars (external public sources) ---
  # Source: GBD 2023 (IHME). CVD remains the leading cause of death globally.
  cvd_deaths_global     = "19.2 million",     # annual CVD deaths, GBD 2023
  cvd_share_noncomm     = "approximately one-third",  # of all global deaths
  # Source: WHO Global Report on Hypertension (2023/2025 update); Lancet 2025
  htn_adults_global     = "1.3 billion",       # adults living with hypertension
  htn_uncontrolled_pct  = "over 75%",          # of adults with HTN not controlled
  htn_undiagnosed_share = "nearly half",        # of adults with HTN undiagnosed
  htn_deaths_annual     = "10.8 million",      # deaths attributable to high SBP (GBD 2023)
  # Source: WHO REPLACE / WHO TFA policy tracker (as of 2024)
  tfa_deaths_annual     = "500,000",            # annual IHD deaths attributable to TFA
  tfa_protected_pop     = "3.4 billion",        # covered by best-practice TFA policies
  tfa_unprotected_pop   = "roughly 5 billion",  # without best-practice protection
  tfa_countries_bpp     = "53",                 # countries with WHO best-practice TFA policy
  tfa_countries_none    = "over 130",           # countries without any TFA policy
  # Source: RTSL program pages (resolvetosavelives.org)
  rtsl_htn_countries    = "30+",                # countries with RTSL HTN program support
  rtsl_htn_enrolled     = "30 million",         # patients enrolled in RTSL-supported HTN programs
  # Source: Lancet 2025 (S0140-6736(25)02208-1) - CVD burden synthesis
  lmic_cvd_share        = "over 80%",           # of CVD deaths occurring in LMICs
  premature_cvd_lmic    = "over three-quarters" # of premature CVD deaths in LMICs
)

saveRDS(sl_pd_scalars, paste0(slides_dir, "sl_pd_scalars.rds"))
message("PD saved: sl_pd_scalars.rds")
```

```{r pd-excel-export, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Dedicated Excel workbook for the donors/partners presentation ---------
# One sheet per figure/table used in the deck.
wb_pd <- createWorkbook()

add_sheet_pd <- function(sheet_name, dat, title_str) {
  addWorksheet(wb_pd, sheet_name)
  writeData(wb_pd, sheet_name, title_str, startRow = 1, startCol = 1)
  addStyle(wb_pd, sheet_name,
           style = createStyle(fontSize = 12, textDecoration = "bold"),
           rows = 1, cols = 1)
  writeData(wb_pd, sheet_name, dat, startRow = 3, startCol = 1,
            headerStyle = createStyle(textDecoration = "bold",
                                      fgFill = "#4B2E83",
                                      fontColour = "white",
                                      border = "Bottom"))
  setColWidths(wb_pd, sheet_name, cols = seq_len(ncol(dat)), widths = "auto")
}

# Fig: Cumulative deaths averted over time (combined)
add_sheet_pd("Fig_Cumulative_Averted",
             as.data.frame(dt_pd_annual),
             "Annual and cumulative deaths averted (HTN + TFA combined), all RTSL countries.")

# Fig: PAHO-merged Pareto chart data
add_sheet_pd("Fig_Pareto_Combined_PAHO",
             as.data.frame(dt_pareto_combined_paho[, .(
               location = as.character(location), htn_averted, tfa_averted,
               deaths_averted, cum_deaths, cum_pct)]),
             "Combined Pareto (top 20 + Rest). LAC countries merged into single PAHO bar.")

# Tbl: Per-country combined totals
add_sheet_pd("Tbl_Country_Combined",
             as.data.frame(dt_pd_country),
             "Per-country combined deaths averted (HTN + TFA, 2017-2050).")

# Tbl: Regional summary (combined)
add_sheet_pd("Tbl_Region_Combined",
             as.data.frame(sl_pd_tab_region),
             "Deaths averted by region (HTN + TFA combined).")

# Tbl: Top-10 countries for presentation
add_sheet_pd("Tbl_Top_Countries",
             as.data.frame(readRDS(paste0(slides_dir, "sl_pd_tab_top_countries.rds"))),
             "Top 10 countries/regions by combined deaths averted (PAHO-merged).")

# Tbl: Grand totals summary used on title impact slide
add_sheet_pd("Tbl_Grand_Totals",
             data.frame(
               Metric = c("Countries", "HTN countries", "TFA countries",
                          "Study period", "Total deaths averted (HTN+TFA)",
                          "Average deaths averted per day"),
               Value  = c(sl_pd_scalars$n_countries, sl_pd_scalars$n_htn_countries,
                          sl_pd_scalars$n_tfa_countries, sl_pd_scalars$study_years,
                          sl_pd_scalars$grand_combined_fmt,
                          sl_pd_scalars$daily_averted_fmt)
             ),
             "Grand totals (combined) used on opening impact slide.")

# Reference: problem-framing scalars with sources
framing_df <- data.frame(
  Variable = c("CVD deaths globally (annual)",
               "CVD share of global deaths",
               "Adults with hypertension globally",
               "Share with uncontrolled HTN",
               "Share with undiagnosed HTN",
               "Deaths attributable to high SBP (annual)",
               "IHD deaths attributable to TFA (annual)",
               "Population covered by best-practice TFA policies",
               "Population without best-practice TFA protection",
               "Countries with WHO best-practice TFA policy",
               "Countries without any TFA policy",
               "RTSL-supported HTN countries",
               "Patients enrolled in RTSL HTN programs",
               "Share of CVD deaths in LMICs",
               "Share of premature CVD deaths in LMICs"),
  Value    = c(sl_pd_scalars$cvd_deaths_global,
               sl_pd_scalars$cvd_share_noncomm,
               sl_pd_scalars$htn_adults_global,
               sl_pd_scalars$htn_uncontrolled_pct,
               sl_pd_scalars$htn_undiagnosed_share,
               sl_pd_scalars$htn_deaths_annual,
               sl_pd_scalars$tfa_deaths_annual,
               sl_pd_scalars$tfa_protected_pop,
               sl_pd_scalars$tfa_unprotected_pop,
               sl_pd_scalars$tfa_countries_bpp,
               sl_pd_scalars$tfa_countries_none,
               sl_pd_scalars$rtsl_htn_countries,
               sl_pd_scalars$rtsl_htn_enrolled,
               sl_pd_scalars$lmic_cvd_share,
               sl_pd_scalars$premature_cvd_lmic),
  Source   = c("GBD 2023 (IHME)",
               "GBD 2023 (IHME)",
               "WHO Global Report on Hypertension 2023; Lancet 2025",
               "WHO Global Report on Hypertension 2023",
               "WHO Global Report on Hypertension 2023",
               "GBD 2023 (IHME)",
               "WHO REPLACE initiative",
               "WHO TFA policy tracker 2024",
               "WHO TFA policy tracker 2024",
               "WHO TFA policy tracker 2024",
               "WHO TFA policy tracker 2024",
               "resolvetosavelives.org/blood-pressure-control",
               "resolvetosavelives.org/blood-pressure-control",
               "Lancet 2025 S0140-6736(25)02208-1",
               "Lancet 2025 S0140-6736(25)02208-1")
)
add_sheet_pd("Ref_Framing_Scalars",
             framing_df,
             "Problem-framing statistics used on opening slides (external sources).")

out_xlsx_pd <- paste0(wd_scen, "presentation_cvh_partners_donors.xlsx")
saveWorkbook(wb_pd, out_xlsx_pd, overwrite = TRUE)
message("Presentation tables exported to: ", out_xlsx_pd)
```

```{r slides-manifest, message=FALSE, warning=FALSE, echo=FALSE}
sl_manifest <- data.table(
  file = c(
    "sl_tab_top15.rds", "sl_dt_cause.rds", "sl_dt_annual.rds",
    "sl_fig_deaths_averted.png", "sl_fig_top15_countries.png",
    "sl_fig_cum_averted.png",
    "sl_bangladesh.rds", "sl_tab_scenario.rds",
    "sl_fig_htn_control_rates.rds", "sl_tbl_tfa_countries.rds",
    "sl_fig_htn_per1m.rds", "sl_fig_tfa_per1m.rds",
    "sl_fig_combined_per1m.rds",
    "sl_tbl_htn_program_vs_country.rds",
    "sl_tbl_countries_by_region.rds",
    "sl_tbl_total_lives_saved.rds",
    "sl_tbl_grand_total_lives_saved.rds",
    "sl_fig_global_cumulative.rds",
    "sl_tbl_bangladesh_summary.rds",
    "sl_fig_bangladesh_cumulative.rds",
    "sl_fig_tfa_timeline.rds",
    "sl_fig_country_map.rds"
  ),
  type = c(rep("rds", 6), rep("png", 3), rep("rds", 16)),
  consumed_by = c(
    "inline scalars", "tab-summary kable", "tab-region kable",
    "fig-top-countries ggplot", "fig-by-cause ggplot", "fig-deaths-averted ggplot",
    "fig-deaths-averted include_graphics",
    "fig-top-countries include_graphics",
    "fig-cum-averted include_graphics",
    "bangladesh-results kable (presentation.Rmd)",
    "scenario-table kable (presentation.Rmd)",
    "A1: HTN control rates figure (presentation.Rmd)",
    "A2: TFA policy timeline table (presentation.Rmd)",
    "HTN per-1M results figure (presentation.Rmd)",
    "TFA per-1M results figure (presentation.Rmd)",
    "Combined per-1M figure (presentation.Rmd)",
    "A3: HTN program vs country table (presentation.Rmd)",
    "A5: Countries by region table (presentation.Rmd)",
    "B1: Total lives saved table (presentation.Rmd)",
    "B4: Grand total lives saved table (presentation.Rmd)",
    "B5: Global cumulative chart (presentation.Rmd)",
    "B2: Bangladesh summary table (presentation.Rmd)",
    "B3: Bangladesh cumulative chart (presentation.Rmd)",
    "TFA timeline bar+line chart (presentation_donors.Rmd)",
    "Country coverage world map (presentation_donors.Rmd)"
  )
)
saveRDS(sl_manifest, paste0(slides_dir, "sl_manifest.rds"))

kable(sl_manifest,
      caption = "Slide deck artefact manifest. All files written to output/slides/.",
      col.names = c("File", "Type", "Consumed by (slides chunk)")) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE, font_size = 11) |>
  row_spec(0, bold = TRUE)

message(sprintf("SLIDE DECK OUTPUTS complete. %d artefacts written to %s",
                nrow(sl_manifest), slides_dir))
```

---

<!-- ======================================================================
     100 MILLION LIVES SLIDES  ---  NEW DATA SOURCE
     Source: data/processed/UWRTSL-100MLives_Output20260204.xlsx,
             sheet "Location And Cause".
     Cumulative deaths delayed 2026-2050; four single interventions plus an
     "All Interventions" combined total. Self-contained: reads the workbook
     directly and does NOT use the model RDS output (data_program/data_bau).
     Builds and saves to output/slides/:
       Task 1: country_stacked_bar.rds, country_summary_table.rds
       Task 2: pareto_paho.rds (+ pareto_paho_data.rds,
               pareto_paho_unclassified.rds)
     Rendered (load-only) by docs/presentation_cvh_partners_donors.Rmd.
     ====================================================================== -->

# 100 Million Lives: Four-Intervention Country Slides

```{r ml100-setup, message=FALSE, warning=FALSE, echo=FALSE}
# ---- FIXED palette, labels, legend order (do not change) --------------------
ml100_label_map <- c(
  "Improved Hypertension Control" = "HTN",
  "Population Sodium Reduction"    = "Sodium",
  "TFA Elimination"               = "TFA",
  "Improved Statin Uptake"        = "Statins"
)
ml100_levels <- c("Statins", "TFA", "Sodium", "HTN")   # legend order (top->bottom)
ml100_cols <- c(
  "HTN"     = "#E7298A",   # pink
  "Sodium"  = "#1F77B4",   # blue
  "TFA"     = "#F58025",   # orange
  "Statins" = "#2CA02C"    # green
)

# Clean, readable country labels for chart axes
ml100_clean_loc <- function(x) {
  repl <- c(
    "Russian Federation"                    = "Russia",
    "Iran (Islamic Republic of)"            = "Iran",
    "Venezuela (Bolivarian Republic of)"    = "Venezuela",
    "Bolivia (Plurinational State of)"      = "Bolivia",
    "Democratic People's Republic of Korea" = "North Korea",
    "Republic of Korea"                     = "South Korea",
    "Democratic Republic of the Congo"      = "DR Congo",
    "United Republic of Tanzania"           = "Tanzania",
    "Lao People's Democratic Republic"      = "Laos",
    "Syrian Arab Republic"                  = "Syria",
    "Viet Nam"                              = "Vietnam",
    "Taiwan (Province of China)"            = "Taiwan",
    "Republic of Moldova"                   = "Moldova"
  )
  out <- unname(repl[x]); ifelse(is.na(out), x, out)
}

# ---- Read the 100M-lives output (self-contained; NOT model RDS) -------------
ml100_path <- paste0(wd_data, "UWRTSL-100MLives_Output20260204.xlsx")
dt_ml100   <- as.data.table(read.xlsx(ml100_path, sheet = "Location And Cause"))
dt_ml100[, int_lbl := ml100_label_map[intervention]]

# ---- Aggregation totals (used for sanity check + summary table) -------------
ml100_tot <- dt_ml100[, .(deaths_delayed = sum(deaths_delayed, na.rm = TRUE)),
                      by = intervention]
ml100_comp_sum  <- ml100_tot[intervention != "All Interventions", sum(deaths_delayed)]
ml100_all_total <- ml100_tot[intervention == "All Interventions", deaths_delayed]

message(sprintf("ML100 totals (M): HTN %.1f | Sodium %.1f | TFA %.1f | Statins %.1f | sum-4 %.1f | All %.1f",
  ml100_tot[intervention == "Improved Hypertension Control", deaths_delayed] / 1e6,
  ml100_tot[intervention == "Population Sodium Reduction",   deaths_delayed] / 1e6,
  ml100_tot[intervention == "TFA Elimination",               deaths_delayed] / 1e6,
  ml100_tot[intervention == "Improved Statin Uptake",        deaths_delayed] / 1e6,
  ml100_comp_sum / 1e6, ml100_all_total / 1e6))
```

## Sanity Check --- Per-Intervention Totals

```{r ml100-sanity, message=FALSE, warning=FALSE, echo=FALSE}
ml100_sanity <- copy(ml100_tot)
ml100_sanity[, Millions := formatC(deaths_delayed / 1e6, format = "f",
                                   digits = 1, big.mark = ",")]
ml100_sanity[, intervention := factor(intervention,
  levels = c("Improved Hypertension Control", "Population Sodium Reduction",
             "Improved Statin Uptake", "TFA Elimination", "All Interventions"))]
setorder(ml100_sanity, intervention)

kable(ml100_sanity[, .(Intervention = intervention,
                       `Deaths delayed (millions)` = Millions)],
      caption = paste0(
        "Sanity check: cumulative deaths delayed (2026–2050) by intervention. ",
        "Sum of the four single interventions = ",
        formatC(ml100_comp_sum / 1e6, format = "f", digits = 1), "M; ",
        "All Interventions (combined) = ",
        formatC(ml100_all_total / 1e6, format = "f", digits = 1), "M. ",
        "The combined estimate is lower than the arithmetic sum because deaths ",
        "preventable by more than one intervention are counted once."),
      align = "lr") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE)
```

## Task 1 --- Country Stacked Bar + Summary Table

```{r ml100-task1, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=6}
# Rank countries by the All-Interventions total (descending)
ml100_rank <- dt_ml100[intervention == "All Interventions",
                       .(location, all_tot = deaths_delayed)]
setorder(ml100_rank, -all_tot)
top20_locs <- head(ml100_rank$location, 20)

# Stack the four single interventions; collapse the remainder into "Rest"
dt_comp <- dt_ml100[intervention != "All Interventions"]
dt_comp[, grp := fifelse(location %in% top20_locs, location, "Rest of countries")]
dt_bar <- dt_comp[, .(deaths_delayed = sum(deaths_delayed, na.rm = TRUE)),
                  by = .(grp, int_lbl)]

# Bar order: top 20 desc, then "Rest of countries" last; largest at top
ord_raw   <- c(ml100_rank[location %in% top20_locs][order(-all_tot), location],
               "Rest of countries")
clean_one <- function(g) if (g == "Rest of countries") "Rest of countries" else ml100_clean_loc(g)
ord_clean <- vapply(ord_raw, clean_one, character(1))
dt_bar[, grp_clean := vapply(grp, clean_one, character(1))]
dt_bar[, grp_clean := factor(grp_clean, levels = rev(ord_clean))]
dt_bar[, int_lbl   := factor(int_lbl, levels = ml100_levels)]

country_stacked_bar <- ggplot(dt_bar,
       aes(x = deaths_delayed / 1e6, y = grp_clean, fill = int_lbl)) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = ml100_cols, breaks = ml100_levels, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = label_number(accuracy = 0.1)) +
  labs(title    = "Where the lives are saved, by country",
       subtitle = "Four single interventions stacked; cumulative 2026–2050",
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position    = "bottom",
        panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title         = element_text(face = "bold"))
print(country_stacked_bar)

# Right-hand summary table: Policy | Lives saved (millions) | % of total.
# % of total = share of the summed single-intervention impact (matches the
# reference slide pattern: HTN 59 / Sodium 29 / Statins 9 / TFA 4).
tab <- ml100_tot[intervention != "All Interventions"]
tab[, int_lbl := ml100_label_map[intervention]]
tab[, pct := deaths_delayed / ml100_comp_sum]
setorder(tab, -deaths_delayed)
country_summary_table <- data.table(
  Policy                   = tab$int_lbl,
  `Lives saved (millions)` = formatC(tab$deaths_delayed / 1e6, format = "f",
                                     digits = 1, big.mark = ","),
  `% of total`             = paste0(round(tab$pct * 100), "%")
)
country_summary_table <- rbind(
  country_summary_table,
  data.table(Policy = "All Interventions",
             `Lives saved (millions)` = formatC(ml100_all_total / 1e6, format = "f",
                                                digits = 1, big.mark = ","),
             `% of total` = "—")
)

saveRDS(country_stacked_bar,   paste0(slides_dir, "country_stacked_bar.rds"))
saveRDS(country_summary_table, paste0(slides_dir, "country_summary_table.rds"))
message("ML100 saved: country_stacked_bar.rds, country_summary_table.rds")
```

## Task 2 --- PAHO-Merged Pareto (Stacked Interventions + Cumulative %)

```{r ml100-task2, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
# Explicit PAHO (Latin America & Caribbean) member vector --- sheet naming.
paho_members <- c(
  "Antigua and Barbuda", "Argentina", "Bahamas", "Barbados", "Belize", "Bolivia",
  "Brazil", "Chile", "Colombia", "Costa Rica", "Cuba", "Dominica",
  "Dominican Republic", "Ecuador", "El Salvador", "Grenada", "Guatemala",
  "Guyana", "Haiti", "Honduras", "Jamaica", "Mexico", "Nicaragua", "Panama",
  "Paraguay", "Peru", "Puerto Rico", "Saint Lucia",
  "Saint Vincent and the Grenadines", "Suriname", "Trinidad and Tobago",
  "Venezuela (Bolivarian Republic of)"
)

# ---- Classification flags so the PAHO membership can be reviewed ------------
sheet_locs   <- unique(dt_ml100$location)
paho_missing <- setdiff(paho_members, sheet_locs)   # vector names absent from sheet
lac_csv_missed <- character(0)                       # CSV-LAC countries the vector omits
if (exists("countries_dt")) {
  lac_csv <- countries_dt[region == "Latin America and Caribbean", unique(location)]
  lac_csv_missed <- setdiff(intersect(lac_csv, sheet_locs), paho_members)
}
ml100_unclassified <- list(
  paho_members_not_in_sheet  = paho_missing,
  csv_LAC_not_in_paho_vector = lac_csv_missed
)
message("ML100 PAHO flags | members not in sheet: ",
        if (length(paho_missing)) paste(paho_missing, collapse = ", ") else "(none)",
        " | CSV-LAC not in vector: ",
        if (length(lac_csv_missed)) paste(lac_csv_missed, collapse = ", ") else "(none)")
saveRDS(ml100_unclassified, paste0(slides_dir, "pareto_paho_unclassified.rds"))

# ---- Merge LAC into a single PAHO entity, then rank -------------------------
dt_p <- dt_ml100[intervention != "All Interventions",
                 .(location, int_lbl, deaths_delayed)]
dt_p[, entity := fifelse(location %in% paho_members,
                         "PAHO (Latin America & Caribbean)", location)]
dt_ent <- dt_p[, .(deaths_delayed = sum(deaths_delayed, na.rm = TRUE)),
               by = .(entity, int_lbl)]

ent_tot <- dt_ent[, .(tot = sum(deaths_delayed)), by = entity]
setorder(ent_tot, -tot)
top20_ent <- head(ent_tot$entity, 20)

dt_ent[, grp := fifelse(entity %in% top20_ent, entity, "Rest of countries")]
dt_grp <- dt_ent[, .(deaths_delayed = sum(deaths_delayed, na.rm = TRUE)),
                 by = .(grp, int_lbl)]

# Order: top 20 entities desc, then "Rest of countries" forced LAST (Pareto tail)
grp_tot <- dt_grp[, .(tot = sum(deaths_delayed)), by = grp]
ord_ent <- c(setdiff(grp_tot[order(-tot), grp], "Rest of countries"),
             "Rest of countries")
clean_ent <- function(g) {
  if (g == "Rest of countries") "Rest of countries"
  else if (g == "PAHO (Latin America & Caribbean)") "PAHO (LAC)"
  else ml100_clean_loc(g)
}
ord_ent_clean <- vapply(ord_ent, clean_ent, character(1))

dt_grp[, grp_clean := vapply(grp, clean_ent, character(1))]
dt_grp[, grp_clean := factor(grp_clean, levels = ord_ent_clean)]
dt_grp[, int_lbl   := factor(int_lbl, levels = ml100_levels)]
dt_grp[, dd_M := deaths_delayed / 1e6]

# Cumulative-% line (computed in plotting order)
dt_line <- dt_grp[, .(tot_M = sum(dd_M)), by = grp_clean]
setorder(dt_line, grp_clean)
dt_line[, cum_M     := cumsum(tot_M)]
dt_line[, cum_pct   := cum_M / sum(tot_M)]
y_max_M <- max(dt_line$tot_M, na.rm = TRUE)
dt_line[, cum_scaled := cum_pct * y_max_M]

# Bake the literal y_max into the sec-axis formula so the saved ggplot object
# is self-contained when loaded in the presentation's environment (the loading
# env has no y_max_M). Captured via substitute() + as.formula().
sec_form <- substitute(~ . / ymp, list(ymp = y_max_M))

pareto_paho <- ggplot() +
  geom_col(data = dt_grp,
           aes(x = grp_clean, y = dd_M, fill = int_lbl), width = 0.80) +
  geom_line(data = dt_line,
            aes(x = grp_clean, y = cum_scaled, group = 1),
            color = "#4B2E83", linewidth = 1) +
  geom_point(data = dt_line,
             aes(x = grp_clean, y = cum_scaled),
             color = "#4B2E83", size = 1.8) +
  scale_fill_manual(values = ml100_cols, breaks = ml100_levels, name = NULL) +
  scale_y_continuous(
    name   = "Deaths delayed (millions)",
    labels = label_number(accuracy = 1),
    expand = expansion(mult = c(0, 0.05)),
    sec.axis = sec_axis(trans = as.formula(sec_form),
                        name = "Cumulative %",
                        labels = percent_format(accuracy = 1))
  ) +
  labs(title    = "Concentration of impact across countries",
       subtitle = "Four interventions stacked; Latin America & Caribbean merged as PAHO; cumulative 2026–2050",
       x = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x      = element_text(angle = 45, hjust = 1),
        legend.position  = "bottom",
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(color = "grey40", size = 9))
print(pareto_paho)

saveRDS(pareto_paho, paste0(slides_dir, "pareto_paho.rds"))
saveRDS(dt_grp,      paste0(slides_dir, "pareto_paho_data.rds"))
message("ML100 saved: pareto_paho.rds, pareto_paho_data.rds")
```

---

*Report generated on `r Sys.time()` using R `r R.version$major`.`r R.version$minor`.*
