---
title: "Sodium Policy Intervention Model: Methods and Projected Health Impact (2017–2050)"
date: "`r Sys.Date()`"
author: "University of Washington · Resolve to Save Lives"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 5,
  fig.width  = 9,
  fig.align  = "center"
)
```

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(DT)
library(scales)
```

```{r paths, message=FALSE, warning=FALSE, echo=FALSE}
# Single source of truth for paths (mirrors the RTSL assessment report). The
# project root is resolved relative to this .Rmd (scenarios/ is one level below
# the repo root) so the report knits on any machine; a hard-coded fall-back is
# used when the file is run chunk-by-chunk outside of knit.
wd <- tryCatch({
  root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
                        winslash = "/", mustWork = TRUE)
  paste0(root, "/")
}, error = function(e)
  "C:/Users/wrgar/OneDrive - UW/02Work/ResolveToSaveLives/SodiumPolicy/github/UWRTSL-sodium-policy/")

wd_data <- paste0(wd, "data/processed/")
wd_outp <- paste0(wd, "output/")
wd_scen <- paste0(wd, "scenarios/")
wd_raw  <- paste0(wd, "data/raw/")
```

```{r helpers, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Number formatters -------------------------------------------------------
fmt_m <- function(x, digits = 2)     # millions
  formatC(round(x / 1e6, digits), format = "f", digits = digits, big.mark = ",")
fmt_k <- function(x, digits = 0)     # thousands
  formatC(round(x / 1e3, digits), format = "f", digits = digits, big.mark = ",")
fmt_pct <- function(x, digits = 1)
  paste0(formatC(100 * x, format = "f", digits = digits), "%")

# ---- Cause labels ------------------------------------------------------------
cause_map <- c(
  ihd     = "Ischemic heart disease",
  istroke = "Ischemic stroke",
  hstroke = "Intracerebral hemorrhage",
  hhd     = "Hypertensive heart disease"
)
CAUSES <- names(cause_map)

cause_cols <- c(
  ihd     = "#56B4E9",
  istroke = "#CC79A7",
  hstroke = "#F0E442",
  hhd     = "#E69F00"
)

# ---- Scenario metadata -------------------------------------------------------
# The intervention model writes a 'scenario' column that is authoritative:
#   baseline + one row per individual intervention + full_package.
# (The 'intervention' column is only a binary arm flag - Baseline vs
#  "Sodium reduction" - and must NOT be used to distinguish scenarios.)
scenario_levels <- c("baseline", "public_procurement", "media_campaign", "fopl",
                     "salt_targets", "fiscal", "lss", "full_package")

scenario_labels <- c(
  baseline           = "Baseline (no intervention)",
  public_procurement = "Public food procurement & service",
  media_campaign     = "Behaviour change / media campaigns",
  fopl               = "Front-of-pack labelling (FOPL)",
  salt_targets       = "Salt targets (reformulation)",
  fiscal             = "Fiscal policy (salt tax)",
  lss                = "Low-sodium salt substitutes (LSS)",
  full_package       = "Full package"
)

# Short labels for compact axes / legends
scenario_labels_short <- c(
  baseline           = "Baseline",
  public_procurement = "Public procurement",
  media_campaign     = "Media campaign",
  fopl               = "FOPL",
  salt_targets       = "Salt targets",
  fiscal             = "Fiscal",
  lss                = "LSS",
  full_package       = "Full package"
)

scenario_cols <- c(
  baseline           = "#999999",
  public_procurement = "#1f78b4",
  media_campaign     = "#33a02c",
  fopl               = "#ff7f00",
  salt_targets       = "#e31a1c",
  fiscal             = "#6a3d9a",
  lss                = "#a6761d",
  full_package       = "#000000"
)

# Non-baseline scenarios, in canonical order (individuals then package)
scenario_nb <- setdiff(scenario_levels, "baseline")

as_scn_factor <- function(x) factor(x, levels = scenario_levels)

# ---- WHO standard population (ages 20-95+) for ASMR -------------------------
full_who_std <- data.table(
  age    = 0:100,
  weight = c(
    rep(0.013, 10), rep(0.014, 10), rep(0.013, 10), rep(0.012, 10),
    rep(0.011, 10), rep(0.009, 10), rep(0.007, 10), rep(0.005, 10),
    rep(0.003, 10), rep(0.001, 10), 0.001
  )
)
who_std <- full_who_std[age >= 20 & age <= 94]
who_std <- rbind(
  who_std,
  data.table(age = 95, weight = full_who_std[age >= 95, sum(weight)])
)
who_std[, weight := weight / sum(weight)]

# Age-standardised rate per 100,000. deaths_col / population_col at single-year
# age, weighted to the WHO world standard, then summed within (year, disagg_var).
calculate_asmr <- function(dt, std_pop = who_std,
                           deaths_col = "deaths_dx",
                           population_col = "Nx",
                           disagg_var = "scenario") {
  dt <- copy(dt)
  dt[, age := as.integer(gsub("\\+", "", as.character(age)))]
  dt <- merge(dt, std_pop, by = "age", all.x = TRUE)
  dt[, rate          := get(deaths_col) / get(population_col)]
  dt[, weighted_rate := rate * weight]
  group_vars <- c("year", disagg_var)
  dt[, .(ASMR = sum(weighted_rate, na.rm = TRUE) * 1e5), by = group_vars]
}

# ---- Deaths-delayed calculator (keyed on 'scenario') -------------------------
# deaths_delayed = baseline deaths - scenario deaths, over an optional set of
# grouping columns and an optional year window. The reference is always the
# 'baseline' scenario. Uses 'dead' (cause-specific, safe to sum over causes).
deaths_delayed_by <- function(dt, by = character(0),
                              years = NULL, ref = "baseline") {
  d <- if (is.null(years)) dt else dt[year %in% years]
  agg  <- d[, .(dead = sum(dead, na.rm = TRUE)), by = c("scenario", by)]
  base <- agg[scenario == ref][, scenario := NULL]
  setnames(base, "dead", "baseline_dead")
  out  <- agg[scenario != ref]
  if (length(by) > 0) {
    out <- merge(out, base, by = by, all.x = TRUE)
  } else {
    out[, baseline_dead := base$baseline_dead]
  }
  out[, deaths_delayed := baseline_dead - dead]
  out[]
}
```

```{r source-logic, message=FALSE, warning=FALSE, echo=FALSE}
# Self-contained copy of the source-split efficacy logic from
# code/07_run_interventions.R (Section 3). Reproduced here verbatim so this
# report can display the source-share, effect-size and salteff tables WITHOUT
# depending on run-time globals - it stands alone on the .rds outputs.

SODIUM_SOURCES     <- c("discretionary", "packaged", "restaurant", "public", "inherent")
TARGETABLE_SOURCES <- setdiff(SODIUM_SOURCES, "inherent")

# Effect-size constants (relative reduction applied to the target source).
# Source: RTSL sodium policy document intervention effect-size tables.
EFF_PROCUREMENT_PUBLIC   <- 0.20   # Public food procurement & service -> 'public'
EFF_MEDIA                <- 0.010  # Media campaigns -> 'discretionary' + 'packaged'
EFF_FOPL_PACKAGED        <- 0.138  # Front-of-pack labelling -> 'packaged'
EFF_SALTTARGETS_PACKAGED <- 0.20   # Salt targets (reformulation) -> 'packaged'
EFF_FISCAL_PACKAGED      <- 0.0   # Fiscal policy (salt tax) -> 'packaged'
EFF_LSS_DISCRETIONARY    <- 0.15   # Low-sodium salt substitutes -> 'discretionary'
# SSaSS-derived (Neal et al., NEJM 2021;385:1067-77): a net -350 mg/day ITT
# sodium reduction against a ~4.3 g/day baseline -> ~0.15 relative reduction of
# discretionary sodium (sensitivity 0.12-0.18; sodium-arm only, no K+ channel).
# LSS reach is restricted to the hypertensive/raised-BP subpopulation in the
# model. It is now a full model component included in the package.

intervention_effects <- data.table(
  intervention = c("public_procurement",
                   "media_campaign", "media_campaign",
                   "fopl", "salt_targets", "fiscal", "lss"),
  source = c("public",
             "discretionary", "packaged",
             "packaged", "packaged", "packaged", "discretionary"),
  effect = c(EFF_PROCUREMENT_PUBLIC,
             EFF_MEDIA, EFF_MEDIA,
             EFF_FOPL_PACKAGED, EFF_SALTTARGETS_PACKAGED,
             EFF_FISCAL_PACKAGED, EFF_LSS_DISCRETIONARY)
)

intervention_registry <- data.table(
  intervention = c("public_procurement", "media_campaign", "fopl",
                   "salt_targets", "fiscal", "lss"),
  label        = c("Public food procurement & service",
                   "Behaviour change / media campaigns",
                   "Front-of-pack labelling (FOPL)",
                   "Salt targets (reformulation)",
                   "Fiscal policy (salt tax)",
                   "Low-sodium salt substitutes (LSS)"),
  # LSS is no longer a placeholder: it is an SSaSS-derived model component that
  # is included in the full package (matches include_placeholders_in_package
  # <- TRUE in code/07_run_interventions.R). No intervention is a placeholder.
  placeholder  = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE)
)

# Per-country sodium source-share table (+ 'default' row), renormalised to 1.0.
# Mirrors build_sodium_source_shares() in code/07_run_interventions.R verbatim:
# all TEN prioritised countries, including Nigeria and Cameroon (provisional
# shares supplied by RTSL/Matti on 8 Jul 2026, bouillon + condiments counted as
# packaged; Nigeria is used as a proxy for Cameroon).
build_sodium_source_shares <- function() {
  shares_raw <- data.table(
    location      = c("Viet Nam", "Philippines", "Bangladesh", "China",
                      "Ethiopia", "India", "Malaysia", "Thailand",
                      "Nigeria", "Cameroon"),
    discretionary = c(0.109, 0.320, 0.738, 0.594, 0.783, 0.738, 0.320, 0.235,
                      0.231, 0.231),
    packaged      = c(0.641, 0.430, 0.062, 0.173, 0.017, 0.062, 0.430, 0.141,
                      0.415, 0.415),
    restaurant    = c(0.100, 0.100, 0.050, 0.083, 0.050, 0.050, 0.100, 0.473,
                      0.205, 0.205),
    public        = c(0.050, 0.050, 0.050, 0.050, 0.050, 0.050, 0.050, 0.050,
                      0.050, 0.050),
    inherent      = c(0.100, 0.100, 0.100, 0.100, 0.100, 0.100, 0.100, 0.100,
                      0.100, 0.100)
  )
  # Malaysia uses the Philippines source profile as a proxy (per document).
  default_row <- shares_raw[, lapply(.SD, mean), .SDcols = SODIUM_SOURCES]
  default_row[, location := "default"]
  setcolorder(default_row, c("location", SODIUM_SOURCES))
  shares <- rbind(shares_raw, default_row, use.names = TRUE)
  rs <- rowSums(shares[, ..SODIUM_SOURCES])
  shares[, (SODIUM_SOURCES) := lapply(.SD, function(x) x / rs), .SDcols = SODIUM_SOURCES]
  shares[]
}

# Collapse a set of interventions to a single total-intake reduction fraction
# (the 'salteff') for one country.
compute_total_efficacy <- function(intervention_names, country,
                                   source_shares, intervention_effects) {
  sources <- SODIUM_SOURCES
  srow <- source_shares[location == country]
  if (nrow(srow) == 0L) srow <- source_shares[location == "default"]
  if (nrow(srow) == 0L) stop("No source shares for '", country, "' and no default.")
  shares_vec <- unlist(srow[1L, ..sources])

  if (is.null(intervention_names) || length(intervention_names) == 0L) {
    total <- 0
    attr(total, "decomposition") <-
      data.table(source = sources, share = shares_vec[sources],
                 composed_effect = 0, contribution = 0)
    return(total)
  }
  eff <- intervention_effects[intervention %in% intervention_names]
  composed <- vapply(sources, function(s) {
    e <- eff[source == s, effect]
    if (length(e) == 0L) 0 else 1 - prod(1 - e)
  }, numeric(1))
  dec <- data.table(source = sources, share = shares_vec[sources],
                    composed_effect = composed[sources],
                    contribution = shares_vec[sources] * composed[sources])
  total <- sum(dec$contribution)
  attr(total, "decomposition") <- dec[]
  total
}

source_shares <- build_sodium_source_shares()

# Colours / labels for the five sodium sources (used by the source-share and
# salteff stacked barplots).
source_cols <- c(
  discretionary = "#4E79A7",
  packaged      = "#F28E2B",
  restaurant    = "#59A14F",
  public        = "#E15759",
  inherent      = "#B07AA1"
)
source_labels <- c(
  discretionary = "Discretionary",
  packaged      = "Packaged",
  restaurant    = "Restaurant",
  public        = "Public",
  inherent      = "Inherent"
)

# The full package. code/07_run_interventions.R sets
# include_placeholders_in_package <- TRUE, so LSS (no longer a placeholder) is
# folded into the package alongside the other five interventions.
include_placeholders_in_package <- TRUE
package_interventions <- if (isTRUE(include_placeholders_in_package)) {
  intervention_registry$intervention
} else {
  intervention_registry[placeholder == FALSE, intervention]
}

# Map each scenario -> its intervention vector (baseline = none).
scenario_interventions <- c(
  setNames(as.list(intervention_registry$intervention),
           intervention_registry$intervention),
  list(baseline = character(0),
       full_package = package_interventions)
)

#' Resolve a country's five source shares to a named numeric vector, falling
#' back to the 'default' row for any country without documented shares (all ten
#' prioritised countries now have their own row). Keeps the barplots robust.
resolve_source_shares <- function(countries,
                                  shares = source_shares) {
  rbindlist(lapply(countries, function(cc) {
    srow <- shares[location == cc]
    if (nrow(srow) == 0L) srow <- shares[location == "default"]
    srow <- copy(srow[1L])
    srow[, location := cc]
    srow[]
  }))
}
```

```{r country-metadata, message=FALSE, warning=FALSE, echo=FALSE}
# Region lookup for optional regional breakdowns.
countries_dt <- tryCatch(
  fread(paste0(wd_data, "Country_groupings_extended.csv")),
  error = function(e) data.table(iso3 = character(), location = character(),
                                 region = character())
)
if (nrow(countries_dt)) {
  countries_dt[location == "United States of America", location := "United States"]
  countries_dt[location == "Vietnam",                  location := "Viet Nam"]
}

# Baseline mean sodium intake (2024) - used only to illustrate the salteff ->
# grams -> SBP chain; the report does not otherwise depend on it.
sodium_baseline <- tryCatch({
  sp <- readRDS(paste0(wd_data, "sodium_policy_scenarios.rds"))
  setDT(sp)
  sp[year == 2024, .(location, sodium_current)]
}, error = function(e) NULL)
```

```{r load-model-output, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Load the per-country intervention outputs ------------------------------
# Written by code/07_run_interventions.R, one file per country:
#   output/out_model/model_output_interventions_<Country>.rds
# Each is a data.table with an authoritative 'scenario' column.
out_model_path <- paste0(wd_outp, "out_model/")
files <- list.files(path = out_model_path,
                    pattern = "^model_output_interventions_.*\\.rds$",
                    full.names = TRUE)

if (length(files) == 0)
  stop(paste0(
    "No 'model_output_interventions_*.rds' files found in:\n  ", out_model_path,
    "\n\nRun code/07_run_interventions.R first (it must be sourced after ",
    "05_build_baseline.R) to generate the per-country intervention outputs."
  ))

data.out <- rbindlist(
  lapply(files, function(f) { dt <- readRDS(f); setDT(dt); dt }),
  use.names = TRUE, fill = TRUE
)
data.out <- data.out[!is.na(location) & !is.na(scenario)]

# Keep only recognised scenarios and impose canonical ordering.
data.out <- data.out[scenario %in% scenario_levels]
data.out[, scenario := as_scn_factor(scenario)]

# Restrict to the four modelled CVD causes.
data.out <- data.out[cause %in% CAUSES]

# Age groups for demographic breakdowns.
data.out[, age_group := cut(age, breaks = c(20, 45, 65, 80, Inf),
                            labels = c("20-44", "45-64", "65-79", "80+"),
                            right = FALSE, include.lowest = TRUE)]

# Region metadata (optional).
if (nrow(countries_dt)) {
  data.out <- merge(data.out, unique(countries_dt[, .(location, region)]),
                    by = "location", all.x = TRUE)
}
if (!"region" %in% names(data.out)) data.out[, region := NA_character_]

# ---- Key counts / dimensions ------------------------------------------------
n_countries <- uniqueN(data.out$location)
year_min    <- min(data.out$year, na.rm = TRUE)
year_max    <- max(data.out$year, na.rm = TRUE)
study_years <- paste0(year_min, "–", year_max)
scenarios_present <- levels(droplevels(data.out$scenario))

# Scale-up window (as configured in 07_run_interventions.R).
SCALEUP_YEAR1 <- 2026
SCALEUP_YEAR2 <- 2030

# Impact-accounting window for the headline results.
impact_y1 <- SCALEUP_YEAR1
impact_y2 <- min(2050, year_max)
impact_window <- paste0(impact_y1, "–", impact_y2)
n_impact_years <- impact_y2 - impact_y1 + 1
```

---

# Background

This report documents the **source-split sodium-policy intervention model**
implemented in [`code/07_run_interventions.R`](../code/07_run_interventions.R)
and presents its baseline and projected outputs. The model estimates the
cardiovascular health impact of reducing population sodium intake through a
menu of policy interventions, projecting outcomes annually for
**`r n_countries` countries** over `r study_years`.

The central question is:

> *How many cardiovascular deaths are delayed by each sodium-reduction policy
> — and by a coordinated policy package — when each policy's documented effect
> is applied only to the sodium sources it can actually reach?*

The engine is a **discrete-time, multi-state (Well → Sick → Dead) cohort
model** calibrated to GBD, run once per **scenario**:

- a **baseline** (no intervention),
- one scenario for each **individual** intervention, and
- a combined **full package**.

Four causes of cardiovascular death are modelled jointly: ischemic heart
disease (`ihd`), ischemic stroke (`istroke`), intracerebral hemorrhage
(`hstroke`), and hypertensive heart disease (`hhd`).

The defining feature of this build is **upstream**: instead of applying a
single blended reduction to a single sodium pool, baseline sodium is split into
**five sources**, and each intervention's documented relative reduction is
applied **only to the source(s) it can affect**. Those per-source reductions
are then collapsed to a single total-intake reduction fraction (`salteff`) that
feeds the unchanged epidemiological pathway (sodium → SBP → CVD incidence →
state transitions). Everything from `salteff` onward is identical to the prior
single-pool model.

```{r headline-precompute, message=FALSE, warning=FALSE, echo=FALSE}
# Headline numbers for the inline narrative below.
dd_total <- deaths_delayed_by(data.out, by = character(0),
                              years = impact_y1:impact_y2)
dd_total[, scenario := as_scn_factor(as.character(scenario))]
setorder(dd_total, scenario)
get_dd <- function(scn) dd_total[as.character(scenario) == scn, deaths_delayed]
pkg_total <- get_dd("full_package")
```

Across `r n_countries` countries and `r impact_window`, the **full package** is
projected to delay approximately **`r fmt_m(pkg_total)` million**
cardiovascular deaths relative to baseline. Individual interventions and the
by-cause decomposition are given in the [results](#main-results-deaths-delayed).

> **A note on terminology.** Because the intervention lowers disease
> *incidence* while case fatality is held at its baseline value, the model
> shifts deaths later in time rather than removing them from the life table
> entirely. We therefore report **deaths delayed** (baseline deaths − scenario
> deaths within the window) rather than "deaths averted".

---

# Model inputs

The report reads only the per-country `.rds` outputs; the inputs below are those
consumed *upstream* by `07_run_interventions.R` (via `05_build_baseline.R`) and
are described here for provenance.

| Input | File | Role |
|---|---|---|
| Baseline transition rates | `*adjusted*.rds` → `b_rates` | Age × sex × year × cause incidence (IR), case fatality (CF), background mortality (BG.mx, BG.mx.all), COVID excess (covid.mx), prevalence and population (Nx). |
| BP distribution | `bp_data6.csv` → `data.in` | Mean/SD systolic BP by age × sex, raised-BP fraction, diabetes prevalence; the baseline sodium column `salt`. |
| Baseline sodium intake | `sodium_policy_scenarios.rds` | 2024 mean sodium intake (`sodium_current`) merged in as `salt`. |
| Sodium source shares | recomputed in-report (also `sodium_source_shares.rds`) | Five-source split of baseline sodium per country. |
| GBD relative risks | `IHME_GBD_2019_RELATIVE_RISKS_*_HTN.xlsx` → `dt_gbd_rr` | Log-linear RR per 10 mmHg SBP, by age × cause; sets bin-specific baseline incidence. |
| ETTEHAD relative risks | `ettehad_rr_bp_reduction_10mmHg.csv`, `ettehad_rr_bp_reduction_effects.xlsx` | RR per 10 mmHg SBP reduction by BP bin × cause; converts the SBP shift to an incidence change. |
| Scale-up schedule | user controls in `07_run_interventions.R` | Linear scale-up from `r SCALEUP_YEAR1` to `r SCALEUP_YEAR2`. |

---

# Methods

## Overall structure

Let $t$ index calendar years and $a$ single-year ages (20–95). For each
country, sex, cause and scenario, the model tracks three states — **Well**,
**Sick** (prevalent disease) and **Dead** — and advances them one year at a
time. Disease dynamics are governed by three age/sex/cause/year-specific rates
carried in `b_rates`:

- $IR$ — incidence rate (Well → Sick),
- $CF$ — case fatality (Sick → Dead from the disease),
- background mortality $\text{BG.mx}$ (applied to the well/at-risk population),
  a disease-envelope term $\text{BG.mx.all}$, and a COVID excess term
  $\text{covid.mx}$ for 2020–2021.

**Initialisation.** At the seed year (2017) and at the entry age (20) the states
are set from GBD prevalence and mortality:

$$
\text{sick} = N_x\,\text{PREV}_{t_0}, \qquad
\text{dead} = N_x\,\text{DIS.mx}_{t_0}, \qquad
\text{well} = N_x\bigl(1 - (\text{PREV}_{t_0} + \text{BG.mx})\bigr),
$$

with population $\text{pop} = N_x$ and $\text{all.mx} = N_x\,\text{DIS.mx}_{t_0}
+ N_x\,\text{BG.mx}$.

**Annual recurrences.** Writing $x_{t-1}$ for last year's value of state $x$ in
the cohort that ages into year $t$, the transitions implemented in
`project.all()` are:

$$
\begin{aligned}
\text{newcases}_t &= \text{well}_{t-1}\cdot IR \\[2pt]
\text{sick}_t     &= \text{sick}_{t-1}\bigl(1 - (CF + \text{BG.mx} + \text{covid.mx})\bigr) + \text{well}_{t-1}\cdot IR \\[2pt]
\text{dead}_t     &= \text{sick}_{t-1}\cdot CF \\[2pt]
\text{pop}_t      &= \text{pop}_{t-1} - \text{all.mx}_{t-1} \\[2pt]
\text{all.mx}_t   &= \sum_{\text{causes}} \text{dead}_t \;+\; \text{pop}_t\cdot \text{BG.mx.all} \;+\; \text{pop}_t\cdot \text{covid.mx} \\[2pt]
\text{well}_t     &= \text{pop}_t - \text{all.mx}_t - \text{sick}_t
\end{aligned}
$$

Every quantity is floored at zero, ages advance by one year per step, and the
loop runs up to 41 annual steps (2017 → 2058). In the current build the baseline
rate panel spans 2017–2050, so populated outputs cover **`r study_years`**.

**Aggregation semantics (important).** The four causes are modelled as
independent Well/Sick/Dead processes that *share* one population. In the
outputs this means **`pop` and `all.mx` are identical across the four cause
rows** (the shared population and its total mortality), while **`sick`,
`well`, `dead` and `newcases` are cause-specific**. Consequently:

- disease **counts** (deaths, incident cases, prevalent cases) are summed over
  causes, but
- the **population denominator** is taken from a *single* cause row (summing
  `pop` over the four causes would quadruple-count it); equivalently,
  $\text{well}_c + \text{sick}_c = \text{pop} - \text{all.mx}$ is the total
  living population and is the same for every cause $c$.

All rates in this report follow those rules.

## Source of sodium

Baseline sodium intake is partitioned into **five mutually exclusive sources**
whose shares sum to 1.0 in every country:

$$
\text{discretionary} + \text{packaged} + \text{restaurant} + \text{public} + \text{inherent} = 1 .
$$

`inherent` sodium (naturally present in unprocessed foods) is **never targeted**
by any intervention. Each policy acts only on the source(s) it can plausibly
reach. Country source shares are taken from the RTSL sodium policy document; a
`default` row (the simple mean of the ten documented countries) is used for
any country without its own row, and **every row is renormalised to sum to
exactly 1.0** (handling documented rounding, e.g. Thailand's shares sum to
0.999). **Malaysia uses the Philippines source profile** as a proxy, and
**Nigeria/Cameroon** use the provisional shares supplied by RTSL/Matti (with
bouillon and condiments counted as packaged; Nigeria is used as a proxy for
Cameroon).

```{r tbl-source-shares, message=FALSE, warning=FALSE, echo=FALSE}
ss_disp <- copy(source_shares)
setnames(ss_disp,
         c("location", SODIUM_SOURCES),
         c("Country", "Discretionary", "Packaged", "Restaurant", "Public", "Inherent"))
num_cols <- c("Discretionary", "Packaged", "Restaurant", "Public", "Inherent")
ss_disp[, (num_cols) := lapply(.SD, function(x) formatC(x, format = "f", digits = 3)),
        .SDcols = num_cols]

kable(ss_disp,
      caption = paste0("Table M1. Per-country sodium source shares (fractions of ",
                       "total intake), renormalised to 1.0. The 'default' row is the ",
                       "mean across the ten documented countries and is used for all ",
                       "other countries."),
      align = c("l", rep("r", 5))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(nrow(ss_disp), italic = TRUE) |>
  footnote(general = paste0(
    "Source: RTSL sodium policy document. 'inherent' sodium is never targeted. ",
    "Malaysia uses the Philippines profile as a proxy."))
```

The same shares are shown below as a per-country stacked bar. Each bar sums to
1.0. All ten prioritised countries now have documented shares (Nigeria and
Cameroon on provisional RTSL/Matti values); any other country would resolve to
the `default` (mean) profile.

```{r fig-source-shares-bar, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# Countries to display: those actually modelled (includes Nigeria & Cameroon),
# with a fall-back to the documented list if outputs are unavailable.
plot_countries <- sort(unique(as.character(data.out$location)))
if (length(plot_countries) == 0)
  plot_countries <- setdiff(source_shares$location, "default")

shares_bar <- resolve_source_shares(plot_countries)
shares_long <- melt(shares_bar,
                    id.vars = "location",
                    measure.vars = SODIUM_SOURCES,
                    variable.name = "source", value.name = "share")
shares_long[, source := factor(source, levels = SODIUM_SOURCES)]

p <- ggplot(shares_long, aes(location, share, fill = source)) +
  geom_col(width = 0.78, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Sodium source shares by country",
       subtitle = "Fraction of total baseline sodium intake, by source (bars sum to 100%)",
       x = NULL, y = "Share of total sodium intake",
       caption = paste0("Documented shares for the ten prioritised countries ",
                        "(Nigeria/Cameroon on provisional RTSL values).")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1))
print(p)
```

## Effect sizes

Each intervention has a documented **relative reduction** applied to its target
source at full implementation. Multiple policies may target the same source
(notably `packaged`).

```{r tbl-effect-sizes, message=FALSE, warning=FALSE, echo=FALSE}
eff_disp <- merge(intervention_effects,
                  intervention_registry[, .(intervention, label, placeholder)],
                  by = "intervention", all.x = TRUE)
eff_disp <- eff_disp[, .(
  Intervention = label,
  `Target source` = source,
  `Relative reduction` = fmt_pct(effect, 1),
  Note = fifelse(intervention == "lss",
                 "SSaSS-derived; hypertensive reach; in package",
                 "")
)]
# canonical order
ord <- c("Public food procurement & service",
         "Behaviour change / media campaigns",
         "Behaviour change / media campaigns",
         "Front-of-pack labelling (FOPL)",
         "Salt targets (reformulation)",
         "Fiscal policy (salt tax)",
         "Low-sodium salt substitutes (LSS)")
eff_disp[, Intervention := factor(Intervention, levels = unique(ord))]
setorder(eff_disp, Intervention, `Target source`)

kable(eff_disp,
      caption = "Table M2. Intervention → source effect sizes (relative reduction of the target source at full implementation).",
      align = c("l", "l", "r", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "Media campaigns act on both discretionary and packaged sodium (1.0% each). ",
    "FOPL, salt targets and fiscal policy all act on packaged sodium and are ",
    "composed multiplicatively when combined. LSS acts on discretionary sodium ",
    "(SSaSS-derived, 15%) and is a full component of the package. Source: RTSL ",
    "sodium policy effect-size tables; LSS from Neal et al., NEJM 2021."))
```

## Low-sodium salt substitutes (LSS)

LSS is no longer a placeholder: it is now a fully specified model component
derived from the **Salt Substitute and Stroke Study** (SSaSS; Neal et al.,
*NEJM* 2021;385:1067–77). In SSaSS, replacing regular salt with a 75% NaCl /
25% KCl substitute delivered an intention-to-treat net reduction of about
**−350 mg/day sodium** against a baseline intake of roughly **4.3 g/day**.
Referred to the discretionary source that a salt substitute can actually reach
(discretionary sodium ≈ 59% of intake in the SSaSS setting), this corresponds
to a **≈ 15% relative reduction of discretionary sodium**
(`EFF_LSS_DISCRETIONARY = 0.15`), which is exactly how LSS enters Table M2. A
**sensitivity range of 0.12–0.18** brackets the point estimate.

Two modelling choices make this a **deliberately conservative lower bound**:

- **Hypertensive-only reach.** Following the SSaSS enrolment criteria
  (prior stroke, or age ≥ 60 with raised blood pressure), the LSS reduction is
  applied **only to the raised-BP subpopulation** (SBP ≥ 140 mmHg bins). In the
  code this is the `lss_only` flag, which zeroes the reduction in
  normotensive BP bins inside `calculate_sodium_impact_etihad()`. This
  restriction is active when LSS is run as its **own** scenario; within the
  **full package** LSS is not isolated, so its discretionary reduction is
  composed with the other policies across all bins (see the caveat in the
  code).
- **No potassium channel.** This model represents the sodium arm only. SSaSS
  also raised urinary potassium by ≈ 803 mg/day, and that potassium increase
  contributed a large share of the trial's observed **−3.34 mmHg** systolic
  effect. Because the model has **no K⁺ pathway**, it captures only the
  sodium-mediated portion of the blood-pressure benefit and therefore
  **understates** the true LSS effect.

## Composition math

For a **package** of interventions, effects on the *same* source are composed
so that overlapping policies never double-count or exceed 100%. For source $s$
with intervention effects $\text{eff}_{i,s}$:

$$
\text{composed}_s \;=\; 1 - \prod_{i}\bigl(1 - \text{eff}_{i,s}\bigr).
$$

The composed per-source effects are then **share-weighted** and summed to give
the single total-intake reduction fraction — exactly the `salteff` the
downstream pathway expects:

$$
\boxed{\;\text{salteff} \;=\; \sum_{s}\, \text{share}_s \cdot \text{composed}_s\;}
$$

**Worked sanity check (asserted in code).** For public procurement in Viet Nam,
only the `public` source is affected (share 0.05, effect 20%), so
$\text{salteff} = 0.20 \times 0.05 = 0.01$.

```{r sanity-check, message=FALSE, warning=FALSE, echo=FALSE}
chk <- as.numeric(compute_total_efficacy("public_procurement", "Viet Nam",
                                         source_shares, intervention_effects))
```

The in-report recomputation returns
$\text{salteff}_{\text{Viet Nam, public procurement}} =$
**`r formatC(chk, format = "f", digits = 4)`**, matching $0.20 \times 0.05$.

## From `salteff` to health impact

Given a country's `salteff`, the intervention reduces sodium intake and — through
blood pressure — CVD incidence. Case fatality is **not** modified.

1. **Target reduction (grams).** With `salt` the baseline sodium column and the
   `"percent"` method used here, the targeted reduction is
   $\text{salt\_target} = \text{salt}\cdot\text{salteff}$, subject to a
   **2 g/day minimum-intake floor**: if $\text{salt} - \text{salt\_target} < 2$
   then $\text{salt\_target} = \text{salt} - 2$.

2. **Linear scale-up.** Between `saltyear1` = `r SCALEUP_YEAR1` and `saltyear2`
   = `r SCALEUP_YEAR2` the reduction ramps linearly, then holds:
   $$
   \text{salt\_reduction}_t =
   \begin{cases}
   0 & t < \text{saltyear1}\\[4pt]
   \text{salt\_target}\cdot\dfrac{t - \text{saltyear1} + 1}{\text{saltyear2}-\text{saltyear1}+1} & \text{saltyear1}\le t \le \text{saltyear2}\\[10pt]
   \text{salt\_target} & t > \text{saltyear2}.
   \end{cases}
   $$

3. **Sodium → SBP (Filippini dose–response).** The SBP reduction depends on
   whether the individual has raised BP (fraction $\text{raisedBP}$):
   $$
   \Delta SBP = \bigl(2.8\cdot\text{raisedBP} + 1.0\cdot(1-\text{raisedBP})\bigr)\cdot \text{salt\_reduction}.
   $$

4. **Baseline bin-specific incidence (GBD).** Within each BP bin $b$, baseline
   incidence is scaled by the GBD log-linear RR per 10 mmHg, normalised so the
   population-average recovers the observed $IR$:
   $$
   IR_{\text{bin},b} = \frac{RR^{\text{GBD}}_{b}\cdot IR}{\alpha},
   \qquad \alpha = \sum_b \text{prob}_b\, RR^{\text{GBD}}_{b},
   $$
   where $RR^{\text{GBD}}_b = \bigl(rr^{10}\bigr)^{(m_b-120)/10}$ for bin
   midpoint $m_b>120$ (else 1) and $\text{prob}_b$ is the share of the age/sex
   group in bin $b$ (from a normal BP distribution).

5. **SBP → incidence (ETTEHAD).** The proportional reduction in bin-specific
   incidence is driven by the ETTEHAD relative risk per 10 mmHg,
   $RR^{10}_{b,c}$, for BP bin $b$ and cause $c$, applied in proportion to the
   SBP reduction; incidence is then averaged over BP bins:
   $$
   IR^{\text{new}}_{c} = \sum_b \text{prob}_b\; IR_{\text{bin},b}\,
   \Bigl(1 - \bigl(1 - RR^{10}_{b,c}\bigr)\cdot 0.1\cdot \Delta SBP\Bigr),
   \qquad
   \text{eff\_ir} = \frac{IR^{\text{new}}_c}{IR_c}.
   $$
   The factor $0.1$ converts $\Delta SBP$ (mmHg) into units of 10 mmHg. The
   ETTEHAD inputs include diabetes-specific effect sizes that the pipeline
   blends by local diabetes prevalence; the applied incidence modifier is the
   per-10-mmHg RR above.

6. **Case fatality unchanged.** $CF^{\text{new}} = CF$ and $\text{eff\_cf}=1$:
   sodium reduction acts on incidence only. Before `saltyear1`,
   $IR^{\text{new}} = IR$.

The modified $IR$ then flows through the same state-transition recurrences given
above.

```{r flow-diagram, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=2.6}
boxes <- data.table(
  x = 1:6,
  label = c("Source shares\n× effect sizes",
            "salteff\nΣ share·composed",
            "ΔSBP\n(Filippini)",
            "ΔIncidence\n(ETTEHAD RR)",
            "State transitions\nWell→Sick→Dead",
            "Deaths\ndelayed")
)
arrows <- data.table(x = (1:5) + 0.42, xend = (2:6) - 0.42, y = 1, yend = 1)

ggplot() +
  geom_tile(data = boxes, aes(x = x, y = 1), width = 0.84, height = 0.7,
            fill = "#e8eef4", color = "#1f78b4", linewidth = 0.6) +
  geom_text(data = boxes, aes(x = x, y = 1, label = label), size = 3.1, lineheight = 0.95) +
  geom_segment(data = arrows, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.16, "cm"), type = "closed"),
               linewidth = 0.5, color = "grey30") +
  coord_cartesian(xlim = c(0.5, 6.5), ylim = c(0.55, 1.45)) +
  labs(title = "Causal chain: sources → salteff → SBP → incidence → deaths delayed") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 11))
```

## Scenarios

The model's user controls (`selected_interventions = "all"`,
`scenario_mode = "both"`) generate a **baseline**, one scenario per
**individual** intervention, and a combined **`full_package`**. With
`include_placeholders_in_package <- TRUE`, the package now comprises **all six
interventions** — `public_procurement + media_campaign + fopl + salt_targets +
fiscal + lss`. LSS is a full member of the package (its SSaSS-derived effect
size is specified above); it also runs as its own individual scenario, where its
reduction is restricted to the hypertensive subpopulation.

```{r tbl-scenarios, message=FALSE, warning=FALSE, echo=FALSE}
scn_tbl <- data.table(
  scenario = scenario_levels,
  interventions = vapply(scenario_levels, function(s) {
    iv <- scenario_interventions[[s]]
    if (length(iv) == 0) "(none — baseline)" else paste(iv, collapse = " + ")
  }, character(1))
)
scn_tbl[, Scenario := scenario_labels[scenario]]
scn_tbl[, present := scenario %in% scenarios_present]
scn_disp <- scn_tbl[, .(Scenario, `Interventions (source-split)` = interventions,
                        `In outputs` = fifelse(present, "yes", "no"))]

kable(scn_disp,
      caption = "Table M3. Scenarios and their constituent interventions.",
      align = c("l", "l", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(which(scn_tbl$scenario %in% c("baseline", "full_package")), bold = TRUE)
```

The resulting `salteff` is **country-specific** because it depends on the local
source shares. Table M4 illustrates this for a few representative countries
(recomputed in-report from Tables M1–M2).

```{r tbl-salteff, message=FALSE, warning=FALSE, echo=FALSE}
ref_countries <- intersect(c("Viet Nam", "Bangladesh", "China", "Thailand", "default"),
                           source_shares$location)
salteff_grid <- rbindlist(lapply(scenario_nb, function(s) {
  iv <- scenario_interventions[[s]]
  vals <- vapply(ref_countries, function(cc)
    as.numeric(compute_total_efficacy(iv, cc, source_shares, intervention_effects)),
    numeric(1))
  data.table(Scenario = scenario_labels[s], t(vals))
}))
setnames(salteff_grid, c("Scenario", ref_countries))
num_cols <- ref_countries
salteff_grid[, (num_cols) := lapply(.SD, function(x) formatC(x, format = "f", digits = 4)),
             .SDcols = num_cols]

kable(salteff_grid,
      caption = paste0("Table M4. Illustrative salteff (fraction of TOTAL sodium ",
                       "reduced at full implementation) by scenario and country."),
      align = c("l", rep("r", length(ref_countries)))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "salteff is largest where the intervention's target source dominates local ",
    "intake (e.g. FOPL/salt-targets/fiscal in packaged-heavy Viet Nam; LSS in ",
    "discretionary-heavy Bangladesh/China). The LSS column is the share-weighted ",
    "discretionary reduction; run as its own scenario LSS reaches only the ",
    "hypertensive subpopulation, so its realised sodium reduction is smaller ",
    "than this illustrative value."))
```

The stacked bars below show the same illustrative `salteff` for **every
modelled country** (all ten prioritised countries, each on its own documented
source profile), one panel per scenario. Each bar's total height is the country's `salteff`
(fraction of *total* sodium reduced at full implementation); the segments
decompose it into the per-source contributions ($\text{share}_s \times
\text{composed}_s$) that sum to that total.

```{r fig-salteff-bar, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7}
# Per-(scenario, country) source decomposition of salteff, for all modelled
# countries. Uses the same source colours as the source-share barplot.
salteff_bar <- rbindlist(lapply(scenario_nb, function(s) {
  iv <- scenario_interventions[[s]]
  rbindlist(lapply(plot_countries, function(cc) {
    te  <- compute_total_efficacy(iv, cc, source_shares, intervention_effects)
    dec <- copy(attr(te, "decomposition"))
    dec[, `:=`(scenario = s, location = cc)]
    dec[]
  }))
}))
salteff_bar[, source   := factor(source, levels = SODIUM_SOURCES)]
salteff_bar[, scn_lbl  := factor(scenario_labels_short[scenario],
                                 levels = scenario_labels_short[scenario_nb])]

p <- ggplot(salteff_bar[scenario!="full_package",], aes(location, contribution, fill = source)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.15) +
  facet_wrap(~ scn_lbl, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Illustrative salteff by scenario and country",
       subtitle = paste0("Fraction of TOTAL sodium reduced at full ",
                         "implementation; stacked by source contribution"),
       x = NULL, y = "salteff (fraction of total sodium reduced)",
       caption = paste0("Bar height = country salteff; note free y-axes across ",
                        "panels. LSS shown at its share-weighted value.")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
print(p)
```

```{r fig-salteff-bar2, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7}
# Per-(scenario, country) source decomposition of salteff, for all modelled
# countries. Uses the same source colours as the source-share barplot.
p <- ggplot(salteff_bar[scenario=="full_package",], aes(location, contribution, fill = source)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.15) +
  facet_wrap(~ scn_lbl, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Illustrative salteff by scenario and country",
       subtitle = paste0("Fraction of TOTAL sodium reduced at full ",
                         "implementation; stacked by source contribution"),
       x = NULL, y = "salteff (fraction of total sodium reduced)",
       caption = paste0("Bar height = country salteff; note free y-axes across ",
                        "panels. LSS shown at its share-weighted value.")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
print(p)
```

---

# Baseline (pre-intervention) time series

The panels below use the **`baseline`** scenario only. Population denominators
use the de-duplicated total population (a single cause row); counts sum
cause-specific values over the four causes. Incidence begins in 2018 (2017 is
the seed year, in which `newcases` = 0 by construction).

```{r baseline-aggregates, message=FALSE, warning=FALSE, echo=FALSE}
base <- data.out[scenario == "baseline"]

# Total (de-duplicated) population per year: take one cause row.
base_pop  <- base[cause == "ihd", .(pop = sum(pop, na.rm = TRUE)), by = year]

# All-cause counts per year.
base_all  <- base[, .(sick = sum(sick, na.rm = TRUE),
                      newcases = sum(newcases, na.rm = TRUE),
                      dead = sum(dead, na.rm = TRUE)), by = year]
base_all  <- merge(base_all, base_pop, by = "year")
base_all[, `:=`(prev_rate = sick / pop * 1000,
                inc_rate  = newcases / pop * 1000,
                mort_rate = dead / pop * 1e5)]

# By-cause counts per year (denominator = total population, same for each cause).
base_cause <- base[, .(sick = sum(sick, na.rm = TRUE),
                       newcases = sum(newcases, na.rm = TRUE),
                       dead = sum(dead, na.rm = TRUE)), by = .(cause, year)]
base_cause <- merge(base_cause, base_pop, by = "year")
base_cause[, `:=`(prev_rate = sick / pop * 1000,
                  inc_rate  = newcases / pop * 1000,
                  mort_rate = dead / pop * 1e5)]
base_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]

hist_cut <- 2024  # solid = historical, dashed = projected
```

## Prevalence over time

```{r baseline-prevalence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
p_all <- ggplot(base_all, aes(year, prev_rate, linetype = year > hist_cut)) +
  geom_line(linewidth = 1.1, color = "#b2182b") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "Prevalent CVD cases per 1,000 population — baseline (all causes)",
       subtitle = paste0(n_countries, " countries; solid = historical, dashed = projected"),
       x = "Year", y = "Prevalent cases per 1,000",
       caption = "Sum of cause-specific prevalent cases / total population.") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
print(p_all)

p_cause <- ggplot(base_cause, aes(year, prev_rate, color = cause_lbl,
                                  linetype = year > hist_cut)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(title = "Prevalent CVD cases per 1,000 population — baseline, by cause",
       x = "Year", y = "Prevalent cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p_cause)
```

## Incidence over time

```{r baseline-incidence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
inc_all   <- base_all[year >= 2018]
inc_cause <- base_cause[year >= 2018]

p_all <- ggplot(inc_all, aes(year, inc_rate, linetype = year > hist_cut)) +
  geom_line(linewidth = 1.1, color = "#2166ac") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "Incident CVD cases per 1,000 population — baseline (all causes)",
       subtitle = "New cases per 1,000 population per year; 2017 seed year omitted",
       x = "Year", y = "Incident cases per 1,000",
       caption = "Sum of cause-specific new cases / total population.") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
print(p_all)

p_cause <- ggplot(inc_cause, aes(year, inc_rate, color = cause_lbl,
                                 linetype = year > hist_cut)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(title = "Incident CVD cases per 1,000 population — baseline, by cause",
       x = "Year", y = "Incident cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p_cause)
```

## Mortality over time

```{r baseline-mortality, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
p_all <- ggplot(base_all, aes(year, dead / 1e6, linetype = year > hist_cut)) +
  geom_line(linewidth = 1.1, color = "#4d4d4d") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — baseline (all causes)",
       subtitle = paste0(n_countries, " countries; millions of deaths per year"),
       x = "Year", y = "Deaths per year (millions)",
       caption = "Sum of cause-specific deaths.") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
print(p_all)

p_cause <- ggplot(base_cause, aes(year, dead / 1e6, color = cause_lbl,
                                  linetype = year > hist_cut)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — baseline, by cause",
       x = "Year", y = "Deaths per year (millions)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p_cause)
```

## Age-standardised mortality rate (ASMR)

```{r baseline-asmr, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# All-cause ASMR: deaths summed over causes; Nx = de-duplicated population.
asmr_all_in <- base[, .(deaths_dx = sum(dead, na.rm = TRUE)), by = .(age, year)]
pop_ay      <- base[cause == "ihd", .(Nx = sum(pop, na.rm = TRUE)), by = .(age, year)]
asmr_all_in <- merge(asmr_all_in, pop_ay, by = c("age", "year"))
asmr_all_in[, grp := "All CVD"]
asmr_all <- calculate_asmr(asmr_all_in, disagg_var = "grp")

# By-cause ASMR: Nx = well+sick (per cause = total living population).
asmr_c_in <- base[, .(deaths_dx = sum(dead, na.rm = TRUE),
                      Nx = sum(well, na.rm = TRUE) + sum(sick, na.rm = TRUE)),
                  by = .(cause, age, year)]
asmr_cause <- calculate_asmr(asmr_c_in, disagg_var = "cause")
asmr_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]

p <- ggplot() +
  geom_line(data = asmr_cause, aes(year, ASMR, color = cause_lbl,
                                   linetype = year > hist_cut), linewidth = 1) +
  geom_line(data = asmr_all, aes(year, ASMR, linetype = year > hist_cut),
            color = "black", linewidth = 1.2) +
  annotate("text", x = year_min + 1, y = max(asmr_all$ASMR),
           label = "All CVD", hjust = 0, vjust = 1, fontface = "bold", size = 3.6) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(title = "Age-standardised CVD mortality rate — baseline",
       subtitle = "Per 100,000, WHO world standard population (ages 20–95+)",
       x = "Year", y = "ASMR per 100,000",
       caption = "Black = all CVD; coloured = by cause. Solid = historical, dashed = projected.") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

---

# Projected time series (all scenarios)

The same three metrics, now overlaying **every scenario** — baseline, each
individual intervention, and the full package — across the projection horizon.
Interventions begin scaling up in `r SCALEUP_YEAR1`, so scenarios coincide
before then and fan out thereafter.

```{r projection-aggregates, message=FALSE, warning=FALSE, echo=FALSE}
# All-cause per scenario x year.
proj_pop <- data.out[cause == "ihd", .(pop = sum(pop, na.rm = TRUE)),
                     by = .(scenario, year)]
proj_all <- data.out[, .(sick = sum(sick, na.rm = TRUE),
                         newcases = sum(newcases, na.rm = TRUE),
                         dead = sum(dead, na.rm = TRUE)),
                     by = .(scenario, year)]
proj_all <- merge(proj_all, proj_pop, by = c("scenario", "year"))
proj_all[, `:=`(prev_rate = sick / pop * 1000,
                inc_rate  = newcases / pop * 1000,
                mort_rate = dead / pop * 1e5)]

# By cause per scenario x year.
proj_cause <- data.out[, .(sick = sum(sick, na.rm = TRUE),
                           newcases = sum(newcases, na.rm = TRUE),
                           dead = sum(dead, na.rm = TRUE)),
                       by = .(scenario, cause, year)]
proj_cause <- merge(proj_cause, proj_pop, by = c("scenario", "year"))
proj_cause[, `:=`(prev_rate = sick / pop * 1000,
                  inc_rate  = newcases / pop * 1000,
                  mort_rate = dead / pop * 1e5)]
proj_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]

scn_scale <- function(name = NULL)
  list(
    scale_color_manual(values = scenario_cols, labels = scenario_labels_short,
                       breaks = scenario_levels, name = name),
    guides(color = guide_legend(nrow = 2, byrow = TRUE))
  )
```

## Prevalence

```{r proj-prevalence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
p <- ggplot(proj_all[year >= 2020], aes(year, prev_rate, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  labs(title = "Prevalent CVD cases per 1,000 population — all scenarios",
       subtitle = "All causes combined", x = "Year", y = "Prevalent cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)

p <- ggplot(proj_cause[year >= 2020], aes(year, prev_rate, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ cause_lbl, scales = "free_y") +
  scn_scale() +
  labs(title = "Prevalent CVD cases per 1,000 population — by cause",
       x = "Year", y = "Prevalent cases per 1,000") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

## Incidence

```{r proj-incidence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
p <- ggplot(proj_all[year >= 2020], aes(year, inc_rate, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  labs(title = "Incident CVD cases per 1,000 population — all scenarios",
       subtitle = "All causes combined", x = "Year", y = "Incident cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)

p <- ggplot(proj_cause[year >= 2020], aes(year, inc_rate, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ cause_lbl, scales = "free_y") +
  scn_scale() +
  labs(title = "Incident CVD cases per 1,000 population — by cause",
       x = "Year", y = "Incident cases per 1,000") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

To isolate the **relative** change each policy drives — stripping out the shared
demographic trend — the figure below re-expresses all-cause incidence per 1,000
as an **index** with the intervention/base year (`r SCALEUP_YEAR1`) set to 100
for every scenario. Scenarios coincide at the base year and then fan out as the
intervention scales up; lines below 100 indicate incidence falling faster than
its base-year level.

```{r proj-incidence-indexed, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
# Index each scenario's all-cause incidence rate to its base-year value = 100.
base_idx_year <- SCALEUP_YEAR1
inc_idx <- proj_all[year >= base_idx_year, .(scenario, year, inc_rate)]
base_inc <- inc_idx[year == base_idx_year, .(scenario, base_val = inc_rate)]
inc_idx  <- merge(inc_idx, base_inc, by = "scenario", all.x = TRUE)
inc_idx  <- inc_idx[!is.na(base_val) & base_val > 0]
inc_idx[, inc_index := inc_rate / base_val * 100]

if (nrow(inc_idx) > 0) {
  p <- ggplot(inc_idx, aes(year, inc_index, color = scenario)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1) +
    scn_scale() +
    labs(title = "Incident CVD cases per 1,000 — indexed to base year",
         subtitle = paste0("All causes; ", base_idx_year,
                          " = 100 for every scenario (relative change over time)"),
         x = "Year", y = paste0("Incidence index (", base_idx_year, " = 100)")) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank())
  print(p)
} else {
  cat("Indexed incidence series unavailable (no base-year data).\n")
}
```

## Mortality

```{r proj-mortality, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
p <- ggplot(proj_all[year >= 2020], aes(year, dead / 1e6, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — all scenarios",
       subtitle = "All causes combined (millions per year)",
       x = "Year", y = "Deaths per year (millions)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)

p <- ggplot(proj_cause[year >= 2020], aes(year, dead / 1e6, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ cause_lbl, scales = "free_y") +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — by cause",
       x = "Year", y = "Deaths per year (millions)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

---

# Main results: deaths delayed {#main-results-deaths-delayed}

The headline outcome is **deaths delayed** relative to baseline, keyed on the
authoritative `scenario` column:

$$
\text{deaths\_delayed}_{\text{scenario}} = \text{deaths}_{\text{baseline}} - \text{deaths}_{\text{scenario}} .
$$

Cumulative totals are summed over **`r impact_window`** (the scale-up starts in
`r SCALEUP_YEAR1`, so earlier years contribute negligibly).

```{r deaths-delayed-compute, message=FALSE, warning=FALSE, echo=FALSE}
# Cumulative deaths delayed over the impact window, by scenario.
dd_cum <- deaths_delayed_by(data.out, by = character(0), years = impact_y1:impact_y2)
dd_cum[, scenario := as_scn_factor(as.character(scenario))]
setorder(dd_cum, scenario)
dd_cum[, `:=`(per_year = deaths_delayed / n_impact_years,
              pct_vs_base = deaths_delayed / baseline_dead)]

# Annual + cumulative time series, by scenario (full horizon).
dd_annual <- deaths_delayed_by(data.out, by = "year")
dd_annual[, scenario := as_scn_factor(as.character(scenario))]
setorder(dd_annual, scenario, year)
dd_annual[, cum_delayed := cumsum(deaths_delayed), by = scenario]

# Helper for inline numbers.
ddc <- function(scn) dd_cum[as.character(scenario) == scn, deaths_delayed]
```

Over `r impact_window`, across `r n_countries` countries:

- The **full package** delays **`r fmt_m(ddc("full_package"))` million** CVD
  deaths (`r fmt_pct(dd_cum[scenario=="full_package", pct_vs_base])` of baseline
  CVD deaths in the window; ~`r fmt_k(dd_cum[scenario=="full_package", per_year])`
  thousand per year).
- Among the packaged-food policies, **salt targets** contribute most
  (`r fmt_m(ddc("salt_targets"))` million), followed by **FOPL**
  (`r fmt_m(ddc("fopl"))` million) and **fiscal policy**
  (`r fmt_m(ddc("fiscal"))` million).
- **Public procurement** (`r fmt_m(ddc("public_procurement"))` million) and
  **media campaigns** (`r fmt_m(ddc("media_campaign"))` million) are smaller.
- **LSS** (SSaSS-derived) delays `r fmt_m(ddc("lss"))` million deaths as its own
  scenario, where its reduction reaches only the hypertensive subpopulation.
  It targets *discretionary* salt, which dominates intake in the most populous
  countries. LSS is now a full component of the **full package** above; because
  the model has no potassium channel, this is a **conservative lower bound** on
  its true effect.

## Table 1 — Cumulative deaths delayed by scenario

```{r table1, message=FALSE, warning=FALSE, echo=FALSE}
tab1 <- dd_cum[as.character(scenario) != "baseline"]
tab1[, Scenario := scenario_labels[as.character(scenario)]]
tab1[, is_pkg := as.character(scenario) == "full_package"]
setorder(tab1, scenario)

tab1_fmt <- tab1[, .(
  Scenario,
  `Interventions` = vapply(as.character(scenario), function(s) {
    iv <- scenario_interventions[[s]]; paste(iv, collapse = " + ")
  }, character(1)),
  `Deaths delayed (millions)` = fmt_m(deaths_delayed, 3),
  `Avg per year (thousands)`  = fmt_k(per_year, 1),
  `% of baseline`             = fmt_pct(pct_vs_base, 2)
)]

kable(tab1_fmt,
      caption = paste0("Table 1. Cumulative deaths delayed by scenario, ",
                       impact_window, ", across ", n_countries, " countries."),
      align = c("l", "l", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(which(tab1$is_pkg), bold = TRUE, background = "#eef2f7") |>
  footnote(general = paste0(
    "Deaths delayed = baseline deaths - scenario deaths, summed over ",
    impact_window, ". '% of baseline' is relative to baseline CVD deaths in ",
    "the same window. LSS (SSaSS-derived) is included in the full package; as ",
    "an individual scenario it reaches only the hypertensive subpopulation."))
```

```{r table1-dt, message=FALSE, warning=FALSE, echo=FALSE}
dt_show <- tab1[, .(
  Scenario,
  `Deaths delayed`   = round(deaths_delayed),
  `Avg per year`     = round(per_year),
  `% of baseline`    = round(100 * pct_vs_base, 2)
)]
datatable(dt_show, rownames = FALSE,
          caption = "Interactive version of Table 1 (absolute counts).",
          options = list(pageLength = 8, dom = "t")) |>
  formatCurrency(c("Deaths delayed", "Avg per year"), currency = "", digits = 0)
```

## Figure 1 — Cumulative deaths delayed by scenario

```{r figure1-bar, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
fig1 <- dd_cum[as.character(scenario) != "baseline"]
fig1[, scn_lbl := factor(scenario_labels_short[as.character(scenario)],
                         levels = rev(scenario_labels_short[scenario_nb]))]

p <- ggplot(fig1, aes(deaths_delayed / 1e6, scn_lbl, fill = scenario)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.4) +
  scale_fill_manual(values = scenario_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), labels = comma) +
  labs(title = paste0("Cumulative deaths delayed by scenario, ", impact_window),
       subtitle = paste0(n_countries, " countries; all CVD causes combined"),
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
print(p)
```

## Figure 2 — Annual deaths delayed over time

```{r figure2-annual, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
p <- ggplot(dd_annual[as.character(scenario) != "baseline" & year >= 2024],
            aes(year, deaths_delayed / 1e3, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "Annual deaths delayed by scenario",
       subtitle = "Thousands of CVD deaths delayed per year vs. baseline",
       x = "Year", y = "Deaths delayed per year (thousands)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

The absolute magnitudes above are dominated by the largest scenarios. Indexing
each scenario's annual deaths-delayed series to a common base year = 100 removes
that scale difference and exposes the **relative change per year** — how quickly
each scenario's annual benefit grows once the intervention takes hold. The base
year is the first post-scale-up year in which every scenario has a non-zero
annual benefit (deaths delayed in the very first scale-up year are ≈ 0 because
mortality responds to the prior year's incidence).

```{r figure2b-annual-indexed, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# Index annual deaths delayed to a base intervention year = 100, per scenario.
dd_nb <- dd_annual[as.character(scenario) != "baseline"]
n_scn <- uniqueN(as.character(dd_nb$scenario))

# Base year: earliest year >= scale-up start where ALL scenarios are non-zero.
ok_years <- dd_nb[year >= SCALEUP_YEAR1 & deaths_delayed > 0, .N,
                  by = year][N == n_scn, year]
base_dd_year <- if (length(ok_years) > 0) min(ok_years) else
  suppressWarnings(min(dd_nb[deaths_delayed > 0, year]))

if (is.finite(base_dd_year)) {
  base_dd <- dd_nb[year == base_dd_year, .(scenario, base_val = deaths_delayed)]
  dd_idx  <- merge(dd_nb[year >= base_dd_year], base_dd, by = "scenario", all.x = TRUE)
  dd_idx  <- dd_idx[!is.na(base_val) & base_val > 0]
  dd_idx[, dd_index := deaths_delayed / base_val * 100]

  p <- ggplot(dd_idx, aes(year, dd_index, color = scenario)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1) +
    scn_scale() +
    scale_y_continuous(labels = comma) +
    labs(title = "Annual deaths delayed — indexed to base intervention year",
         subtitle = paste0(base_dd_year,
                          " = 100 for every scenario (relative change per year)"),
         x = "Year", y = paste0("Deaths-delayed index (", base_dd_year, " = 100)")) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank())
  print(p)
} else {
  cat("Indexed deaths-delayed series unavailable (no non-zero base year).\n")
}
```

## Figure 3 — Cumulative deaths delayed over time

```{r figure3-cumulative, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
p <- ggplot(dd_annual[as.character(scenario) != "baseline"],
            aes(year, cum_delayed / 1e6, color = scenario)) +
  geom_line(linewidth = 1.1) +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "Cumulative deaths delayed by scenario",
       subtitle = "Millions of CVD deaths delayed vs. baseline (running total)",
       x = "Year", y = "Cumulative deaths delayed (millions)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

## Figure 4 & Table 2 — Full package, by cause

```{r figure4-cause, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
dd_cause <- deaths_delayed_by(data.out, by = "cause", years = impact_y1:impact_y2)
dd_cause <- dd_cause[as.character(scenario) == "full_package"]
dd_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]
setorder(dd_cause, -deaths_delayed)

p <- ggplot(dd_cause, aes(deaths_delayed / 1e6,
                          reorder(cause_lbl, deaths_delayed), fill = cause)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.4) +
  scale_fill_manual(values = cause_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), labels = comma) +
  labs(title = paste0("Full package: deaths delayed by cause, ", impact_window),
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
print(p)
```

```{r table2-cause, message=FALSE, warning=FALSE, echo=FALSE}
# Deaths delayed by scenario x cause (wide), impact window.
dd_sc <- deaths_delayed_by(data.out, by = "cause", years = impact_y1:impact_y2)
dd_sc[, scenario := as_scn_factor(as.character(scenario))]
dd_sc[, cause_lbl := cause_map[cause]]
wide <- dcast(dd_sc, scenario ~ cause_lbl, value.var = "deaths_delayed")
wide <- wide[as.character(scenario) != "baseline"]
setorder(wide, scenario)
wide[, Scenario := scenario_labels[as.character(scenario)]]
setcolorder(wide, c("Scenario", unname(cause_map)))
wide[, scenario := NULL]
for (cc in unname(cause_map))
  wide[, (cc) := fmt_m(get(cc), 3)]

kable(wide,
      caption = paste0("Table 2. Deaths delayed (millions) by scenario and cause, ",
                       impact_window, "."),
      align = c("l", rep("r", length(cause_map)))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  add_header_above(setNames(c(1L, length(cause_map)),
                            c(" ", "Deaths delayed, millions"))) |>
  row_spec(0, bold = TRUE)
```

## Country-level breakdown (full package)

```{r figure5-country, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=6.5}
dd_country <- deaths_delayed_by(data.out, by = "location", years = impact_y1:impact_y2)
dd_country <- dd_country[as.character(scenario) == "full_package"]
setorder(dd_country, -deaths_delayed)

topn <- head(dd_country, 20)
p <- ggplot(topn, aes(deaths_delayed / 1e6, reorder(location, deaths_delayed))) +
  geom_col(fill = "#1f78b4", width = 0.72) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.10)), labels = comma) +
  labs(title = "Full package: top 20 countries by deaths delayed",
       subtitle = impact_window,
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
print(p)
```

```{r table3-country-dt, message=FALSE, warning=FALSE, echo=FALSE}
ctab <- dd_country[, .(
  Country = location,
  `Deaths delayed`  = round(deaths_delayed),
  `% of baseline`   = round(100 * deaths_delayed / baseline_dead, 2)
)]
setorder(ctab, -`Deaths delayed`)
datatable(ctab, rownames = FALSE,
          caption = paste0("All ", n_countries,
                           " countries: full-package deaths delayed, ",
                           impact_window, " (searchable / sortable)."),
          options = list(pageLength = 12, order = list(list(1, "desc")))) |>
  formatCurrency("Deaths delayed", currency = "", digits = 0)
```

## Regional breakdown (full package)

```{r figure6-region, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
if (data.out[, any(!is.na(region))]) {
  dd_region <- deaths_delayed_by(data.out[!is.na(region)], by = "region",
                                 years = impact_y1:impact_y2)
  dd_region <- dd_region[as.character(scenario) == "full_package"]
  setorder(dd_region, -deaths_delayed)
  p <- ggplot(dd_region, aes(deaths_delayed / 1e6, reorder(region, deaths_delayed))) +
    geom_col(fill = "#6a3d9a", width = 0.7) +
    geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.2) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.14)), labels = comma) +
    labs(title = "Full package: deaths delayed by region",
         subtitle = impact_window, x = "Deaths delayed (millions)", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
  print(p)
} else {
  cat("Region metadata unavailable (Country_groupings_extended.csv not found); ",
      "regional breakdown skipped.\n")
}
```

---

# Limitations and interpretation

- **Deaths delayed, not averted.** Case fatality is held at baseline, so the
  intervention lowers incidence and shifts mortality later; cumulative totals
  depend on the accounting window.
- **Effect sizes are point estimates.** The per-source reductions (Table M2)
  are deterministic; no probabilistic sensitivity analysis is run here.
- **LSS is a conservative lower bound.** The LSS effect is SSaSS-derived
  (≈ 15% of discretionary sodium; sensitivity 0.12–0.18) and models the
  **sodium arm only**. Because there is **no potassium channel**, the large
  K⁺-driven share of SSaSS's observed −3.34 mmHg SBP effect is omitted, so the
  modelled LSS benefit understates the trial. Its reduction is also restricted
  to the hypertensive subpopulation when run as an individual scenario.
- **Source shares** are documented for the ten prioritised countries, but the
  Nigeria/Cameroon values are **provisional** (RTSL/Matti working estimates, with
  Nigeria used as a proxy for Cameroon) and several others rest on single-study
  or older survey data; any country outside this set would fall back to the
  `default` (mean) profile.
- **Independent-cause structure.** The four causes share one population but are
  modelled independently, so "all-cause" prevalence/incidence sum cause-specific
  cases and may count comorbid individuals more than once.
- **Diabetes weighting.** ETTEHAD diabetes-specific effect sizes are prepared in
  the pipeline; the applied incidence modifier is the per-10-mmHg RR.

---

# Reproducibility

```{r session-info, message=FALSE, warning=FALSE, echo=FALSE}
cat("Report generated:", format(Sys.time()), "\n")
cat("Model outputs:    ", length(files), "country files from\n                  ",
    out_model_path, "\n")
cat("Countries:        ", n_countries, "\n")
cat("Scenarios:        ", paste(scenarios_present, collapse = ", "), "\n")
cat("Year range:       ", study_years, "\n\n")
sessionInfo()
```

---

# Deck artifacts (slide-deck export) {#deck-export}

This final section exists purely to hand computed elements to the Beamer slide
deck `docs/UWRTSL-sodium-policy.Rmd`. **All computation for the deck happens
here** (the deck itself performs none): every figure is written as a PNG and
every table/number is written into a single `deck_data.rds` bundle. The deck
loads that bundle and includes the PNGs, so the two documents can never drift.
Artifacts are written to `docs/deck_assets/`.

```{r deck-export, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# ---- Output location --------------------------------------------------------
deck_dir <- file.path(wd, "docs", "deck_assets")
dir.create(deck_dir, recursive = TRUE, showWarnings = FALSE)

save_fig <- function(plot, name, w, h) {
  ggplot2::ggsave(filename = file.path(deck_dir, name), plot = plot,
                  width = w, height = h, units = "in", dpi = 300, bg = "white")
  invisible(name)
}

# Pre-render a data.frame to a booktabs LaTeX tabular, wrapped in \resizebox so
# it can never overflow a 16:9 Beamer frame. escape = TRUE handles %, & etc.
mk_tex <- function(df, align = NULL) {
  if (is.null(align)) align <- c("l", rep("r", ncol(df) - 1L))
  body <- knitr::kable(df, format = "latex", booktabs = TRUE, align = align,
                       linesep = "", escape = TRUE, row.names = FALSE)
  paste0("\\resizebox{\\textwidth}{!}{%\n", as.character(body), "\n}")
}

base_year <- SCALEUP_YEAR1
lab_idx    <- paste0("Indexed (", base_year, " = 100)")

# ---- FIG: sodium source shares by country -----------------------------------
p_ss <- ggplot(shares_long, aes(location, share, fill = source)) +
  geom_col(width = 0.78, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "Share of total sodium intake") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1))
save_fig(p_ss, "fig_source_shares.png", 10, 4.7)

# ---- FIG: illustrative salteff by scenario (individual policies) -------------
p_se <- ggplot(salteff_bar[scenario != "full_package"],
               aes(location, contribution, fill = source)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.15) +
  facet_wrap(~ scn_lbl, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "salteff (fraction of total sodium reduced)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
save_fig(p_se, "fig_salteff.png", 10.5, 5.6)

# ---- FIG: full-package salteff by country ------------------------------------
p_sepkg <- ggplot(salteff_bar[scenario == "full_package"],
                  aes(location, contribution, fill = source)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Full-package salteff") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1))
save_fig(p_sepkg, "fig_salteff_pkg.png", 9.5, 4.7)

# ---- FIG: causal-chain flow diagram -----------------------------------------
p_flow <- ggplot() +
  geom_tile(data = boxes, aes(x = x, y = 1), width = 0.84, height = 0.7,
            fill = "#e8eef4", color = "#1f78b4", linewidth = 0.6) +
  geom_text(data = boxes, aes(x = x, y = 1, label = label), size = 3.4,
            lineheight = 0.95) +
  geom_segment(data = arrows, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.16, "cm"), type = "closed"),
               linewidth = 0.5, color = "grey30") +
  coord_cartesian(xlim = c(0.5, 6.5), ylim = c(0.55, 1.45)) +
  theme_void(base_size = 12)
save_fig(p_flow, "fig_flow.png", 11, 2.8)

# ---- FIG: baseline age-standardised mortality rate --------------------------
p_asmr <- ggplot() +
  geom_line(data = asmr_cause, aes(year, ASMR, color = cause_lbl,
                                   linetype = year > hist_cut), linewidth = 1) +
  geom_line(data = asmr_all, aes(year, ASMR, linetype = year > hist_cut),
            color = "black", linewidth = 1.2) +
  annotate("text", x = year_min + 1, y = max(asmr_all$ASMR),
           label = "All CVD", hjust = 0, vjust = 1, fontface = "bold", size = 3.6) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"),
                        guide = "none") +
  labs(x = "Year", y = "ASMR per 100,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
save_fig(p_asmr, "fig_baseline_asmr.png", 9.5, 5)

# ---- FIG: projected incidence (level + indexed) -----------------------------
inc_lvl <- proj_all[year >= 2020, .(scenario, year, value = inc_rate)][
  , panel := "Incident cases per 1,000"]
inc_ix  <- inc_idx[, .(scenario, year, value = inc_index)][, panel := lab_idx]
inc_comb <- rbind(inc_lvl, inc_ix)
inc_comb[, panel := factor(panel, levels = c("Incident cases per 1,000", lab_idx))]
hl_inc <- data.table(panel = factor(lab_idx, levels = levels(inc_comb$panel)),
                     yy = 100)
p_inc <- ggplot(inc_comb, aes(year, value, colour = scenario)) +
  geom_hline(data = hl_inc, aes(yintercept = yy), linetype = "dashed",
             colour = "grey60", inherit.aes = FALSE) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_colour_manual(values = scenario_cols, labels = scenario_labels_short,
                      breaks = scenario_levels, name = NULL) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = "Year", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
save_fig(p_inc, "fig_incidence.png", 10.5, 4.7)

# ---- FIG: projected deaths (per 1,000 + indexed) ----------------------------
proj_all[, dead_p1000 := dead / pop * 1000]
d_lvl  <- proj_all[year >= 2020, .(scenario, year, value = dead_p1000)][
  , panel := "Deaths per 1,000"]
base_d <- proj_all[year == base_year, .(scenario, bv = dead_p1000)]
d_ix   <- merge(proj_all[year >= base_year, .(scenario, year, dead_p1000)],
                base_d, by = "scenario")
d_ix   <- d_ix[!is.na(bv) & bv > 0]
d_ix[, value := dead_p1000 / bv * 100]
d_ix2  <- d_ix[, .(scenario, year, value)][, panel := lab_idx]
d_comb <- rbind(d_lvl, d_ix2)
d_comb[, panel := factor(panel, levels = c("Deaths per 1,000", lab_idx))]
hl_d <- data.table(panel = factor(lab_idx, levels = levels(d_comb$panel)), yy = 100)
p_d <- ggplot(d_comb, aes(year, value, colour = scenario)) +
  geom_hline(data = hl_d, aes(yintercept = yy), linetype = "dashed",
             colour = "grey60", inherit.aes = FALSE) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_colour_manual(values = scenario_cols, labels = scenario_labels_short,
                      breaks = scenario_levels, name = NULL) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = "Year", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
save_fig(p_d, "fig_deaths.png", 10.5, 4.7)

# ---- FIG: deaths delayed by scenario (headline result) ----------------------
fig_res <- dd_cum[as.character(scenario) != "baseline"]
fig_res[, scn_lbl := factor(scenario_labels_short[as.character(scenario)],
                            levels = rev(scenario_labels_short[scenario_nb]))]
p_res <- ggplot(fig_res, aes(deaths_delayed / 1e6, scn_lbl, fill = scenario)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.6) +
  scale_fill_manual(values = scenario_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14)), labels = comma) +
  labs(x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
save_fig(p_res, "fig_results_bar.png", 9, 4.8)

# ---- FIG: full package, deaths delayed by cause -----------------------------
p_bc <- ggplot(dd_cause, aes(deaths_delayed / 1e6,
                             reorder(cause_lbl, deaths_delayed), fill = cause)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.6) +
  scale_fill_manual(values = cause_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14)), labels = comma) +
  labs(x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
save_fig(p_bc, "fig_bycause.png", 8.5, 3.9)

# ---- FIG: full package, deaths delayed by country ---------------------------
dd_ctry <- copy(dd_country)
setorder(dd_ctry, -deaths_delayed)
p_ct <- ggplot(dd_ctry, aes(deaths_delayed / 1e6, reorder(location, deaths_delayed))) +
  geom_col(fill = "#1f78b4", width = 0.72) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)), labels = comma) +
  labs(x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
save_fig(p_ct, "fig_countries.png", 8.5, 5)

# ---- TABLES (pre-rendered LaTeX) --------------------------------------------
# Effect sizes (Table M2).
tef <- as.data.frame(eff_disp)
names(tef) <- c("Intervention", "Source", "Applied reduction", "Note")
tef$Intervention <- as.character(tef$Intervention)
tex_effect_sizes <- mk_tex(tef, align = c("l", "l", "r", "l"))

# Illustrative salteff by scenario x country (Table M4).
tsl <- as.data.frame(salteff_grid)
tex_salteff <- mk_tex(tsl, align = c("l", rep("r", ncol(tsl) - 1L)))

# Source shares (Table M1).
tss <- as.data.frame(ss_disp)
tex_source_shares <- mk_tex(tss, align = c("l", rep("r", ncol(tss) - 1L)))

# Headline results (compact Table 1).
tr <- dd_cum[as.character(scenario) != "baseline"]
tr[, scenario := factor(as.character(scenario), levels = scenario_levels)]
setorder(tr, scenario)
tres <- data.frame(
  Scenario                    = unname(scenario_labels[as.character(tr$scenario)]),
  `Deaths delayed (millions)` = fmt_m(tr$deaths_delayed, 3),
  `Avg per year (thousands)`  = fmt_k(tr$per_year, 1),
  `% of baseline`             = fmt_pct(tr$pct_vs_base, 2),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
tex_results <- mk_tex(tres, align = c("l", "r", "r", "r"))

# ---- NUMBERS ----------------------------------------------------------------
pkg_row <- dd_cum[as.character(scenario) == "full_package"]
vn_sanity <- as.numeric(compute_total_efficacy("public_procurement", "Viet Nam",
                                               source_shares, intervention_effects))

nums <- list(
  n_countries    = n_countries,
  study_years    = study_years,
  impact_window  = impact_window,
  scaleup1       = SCALEUP_YEAR1,
  scaleup2       = SCALEUP_YEAR2,
  horizon_end    = impact_y2,
  # headline deaths delayed (millions, formatted strings)
  pkg_m          = fmt_m(ddc("full_package"), 2),
  pkg_per_year_k = fmt_k(pkg_row$per_year, 0),
  pkg_pct        = fmt_pct(pkg_row$pct_vs_base, 1),
  salt_targets_m = fmt_m(ddc("salt_targets"), 2),
  fopl_m         = fmt_m(ddc("fopl"), 2),
  fiscal_m       = fmt_m(ddc("fiscal"), 2),
  public_m       = fmt_m(ddc("public_procurement"), 2),
  media_m        = fmt_m(ddc("media_campaign"), 2),
  lss_m          = fmt_m(ddc("lss"), 2),
  # applied per-source effect sizes (must mirror EFF_* in 07_run_interventions.R)
  eff_public       = fmt_pct(EFF_PROCUREMENT_PUBLIC, 0),
  eff_media        = fmt_pct(EFF_MEDIA, 1),
  eff_fopl         = fmt_pct(EFF_FOPL_PACKAGED, 1),
  eff_salttargets  = fmt_pct(EFF_SALTTARGETS_PACKAGED, 0),
  eff_fiscal       = fmt_pct(EFF_FISCAL_PACKAGED, 0),
  eff_lss          = fmt_pct(EFF_LSS_DISCRETIONARY, 0),
  lss_sens         = "0.12--0.18",
  vn_sanity        = formatC(vn_sanity, format = "f", digits = 4),
  # worked composition example: FOPL + salt targets on 'packaged'
  compose_fopl_salt = fmt_pct(1 - (1 - EFF_FOPL_PACKAGED) * (1 - EFF_SALTTARGETS_PACKAGED), 1),
  # LSS / SSaSS trial constants (surfaced so the deck states no raw numbers itself)
  ssass_na_mgday   = "-350 mg/day",
  ssass_base_gday  = "4.3 g/day",
  ssass_sbp_mmhg   = "-3.34 mmHg",
  ssass_k_mgday    = "+803 mg/day",
  lss_ratio        = "75% NaCl / 25% KCl"
)

figs <- list(
  source_shares = "fig_source_shares.png",
  salteff       = "fig_salteff.png",
  salteff_pkg   = "fig_salteff_pkg.png",
  flow          = "fig_flow.png",
  baseline_asmr = "fig_baseline_asmr.png",
  incidence     = "fig_incidence.png",
  deaths        = "fig_deaths.png",
  results_bar   = "fig_results_bar.png",
  bycause       = "fig_bycause.png",
  countries     = "fig_countries.png"
)

tex <- list(
  effect_sizes  = tex_effect_sizes,
  salteff       = tex_salteff,
  source_shares = tex_source_shares,
  results       = tex_results
)

deck <- list(nums = nums, figs = figs, tex = tex)
saveRDS(deck, file = file.path(deck_dir, "deck_data.rds"))

message("Deck artifacts written to: ", deck_dir)
message("  figures: ", paste(unlist(figs), collapse = ", "))
message("  bundle : deck_data.rds")
```
