---
title: "Sodium Policy Intervention Model: Methods and Projected Health Impact (2017–2050)"
date: "`r Sys.Date()`"
author: "University of Washington · Resolve to Save Lives"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 5,
  fig.width  = 9,
  fig.align  = "center"
)
```

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(DT)
library(scales)
library(readxl)   # benchmark import (100 Million Lives Saved workbook)
```

```{r paths, message=FALSE, warning=FALSE, echo=FALSE}
# Portable project-root resolution (Task 4j: no hard-coded machine path). The
# repo root is reports/'s parent; resolved relative to this .Rmd when knitting,
# and by ascending from the working directory when run chunk-by-chunk. Fails
# informatively if it cannot be found.
.find_root <- function(start) {
  d <- normalizePath(start, winslash = "/", mustWork = FALSE)
  for (i in 1:6) {
    if (dir.exists(file.path(d, "data", "processed")) &&
        dir.exists(file.path(d, "code"))) return(d)
    parent <- dirname(d); if (parent == d) break; d <- parent
  }
  NA_character_
}
wd <- tryCatch(
  normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
                winslash = "/", mustWork = TRUE),
  error = function(e) .find_root(getwd()))
if (is.na(wd) || !dir.exists(file.path(wd, "data", "processed")))
  stop("Could not locate the project root (a dir containing data/processed/ and code/). ",
       "Set `wd` manually or knit from within the repo.")
wd <- paste0(sub("/+$", "", wd), "/")

wd_data <- paste0(wd, "data/processed/")
wd_outp <- paste0(wd, "output/")
wd_scen <- paste0(wd, "scenarios/")
wd_raw  <- paste0(wd, "data/raw/")
```

```{r helpers, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Number formatters -------------------------------------------------------
fmt_m <- function(x, digits = 2)     # millions
  formatC(round(x / 1e6, digits), format = "f", digits = digits, big.mark = ",")
fmt_k <- function(x, digits = 0)     # thousands
  formatC(round(x / 1e3, digits), format = "f", digits = digits, big.mark = ",")
fmt_pct <- function(x, digits = 1)
  paste0(formatC(100 * x, format = "f", digits = digits), "%")

# ---- Cause labels ------------------------------------------------------------
cause_map <- c(
  ihd     = "Ischemic heart disease",
  istroke = "Ischemic stroke",
  hstroke = "Intracerebral hemorrhage",
  hhd     = "Hypertensive heart disease"
)
CAUSES <- names(cause_map)

cause_cols <- c(
  ihd     = "#56B4E9",
  istroke = "#CC79A7",
  hstroke = "#F0E442",
  hhd     = "#E69F00"
)

# ---- Scenario metadata -------------------------------------------------------
# NOTE: scenario_levels / scenario_labels / scenario_cols / as_scn_factor are NO
# LONGER hardcoded here. They are derived from run_config.rds (the pipeline's
# resolved scenario registry) intersected with the scenarios actually present in
# the outputs, in the `scenario-metadata` chunk AFTER run_config + outputs load
# (Task 4a). This guarantees the report's scenario set reconciles with production
# (na_k_sbp LSS variants, the SSaSS benchmark, fiscal sensitivity variants, and
# full-package membership per include_fiscal_in_package) instead of a stale
# hand-maintained vector.

# ---- WHO standard population (ages 20-95+) for ASMR -------------------------
full_who_std <- data.table(
  age    = 0:100,
  weight = c(
    rep(0.013, 10), rep(0.014, 10), rep(0.013, 10), rep(0.012, 10),
    rep(0.011, 10), rep(0.009, 10), rep(0.007, 10), rep(0.005, 10),
    rep(0.003, 10), rep(0.001, 10), 0.001
  )
)
who_std <- full_who_std[age >= 20 & age <= 94]
who_std <- rbind(
  who_std,
  data.table(age = 95, weight = full_who_std[age >= 95, sum(weight)])
)
who_std[, weight := weight / sum(weight)]

# Age-standardised rate per 100,000. deaths_col / population_col at single-year
# age, weighted to the WHO world standard, then summed within (year, disagg_var).
calculate_asmr <- function(dt, std_pop = who_std,
                           deaths_col = "deaths_dx",
                           population_col = "Nx",
                           disagg_var = "scenario") {
  dt <- copy(dt)
  dt[, age := as.integer(gsub("\\+", "", as.character(age)))]
  dt <- merge(dt, std_pop, by = "age", all.x = TRUE)
  dt[, rate          := get(deaths_col) / get(population_col)]
  dt[, weighted_rate := rate * weight]
  group_vars <- c("year", disagg_var)
  dt[, .(ASMR = sum(weighted_rate, na.rm = TRUE) * 1e5), by = group_vars]
}

# ---- Deaths-delayed calculator (keyed on 'scenario') -------------------------
# deaths_delayed = baseline deaths - scenario deaths, over an optional set of
# grouping columns and an optional year window. The reference is always the
# 'baseline' scenario. Uses 'dead' (cause-specific, safe to sum over causes).
deaths_delayed_by <- function(dt, by = character(0),
                              years = NULL, ref = "baseline") {
  d <- if (is.null(years)) dt else dt[year %in% years]
  agg  <- d[, .(dead = sum(dead, na.rm = TRUE)), by = c("scenario", by)]
  base <- agg[scenario == ref][, scenario := NULL]
  setnames(base, "dead", "baseline_dead")
  out  <- agg[scenario != ref]
  if (length(by) > 0) {
    out <- merge(out, base, by = by, all.x = TRUE)
  } else {
    out[, baseline_dead := base$baseline_dead]
  }
  out[, deaths_delayed := baseline_dead - dead]
  out[]
}
```

```{r load-artifacts, message=FALSE, warning=FALSE, echo=FALSE}
# Task 4f: the report NO LONGER re-implements the model. It CONSUMES the pipeline
# artifacts written by code/07_run_interventions.R (plus 01/03). Everything the
# report must not re-guess is read from run_config.rds. Missing artifacts fail
# informatively (Task 4j).
.need <- function(f, what) {
  p <- paste0(wd_data, f)
  if (!file.exists(p)) stop(sprintf(
    "Required pipeline artifact missing:\n  %s\n  (%s)\n  Run code/07_run_interventions.R first.",
    p, what))
  x <- readRDS(p); if (is.data.frame(x)) setDT(x); x
}
run_config            <- .need("run_config.rds",            "run configuration + scenario registry")
RC                    <- run_config
source_shares_by_year <- .need("source_shares_by_year.rds", "Task 3 year-specific source shares")
lss_audit             <- .need("lss_audit.rds",             "Task 1e LSS mechanistic audit")
fiscal_audit          <- .need("fiscal_audit.rds",          "Task 2 fiscal pathway audit")
htn_eligibility       <- .need("htn_eligibility.rds",       "Task 1a NCD-RisC eligibility")
baseline_potassium    <- .need("baseline_potassium.rds",    "Task 1a baseline potassium")
packaged_trend_diag   <- tryCatch(readRDS(paste0(wd_data, "packaged_trend_diagnostic.rds")),
                                  error = function(e) NULL)   # optional

# ---- Authoritative source-split inputs (NOT re-derived here) ----------------
SODIUM_SOURCES       <- RC$sodium_sources
intervention_effects <- as.data.table(RC$intervention_effects)   # published effect sizes
source_shares        <- as.data.table(RC$source_shares_base)     # documented base-year shares

# Presentation-only colours/labels for the five sources.
source_cols   <- c(discretionary = "#4E79A7", packaged = "#F28E2B",
                   restaurant = "#59A14F", public = "#E15759", inherent = "#B07AA1")
source_labels <- c(discretionary = "Discretionary", packaged = "Packaged",
                   restaurant = "Restaurant", public = "Public", inherent = "Inherent")

# DISPLAY-ONLY salteff resolver: the pipeline's documented composition formula
# [salteff = sum_s share_s x (1 - prod_i(1 - eff_i,s))] fed by the PUBLISHED
# effect sizes (RC$intervention_effects) and shares. This presents the SAME
# formula the model uses; all inputs are authoritative, so it cannot diverge from
# production (Task 4f: no second model).
compute_total_efficacy <- function(intervention_names, country,
                                    shares = source_shares, effects = intervention_effects) {
  srow <- shares[location == country]
  if (nrow(srow) == 0L) srow <- shares[location == "default"]
  if (nrow(srow) == 0L) stop("No source shares for '", country, "'.")
  sv <- unlist(srow[1L, ..SODIUM_SOURCES])
  if (is.null(intervention_names) || length(intervention_names) == 0L) {
    t <- 0
    attr(t, "decomposition") <- data.table(source = SODIUM_SOURCES, share = sv,
                                            composed_effect = 0, contribution = 0)
    return(t)
  }
  eff  <- effects[intervention %in% intervention_names]
  comp <- vapply(SODIUM_SOURCES, function(s) {
    e <- eff[source == s, effect]; if (!length(e)) 0 else 1 - prod(1 - e) }, numeric(1))
  dec <- data.table(source = SODIUM_SOURCES, share = sv,
                    composed_effect = comp[SODIUM_SOURCES],
                    contribution = sv * comp[SODIUM_SOURCES])
  t <- sum(dec$contribution); attr(t, "decomposition") <- dec[]; t
}

# Year-specific salteff for display (reads the published YEAR-share table).
report_salteff_year <- function(intervention_names, country, yr) {
  ssy <- source_shares_by_year[location == country & year == yr]
  if (!nrow(ssy)) ssy <- source_shares_by_year[location == "default" & year == yr]
  compute_total_efficacy(intervention_names, country,
                         shares = ssy[, c("location", SODIUM_SOURCES), with = FALSE])
}

resolve_source_shares <- function(countries, shares = source_shares) {
  rbindlist(lapply(countries, function(cc) {
    srow <- shares[location == cc]; if (!nrow(srow)) srow <- shares[location == "default"]
    srow <- copy(srow[1L]); srow[, location := cc]; srow[] }))
}

# LSS config from run_config (na_k_sbp PRIMARY; SSaSS a benchmark). The SSaSS
# trial RRs are surfaced for the benchmark section / deck only -- NOT as the sole
# primary LSS model.
LSS_COVERAGE_ALL  <- RC$lss$coverage_all
LSS_RR_STROKE_INC <- 0.90            # SSaSS nonfatal-stroke incidence RR (benchmark)
LSS_CF_MULT       <- 0.77 / 0.90     # SSaSS conditional stroke CF multiplier (= 0.856)
LSS_STROKE_TOTAL  <- 0.86            # SSaSS total-stroke RR (= 0.90 x 0.856)
```

```{r country-metadata, message=FALSE, warning=FALSE, echo=FALSE}
# Region lookup for optional regional breakdowns.
countries_dt <- tryCatch(
  fread(paste0(wd_data, "Country_groupings_extended.csv")),
  error = function(e) data.table(iso3 = character(), location = character(),
                                 region = character())
)
if (nrow(countries_dt)) {
  countries_dt[location == "United States of America", location := "United States"]
  countries_dt[location == "Vietnam",                  location := "Viet Nam"]
}

# Baseline mean sodium intake (2024) - used only to illustrate the salteff ->
# grams -> SBP chain; the report does not otherwise depend on it.
sodium_baseline <- tryCatch({
  sp <- readRDS(paste0(wd_data, "sodium_policy_scenarios.rds"))
  setDT(sp)
  sp[year == 2024, .(location, sodium_current)]
}, error = function(e) NULL)
```

```{r load-model-output, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Load the per-country intervention outputs ------------------------------
# Written by code/07_run_interventions.R, one file per country:
#   output/out_model/model_output_interventions_<Country>.rds
# Each is a data.table with an authoritative 'scenario' column.
out_model_path <- paste0(wd_outp, "out_model/")
files <- list.files(path = out_model_path,
                    pattern = "^model_output_interventions_.*\\.rds$",
                    full.names = TRUE)

if (length(files) == 0)
  stop(paste0(
    "No 'model_output_interventions_*.rds' files found in:\n  ", out_model_path,
    "\n\nRun code/07_run_interventions.R first (it must be sourced after ",
    "05_build_baseline.R) to generate the per-country intervention outputs."
  ))

data.out <- rbindlist(
  lapply(files, function(f) { dt <- readRDS(f); setDT(dt); dt }),
  use.names = TRUE, fill = TRUE
)
data.out <- data.out[!is.na(location) & !is.na(scenario)]

# ---- Scenario registry, DERIVED from run_config (Task 4a) -------------------
# The scenario set, labels, methods and full-package membership all come from
# the pipeline's resolved registry (RC$scenarios) intersected with the scenarios
# actually present in the outputs. No hardcoded vector. Scenarios in the registry
# but absent from the outputs are noted; scenarios in the outputs but not the
# registry are dropped with a warning.
scn_reg <- as.data.table(RC$scenarios)
present <- unique(as.character(data.out$scenario))
scenario_levels <- scn_reg$scenario[scn_reg$scenario %in% present]   # canonical order
missing_from_out <- setdiff(scn_reg$scenario, present)
extra_in_out     <- setdiff(present, scn_reg$scenario)
if (length(extra_in_out))
  warning("Output scenarios not in run_config registry (dropped): ",
          paste(extra_in_out, collapse = ", "))

scenario_labels <- setNames(scn_reg$label, scn_reg$scenario)
scenario_labels[is.na(scenario_labels) | scenario_labels == ""] <-
  names(scenario_labels)[is.na(scenario_labels) | scenario_labels == ""]

# Compact labels for axes / legends.
short_lab <- function(s) {
  m <- c(baseline = "Baseline", public_procurement = "Public procurement",
         fopl = "FOPL", salt_targets = "Salt targets", full_package = "Full package",
         fiscal = "Fiscal", fiscal_low = "Fiscal (low)", fiscal_base = "Fiscal (base)",
         fiscal_high = "Fiscal (high)",
         lss_s2 = "LSS s2 all (Na/K)", lss_s4 = "LSS s4 dx (Na/K)",
         lss_s5 = "LSS s5 tx (Na/K)", lss_s4_ssass = "LSS s4 SSaSS",
         lss_s5_ssass = "LSS s5 SSaSS")
  ifelse(s %in% names(m), m[s], s)
}
scenario_labels_short <- setNames(short_lab(scenario_levels), scenario_levels)

# Colours: fixed hues for the core families, generated for any extras.
.base_cols <- c(baseline = "#999999", public_procurement = "#1f78b4", fopl = "#ff7f00",
                salt_targets = "#e31a1c", fiscal = "#6a3d9a", fiscal_low = "#9e9ac8",
                fiscal_base = "#6a3d9a", fiscal_high = "#3f007d",
                lss_s2 = "#8c510a", lss_s4 = "#bf812d", lss_s5 = "#dfc27d",
                lss_s4_ssass = "#01665e", lss_s5_ssass = "#5ab4ac", full_package = "#000000")
scenario_cols <- .base_cols[scenario_levels]
if (any(is.na(scenario_cols))) {
  miss <- scenario_levels[is.na(scenario_cols)]
  scenario_cols[is.na(scenario_cols)] <- scales::hue_pal()(length(miss))
}
names(scenario_cols) <- scenario_levels

as_scn_factor <- function(x) factor(as.character(x), levels = scenario_levels)
scenario_nb   <- setdiff(scenario_levels, "baseline")

# scenario -> intervention vector (baseline = none), from the registry.
scenario_interventions <- setNames(lapply(scn_reg$scenario, function(s) {
  iv <- scn_reg[scenario == s, interventions]
  if (is.na(iv) || iv == "") character(0) else trimws(strsplit(iv, "\\+")[[1]])
}), scn_reg$scenario)
package_interventions <- scenario_interventions[["full_package"]]

# Keep only registry scenarios present in outputs; impose canonical order.
data.out <- data.out[scenario %in% scenario_levels]
data.out[, scenario := as_scn_factor(scenario)]

# Restrict to the four modelled CVD causes.
data.out <- data.out[cause %in% CAUSES]

# Age groups for demographic breakdowns.
data.out[, age_group := cut(age, breaks = c(20, 45, 65, 80, Inf),
                            labels = c("20-44", "45-64", "65-79", "80+"),
                            right = FALSE, include.lowest = TRUE)]

# Region metadata (optional).
if (nrow(countries_dt)) {
  data.out <- merge(data.out, unique(countries_dt[, .(location, region)]),
                    by = "location", all.x = TRUE)
}
if (!"region" %in% names(data.out)) data.out[, region := NA_character_]

# ---- Key counts / dimensions ------------------------------------------------
n_countries <- uniqueN(data.out$location)
year_min    <- min(data.out$year, na.rm = TRUE)
year_max    <- max(data.out$year, na.rm = TRUE)
study_years <- paste0(year_min, "–", year_max)
scenarios_present <- levels(droplevels(data.out$scenario))

# Scale-up + impact window from run_config (Task 4j: configured years, not hardcoded).
SCALEUP_YEAR1 <- RC$scaleup_window[1]
SCALEUP_YEAR2 <- RC$scaleup_window[2]
impact_years  <- RC$impact_years
impact_y1 <- SCALEUP_YEAR1
impact_y2 <- min(RC$analysis_years[2], year_max)
impact_window <- paste0(impact_y1, "–", impact_y2)
n_impact_years <- impact_y2 - impact_y1 + 1

# Illustrative-salteff display set (the DISTINCT policies, LSS shown once as its
# sodium-displacement composition; fiscal at its base case). Used by the methods
# salteff tables/figures only.
salteff_display <- list(
  public_procurement = "public_procurement", fopl = "fopl",
  salt_targets = "salt_targets", fiscal = "fiscal", lss = "lss",
  full_package = package_interventions)
salteff_display <- salteff_display[vapply(salteff_display, function(iv)
  all(iv %in% intervention_effects$intervention) || length(iv) == 0, logical(1))]
salteff_scenarios     <- names(salteff_display)
salteff_labels_short  <- c(public_procurement = "Public procurement", fopl = "FOPL",
  salt_targets = "Salt targets", fiscal = "Fiscal (base)",
  lss = "LSS (Na displacement)", full_package = "Full package")[salteff_scenarios]
```

---

# Background

This report documents the **source-split sodium-policy intervention model**
implemented in [`code/07_run_interventions.R`](../code/07_run_interventions.R)
and presents its baseline and projected outputs. The model estimates the
cardiovascular health impact of reducing population sodium intake through a
menu of policy interventions, projecting outcomes annually for
**`r n_countries` countries** over `r study_years`.

The central question is:

> *How many cardiovascular deaths are delayed by each sodium-reduction policy
> — and by a coordinated policy package — when each policy's documented effect
> is applied only to the sodium sources it can actually reach?*

The engine is a **discrete-time, multi-state (Well → Sick → Dead) cohort
model** calibrated to GBD, run once per **scenario**:

- a **baseline** (no intervention),
- one scenario for each **individual** intervention, and
- a combined **full package**.

Four causes of cardiovascular death are modelled jointly: ischemic heart
disease (`ihd`), ischemic stroke (`istroke`), intracerebral hemorrhage
(`hstroke`), and hypertensive heart disease (`hhd`).

The defining feature of this build is **upstream**: instead of applying a
single blended reduction to a single sodium pool, baseline sodium is split into
**five sources**, and each intervention's documented relative reduction is
applied **only to the source(s) it can affect**. Those per-source reductions
are then collapsed to a single total-intake reduction fraction (`salteff`) that
feeds the unchanged epidemiological pathway (sodium → SBP → CVD incidence →
state transitions). Everything from `salteff` onward is identical to the prior
single-pool model.

```{r headline-precompute, message=FALSE, warning=FALSE, echo=FALSE}
# Headline numbers for the inline narrative below.
dd_total <- deaths_delayed_by(data.out, by = character(0),
                              years = impact_y1:impact_y2)
dd_total[, scenario := as_scn_factor(as.character(scenario))]
setorder(dd_total, scenario)
get_dd <- function(scn) dd_total[as.character(scenario) == scn, deaths_delayed]
pkg_total <- get_dd("full_package")
```

Across `r n_countries` countries and `r impact_window`, the **full package** is
projected to delay approximately **`r fmt_m(pkg_total)` million**
cardiovascular deaths relative to baseline. Individual interventions and the
by-cause decomposition are given in the [results](#main-results-deaths-delayed).

> **A note on terminology.** For the **sodium→SBP** interventions (public
> procurement, FOPL, salt targets, fiscal, and the **primary Na/K→SBP LSS**
> method), the intervention lowers disease *incidence* while case fatality is
> held at its baseline value, so the model shifts deaths later in time rather
> than removing them from the life table entirely. We therefore report **deaths
> delayed** (baseline deaths − scenario deaths within the window) rather than
> "deaths averted". **The exception is the optional SSaSS trial-RR benchmark**,
> which also lowers stroke **case fatality** (multiplier 0.856) and so delivers
> some genuinely *averted* deaths; we still report everything under the common
> "deaths delayed" heading for comparability.

---

# Model inputs

This report **consumes the pipeline's published artifacts** (it does not re-run
or re-implement the model). The run configuration and every derived input are
read from `data/processed/`; the modelling inputs are consumed *upstream* by
`07_run_interventions.R` (via `05_build_baseline.R`) and are described here for
provenance.

| Input | File | Role |
|---|---|---|
| Run configuration | `run_config.rds` | LSS method + Na/K constants, HTN source year, fiscal sensitivity + package membership, Euromonitor window/base year/ceiling/post-2030, scale-up + impact years, and the **resolved scenario registry**. The report reads these rather than re-guessing. |
| Baseline transition rates | `*adjusted*.rds` → `b_rates` | Age × sex × year × cause incidence (IR), case fatality (CF), background mortality (BG.mx, BG.mx.all), COVID excess (covid.mx), prevalence and population (Nx). |
| BP distribution | `bp_data6.csv` → `data.in` | Mean/SD systolic BP by age × sex, raised-BP fraction, diabetes prevalence; the baseline sodium column `salt` (g **sodium**/day). |
| Baseline sodium intake | `sodium_policy_scenarios.rds` | 2024 mean sodium intake (`sodium_current`) merged in as `salt`. |
| Sodium source shares (year-specific) | `source_shares_by_year.rds` | Five-source split of baseline sodium per country **and year** (packaged share follows the Euromonitor trend; the other four renormalise; composition-only). |
| NCD-RisC hypertension eligibility | `htn_eligibility.rds` | Age-standardised 30–79 diagnosed/treated proportions → population eligibility shares that size the LSS diagnosed/treated scenarios (Task 1a). |
| Baseline potassium intake | `baseline_potassium.rds` | Country × sex baseline K intake (Reddin 2023; region-mean fallback for Viet Nam & Ethiopia) feeding the LSS potassium channel. |
| Euromonitor packaged trend | `Euromonitor_trends_RTSL_v002.xlsx` → `packaged_food_trends.rds` | Matti's pre-combined per-country sodium-source packaged growth rate (+ 95% CI) driving the year-specific shares. |
| LSS mechanistic audit | `lss_audit.rds` | Per country × sex × scenario Na/K→SBP pathway quantities (the single source of truth for the LSS methods/results tables). |
| Fiscal audit | `fiscal_audit.rds` | Fiscal pathway parameters + resulting packaged/total-sodium change per sensitivity variant (**exploratory**). |
| GBD relative risks | `IHME_GBD_2019_RELATIVE_RISKS_*_HTN.xlsx` → `dt_gbd_rr` | Log-linear RR per 10 mmHg SBP, by age × cause; sets bin-specific baseline incidence. |
| ETTEHAD relative risks | `ettehad_rr_bp_reduction_10mmHg.csv`, `ettehad_rr_bp_reduction_effects.xlsx` | RR per 10 mmHg SBP reduction by BP bin × cause; converts the SBP shift to an incidence change. |
| Scale-up schedule | `run_config.rds` | Linear scale-up from `r SCALEUP_YEAR1` to `r SCALEUP_YEAR2`. |

---

# Methods

## Overall structure

Let $t$ index calendar years and $a$ single-year ages (20–95). For each
country, sex, cause and scenario, the model tracks three states — **Well**,
**Sick** (prevalent disease) and **Dead** — and advances them one year at a
time. Disease dynamics are governed by three age/sex/cause/year-specific rates
carried in `b_rates`:

- $IR$ — incidence rate (Well → Sick),
- $CF$ — case fatality (Sick → Dead from the disease),
- background mortality $\text{BG.mx}$ (applied to the well/at-risk population),
  a disease-envelope term $\text{BG.mx.all}$, and a COVID excess term
  $\text{covid.mx}$ for 2020–2021.

**Initialisation.** At the seed year (2017) and at the entry age (20) the states
are set from GBD prevalence and mortality:

$$
\text{sick} = N_x\,\text{PREV}_{t_0}, \qquad
\text{dead} = N_x\,\text{DIS.mx}_{t_0}, \qquad
\text{well} = N_x\bigl(1 - (\text{PREV}_{t_0} + \text{BG.mx})\bigr),
$$

with population $\text{pop} = N_x$ and $\text{all.mx} = N_x\,\text{DIS.mx}_{t_0}
+ N_x\,\text{BG.mx}$.

**Annual recurrences.** Writing $x_{t-1}$ for last year's value of state $x$ in
the cohort that ages into year $t$, the transitions implemented in
`project.all()` are:

$$
\begin{aligned}
\text{newcases}_t &= \text{well}_{t-1}\cdot IR \\[2pt]
\text{sick}_t     &= \text{sick}_{t-1}\bigl(1 - (CF + \text{BG.mx} + \text{covid.mx})\bigr) + \text{well}_{t-1}\cdot IR \\[2pt]
\text{dead}_t     &= \text{sick}_{t-1}\cdot CF \\[2pt]
\text{pop}_t      &= \text{pop}_{t-1} - \text{all.mx}_{t-1} \\[2pt]
\text{all.mx}_t   &= \sum_{\text{causes}} \text{dead}_t \;+\; \text{pop}_t\cdot \text{BG.mx.all} \;+\; \text{pop}_t\cdot \text{covid.mx} \\[2pt]
\text{well}_t     &= \text{pop}_t - \text{all.mx}_t - \text{sick}_t
\end{aligned}
$$

Every quantity is floored at zero, ages advance by one year per step, and the
loop runs up to 41 annual steps (2017 → 2058). In the current build the baseline
rate panel spans 2017–2050, so populated outputs cover **`r study_years`**.

**Aggregation semantics (important).** The four causes are modelled as
independent Well/Sick/Dead processes that *share* one population. In the
outputs this means **`pop` and `all.mx` are identical across the four cause
rows** (the shared population and its total mortality), while **`sick`,
`well`, `dead` and `newcases` are cause-specific**. Consequently:

- disease **counts** (deaths, incident cases, prevalent cases) are summed over
  causes, but
- the **population denominator** is taken from a *single* cause row (summing
  `pop` over the four causes would quadruple-count it); equivalently,
  $\text{well}_c + \text{sick}_c = \text{pop} - \text{all.mx}$ is the total
  living population and is the same for every cause $c$.

All rates in this report follow those rules.

## Source of sodium

Baseline sodium intake is partitioned into **five mutually exclusive sources**
whose shares sum to 1.0 in every country:

$$
\text{discretionary} + \text{packaged} + \text{restaurant} + \text{public} + \text{inherent} = 1 .
$$

`inherent` sodium (naturally present in unprocessed foods) is **never targeted**
by any intervention. Each policy acts only on the source(s) it can plausibly
reach. Country source shares are taken from the RTSL sodium policy document; a
`default` row (the simple mean of the ten documented countries) is used for
any country without its own row, and **every row is renormalised to sum to
exactly 1.0**. **Malaysia uses the Philippines source profile** as a proxy, and
**Nigeria/Cameroon** use the provisional shares supplied by RTSL/Matti (with
bouillon and condiments counted as packaged; Nigeria is used as a proxy for
Cameroon).

These shares are now **year-specific**: the packaged share follows the
Euromonitor trend (next section) and the other four renormalise each year so the
five still sum to 1 in every country-year — a change in *composition* only
(total sodium intake is unchanged). Table M1 below shows the **base-year
(`r RC$euromonitor$base_year`)** shares consumed from `source_shares_by_year.rds`;
the report does not recompute them.

```{r tbl-source-shares, message=FALSE, warning=FALSE, echo=FALSE}
ss_disp <- copy(source_shares)
setnames(ss_disp,
         c("location", SODIUM_SOURCES),
         c("Country", "Discretionary", "Packaged", "Restaurant", "Public", "Inherent"))
num_cols <- c("Discretionary", "Packaged", "Restaurant", "Public", "Inherent")
ss_disp[, (num_cols) := lapply(.SD, function(x) formatC(x, format = "f", digits = 3)),
        .SDcols = num_cols]

kable(ss_disp,
      caption = paste0("Table M1. Per-country sodium source shares (fractions of ",
                       "total intake), renormalised to 1.0. The 'default' row is the ",
                       "mean across the ten documented countries and is used for all ",
                       "other countries."),
      align = c("l", rep("r", 5))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(nrow(ss_disp), italic = TRUE) |>
  footnote(general = paste0(
    "Source: RTSL sodium policy document. 'inherent' sodium is never targeted. ",
    "Malaysia uses the Philippines profile as a proxy."))
```

The same shares are shown below as a per-country stacked bar. Each bar sums to
1.0. All ten prioritised countries now have documented shares (Nigeria and
Cameroon on provisional RTSL/Matti values); any other country would resolve to
the `default` (mean) profile.

```{r fig-source-shares-bar, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# Countries to display: those actually modelled (includes Nigeria & Cameroon),
# with a fall-back to the documented list if outputs are unavailable.
plot_countries <- sort(unique(as.character(data.out$location)))
if (length(plot_countries) == 0)
  plot_countries <- setdiff(source_shares$location, "default")

shares_bar <- resolve_source_shares(plot_countries)
shares_long <- melt(shares_bar,
                    id.vars = "location",
                    measure.vars = SODIUM_SOURCES,
                    variable.name = "source", value.name = "share")
shares_long[, source := factor(source, levels = SODIUM_SOURCES)]

p <- ggplot(shares_long, aes(location, share, fill = source)) +
  geom_col(width = 0.78, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Sodium source shares by country",
       subtitle = "Fraction of total baseline sodium intake, by source (bars sum to 100%)",
       x = NULL, y = "Share of total sodium intake",
       caption = paste0("Documented shares for the ten prioritised countries ",
                        "(Nigeria/Cameroon on provisional RTSL values).")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1))
print(p)
```

## Packaged-food consumption trend (Euromonitor)

Packaged sodium is **no longer assumed constant**, and the trend is now
**country- and year-specific**. It comes from Matti's updated Euromonitor
workbook (`Euromonitor_trends_RTSL_v002.xlsx`), which supplies a **pre-combined
per-geography trend** — one row per country (`series_category = "Combined
selected sodium-sources categories"`) that already merges the sodium-relevant
categories (sauces/condiments, soups, ready meals, processed meat, savoury
snacks, cheese, baked goods, **rice/pasta/noodles**, etc.) into a single annual
percent change **with a 95% confidence interval**. We consume that row directly
(window = **`r RC$euromonitor$window`**, `r RC$euromonitor$base_year`–2030),
rather than re-deriving a median over leaf categories.

The packaged share in year $y$ is
$\text{packaged}(y) = \text{packaged}_{\text{base}}\,(1+g)^{\,y-\text{base}}$,
capped at a ceiling of **`r fmt_pct(RC$euromonitor$ceiling, 0)`**; the other four
sources renormalise each year so all five sum to 1. After 2030 the rule is
**`r RC$euromonitor$post2030`**. This is a change in *composition* only — total
sodium intake is unchanged (`composition_only = `r RC$euromonitor$composition_only``).
The trend input is built in [`03_clean_inputs.R`](../code/03_clean_inputs.R) and
the year-indexed shares in [`07_run_interventions.R`](../code/07_run_interventions.R)
(`source_shares_by_year.rds`).

```{r packaged-trend, message=FALSE, warning=FALSE, echo=FALSE}
# Country x selected-year packaged share, growth (+CI), ceiling, post-2030 rule
# -- consumed from the pipeline's trend diagnostic (no re-derivation).
if (!is.null(packaged_trend_diag)) {
  pd <- as.data.table(packaged_trend_diag)[order(-pkg_2030)]
  pkg_trend_tbl <- pd[, .(
    Country               = location,
    `Growth g (%/yr)`     = formatC(100 * g, format = "f", digits = 2),
    `95% CI`              = ifelse(is.na(g_lcl), "—",
                               paste0(formatC(100*g_lcl,format="f",digits=2), "–",
                                      formatC(100*g_ucl,format="f",digits=2))),
    `Packaged base`       = fmt_pct(pkg_base, 1),
    `Packaged 2030`       = fmt_pct(pkg_2030, 1),
    `Packaged 2050`       = fmt_pct(pkg_2050, 1))]
  pkg_apc_median <- fmt_pct(median(pd$g, na.rm = TRUE), 1)
  pkg_apc_range  <- paste0(fmt_pct(min(pd$g, na.rm = TRUE), 1), "–",
                           fmt_pct(max(pd$g, na.rm = TRUE), 1))
  kable(pkg_trend_tbl,
        caption = paste0("Table M2b. Euromonitor v002 pre-combined packaged growth ",
                         "(g, %/yr, +95% CI) and the year-specific packaged share at ",
                         "base year (", RC$euromonitor$base_year, "), 2030 and 2050. ",
                         "Ceiling = ", fmt_pct(RC$euromonitor$ceiling,0),
                         "; post-2030 = ", RC$euromonitor$post2030, "."),
        align = c("l", "r", "r", "r", "r", "r")) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 13) |>
    row_spec(0, bold = TRUE) |>
    footnote(general = paste0(
      "g = Euromonitor v002 combined per-geography APC (window = ",
      RC$euromonitor$window, "). Applied as (1+g)^(year-base) to the packaged ",
      "share, capped at the ceiling; the other four sources renormalise. ",
      "Composition-only: total sodium intake unchanged. Source: Euromonitor ",
      "(Marklund v002)."))
} else {
  pkg_trend_tbl <- NULL; pkg_apc_median <- "n/a"; pkg_apc_range <- "n/a"
  cat("packaged_trend_diagnostic.rds not found; packaged-trend table skipped.\n")
}

# Verify shares sum to 1 in EVERY country-year (consume source_shares_by_year).
.ssy_ck <- source_shares_by_year[, .(s = sum(.SD)), .SDcols = SODIUM_SOURCES,
                                 by = .(location, year)]
share_sum_ok <- max(abs(.ssy_ck$s - 1)) < 1e-6
```

`r if (exists("share_sum_ok")) paste0("**Share integrity check:** all five source shares sum to 1 in every country-year (max deviation ", formatC(max(abs(.ssy_ck$s-1)), format="e", digits=1), ").")`

```{r fig-packaged-over-time, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# Packaged share over time by country (Task 4e).
pk <- source_shares_by_year[location %in% RC$priority_countries &
                            year >= RC$euromonitor$base_year & year <= impact_y2]
fig_pkg_time <- ggplot(pk, aes(year, packaged, colour = location)) +
  geom_hline(yintercept = RC$euromonitor$ceiling, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Packaged share of sodium over time, by country",
       subtitle = paste0("Euromonitor v002 trend; ceiling ",
                         fmt_pct(RC$euromonitor$ceiling,0), "; post-2030 = ",
                         RC$euromonitor$post2030, " (dashed = ceiling)"),
       x = NULL, y = "Packaged share of total sodium", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right", plot.title = element_text(face = "bold"))
print(fig_pkg_time)
```

## Effect sizes

The **composition-based** interventions (public food procurement, FOPL, salt
targets, and the exploratory fiscal policy) have a documented **relative
reduction** applied to their target source at full implementation; when several
target the same source (notably `packaged`) they compose multiplicatively
($1-\prod_i(1-\text{eff}_i)$). **LSS is different**: it is a mechanistic
Na/K→SBP model (see below), not a single source effect size — the discretionary
sodium it displaces is derived from the LSS composition, and its potassium
channel is added on top. The effect sizes below are read from the pipeline's
published table (`run_config.rds`); the fiscal effect shown is the **base case
of an exploratory pathway** (next section), not a confirmed value.

```{r tbl-effect-sizes, message=FALSE, warning=FALSE, echo=FALSE}
.eff_lab <- c(public_procurement = "Public food procurement & service",
              media_campaign = "Behaviour change / media campaigns",
              fopl = "Front-of-pack labelling (FOPL)",
              salt_targets = "Salt targets (reformulation)",
              fiscal = "Fiscal policy — salt tax (EXPLORATORY, base case)",
              lss = "Low-sodium salt substitutes (mechanistic; see LSS section)")
eff_disp <- intervention_effects[intervention %in% names(.eff_lab)]
eff_disp <- eff_disp[, .(
  Intervention = .eff_lab[intervention],
  `Target source` = source,
  `Relative reduction` = fmt_pct(effect, 1),
  Note = fcase(
    intervention == "lss",            "per-discretionary-user Na displacement (KCl x adherence x uptake); K channel adds SBP effect",
    intervention == "fiscal",         "EXPLORATORY base case; see fiscal pathway + sensitivity",
    intervention == "media_campaign", "retained in registry; excluded from reported scenarios",
    default = ""))]
setorder(eff_disp, Intervention, `Target source`)

kable(eff_disp,
      caption = "Table M2. Intervention → source effect sizes (relative reduction of the target source at full implementation).",
      align = c("l", "l", "r", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "FOPL, salt targets and fiscal policy act on packaged sodium and compose ",
    "multiplicatively. Fiscal is EXPLORATORY (a tax->pass-through->elasticity ",
    "pathway; not '10% tax => 10% reduction' and not 'all packaged taxable'). ",
    "The LSS row is the per-discretionary-user sodium DISPLACEMENT that sizes ",
    "salteff under the primary Na/K-SBP method; the potassium-mediated SBP ",
    "effect is added inside the model and is NOT captured by this single ",
    "number. Media is excluded from reported scenarios. Source: RTSL effect-size ",
    "tables; run_config.rds."))
```

## Low-sodium salt substitutes (LSS) — primary Na/K→SBP method

The **primary** LSS method is a mechanistic **sodium + potassium → SBP** model
(method `r sprintf("\x60%s\x60", RC$lss$method)`), following the structure of
**Huang et al. 2026** and using the **Filippini (2020)** potassium→blood-pressure
dose-response. Switching discretionary table salt to a
`r fmt_pct(RC$lss$nacl_fraction,0)` NaCl / `r fmt_pct(RC$lss$kcl_fraction,0)` KCl
substitute both **cuts sodium** and **adds potassium**; both channels lower SBP,
and the combined SBP change feeds the same ETIHAD/GBD SBP→cause machinery as
every other sodium policy (so LSS here affects **all four CVD causes**, not only
stroke).

**This implementation is benefit-only.** Unlike Huang et al., this disease model
has **no CKD / hyperkalaemia state**, so LSS harms are **not** modelled and the
LSS benefit is **not offset** (flagged in Limitations and recorded in the audit).

The per-eligible-and-adherent-person pathway is:

1. **Composition.** LSS is `r fmt_pct(RC$lss$nacl_fraction,0)`/`r fmt_pct(RC$lss$kcl_fraction,0)`
   NaCl/KCl by mass. Discretionary **sodium** is cut by the KCl fraction, scaled
   by uptake × adherence (SSaSS yr-5 adherence ≈ `r formatC(RC$lss$adherence, format="f", digits=2)`;
   uptake = `r formatC(RC$lss$uptake, format="f", digits=2)`).
2. **Sodium channel.** The displaced discretionary sodium feeds the **existing
   Filippini Na→SBP dose-response** (2.8 mmHg/g raised-BP, 1.0 mmHg/g normal-BP),
   subject to the WHO 2 g/day floor.
3. **Potassium channel.** Baseline K intake (g/day, Reddin 2023) and the K added
   by LSS are converted to **achieved excretion (mmol/day)** — mg↔mmol via
   `r RC$lss$k_mg_per_mmol` mg/mmol, intake→excretion via ÷`r RC$lss$k_intake_to_excretion`
   — and mapped to an SBP change with the Filippini U-shaped achieved-excretion
   curve (reference 90 mmol/day), evaluated at baseline vs post-LSS excretion and
   differenced. The effect is modulated by baseline sodium (stronger at higher
   sodium) and **attenuated in normotensives** (factor
   `r formatC(RC$lss$nonhtn_factor, format="f", digits=2)`;
   `r if (isTRUE(RC$lss$nonhtn_attenuation)) "enabled" else "disabled"`).
4. **Combine.** The Na- and K-mediated SBP changes are combined
   **`r if (isTRUE(RC$lss$additive)) "additively" else "non-additively"`**
   (additivity factor `r formatC(RC$lss$additivity_factor, format="f", digits=2)`;
   a companion factor lets the Huang "100%→80%" sensitivity be a one-line change),
   and the combined ΔSBP drives IHD, stroke (ischemic + hemorrhagic) and HHD.

```{r lss-constants, message=FALSE, warning=FALSE, echo=FALSE}
lss_const <- data.table(
  Parameter = c("NaCl fraction", "KCl fraction", "Uptake", "Adherence",
                "Additive?", "Additivity factor", "Non-HTN K attenuation",
                "mg per mmol K", "Intake→excretion factor",
                "Whole-population coverage (s2)", "Harms modelled"),
  Value = c(fmt_pct(RC$lss$nacl_fraction,0), fmt_pct(RC$lss$kcl_fraction,0),
            formatC(RC$lss$uptake,format="f",digits=2), formatC(RC$lss$adherence,format="f",digits=2),
            as.character(RC$lss$additive), formatC(RC$lss$additivity_factor,format="f",digits=2),
            paste0(RC$lss$nonhtn_attenuation, " (x", formatC(RC$lss$nonhtn_factor,format="f",digits=2), ")"),
            as.character(RC$lss$k_mg_per_mmol), as.character(RC$lss$k_intake_to_excretion),
            formatC(RC$lss$coverage_all,format="f",digits=2), RC$lss$harms_modelled))
kable(lss_const, caption = "Table M3a. LSS Na/K→SBP parameters (from run_config.rds).",
      align = c("l","r")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE)
```

### NCD-RisC hypertension eligibility (LSS scenarios 4 & 5)

The diagnosed/treated LSS scenarios are sized by **NCD-RisC (Lancet 2021)**
population eligibility, **replacing the earlier BP ≥ 140 proxy**. These are
**age-standardised 30–79** estimates (NOT age-specific), annual to
`r RC$htn_source_year`, carried forward for later model years, by country and
sex. The eligible **population** shares are

$$
P(\text{diagnosed in pop}) = P(\text{HTN})\times P(\text{dx}\mid\text{HTN}),
\qquad
P(\text{treated in pop}) = P(\text{HTN})\times P(\text{tx}\mid\text{HTN}),
$$

used as scenario-4 and scenario-5 coverage respectively (scenario 2 uses the
whole-population toggle). 95% uncertainty intervals are carried through; the
population-share UI is a transparent product of the prevalence and conditional
bounds (not a formally propagated CI).

```{r ncdrisc-eligibility, message=FALSE, warning=FALSE, echo=FALSE}
he <- as.data.table(htn_eligibility)[location %in% RC$priority_countries]
# Knit-visible integrity check (Task 4c): every modelled country x sex present.
need <- CJ(location = RC$priority_countries, sex = c("Male","Female"))
miss_elig <- need[!he[!is.na(diagnosed_pop) & !is.na(treated_pop), .(location, sex)],
                  on = c("location","sex")]
he_tbl <- he[order(location, sex), .(
  Country = location, Sex = sex, `Source yr` = source_year,
  `HTN prev` = fmt_pct(htn_prev,1),
  `Dx | HTN` = fmt_pct(diagnosed_cond,1), `Tx | HTN` = fmt_pct(treated_cond,1),
  `Diagnosed pop` = fmt_pct(diagnosed_pop,1), `Treated pop` = fmt_pct(treated_pop,1))]
kable(he_tbl, caption = paste0("Table M3b. NCD-RisC hypertension eligibility by ",
      "country × sex (age-standardised 30–79, ", RC$htn_source_year,
      "). Diagnosed/treated pop shares size LSS scenarios 4/5."),
      align = c("l","l", rep("r",6))) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 12) |>
  row_spec(0, bold = TRUE)
```

`r if (nrow(miss_elig) > 0) sprintf("⚠️ **Missing eligibility inputs** for: %s. LSS scenarios 4/5 cannot be sized for these — investigate before using results.", paste(sprintf("%s/%s", miss_elig$location, miss_elig$sex), collapse=", ")) else "All modelled country × sex combinations have NCD-RisC eligibility inputs."`

### Baseline potassium intake

Baseline potassium is from **Reddin et al. 2023** (country × sex, g/day). Viet Nam
and Ethiopia are absent from the source and inherit their **GBD region mean**
(Southeast Asia; Eastern Sub-Saharan Africa), flagged below.

```{r baseline-potassium, message=FALSE, warning=FALSE, echo=FALSE}
bk <- as.data.table(baseline_potassium)[location %in% RC$priority_countries]
bk_tbl <- bk[order(location, sex), .(
  Country = location, Sex = sex,
  `K intake (g/d)` = formatC(k_intake_g, format="f", digits=2),
  `95% CI` = paste0(formatC(k_lcl,format="f",digits=2), "–", formatC(k_ucl,format="f",digits=2)),
  Source = fifelse(source == "reddin2023_region_fallback", "region-mean fallback", "Reddin 2023"))]
kable(bk_tbl, caption = "Table M3c. Baseline potassium intake by country × sex (Reddin 2023; region-mean fallback flagged).",
      align = c("l","l","r","r","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 12) |>
  row_spec(0, bold = TRUE)
```

### LSS mechanistic quantities (audit)

Table M3d shows the full Na/K→SBP pathway quantities for the **whole-population
scenario (s2)** by country × sex, consumed from `lss_audit.rds` (the single
source of truth). The diagnosed/treated scenarios (s4/s5) differ only in the
eligible-population coverage and are in the same audit file.

```{r lss-audit-table, message=FALSE, warning=FALSE, echo=FALSE}
la <- as.data.table(lss_audit)
la_s2 <- la[grepl("^lss_s2", scenario) & method == "na_k_sbp"][order(location, sex)]
if (nrow(la_s2)) {
  la_tbl <- la_s2[, .(
    Country = location, Sex = sex,
    `Base Na (g)` = formatC(baseline_sodium_g, format="f", digits=2),
    `Base K (g)`  = formatC(baseline_potassium_g, format="f", digits=2),
    `Disc share`  = fmt_pct(discretionary_share,0),
    `Cov`         = fmt_pct(eligible_pop_share,0),
    `Na displ (g)` = formatC(sodium_displaced_g, format="f", digits=3),
    `K added (g)`  = formatC(potassium_added_g, format="f", digits=3),
    `ΔSBP Na` = formatC(sbp_delta_na_mmHg, format="f", digits=2),
    `ΔSBP K`  = formatC(sbp_delta_k_mmHg, format="f", digits=2),
    `ΔSBP tot` = formatC(sbp_delta_combined_mmHg, format="f", digits=2))]
  kable(la_tbl, caption = paste0("Table M3d. LSS Na/K→SBP pathway (scenario 2, whole-population) ",
        "by country × sex. Harms modelled: none (no CKD channel). Causes affected: all four CVD."),
        align = c("l","l", rep("r",8))) |>
    kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 11) |>
    row_spec(0, bold = TRUE)
} else cat("No na_k_sbp LSS rows in lss_audit.rds.\n")
```

### SSaSS trial-RR benchmark (sensitivity, not combined)

For **SSaSS-comparable populations** (diagnosed/treated hypertension) the report
also reports an **SSaSS trial-RR benchmark** (method `ssass_trial_rr`), if it was
run (scenarios `lss_s4_ssass` / `lss_s5_ssass`). It applies the SSaSS stroke
relative risks directly on `(IR, CF)`:

- **stroke incidence RR `r formatC(LSS_RR_STROKE_INC, format="f", digits=2)`** (ischemic + hemorrhagic), and
- a **conditional case-fatality multiplier `r formatC(LSS_CF_MULT, format="f", digits=3)`**
  (= 0.77/0.90; total-stroke RR `r formatC(LSS_STROKE_TOTAL, format="f", digits=2)` = 0.90 × 0.856, not applied separately).

Non-stroke causes are unaffected. **This benchmark is NOT combined with the
Na/K→SBP method** — they are distinct scenarios driven by a single pathway each
(the model enforces one pathway per cause), and SSaSS RRs are **not applicable to
the whole population**, so no `lss_s2` benchmark exists (guarded in code). The
comparison of the two methods is in the Results section.

### How this LSS model compares with Huang et al. (2026)

Our mechanistic LSS model follows the **structure** of Huang et al. 2026
\autocite{huang2026} — the sodium **and** potassium → SBP → disease pathway,
75/25 NaCl/KCl, Filippini (2020) potassium dose-response, Reddin (2023) baseline
potassium — but is deliberately **narrower in scope**. It is embedded in our
*dynamic* 10-country CVD state-transition model (not a static 193-country
comparative risk assessment), covers our four SBP-mediated CVD causes (not
Huang's broad CVD + CKD basket), and is **benefit-only**: unlike Huang, we do not
model the hyperkalaemia/CKD harm channel. Table 3a positions the two side by
side. The differences are choices of scope and integration, **not** corrections
to Huang.

```{r huang-comparison, message=FALSE, warning=FALSE, echo=FALSE}
lc <- RC$lss
huang_cmp <- data.frame(
  Component = c(
    "Modelling approach", "Geographic scope", "LSS formulation",
    "Sodium → SBP", "Potassium → SBP", "Baseline potassium",
    "Population scenarios", "Hypertension targeting", "CVD outcomes",
    "CKD / hyperkalaemia harms", "Uncertainty", "Role of SSaSS trial"),
  `Huang et al. 2026` = c(
    "Static comparative risk assessment (SBP-distribution shift); one time point",
    "193 countries/territories",
    "75% NaCl / 25% KCl (15/85 in sensitivity)",
    "Mozaffarian 2014 meta-regression, by age and hypertension status; WHO 2 g/day floor",
    "Filippini 2020 non-linear; stronger at higher baseline sodium; additive (80% in a sensitivity)",
    "Reddin 2023 (global mean ≈ 2.3 g/day)",
    "5: all salt / discretionary / non-discretionary / diagnosed HTN / treated HTN",
    "NCD-RisC 2019; 80% uptake (20% excluded for hyperkalaemia risk)",
    "Broad CVD basket (IHD; ischaemic, intracerebral & subarachnoid stroke; HHD; AF; PAD …) + CKD",
    "Modelled as a NET effect for population-wide scenarios (benefit − hyperkalaemia harm)",
    "Probabilistic (1000 Monte Carlo) + 11 deterministic sensitivity analyses",
    "Composition & adherence only (NOT the SBP effect size)"),
  `Current UW/RTSL model` = c(
    "Dynamic Well→Sick→Dead state-transition; annual projection to 2050",
    paste0(n_countries, " RTSL priority countries"),
    paste0(fmt_pct(lc$nacl_fraction, 0), " NaCl / ", fmt_pct(lc$kcl_fraction, 0), " KCl"),
    "Filippini Na→SBP (2.8 mmHg/g raised-BP; 1.0 mmHg/g normal-BP); WHO 2 g/day floor",
    paste0("Filippini 2020 achieved-excretion U-curve; baseline-Na modulation; non-HTN ×",
           formatC(lc$nonhtn_factor, format = "f", digits = 2), "; additive"),
    "Reddin 2023 (country × sex; region-mean fallback: Viet Nam, Ethiopia)",
    "3: whole-population discretionary (s2), diagnosed HTN (s4), treated HTN (s5) — Huang's 2/4/5",
    paste0("NCD-RisC ", RC$htn_source_year,
           " diagnosed/treated population shares; uptake ",
           formatC(lc$uptake, format = "f", digits = 2), " × adherence ",
           formatC(lc$adherence, format = "f", digits = 2)),
    "4 causes: IHD, ischaemic stroke, intracerebral haemorrhage, HHD (all via SBP)",
    paste0(lc$harms_modelled, " — benefit-only"),
    "Point estimates; SSaSS trial-RR as a sensitivity benchmark; input CIs shown, not propagated",
    "Optional direct stroke-RR benchmark (HTN scenarios only); separate pathway, never combined"),
  check.names = FALSE, stringsAsFactors = FALSE)

kable(huang_cmp,
      caption = paste0("Table 3a. LSS mechanistic model vs Huang et al. (2026). ",
                       "Both share the Na+K→SBP structure, 75/25 formulation, ",
                       "Filippini 2020 K dose-response and Reddin 2023 baseline K; ",
                       "the differences are scope/integration, not corrections."),
      align = c("l", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1, bold = TRUE)
```

## LSS: Huang et al. (2026) vs. our current implementation

\scriptsize

\begin{columns}[T]

\begin{column}{0.49\textwidth}
\textbf{Huang et al.: static comparative risk assessment}

\[
Na^1 = Na^0 - 0.25\,Na^{R},
\qquad
K^1 = K^0 + \Delta K
\]

\[
\Delta SBP
=
\Delta SBP_{Na}
+
\Delta SBP_{K}
\]

Post-LSS SBP distribution is shifted:

\[
X^1 \sim N(\mu+\Delta SBP,\sigma^2)
\]

Disease impact is calculated using the full RR function:

\[
PIF
=
1-
\frac{E_{P^1}[RR(X)]}
     {E_{P^0}[RR(X)]}
\]

\[
D^{averted}=PIF\times D^0
\]

\textbf{Net effect also subtracts hyperkalaemia harm:}

\[
D^{net}
=
D^{averted}_{SBP}
-
D^{hyperK}_{CKD}
\]

\end{column}

\begin{column}{0.49\textwidth}
\textbf{UW--RTSL: dynamic state-transition implementation}

LSS displaces discretionary sodium:

\[
\Delta Na
=
Na^0\,q_D(0.25)(0.92)(u)(c_t)
\]

and adds potassium stoichiometrically:

\[
\Delta K
=
\Delta Na
\frac{0.524}{0.393}
\approx1.33\,\Delta Na
\]

SBP reduction is:

\[
\Delta SBP_{Na}
=
[2.8H+1.0(1-H)]\Delta Na
\]

\[
\Delta SBP
=
\Delta SBP_{Na}
+
m_{Na}\,m_H\,\Delta SBP_K
\]

where $\Delta SBP_K$ comes from the Filippini K-excretion curve.

Rather than a PIF, $\Delta SBP$ modifies annual incidence:

\[
IR'_{b,c}
=
IR_{b,c}
\left[
1-(1-RR^{10}_{b,c})
\frac{\Delta SBP}{10}
\right]
\]

and the new incidence enters the annual population/state-transition model:

\[
Pop_t \rightarrow Cases_t \rightarrow Deaths_t
\]

\textbf{CF unchanged:}\quad $CF'=CF$.

\end{column}
\end{columns}

\medskip

\begin{alertblock}{Key interpretation}
\scriptsize
The biological pathway is intentionally similar --- \textbf{LSS $\to$ lower Na + higher K $\to$ lower SBP $\to$ lower CVD risk} --- but the mathematical health-impact engine differs. Huang applies a \textbf{distributional PIF to current disease burden} and explicitly models CKD/hyperkalaemia harms; our model applies the Na/K-induced SBP change to \textbf{cause-specific incidence inside a dynamic 2026--2050 state-transition model}. The current implementation is therefore \textbf{benefit-only for LSS} and does not reproduce Huang's PIF or hyperkalaemia module.
\end{alertblock}

## Fiscal policy (exploratory)

Fiscal policy is modelled as an **explicit, still-simple pathway** and is labelled
**exploratory/provisional** throughout. It replaces the earlier single opaque
multiplier; it does **not** imply "a 10% tax ⇒ 10% sodium reduction", nor that
**all** packaged sodium is taxable. The chain is

$$
\text{tax rate} \to \text{pass-through} \to \text{price change} \to
\text{elasticity-based quantity response} \to \text{substitution} \to
\Delta\text{taxable packaged sodium} \to \Delta\text{total packaged sodium},
$$

with the last step multiplying by a **taxable-packaged share < 1** (the fraction
of packaged sodium in taxed nutrient-profile categories). The base case is
anchored on **Saxena et al. 2025** (Philippines): a 20% tax → ~10% relative
reduction in *taxed* packaged sodium, consumer (quantity) response dominant,
reformulation minor. Applying broad food-category price elasticities to
high-sodium foods specifically is uncertain (a peer-review caveat). The sensitivity
set brackets the base case. All values are read from `fiscal_audit.rds`.

```{r fiscal-audit-table, message=FALSE, warning=FALSE, echo=FALSE}
fa <- as.data.table(fiscal_audit)
fa_tbl <- fa[, .(
  Variant = variant,
  `Tax`         = fmt_pct(tax_rate, 0),
  `Pass-thru`   = fmt_pct(pass_through, 0),
  `Price Δ`     = fmt_pct(price_change, 0),
  `Elasticity`  = formatC(elasticity, format = "f", digits = 2),
  `Qty response`= fmt_pct(quantity_response, 1),
  `Taxable share` = fmt_pct(taxable_share, 0),
  `Substitution`  = fmt_pct(substitution, 0),
  `Reformulation` = fmt_pct(reformulation, 0),
  `Δ taxed packaged`  = fmt_pct(taxed_packaged_reduction, 1),
  `Δ total packaged`  = fmt_pct(packaged_reduction, 1))]
kable(fa_tbl, caption = paste0("Table M5. Fiscal policy pathway by sensitivity ",
      "variant (EXPLORATORY). 'Δ total packaged' = reduction in TOTAL packaged ",
      "sodium = Δ taxed packaged × taxable share. Ramp: ", RC$scaleup_window[1],
      "–", RC$scaleup_window[2], ". Source: Saxena et al. 2025; fiscal_audit.rds."),
      align = c("l", rep("r", 10))) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 11) |>
  row_spec(0, bold = TRUE)
```

The full package uses the **base** fiscal effect if
`include_fiscal_in_package = `r RC$fiscal$include_in_package``; the low/base/high
variants also run as their own **individual** scenarios (`fiscal_low`,
`fiscal_base`, `fiscal_high`) so the sensitivity is visible in the results.

### How this fiscal model compares with Saxena et al. (2025)

Saxena et al. 2025 \autocite{saxena2025} — the Philippines tax modelling study
that anchors our base case — is a full behavioural-economic evaluation. Our
implementation intentionally **collapses** its structure into a transparent,
transferable multi-country approximation for the Bloomberg exercise: one own-price
elasticity, one pass-through, one taxable-packaged share, and flat
substitution/reformulation offsets, feeding the same sodium → SBP → CVD engine as
every other policy. Table M5b positions the two side by side. Saxena provides the
conceptual and empirical **anchor** (its base case is a 20% tax with ~100%
pass-through giving ~10% lower taxed-food purchases — the calibration point for
our base variant); it does **not** provide a generic elasticity that can simply be
transported to all countries.

```{r saxena-comparison, message=FALSE, warning=FALSE, echo=FALSE}
fb   <- RC$fiscal$base
fsen <- RC$fiscal$sensitivity
faB  <- as.data.table(fiscal_audit)[variant %in% c("fiscal_base", "fiscal")][1]
rng  <- function(x) paste0(x[1], " / ", x[2], " / ", x[3])
saxena_cmp <- data.frame(
  Component = c(
    "Setting & model", "Time horizon", "Tax rate", "What is taxed",
    "Pass-through", "Demand response", "Wealth / equity",
    "Substitution & reformulation", "Packaged-food trend", "Health pathway",
    "Diseases modelled", "Sodium reduction", "Economic evaluation", "Uncertainty"),
  `Saxena et al. 2025 (Philippines)` = c(
    "Philippines; proportional multistate life-table (Markov) cohort",
    "20 years (results at 5/10/20 y)",
    "20% excise (40% in sensitivity)",
    "WHO WPRO nutrient-profile thresholds; 17 packaged categories; sodium + sugar",
    "100% base (50% / 150% in sensitivity)",
    "QUAIDS demand system; own- AND cross-price elasticities",
    "Wealth-quintile-specific elasticities; equity outcomes reported",
    "Within-category taxed↔untaxed substitution; endogenous threshold-distance reformulation",
    "Euromonitor 2008–19, dynamic; total intake held via offset assumption",
    "Sodium→SBP→CVD AND sugar→BMI→type-2 diabetes",
    "Type-2 diabetes, IHD, ischaemic stroke (incidence only)",
    "≈ 100 mg/day (≈ 10% relative) at year 20",
    "Full: HALYs, health-care savings, tax revenue (~US$13 B), ICER, equity",
    "Probabilistic (1000 Monte Carlo) + 5 deterministic scenarios"),
  `Current UW/RTSL fiscal model` = c(
    "Multi-country; embedded in the shared Well→Sick→Dead state-transition engine",
    paste0("Annual to 2050 (scale-up ", SCALEUP_YEAR1, "–", SCALEUP_YEAR2, ")"),
    paste0("base ", fmt_pct(fb$tax_rate, 0), " (low/base/high ",
           rng(sapply(list(fsen$fiscal_low, fsen$fiscal_base, fsen$fiscal_high),
                      function(z) fmt_pct(z$tax_rate, 0))), ")"),
    paste0("Flat taxable-packaged share (base ", fmt_pct(fb$taxable_share, 0),
           "); NO item-level nutrient profiling"),
    paste0("base ", fmt_pct(fb$pass_through, 0), " (low/base/high ",
           rng(sapply(list(fsen$fiscal_low, fsen$fiscal_base, fsen$fiscal_high),
                      function(z) fmt_pct(z$pass_through, 0))), ")"),
    paste0("Single own-price elasticity (low/base/high ",
           rng(sapply(list(fsen$fiscal_low, fsen$fiscal_base, fsen$fiscal_high),
                      function(z) formatC(z$elasticity, format = "f", digits = 2))),
           "); NO cross-price terms"),
    "None — a single national response (no wealth stratification)",
    paste0("Flat offsets only (base substitution ", fmt_pct(fb$substitution, 0),
           ", reformulation ", fmt_pct(fb$reformulation, 0),
           "; varied in low/high)"),
    "Euromonitor packaged trend; composition-only, total intake held",
    "Sodium→SBP→CVD only (no sugar / BMI / diabetes pathway)",
    "IHD, ischaemic stroke, intracerebral haemorrhage, HHD (incidence only)",
    paste0("base ≈ ", fmt_pct(faB$taxed_packaged_reduction, 0),
           " of taxed packaged → ", fmt_pct(faB$packaged_reduction, 1),
           " of packaged sodium"),
    "Health impact only — no costing, tax revenue or ICER (yet)",
    "Deterministic low / base / high variants only"),
  check.names = FALSE, stringsAsFactors = FALSE)

kable(saxena_cmp,
      caption = paste0("Table M5b. Fiscal model vs Saxena et al. (2025). ",
                       "Saxena is the conceptual/empirical anchor; ours is a ",
                       "deliberately simpler, transferable approximation. Values in ",
                       "the right column are read from run_config.rds / fiscal_audit.rds."),
      align = c("l", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1, bold = TRUE)
```

**Elements of Saxena et al. that our model does *not* reproduce:** the full
QUAIDS demand system; category-specific own- and cross-price elasticities;
wealth-quintile heterogeneity and equity outcomes; item/category-specific tax
eligibility (we use one flat taxable-packaged share); endogenous, product-level
reformulation; the sugar→BMI→diabetes pathway; and the full economic evaluation
(tax revenue, implementation cost, ICER). These are deliberate simplifications
for a transparent multi-country exercise, not oversights — and each is a candidate
for future refinement if RTSL prioritises fiscal policy.

## Composition math

For a **package** of interventions, effects on the *same* source are composed
so that overlapping policies never double-count or exceed 100%. For source $s$
with intervention effects $\text{eff}_{i,s}$:

$$
\text{composed}_s \;=\; 1 - \prod_{i}\bigl(1 - \text{eff}_{i,s}\bigr).
$$

The composed per-source effects are then **share-weighted** and summed to give
the single total-intake reduction fraction — exactly the `salteff` the
downstream pathway expects:

$$
\boxed{\;\text{salteff} \;=\; \sum_{s}\, \text{share}_s \cdot \text{composed}_s\;}
$$

**Worked sanity check.** For public procurement, only the `public` source is
affected (effect `r fmt_pct(intervention_effects[intervention=="public_procurement", effect],0)`),
so at base year $\text{salteff} = 0.20 \times \text{public share}$.

```{r sanity-check, message=FALSE, warning=FALSE, echo=FALSE}
chk        <- as.numeric(compute_total_efficacy("public_procurement", "Viet Nam"))
vn_public  <- source_shares[location == "Viet Nam", public]
```

For Viet Nam the base-year `public` share is
**`r formatC(vn_public, format="f", digits=3)`**, so
$\text{salteff}_{\text{Viet Nam, public procurement}} = 0.20 \times
`r formatC(vn_public, format="f", digits=3)` =$
**`r formatC(chk, format = "f", digits = 4)`** (the in-report recomputation of the
pipeline's formula). Note `salteff` is now **year-specific** for packaged-source
policies — the value above is the base-year (`r RC$euromonitor$base_year`) figure.

## From `salteff` to health impact

Given a country's `salteff`, the sodium→SBP interventions reduce sodium intake
and — through blood pressure — CVD incidence, with case fatality **not** modified
(steps 1–6 below). The **primary LSS** scenarios use the mechanistic Na/K→SBP
method: the combined (sodium + potassium) SBP change flows through the SAME
sodium→SBP machinery (affecting all four causes). The optional **SSaSS trial-RR
benchmark** scenarios instead modify stroke **incidence and case fatality**
directly (see the LSS section) and bypass the sodium→SBP cut for stroke; the two
LSS methods are never combined.

1. **Target reduction (grams).** With `salt` the baseline sodium column and the
   `"percent"` method used here, the targeted reduction is
   $\text{salt\_target} = \text{salt}\cdot\text{salteff}$, subject to a
   **2 g/day minimum-intake floor**: if $\text{salt} - \text{salt\_target} < 2$
   then $\text{salt\_target} = \text{salt} - 2$.

2. **Linear scale-up.** Between `saltyear1` = `r SCALEUP_YEAR1` and `saltyear2`
   = `r SCALEUP_YEAR2` the reduction ramps linearly, then holds:
   $$
   \text{salt\_reduction}_t =
   \begin{cases}
   0 & t < \text{saltyear1}\\[4pt]
   \text{salt\_target}\cdot\dfrac{t - \text{saltyear1} + 1}{\text{saltyear2}-\text{saltyear1}+1} & \text{saltyear1}\le t \le \text{saltyear2}\\[10pt]
   \text{salt\_target} & t > \text{saltyear2}.
   \end{cases}
   $$

3. **Sodium → SBP (Filippini dose–response).** The SBP reduction depends on
   whether the individual has raised BP (fraction $\text{raisedBP}$):
   $$
   \Delta SBP = \bigl(2.8\cdot\text{raisedBP} + 1.0\cdot(1-\text{raisedBP})\bigr)\cdot \text{salt\_reduction}.
   $$

4. **Baseline bin-specific incidence (GBD).** Within each BP bin $b$, baseline
   incidence is scaled by the GBD log-linear RR per 10 mmHg, normalised so the
   population-average recovers the observed $IR$:
   $$
   IR_{\text{bin},b} = \frac{RR^{\text{GBD}}_{b}\cdot IR}{\alpha},
   \qquad \alpha = \sum_b \text{prob}_b\, RR^{\text{GBD}}_{b},
   $$
   where $RR^{\text{GBD}}_b = \bigl(rr^{10}\bigr)^{(m_b-120)/10}$ for bin
   midpoint $m_b>120$ (else 1) and $\text{prob}_b$ is the share of the age/sex
   group in bin $b$ (from a normal BP distribution).

5. **SBP → incidence (ETTEHAD).** The proportional reduction in bin-specific
   incidence is driven by the ETTEHAD relative risk per 10 mmHg,
   $RR^{10}_{b,c}$, for BP bin $b$ and cause $c$, applied in proportion to the
   SBP reduction; incidence is then averaged over BP bins:
   $$
   IR^{\text{new}}_{c} = \sum_b \text{prob}_b\; IR_{\text{bin},b}\,
   \Bigl(1 - \bigl(1 - RR^{10}_{b,c}\bigr)\cdot 0.1\cdot \Delta SBP\Bigr),
   \qquad
   \text{eff\_ir} = \frac{IR^{\text{new}}_c}{IR_c}.
   $$
   The factor $0.1$ converts $\Delta SBP$ (mmHg) into units of 10 mmHg. The
   ETTEHAD inputs include diabetes-specific effect sizes that the pipeline
   blends by local diabetes prevalence; the applied incidence modifier is the
   per-10-mmHg RR above.

6. **Case fatality unchanged (sodium→SBP path).** $CF^{\text{new}} = CF$ and
   $\text{eff\_cf}=1$: the sodium→SBP interventions (including the **primary
   Na/K→SBP LSS** method) act on incidence only. Before `saltyear1`,
   $IR^{\text{new}} = IR$.

**Primary LSS (Na/K→SBP).** The LSS scenarios add a potassium-mediated SBP
change to the sodium-mediated one (step 3) and run the **combined** ΔSBP through
steps 4–5, so LSS affects incidence for **all four causes** via SBP (case
fatality unchanged, like the other sodium→SBP policies).

**SSaSS trial-RR benchmark exception.** For the optional benchmark scenarios
only, stroke incidence is scaled by the SSaSS RR
(`r formatC(LSS_RR_STROKE_INC, format="f", digits=2)`) and stroke **case
fatality** by the conditional multiplier `r formatC(LSS_CF_MULT, format="f", digits=3)`
(both blended toward 1 by the reached fraction and ramped); non-stroke causes are
unchanged. So for the benchmark, $\text{eff\_cf}\neq 1$ on stroke. The two LSS
methods are never combined.

The modified $IR$ (and, for the SSaSS benchmark, $CF$) then flow through the
same state-transition recurrences given above.

```{r flow-diagram, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=2.6}
boxes <- data.table(
  x = 1:6,
  label = c("Source shares\n× effect sizes",
            "salteff\nΣ share·composed",
            "ΔSBP\n(Filippini)",
            "ΔIncidence\n(ETTEHAD RR)",
            "State transitions\nWell→Sick→Dead",
            "Deaths\ndelayed")
)
arrows <- data.table(x = (1:5) + 0.42, xend = (2:6) - 0.42, y = 1, yend = 1)

ggplot() +
  geom_tile(data = boxes, aes(x = x, y = 1), width = 0.84, height = 0.7,
            fill = "#e8eef4", color = "#1f78b4", linewidth = 0.6) +
  geom_text(data = boxes, aes(x = x, y = 1, label = label), size = 3.1, lineheight = 0.95) +
  geom_segment(data = arrows, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.16, "cm"), type = "closed"),
               linewidth = 0.5, color = "grey30") +
  coord_cartesian(xlim = c(0.5, 6.5), ylim = c(0.55, 1.45)) +
  labs(title = "Causal chain: sources → salteff → SBP → incidence → deaths delayed") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 11))
```

## Scenarios

The scenario set is read from the pipeline's resolved registry (`run_config.rds`)
and intersected with the outputs (Table M3). The **`full_package`** membership is
`r paste(package_interventions, collapse = " + ")` — driven by the pipeline's
`include_fiscal_in_package = `r RC$fiscal$include_in_package`` (fiscal is
`r if (isTRUE(RC$fiscal$include_in_package)) "included" else "excluded"`). **LSS**
is not folded into the package; it runs as separate scenarios under the primary
**Na/K→SBP** method (whole-population, diagnosed-HTN, treated-HTN), plus the
optional SSaSS trial-RR benchmark for the HTN variants (see the LSS section).
**Media** and the fiscal sensitivity variants are handled per the registry.

```{r tbl-scenarios, message=FALSE, warning=FALSE, echo=FALSE}
scn_tbl <- data.table(
  scenario = scenario_levels,
  interventions = vapply(scenario_levels, function(s) {
    iv <- scenario_interventions[[s]]
    if (length(iv) == 0) "(none — baseline)" else paste(iv, collapse = " + ")
  }, character(1))
)
scn_tbl[, Scenario := scenario_labels[scenario]]
scn_tbl[, present := scenario %in% scenarios_present]
scn_disp <- scn_tbl[, .(Scenario, `Interventions (source-split)` = interventions,
                        `In outputs` = fifelse(present, "yes", "no"))]

kable(scn_disp,
      caption = "Table M3. Scenarios and their constituent interventions.",
      align = c("l", "l", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(which(scn_tbl$scenario %in% c("baseline", "full_package")), bold = TRUE)
```

The resulting `salteff` is **country-specific** because it depends on the local
source shares, and now **year-specific** (packaged shares trend). Table M4
illustrates the **base-year** value for a few representative countries, computed
from the pipeline's published base-year shares and effect sizes (`run_config.rds`)
— the same composition formula production uses, not a second model.

```{r tbl-salteff, message=FALSE, warning=FALSE, echo=FALSE}
ref_countries <- intersect(c("Viet Nam", "Bangladesh", "China", "Thailand", "default"),
                           source_shares$location)
salteff_grid <- rbindlist(lapply(salteff_scenarios, function(s) {
  iv <- scenario_interventions[[s]]
  vals <- vapply(ref_countries, function(cc)
    as.numeric(compute_total_efficacy(iv, cc, source_shares, intervention_effects)),
    numeric(1))
  data.table(Scenario = salteff_labels_short[s], t(vals))
}))
setnames(salteff_grid, c("Scenario", ref_countries))
num_cols <- ref_countries
salteff_grid[, (num_cols) := lapply(.SD, function(x) formatC(x, format = "f", digits = 4)),
             .SDcols = num_cols]

kable(salteff_grid,
      caption = paste0("Table M4. Illustrative salteff (fraction of TOTAL sodium ",
                       "reduced at full implementation) by scenario and country."),
      align = c("l", rep("r", length(ref_countries)))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "salteff is largest where the intervention's target source dominates local ",
    "intake. The LSS (Na displacement) column is the share-weighted discretionary ",
    "SODIUM reduction that sizes the Na channel of the primary Na/K->SBP method; ",
    "it is NOT the full LSS health effect (the potassium-mediated SBP change is ",
    "added inside the model and is not captured by this salteff). Base-year ",
    "values; salteff is year-specific for packaged-source policies."))
```

The stacked bars below show the same illustrative `salteff` for **every
modelled country** (all ten prioritised countries, each on its own documented
source profile), one panel per scenario. Each bar's total height is the country's `salteff`
(fraction of *total* sodium reduced at full implementation); the segments
decompose it into the per-source contributions ($\text{share}_s \times
\text{composed}_s$) that sum to that total.

```{r fig-salteff-bar, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7}
# Per-(scenario, country) source decomposition of salteff, for all modelled
# countries. Uses the same source colours as the source-share barplot.
salteff_bar <- rbindlist(lapply(salteff_scenarios, function(s) {
  iv <- scenario_interventions[[s]]
  rbindlist(lapply(plot_countries, function(cc) {
    te  <- compute_total_efficacy(iv, cc, source_shares, intervention_effects)
    dec <- copy(attr(te, "decomposition"))
    dec[, `:=`(scenario = s, location = cc)]
    dec[]
  }))
}))
salteff_bar[, source   := factor(source, levels = SODIUM_SOURCES)]
salteff_bar[, scn_lbl  := factor(salteff_labels_short[scenario],
                                 levels = salteff_labels_short[salteff_scenarios])]

p <- ggplot(salteff_bar[scenario!="full_package",], aes(location, contribution, fill = source)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.15) +
  facet_wrap(~ scn_lbl, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Illustrative salteff by scenario and country",
       subtitle = paste0("Fraction of TOTAL sodium reduced at full ",
                         "implementation; stacked by source contribution"),
       x = NULL, y = "salteff (fraction of total sodium reduced)",
       caption = paste0("Bar height = country salteff; note free y-axes across ",
                        "panels. LSS shown at its share-weighted value.")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
print(p)
```

```{r fig-salteff-bar2, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=7}
# Per-(scenario, country) source decomposition of salteff, for all modelled
# countries. Uses the same source colours as the source-share barplot.
p <- ggplot(salteff_bar[scenario=="full_package",], aes(location, contribution, fill = source)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.15) +
  facet_wrap(~ scn_lbl, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Illustrative salteff by scenario and country",
       subtitle = paste0("Fraction of TOTAL sodium reduced at full ",
                         "implementation; stacked by source contribution"),
       x = NULL, y = "salteff (fraction of total sodium reduced)",
       caption = paste0("Bar height = country salteff; note free y-axes across ",
                        "panels. LSS shown at its share-weighted value.")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
print(p)
```

---

# Baseline (pre-intervention) time series

The panels below use the **`baseline`** scenario only. Population denominators
use the de-duplicated total population (a single cause row); counts sum
cause-specific values over the four causes. Incidence begins in 2018 (2017 is
the seed year, in which `newcases` = 0 by construction).

```{r baseline-aggregates, message=FALSE, warning=FALSE, echo=FALSE}
base <- data.out[scenario == "baseline"]

# Total (de-duplicated) population per year: take one cause row.
base_pop  <- base[cause == "ihd", .(pop = sum(pop, na.rm = TRUE)), by = year]

# All-cause counts per year.
base_all  <- base[, .(sick = sum(sick, na.rm = TRUE),
                      newcases = sum(newcases, na.rm = TRUE),
                      dead = sum(dead, na.rm = TRUE)), by = year]
base_all  <- merge(base_all, base_pop, by = "year")
base_all[, `:=`(prev_rate = sick / pop * 1000,
                inc_rate  = newcases / pop * 1000,
                mort_rate = dead / pop * 1e5)]

# By-cause counts per year (denominator = total population, same for each cause).
base_cause <- base[, .(sick = sum(sick, na.rm = TRUE),
                       newcases = sum(newcases, na.rm = TRUE),
                       dead = sum(dead, na.rm = TRUE)), by = .(cause, year)]
base_cause <- merge(base_cause, base_pop, by = "year")
base_cause[, `:=`(prev_rate = sick / pop * 1000,
                  inc_rate  = newcases / pop * 1000,
                  mort_rate = dead / pop * 1e5)]
base_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]

hist_cut <- 2024  # solid = historical, dashed = projected
```

## Prevalence over time

```{r baseline-prevalence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
p_all <- ggplot(base_all, aes(year, prev_rate, linetype = year > hist_cut)) +
  geom_line(linewidth = 1.1, color = "#b2182b") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "Prevalent CVD cases per 1,000 population — baseline (all causes)",
       subtitle = paste0(n_countries, " countries; solid = historical, dashed = projected"),
       x = "Year", y = "Prevalent cases per 1,000",
       caption = "Sum of cause-specific prevalent cases / total population.") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
print(p_all)

p_cause <- ggplot(base_cause, aes(year, prev_rate, color = cause_lbl,
                                  linetype = year > hist_cut)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(title = "Prevalent CVD cases per 1,000 population — baseline, by cause",
       x = "Year", y = "Prevalent cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p_cause)
```

## Incidence over time

```{r baseline-incidence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
inc_all   <- base_all[year >= 2018]
inc_cause <- base_cause[year >= 2018]

p_all <- ggplot(inc_all, aes(year, inc_rate, linetype = year > hist_cut)) +
  geom_line(linewidth = 1.1, color = "#2166ac") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "Incident CVD cases per 1,000 population — baseline (all causes)",
       subtitle = "New cases per 1,000 population per year; 2017 seed year omitted",
       x = "Year", y = "Incident cases per 1,000",
       caption = "Sum of cause-specific new cases / total population.") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
print(p_all)

p_cause <- ggplot(inc_cause, aes(year, inc_rate, color = cause_lbl,
                                 linetype = year > hist_cut)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(title = "Incident CVD cases per 1,000 population — baseline, by cause",
       x = "Year", y = "Incident cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p_cause)
```

## Mortality over time

```{r baseline-mortality, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.5}
p_all <- ggplot(base_all, aes(year, dead / 1e6, linetype = year > hist_cut)) +
  geom_line(linewidth = 1.1, color = "#4d4d4d") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — baseline (all causes)",
       subtitle = paste0(n_countries, " countries; millions of deaths per year"),
       x = "Year", y = "Deaths per year (millions)",
       caption = "Sum of cause-specific deaths.") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
print(p_all)

p_cause <- ggplot(base_cause, aes(year, dead / 1e6, color = cause_lbl,
                                  linetype = year > hist_cut)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — baseline, by cause",
       x = "Year", y = "Deaths per year (millions)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p_cause)
```

## Age-standardised mortality rate (ASMR)

```{r baseline-asmr, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# All-cause ASMR: deaths summed over causes; Nx = de-duplicated population.
asmr_all_in <- base[, .(deaths_dx = sum(dead, na.rm = TRUE)), by = .(age, year)]
pop_ay      <- base[cause == "ihd", .(Nx = sum(pop, na.rm = TRUE)), by = .(age, year)]
asmr_all_in <- merge(asmr_all_in, pop_ay, by = c("age", "year"))
asmr_all_in[, grp := "All CVD"]
asmr_all <- calculate_asmr(asmr_all_in, disagg_var = "grp")

# By-cause ASMR: Nx = well+sick (per cause = total living population).
asmr_c_in <- base[, .(deaths_dx = sum(dead, na.rm = TRUE),
                      Nx = sum(well, na.rm = TRUE) + sum(sick, na.rm = TRUE)),
                  by = .(cause, age, year)]
asmr_cause <- calculate_asmr(asmr_c_in, disagg_var = "cause")
asmr_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]

p <- ggplot() +
  geom_line(data = asmr_cause, aes(year, ASMR, color = cause_lbl,
                                   linetype = year > hist_cut), linewidth = 1) +
  geom_line(data = asmr_all, aes(year, ASMR, linetype = year > hist_cut),
            color = "black", linewidth = 1.2) +
  annotate("text", x = year_min + 1, y = max(asmr_all$ASMR),
           label = "All CVD", hjust = 0, vjust = 1, fontface = "bold", size = 3.6) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"), guide = "none") +
  labs(title = "Age-standardised CVD mortality rate — baseline",
       subtitle = "Per 100,000, WHO world standard population (ages 20–95+)",
       x = "Year", y = "ASMR per 100,000",
       caption = "Black = all CVD; coloured = by cause. Solid = historical, dashed = projected.") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

---

# Projected time series (all scenarios)

The same three metrics, now overlaying **every scenario** — baseline, each
individual intervention, and the full package — across the projection horizon.
Interventions begin scaling up in `r SCALEUP_YEAR1`, so scenarios coincide
before then and fan out thereafter.

```{r projection-aggregates, message=FALSE, warning=FALSE, echo=FALSE}
# All-cause per scenario x year.
proj_pop <- data.out[cause == "ihd", .(pop = sum(pop, na.rm = TRUE)),
                     by = .(scenario, year)]
proj_all <- data.out[, .(sick = sum(sick, na.rm = TRUE),
                         newcases = sum(newcases, na.rm = TRUE),
                         dead = sum(dead, na.rm = TRUE)),
                     by = .(scenario, year)]
proj_all <- merge(proj_all, proj_pop, by = c("scenario", "year"))
proj_all[, `:=`(prev_rate = sick / pop * 1000,
                inc_rate  = newcases / pop * 1000,
                mort_rate = dead / pop * 1e5)]

# By cause per scenario x year.
proj_cause <- data.out[, .(sick = sum(sick, na.rm = TRUE),
                           newcases = sum(newcases, na.rm = TRUE),
                           dead = sum(dead, na.rm = TRUE)),
                       by = .(scenario, cause, year)]
proj_cause <- merge(proj_cause, proj_pop, by = c("scenario", "year"))
proj_cause[, `:=`(prev_rate = sick / pop * 1000,
                  inc_rate  = newcases / pop * 1000,
                  mort_rate = dead / pop * 1e5)]
proj_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]

scn_scale <- function(name = NULL)
  list(
    scale_color_manual(values = scenario_cols, labels = scenario_labels_short,
                       breaks = scenario_levels, name = name),
    guides(color = guide_legend(nrow = 2, byrow = TRUE))
  )
```

## Prevalence

```{r proj-prevalence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
p <- ggplot(proj_all[year >= 2020], aes(year, prev_rate, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  labs(title = "Prevalent CVD cases per 1,000 population — all scenarios",
       subtitle = "All causes combined", x = "Year", y = "Prevalent cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)

p <- ggplot(proj_cause[year >= 2020], aes(year, prev_rate, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ cause_lbl, scales = "free_y") +
  scn_scale() +
  labs(title = "Prevalent CVD cases per 1,000 population — by cause",
       x = "Year", y = "Prevalent cases per 1,000") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

## Incidence

```{r proj-incidence, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
p <- ggplot(proj_all[year >= 2020], aes(year, inc_rate, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  labs(title = "Incident CVD cases per 1,000 population — all scenarios",
       subtitle = "All causes combined", x = "Year", y = "Incident cases per 1,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)

p <- ggplot(proj_cause[year >= 2020], aes(year, inc_rate, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ cause_lbl, scales = "free_y") +
  scn_scale() +
  labs(title = "Incident CVD cases per 1,000 population — by cause",
       x = "Year", y = "Incident cases per 1,000") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

To isolate the **relative** change each policy drives — stripping out the shared
demographic trend — the figure below re-expresses all-cause incidence per 1,000
as an **index** with the intervention/base year (`r SCALEUP_YEAR1`) set to 100
for every scenario. Scenarios coincide at the base year and then fan out as the
intervention scales up; lines below 100 indicate incidence falling faster than
its base-year level.

```{r proj-incidence-indexed, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
# Index each scenario's all-cause incidence rate to its base-year value = 100.
base_idx_year <- SCALEUP_YEAR1
inc_idx <- proj_all[year >= base_idx_year, .(scenario, year, inc_rate)]
base_inc <- inc_idx[year == base_idx_year, .(scenario, base_val = inc_rate)]
inc_idx  <- merge(inc_idx, base_inc, by = "scenario", all.x = TRUE)
inc_idx  <- inc_idx[!is.na(base_val) & base_val > 0]
inc_idx[, inc_index := inc_rate / base_val * 100]

if (nrow(inc_idx) > 0) {
  p <- ggplot(inc_idx, aes(year, inc_index, color = scenario)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1) +
    scn_scale() +
    labs(title = "Incident CVD cases per 1,000 — indexed to base year",
         subtitle = paste0("All causes; ", base_idx_year,
                          " = 100 for every scenario (relative change over time)"),
         x = "Year", y = paste0("Incidence index (", base_idx_year, " = 100)")) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank())
  print(p)
} else {
  cat("Indexed incidence series unavailable (no base-year data).\n")
}
```

## Mortality

```{r proj-mortality, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=4.8}
p <- ggplot(proj_all[year >= 2020], aes(year, dead / 1e6, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — all scenarios",
       subtitle = "All causes combined (millions per year)",
       x = "Year", y = "Deaths per year (millions)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)

p <- ggplot(proj_cause[year >= 2020], aes(year, dead / 1e6, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ cause_lbl, scales = "free_y") +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "CVD deaths per year — by cause",
       x = "Year", y = "Deaths per year (millions)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

---

# Main results: deaths delayed {#main-results-deaths-delayed}

The headline outcome is **deaths delayed** relative to baseline, keyed on the
authoritative `scenario` column:

$$
\text{deaths\_delayed}_{\text{scenario}} = \text{deaths}_{\text{baseline}} - \text{deaths}_{\text{scenario}} .
$$

Cumulative totals are summed over **`r impact_window`** (the scale-up starts in
`r SCALEUP_YEAR1`, so earlier years contribute negligibly).

```{r deaths-delayed-compute, message=FALSE, warning=FALSE, echo=FALSE}
# Cumulative deaths delayed over the impact window, by scenario.
dd_cum <- deaths_delayed_by(data.out, by = character(0), years = impact_y1:impact_y2)
dd_cum[, scenario := as_scn_factor(as.character(scenario))]
setorder(dd_cum, scenario)
dd_cum[, `:=`(per_year = deaths_delayed / n_impact_years,
              pct_vs_base = deaths_delayed / baseline_dead)]

# Annual + cumulative time series, by scenario (full horizon).
dd_annual <- deaths_delayed_by(data.out, by = "year")
dd_annual[, scenario := as_scn_factor(as.character(scenario))]
setorder(dd_annual, scenario, year)
dd_annual[, cum_delayed := cumsum(deaths_delayed), by = scenario]

# Helpers for inline numbers (robust to scenarios absent from this run).
ddc <- function(scn) { v <- dd_cum[as.character(scenario) == scn, deaths_delayed]
  if (length(v) == 0) NA_real_ else v }
mm  <- function(scn) { v <- ddc(scn); if (is.na(v)) "n/a" else fmt_m(v) }
.fiscal_scn <- if ("fiscal_base" %in% scenarios_present) "fiscal_base" else
               if ("fiscal" %in% scenarios_present) "fiscal" else NA_character_
```

Over `r impact_window`, across `r n_countries` countries:

- The **full package** (`r paste(package_interventions, collapse = " + ")`)
  delays **`r mm("full_package")` million** CVD deaths
  (`r fmt_pct(dd_cum[scenario=="full_package", pct_vs_base])` of baseline
  CVD deaths in the window; ~`r fmt_k(dd_cum[scenario=="full_package", per_year])`
  thousand per year). LSS is **not** in the package (it runs separately);
  fiscal membership follows `include_fiscal_in_package = `r RC$fiscal$include_in_package``.
- Among the packaged-food policies, **salt targets** contribute most
  (`r mm("salt_targets")` million), followed by **FOPL** (`r mm("fopl")` million)
  and the **exploratory fiscal policy** (base case
  `r if (!is.na(.fiscal_scn)) mm(.fiscal_scn) else "n/a"` million — see the
  fiscal section; low/high variants bracket this).
- **Public procurement** delays `r mm("public_procurement")` million
  (provisional; public share under revision). Media is excluded.
- **LSS — primary Na/K→SBP** runs as three scenarios: whole-population
  **s2** `r mm("lss_s2")` million, diagnosed-HTN **s4** `r mm("lss_s4")` million,
  treated-HTN **s5** `r mm("lss_s5")` million. These combine the sodium- and
  potassium-mediated SBP effects across all four CVD causes. The **SSaSS
  trial-RR benchmark** (stroke-only), where run, gives s4
  `r mm("lss_s4_ssass")` and s5 `r mm("lss_s5_ssass")` million — compared head-to-head
  in the Na/K-vs-SSaSS table below. Coverage is from NCD-RisC population
  eligibility (no BP≥140 proxy).

## Table 1 — Cumulative deaths delayed by scenario

```{r table1, message=FALSE, warning=FALSE, echo=FALSE}
tab1 <- dd_cum[as.character(scenario) != "baseline"]
tab1[, Scenario := scenario_labels[as.character(scenario)]]
tab1[, is_pkg := as.character(scenario) == "full_package"]
setorder(tab1, scenario)

tab1_fmt <- tab1[, .(
  Scenario,
  `Interventions` = vapply(as.character(scenario), function(s) {
    iv <- scenario_interventions[[s]]; paste(iv, collapse = " + ")
  }, character(1)),
  `Deaths delayed (millions)` = fmt_m(deaths_delayed, 3),
  `Avg per year (thousands)`  = fmt_k(per_year, 1),
  `% of baseline`             = fmt_pct(pct_vs_base, 2)
)]

kable(tab1_fmt,
      caption = paste0("Table 1. Cumulative deaths delayed by scenario, ",
                       impact_window, ", across ", n_countries, " countries."),
      align = c("l", "l", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(0, bold = TRUE) |>
  row_spec(which(tab1$is_pkg), bold = TRUE, background = "#eef2f7") |>
  footnote(general = paste0(
    "Deaths delayed = baseline deaths - scenario deaths, summed over ",
    impact_window, ". '% of baseline' is relative to baseline CVD deaths in ",
    "the same window. Full-package membership follows include_fiscal_in_package = ",
    RC$fiscal$include_in_package, " (LSS never in the package). LSS scenarios use ",
    "the primary Na/K->SBP method (all four causes via SBP); the SSaSS trial-RR ",
    "benchmark rows (if present) are stroke-only and also lower stroke case ",
    "fatality. Fiscal is EXPLORATORY; public-share and coverage assumptions are ",
    "provisional."))
```

```{r table1-dt, message=FALSE, warning=FALSE, echo=FALSE}
dt_show <- tab1[, .(
  Scenario,
  `Deaths delayed`   = round(deaths_delayed),
  `Avg per year`     = round(per_year),
  `% of baseline`    = round(100 * pct_vs_base, 2)
)]
datatable(dt_show, rownames = FALSE,
          caption = "Interactive version of Table 1 (absolute counts).",
          options = list(pageLength = 8, dom = "t")) |>
  formatCurrency(c("Deaths delayed", "Avg per year"), currency = "", digits = 0)
```

## Figure 1 — Cumulative deaths delayed by scenario

```{r figure1-bar, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
fig1 <- dd_cum[as.character(scenario) != "baseline"]
fig1[, scn_lbl := factor(scenario_labels_short[as.character(scenario)],
                         levels = rev(scenario_labels_short[scenario_nb]))]

p <- ggplot(fig1, aes(deaths_delayed / 1e6, scn_lbl, fill = scenario)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.4) +
  scale_fill_manual(values = scenario_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), labels = comma) +
  labs(title = paste0("Cumulative deaths delayed by scenario, ", impact_window),
       subtitle = paste0(n_countries, " countries; all CVD causes combined"),
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
print(p)
```

## Figure 2 — Annual deaths delayed over time

```{r figure2-annual, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
p <- ggplot(dd_annual[as.character(scenario) != "baseline" & year >= 2024],
            aes(year, deaths_delayed / 1e3, color = scenario)) +
  geom_line(linewidth = 1) +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "Annual deaths delayed by scenario",
       subtitle = "Thousands of CVD deaths delayed per year vs. baseline",
       x = "Year", y = "Deaths delayed per year (thousands)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

The absolute magnitudes above are dominated by the largest scenarios. Indexing
each scenario's annual deaths-delayed series to a common base year = 100 removes
that scale difference and exposes the **relative change per year** — how quickly
each scenario's annual benefit grows once the intervention takes hold. The base
year is the first post-scale-up year in which every scenario has a non-zero
annual benefit (deaths delayed in the very first scale-up year are ≈ 0 because
mortality responds to the prior year's incidence).

```{r figure2b-annual-indexed, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
# Index annual deaths delayed to a base intervention year = 100, per scenario.
dd_nb <- dd_annual[as.character(scenario) != "baseline"]
n_scn <- uniqueN(as.character(dd_nb$scenario))

# Base year: earliest year >= scale-up start where ALL scenarios are non-zero.
ok_years <- dd_nb[year >= SCALEUP_YEAR1 & deaths_delayed > 0, .N,
                  by = year][N == n_scn, year]
base_dd_year <- if (length(ok_years) > 0) min(ok_years) else
  suppressWarnings(min(dd_nb[deaths_delayed > 0, year]))

if (is.finite(base_dd_year)) {
  base_dd <- dd_nb[year == base_dd_year, .(scenario, base_val = deaths_delayed)]
  dd_idx  <- merge(dd_nb[year >= base_dd_year], base_dd, by = "scenario", all.x = TRUE)
  dd_idx  <- dd_idx[!is.na(base_val) & base_val > 0]
  dd_idx[, dd_index := deaths_delayed / base_val * 100]

  p <- ggplot(dd_idx, aes(year, dd_index, color = scenario)) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "grey60") +
    geom_line(linewidth = 1) +
    scn_scale() +
    scale_y_continuous(labels = comma) +
    labs(title = "Annual deaths delayed — indexed to base intervention year",
         subtitle = paste0(base_dd_year,
                          " = 100 for every scenario (relative change per year)"),
         x = "Year", y = paste0("Deaths-delayed index (", base_dd_year, " = 100)")) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank())
  print(p)
} else {
  cat("Indexed deaths-delayed series unavailable (no non-zero base year).\n")
}
```

## Figure 3 — Cumulative deaths delayed over time

```{r figure3-cumulative, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11, fig.height=5}
p <- ggplot(dd_annual[as.character(scenario) != "baseline"],
            aes(year, cum_delayed / 1e6, color = scenario)) +
  geom_line(linewidth = 1.1) +
  scn_scale() +
  scale_y_continuous(labels = comma) +
  labs(title = "Cumulative deaths delayed by scenario",
       subtitle = "Millions of CVD deaths delayed vs. baseline (running total)",
       x = "Year", y = "Cumulative deaths delayed (millions)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())
print(p)
```

## Figure 4 & Table 2 — Full package, by cause

```{r figure4-cause, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=5}
dd_cause <- deaths_delayed_by(data.out, by = "cause", years = impact_y1:impact_y2)
dd_cause <- dd_cause[as.character(scenario) == "full_package"]
dd_cause[, cause_lbl := factor(cause_map[cause], levels = unname(cause_map))]
setorder(dd_cause, -deaths_delayed)

p <- ggplot(dd_cause, aes(deaths_delayed / 1e6,
                          reorder(cause_lbl, deaths_delayed), fill = cause)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.4) +
  scale_fill_manual(values = cause_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12)), labels = comma) +
  labs(title = paste0("Full package: deaths delayed by cause, ", impact_window),
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
print(p)
```

```{r table2-cause, message=FALSE, warning=FALSE, echo=FALSE}
# Deaths delayed by scenario x cause (wide), impact window.
dd_sc <- deaths_delayed_by(data.out, by = "cause", years = impact_y1:impact_y2)
dd_sc[, scenario := as_scn_factor(as.character(scenario))]
dd_sc[, cause_lbl := cause_map[cause]]
wide <- dcast(dd_sc, scenario ~ cause_lbl, value.var = "deaths_delayed")
wide <- wide[as.character(scenario) != "baseline"]
setorder(wide, scenario)
wide[, Scenario := scenario_labels[as.character(scenario)]]
setcolorder(wide, c("Scenario", unname(cause_map)))
wide[, scenario := NULL]
for (cc in unname(cause_map))
  wide[, (cc) := fmt_m(get(cc), 3)]

kable(wide,
      caption = paste0("Table 2. Deaths delayed (millions) by scenario and cause, ",
                       impact_window, "."),
      align = c("l", rep("r", length(cause_map)))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 13) |>
  add_header_above(setNames(c(1L, length(cause_map)),
                            c(" ", "Deaths delayed, millions"))) |>
  row_spec(0, bold = TRUE)
```

## Country-level breakdown (full package)

```{r figure5-country, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9, fig.height=6.5}
dd_country <- deaths_delayed_by(data.out, by = "location", years = impact_y1:impact_y2)
dd_country <- dd_country[as.character(scenario) == "full_package"]
setorder(dd_country, -deaths_delayed)

topn <- head(dd_country, 20)
p <- ggplot(topn, aes(deaths_delayed / 1e6, reorder(location, deaths_delayed))) +
  geom_col(fill = "#1f78b4", width = 0.72) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.10)), labels = comma) +
  labs(title = "Full package: top 20 countries by deaths delayed",
       subtitle = impact_window,
       x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
print(p)
```

```{r table3-country-dt, message=FALSE, warning=FALSE, echo=FALSE}
ctab <- dd_country[, .(
  Country = location,
  `Deaths delayed`  = round(deaths_delayed),
  `% of baseline`   = round(100 * deaths_delayed / baseline_dead, 2)
)]
setorder(ctab, -`Deaths delayed`)
datatable(ctab, rownames = FALSE,
          caption = paste0("All ", n_countries,
                           " countries: full-package deaths delayed, ",
                           impact_window, " (searchable / sortable)."),
          options = list(pageLength = 12, order = list(list(1, "desc")))) |>
  formatCurrency("Deaths delayed", currency = "", digits = 0)
```

## Regional breakdown (full package)

```{r figure6-region, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
if (data.out[, any(!is.na(region))]) {
  dd_region <- deaths_delayed_by(data.out[!is.na(region)], by = "region",
                                 years = impact_y1:impact_y2)
  dd_region <- dd_region[as.character(scenario) == "full_package"]
  setorder(dd_region, -deaths_delayed)
  p <- ggplot(dd_region, aes(deaths_delayed / 1e6, reorder(region, deaths_delayed))) +
    geom_col(fill = "#6a3d9a", width = 0.7) +
    geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.2) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.14)), labels = comma) +
    labs(title = "Full package: deaths delayed by region",
         subtitle = impact_window, x = "Deaths delayed (millions)", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
  print(p)
} else {
  cat("Region metadata unavailable (Country_groupings_extended.csv not found); ",
      "regional breakdown skipped.\n")
}
```

---

# LSS: primary Na/K–SBP vs SSaSS benchmark

The two LSS methods are **distinct interventions on distinct populations**, not
interchangeable estimates of the same thing:

- **Na/K–SBP (primary)** models sodium displacement **and** potassium addition
  through SBP, affecting **all four CVD causes**; it applies to whole-population
  (s2) and to NCD-RisC diagnosed/treated hypertensives (s4/s5).
- **SSaSS trial-RR (benchmark)** applies the SSaSS **stroke** relative risks
  (incidence + case fatality) and is valid only for SSaSS-comparable
  (diagnosed/treated HTN) populations — never the whole population.

```{r nak-vs-ssass, message=FALSE, warning=FALSE, echo=FALSE}
# Deaths delayed by scenario (impact window) joined to the LSS audit composition,
# for the HTN variants where both methods exist.
dd_lss <- deaths_delayed_by(data.out, by = character(0), years = impact_y1:impact_y2)
dd_lss <- dd_lss[grepl("^lss", as.character(scenario)),
                 .(scenario = as.character(scenario), deaths_delayed)]
la <- as.data.table(lss_audit)
la_sc <- la[, .(elig = eligibility_group[1], cov = mean(eligible_pop_share),
                na_disp = mean(sodium_displaced_g), k_add = mean(potassium_added_g),
                dsbp = mean(sbp_delta_combined_mmHg)), by = .(scenario, method)]
cmp <- merge(la_sc, dd_lss, by = "scenario", all.x = TRUE)
cmp[, `:=`(Scenario = scenario_labels[scenario])]
setorder(cmp, method, scenario)
cmp_tbl <- cmp[, .(
  Scenario, Method = method, `Eligibility` = elig,
  `Elig. pop share` = fmt_pct(cov, 1),
  `Na displaced (g)` = formatC(na_disp, format = "f", digits = 3),
  `K increase (g)`   = formatC(k_add,   format = "f", digits = 3),
  `Mean ΔSBP (mmHg)` = ifelse(is.na(dsbp), "n/a (trial RR)", formatC(dsbp, format = "f", digits = 2)),
  `Deaths delayed (M)` = ifelse(is.na(deaths_delayed), "n/a", fmt_m(deaths_delayed)))]
kable(cmp_tbl, caption = paste0("Table 3b. LSS Na/K–SBP vs SSaSS trial-RR ",
      "benchmark: eligibility, exposure change, SBP change and deaths delayed ",
      "(", impact_window, "). The methods are NOT identical interventions — they ",
      "differ in pathway (all-CVD SBP vs stroke-only RR) and are shown side by side, ",
      "not merged."),
      align = c("l","l","l", rep("r",5))) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 11) |>
  row_spec(0, bold = TRUE) |>
  footnote(general = paste0(
    "ΔSBP is 'n/a (trial RR)' for the SSaSS benchmark because its effect is a ",
    "stroke relative risk, not an SBP change. Na displaced / K increase are the ",
    "same composition for both methods; the health effect differs by pathway."))
```

## Fiscal policy — sensitivity results (exploratory)

```{r fiscal-results, message=FALSE, warning=FALSE, echo=FALSE}
fisc_scn <- scenarios_present[grepl("^fiscal", scenarios_present)]
if (length(fisc_scn)) {
  fr <- dd_cum[as.character(scenario) %in% fisc_scn]
  fr_tbl <- fr[, .(Scenario = scenario_labels[as.character(scenario)],
                   `Deaths delayed (M)` = fmt_m(deaths_delayed, 3),
                   `% of baseline` = fmt_pct(pct_vs_base, 2))]
  kable(fr_tbl, caption = paste0("Table 3c. Fiscal policy deaths delayed by ",
        "sensitivity variant (", impact_window, ") — EXPLORATORY."),
        align = c("l","r","r")) |>
    kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 12) |>
    row_spec(0, bold = TRUE)
} else cat("No fiscal scenarios present in outputs.\n")
```

Fiscal results are **exploratory**: the spread across `fiscal_low`/`base`/`high`
reflects genuine uncertainty in tax rate, pass-through, elasticity, taxable share
and substitution (Table M5) — they are **not** a confirmed estimate.

---

# Benchmarking against the "100 Million Lives Saved" model {#benchmarking}

As an external sanity check, we compare this build's headline result — the
**full package** of sodium policies — against the previous RTSL/UW
**"100 Million Lives Saved"** estimate for a **30% reduction in population
sodium intake**. The previous-model figures are read from
[`data/raw/UWRTSL-100MLives_Output20260204.xlsx`](../data/raw/UWRTSL-100MLives_Output20260204.xlsx)
(sheet `Location And Cause`, intervention `Population Sodium Reduction`). The
comparison is restricted to the **same selected countries** this report models,
so it tracks the country selection automatically.

**This is a contextual comparison, not a like-for-like validation.** The two
estimates are close in construction but *not* equivalent, and we do not adjust
either to force agreement:

- **Exposure change differs.** The previous scenario applies a **uniform 30%
  cut to all population sodium**. The current full package applies each policy's
  documented effect **only to the sodium sources it can reach** (public
  procurement + FOPL + salt targets + fiscal), giving a country-specific
  total-intake reduction (`salteff`) that is generally **well below 30%**. The
  current package is therefore expected to delay **fewer** deaths than a flat
  30% cut — the gap is a feature of the policy-reach logic, not an error.
- **Outcome & causes.** Both report **deaths delayed** relative to a
  no-intervention baseline over the same four CVD causes (IHD, ischemic stroke,
  intracerebral hemorrhage, HHD), so the outcome definition is aligned.
- **Baselines & horizon differ.** The two models use different baseline death
  counts (different populations, age ranges and projection horizons), so we
  report deaths delayed both in **absolute** terms and as a **percent of each
  model's own baseline** — the percent is the more comparable metric. The
  previous model's cumulative horizon is close to, but not identical to, this
  build's `r impact_window` window.
- **Country names** are harmonised onto this model's spellings before matching
  (all ten priority countries match exactly; e.g. `Viet Nam`).

```{r benchmark-compute, message=FALSE, warning=FALSE, echo=FALSE}
# ---- Benchmark: current full package vs 100M-Lives 30% sodium reduction ------
# Previous-model results live in the 100 Million Lives Saved workbook, sheet
# "Location And Cause". The comparison row is the "Population Sodium Reduction"
# intervention, which in that model represents the WHO 30%-relative-reduction
# sodium target. We compare it to THIS build's full package (reusing dd_country
# computed above; the model itself is never re-run here).
PREV_XLSX       <- paste0(wd_raw, "UWRTSL-100MLives_Output20260204.xlsx")
PREV_SHEET      <- "Location And Cause"
PREV_SODIUM_INT <- "Population Sodium Reduction"   # = 30% population sodium reduction

# Harmonise previous-model country names onto this model's spellings. The ten
# priority countries already match; this guards other spellings for robustness.
harmonize_prev_location <- function(x) {
  x <- as.character(x)
  x[x == "Vietnam"]                  <- "Viet Nam"
  x[x == "United States of America"] <- "United States"
  x
}

# Current model: full-package deaths delayed by country over the impact window.
bench_current <- dd_country[, .(location,
                                dd_current   = deaths_delayed,
                                base_current = baseline_dead)]

# Selected countries = whatever the model was run for; the benchmark filters the
# previous-model table to exactly these, so it updates with the selection.
sel_countries <- sort(unique(as.character(bench_current$location)))

benchmark <- tryCatch({
  prev <- as.data.table(readxl::read_excel(PREV_XLSX, sheet = PREV_SHEET))
  prev[, location := harmonize_prev_location(location)]
  prev <- prev[intervention == PREV_SODIUM_INT & location %in% sel_countries,
               .(location,
                 dd_prev   = as.numeric(deaths_delayed),
                 base_prev = as.numeric(baseline_deaths))]
  b <- merge(bench_current, prev, by = "location", all.x = TRUE)
  b[, `:=`(pct_current = dd_current / base_current,
           pct_prev    = dd_prev    / base_prev,
           ratio       = dd_current / dd_prev)]
  setorder(b, -dd_prev)
  b[]
}, error = function(e) {
  message("Benchmark workbook could not be read: ", conditionMessage(e))
  NULL
})

benchmark_available <- !is.null(benchmark) && nrow(benchmark) > 0 &&
  any(!is.na(benchmark$dd_prev))

# Selected-country totals (used inline, in the deck bundle, and for the plot).
if (benchmark_available) {
  bench_cur_tot   <- sum(benchmark$dd_current, na.rm = TRUE)
  bench_prev_tot  <- sum(benchmark$dd_prev,    na.rm = TRUE)
  bench_ratio_tot <- bench_cur_tot / bench_prev_tot
} else {
  bench_cur_tot <- bench_prev_tot <- bench_ratio_tot <- NA_real_
}

# Long form for the grouped bar plot (built once, reused by the deck export).
bench_model_labs <- c(dd_current = "Current full package",
                      dd_prev    = "Previous 30% sodium reduction")
bench_cols <- c(`Current full package`         = "#1f78b4",
                `Previous 30% sodium reduction` = "#b2182b")
if (benchmark_available) {
  bench_long <- melt(benchmark[, .(location, dd_current, dd_prev)],
                     id.vars = "location", variable.name = "model",
                     value.name = "dd")
  bench_long[, model := factor(bench_model_labs[as.character(model)],
                               levels = unname(bench_model_labs))]
  bench_long <- bench_long[!is.na(dd)]
} else {
  bench_long <- NULL
}
```

```{r benchmark-availability-note, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
if (!benchmark_available)
  cat("**Benchmark workbook unavailable — the comparison below is skipped.** ",
      "Place `UWRTSL-100MLives_Output20260204.xlsx` in `data/raw/` and re-knit.\n")
```

Over `r impact_window`, across the `r length(sel_countries)` selected countries,
the current **full package** delays **`r fmt_m(bench_cur_tot)` million** CVD
deaths, versus **`r fmt_m(bench_prev_tot)` million** for the previous flat-30%
sodium-reduction scenario — a ratio of about
**`r formatC(bench_ratio_tot, format = "f", digits = 2)`×**. The current
package delivering a smaller total is expected: its source-limited `salteff` is
generally well below a uniform 30% intake cut.

## Benchmark table

```{r benchmark-table, message=FALSE, warning=FALSE, echo=FALSE}
if (benchmark_available) {
  bt  <- copy(benchmark)
  setorder(bt, -dd_prev)
  tot <- data.table(
    location     = "Total (selected countries)",
    dd_current   = sum(bt$dd_current,   na.rm = TRUE),
    base_current = sum(bt$base_current, na.rm = TRUE),
    dd_prev      = sum(bt$dd_prev,      na.rm = TRUE),
    base_prev    = sum(bt$base_prev,    na.rm = TRUE))
  tot[, `:=`(pct_current = dd_current / base_current,
             pct_prev    = dd_prev    / base_prev,
             ratio       = dd_current / dd_prev)]
  bt_all <- rbind(bt, tot, fill = TRUE)

  bt_fmt <- bt_all[, .(
    Country                              = location,
    `Current: deaths delayed (M)`        = fmt_m(dd_current, 3),
    `Current: % of baseline`             = fmt_pct(pct_current, 2),
    `Previous 30%: deaths delayed (M)`   = fmt_m(dd_prev, 3),
    `Previous 30%: % of baseline`        = fmt_pct(pct_prev, 2),
    `Current ÷ previous`                 = formatC(ratio, format = "f", digits = 2)
  )]

  kable(bt_fmt,
        caption = paste0("Table 4. Benchmark — current full package vs. the ",
                         "100 Million Lives Saved 30% population sodium-reduction ",
                         "estimate, ", impact_window, ", ", nrow(bt),
                         " selected countries."),
        align = c("l", rep("r", 5))) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE, font_size = 13) |>
    add_header_above(c(" " = 1, "Current model (full package)" = 2,
                       "Previous model (30% sodium)" = 2, " " = 1)) |>
    row_spec(0, bold = TRUE) |>
    row_spec(nrow(bt_fmt), bold = TRUE, background = "#eef2f7") |>
    footnote(general = paste0(
      "Previous-model figures: UWRTSL-100MLives_Output20260204.xlsx (sheet ",
      "'Location And Cause', intervention 'Population Sodium Reduction'). ",
      "'% of baseline' is each model's deaths delayed relative to ITS OWN ",
      "baseline CVD deaths, which is the more comparable metric because the ",
      "two models use different baselines and horizons. This is a contextual ",
      "comparison: the previous scenario applies a uniform 30% sodium cut, ",
      "whereas the full package applies source-limited policy effects, so the ",
      "current total is expected to be lower."))
} else {
  cat("Benchmark workbook unavailable; Table 4 skipped.\n")
}
```

## Figure 7 — Deaths delayed by country: full package vs. 30% sodium reduction

```{r benchmark-fig, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6}
if (benchmark_available) {
  p_bench <- ggplot(bench_long,
                    aes(dd / 1e6, reorder(location, dd), fill = model)) +
    geom_col(width = 0.7, position = position_dodge(width = 0.78)) +
    geom_text(aes(label = fmt_m(dd, 2)), position = position_dodge(width = 0.78),
              hjust = -0.1, size = 2.9) +
    scale_fill_manual(values = bench_cols, name = NULL) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.16)), labels = comma) +
    labs(title = paste0("Deaths delayed: current full package vs. previous 30% ",
                        "sodium reduction"),
         subtitle = paste0(length(sel_countries), " selected countries; ",
                          impact_window, "; all CVD causes"),
         x = "Deaths delayed (millions)", y = NULL,
         caption = paste0("Previous = 100 Million Lives Saved, 'Population Sodium ",
                          "Reduction' (30% intake cut). Contextual comparison; ",
                          "scenarios and horizons are not fully equivalent.")) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
          panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
  print(p_bench)
} else {
  cat("Benchmark workbook unavailable; Figure 7 skipped.\n")
}
```

**Interpretation.** The two models rank the selected countries similarly —
China and India dominate the absolute deaths delayed under both, tracking their
share of the regional CVD burden. The **previous flat-30% scenario delays more
deaths** than the current full package in essentially every country, consistent
with its larger (uniform) exposure change; the current package's benefit is
bounded by how much of each country's sodium its four policies can actually
reach. Where the current package closes much of the gap (a higher
current ÷ previous ratio), local intake is concentrated in the **packaged**
source that FOPL, salt targets and fiscal policy jointly target. Because the
scenarios and horizons are **not fully equivalent**, these numbers should be
read as a **contextual cross-check** on order of magnitude and country ranking,
not as a validation target.

---

# Limitations and interpretation

- **Deaths delayed, not averted (sodium→SBP policies).** For public procurement,
  FOPL, salt targets, fiscal, and the **primary Na/K→SBP LSS** method, case
  fatality is held at baseline, so these lower incidence and shift mortality
  later; cumulative totals depend on the accounting window. Only the optional
  **SSaSS trial-RR benchmark** lowers stroke case fatality (averting, not only
  delaying, some stroke deaths).
- **NCD-RisC eligibility is age-standardised 30–79, not age-specific.** The
  diagnosed/treated population shares that size LSS scenarios 4/5 are single
  age-standardised values applied across all modelled ages, and the **post-2019
  proportions are carried forward** unchanged. This misstates eligibility at the
  age extremes and ignores any post-2019 diagnosis/treatment gains.
- **LSS uptake and adherence are uncertain.** Uptake (adoption among the eligible)
  and adherence (SSaSS yr-5 ≈ 0.92) are assumptions; the whole-population coverage
  (s2) is a manual toggle. Household spillover is not modelled.
- **K→SBP evidence is sparser than Na→SBP, and additivity is assumed.** The
  potassium channel uses the Filippini (2020) achieved-excretion dose-response
  with unit-conversion assumptions (39.1 mg/mmol; intake→excretion ÷1.3) and a
  baseline-sodium modulation; the sodium- and potassium-mediated SBP changes are
  **assumed additive** (a factor lets this be relaxed). The K evidence base is
  thinner than for sodium.
- **LSS is benefit-only — no hyperkalaemia/CKD harm channel.** Unlike Huang et
  al., this disease model has no CKD state, so LSS benefits are **not offset** by
  potential renal harms. LSS effects should be read as upper-ish estimates in
  populations where hyperkalaemia risk matters.
- **Baseline potassium** is from Reddin 2023 with a **region-mean fallback** for
  countries absent from the source (Viet Nam → Southeast Asia; Ethiopia → Eastern
  Sub-Saharan Africa), so those two rest on regional, not national, data.
- **Na/K–SBP vs SSaSS applicability.** The two LSS methods are not
  interchangeable: SSaSS RRs apply only to SSaSS-comparable (diagnosed/treated
  HTN) populations and stroke outcomes, whereas Na/K–SBP covers all four causes
  via SBP. They are reported side by side, never combined, and no single pathway
  drives a cause twice.
- **hstroke incidence quirk (pre-existing).** All sodium→SBP scenarios (including
  Na/K–SBP LSS) inherit a known disease-model decomposition quirk that inflates
  hemorrhagic-stroke *incidence* effects; the SSaSS benchmark path was corrected
  for it, but fixing the root cause is out of scope here.
- **Euromonitor trend: category definitions, extrapolation, composition-only.**
  The packaged trend uses Matti's pre-combined v002 category set; the post-2030
  rule (`r RC$euromonitor$post2030`) and the ceiling
  (`r fmt_pct(RC$euromonitor$ceiling,0)`) are **extrapolation assumptions**, and
  the trend reweights *composition* only — total sodium intake is held fixed.
- **Fiscal policy is exploratory.** Pass-through, own-price elasticity, taxable
  share and substitution are uncertain and applying broad-category elasticities
  to high-sodium foods is contested; fiscal results are provisional, not
  confirmed. Its package membership follows `include_fiscal_in_package`.
- **Uncertainty not fully propagated.** Effect sizes, coverage, K→SBP anchors and
  the Euromonitor growth CI are largely used as point estimates; no probabilistic
  sensitivity analysis is run (the fiscal and Euromonitor CIs are shown but not
  propagated into deaths-delayed).
- **Source shares** for the ten prioritised countries are documented but several
  (notably Nigeria/Cameroon, with Nigeria a proxy for Cameroon) are provisional;
  any country outside the set falls back to the `default` (mean) profile. Public
  procurement uses a stale school-inclusive 5% share pending adults-only inputs.
- **Independent-cause structure.** The four causes share one population but are
  modelled independently, so "all-cause" prevalence/incidence sum cause-specific
  cases and may count comorbid individuals more than once.

---

# Reproducibility

```{r session-info, message=FALSE, warning=FALSE, echo=FALSE}
cat("Report generated:", format(Sys.time()), "\n")
cat("Model outputs:    ", length(files), "country files from\n                  ",
    out_model_path, "\n")
cat("Countries:        ", n_countries, "\n")
cat("Scenarios:        ", paste(scenarios_present, collapse = ", "), "\n")
cat("Year range:       ", study_years, "\n\n")
sessionInfo()
```

---

# Deck artifacts (slide-deck export) {#deck-export}

This final section exists purely to hand computed elements to the Beamer slide
deck `docs/UWRTSL-sodium-policy.Rmd`. **All computation for the deck happens
here** (the deck itself performs none): every figure is written as a PNG and
every table/number is written into a single `deck_data.rds` bundle. The deck
loads that bundle and includes the PNGs, so the two documents can never drift.
Artifacts are written to `docs/deck_assets/`.

```{r deck-export, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# ---- Output location --------------------------------------------------------
deck_dir <- file.path(wd, "docs", "deck_assets")
dir.create(deck_dir, recursive = TRUE, showWarnings = FALSE)

save_fig <- function(plot, name, w, h) {
  ggplot2::ggsave(filename = file.path(deck_dir, name), plot = plot,
                  width = w, height = h, units = "in", dpi = 300, bg = "white")
  invisible(name)
}

# Pre-render a data.frame to a booktabs LaTeX tabular, wrapped in \resizebox so
# it can never overflow a 16:9 Beamer frame. escape = TRUE handles %, & etc.
mk_tex <- function(df, align = NULL) {
  if (is.null(align)) align <- c("l", rep("r", ncol(df) - 1L))
  body <- knitr::kable(df, format = "latex", booktabs = TRUE, align = align,
                       linesep = "", escape = TRUE, row.names = FALSE)
  paste0("\\resizebox{\\textwidth}{!}{%\n", as.character(body), "\n}")
}

base_year <- SCALEUP_YEAR1
lab_idx    <- paste0("Indexed (", base_year, " = 100)")

# ---- FIG: sodium source shares by country -----------------------------------
p_ss <- ggplot(shares_long, aes(location, share, fill = source)) +
  geom_col(width = 0.78, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "Share of total sodium intake") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1))
save_fig(p_ss, "fig_source_shares.png", 10, 4.7)

# ---- FIG: illustrative salteff by scenario (individual policies) -------------
p_se <- ggplot(salteff_bar[scenario != "full_package"],
               aes(location, contribution, fill = source)) +
  geom_col(width = 0.8, colour = "white", linewidth = 0.15) +
  facet_wrap(~ scn_lbl, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "salteff (fraction of total sodium reduced)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
save_fig(p_se, "fig_salteff.png", 10.5, 5.6)

# ---- FIG: full-package salteff by country ------------------------------------
p_sepkg <- ggplot(salteff_bar[scenario == "full_package"],
                  aes(location, contribution, fill = source)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = source_cols, labels = source_labels,
                    breaks = SODIUM_SOURCES, name = "Sodium source") +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Full-package salteff") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1))
save_fig(p_sepkg, "fig_salteff_pkg.png", 9.5, 4.7)

# ---- FIG: causal-chain flow diagram -----------------------------------------
p_flow <- ggplot() +
  geom_tile(data = boxes, aes(x = x, y = 1), width = 0.84, height = 0.7,
            fill = "#e8eef4", color = "#1f78b4", linewidth = 0.6) +
  geom_text(data = boxes, aes(x = x, y = 1, label = label), size = 3.4,
            lineheight = 0.95) +
  geom_segment(data = arrows, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.16, "cm"), type = "closed"),
               linewidth = 0.5, color = "grey30") +
  coord_cartesian(xlim = c(0.5, 6.5), ylim = c(0.55, 1.45)) +
  theme_void(base_size = 12)
save_fig(p_flow, "fig_flow.png", 11, 2.8)

# ---- FIG: LSS Na/K causal pathway (deck Slide A) ----------------------------
mk_flow <- function(labels, fill = "#e8f2ea", border = "#59A14F") {
  n <- length(labels)
  bx <- data.table(x = seq_len(n), label = labels)
  ar <- data.table(x = (seq_len(n - 1)) + 0.44, xend = (2:n) - 0.44, y = 1, yend = 1)
  ggplot() +
    geom_tile(data = bx, aes(x = x, y = 1), width = 0.86, height = 0.7,
              fill = fill, color = border, linewidth = 0.7) +
    geom_text(data = bx, aes(x = x, y = 1, label = label), size = 3.2, lineheight = 0.95) +
    geom_segment(data = ar, aes(x = x, xend = xend, y = y, yend = yend),
                 arrow = arrow(length = unit(0.16, "cm"), type = "closed"),
                 linewidth = 0.5, color = "grey30") +
    coord_cartesian(xlim = c(0.5, n + 0.5), ylim = c(0.55, 1.45)) +
    theme_void(base_size = 12)
}
p_lss_flow <- mk_flow(c(
  "LSS adoption\n75% NaCl / 25% KCl",
  "↓ discretionary Na\n+ ↑ K (stoichiometric)",
  "↓ SBP\nΔSBP(Na) + ΔSBP(K)",
  "↓ CVD incidence\nall 4 causes (via SBP)",
  "Deaths\ndelayed"))
save_fig(p_lss_flow, "fig_lss_pathway.png", 11, 2.2)

# ---- FIG: fiscal tax->price->demand->sodium pathway (deck fiscal slide) ------
p_fiscal_flow <- mk_flow(c(
  "Tax rate",
  "Price ↑\n(× pass-through)",
  "Quantity ↓\n(× elasticity)",
  "↓ taxed packaged\n(× taxable share)",
  "↓ packaged\nsodium",
  "↓ total Na →\nSBP → CVD"),
  fill = "#efe8f4", border = "#6a3d9a")
save_fig(p_fiscal_flow, "fig_fiscal_pathway.png", 11.5, 2.2)

# ---- FIG: baseline age-standardised mortality rate --------------------------
p_asmr <- ggplot() +
  geom_line(data = asmr_cause, aes(year, ASMR, color = cause_lbl,
                                   linetype = year > hist_cut), linewidth = 1) +
  geom_line(data = asmr_all, aes(year, ASMR, linetype = year > hist_cut),
            color = "black", linewidth = 1.2) +
  annotate("text", x = year_min + 1, y = max(asmr_all$ASMR),
           label = "All CVD", hjust = 0, vjust = 1, fontface = "bold", size = 3.6) +
  scale_color_manual(values = setNames(cause_cols, cause_map), name = "Cause") +
  scale_linetype_manual(values = c(`FALSE` = "solid", `TRUE` = "dashed"),
                        guide = "none") +
  labs(x = "Year", y = "ASMR per 100,000") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
save_fig(p_asmr, "fig_baseline_asmr.png", 9.5, 5)

# ---- FIG: projected incidence (level + indexed) -----------------------------
inc_lvl <- proj_all[year >= 2020, .(scenario, year, value = inc_rate)][
  , panel := "Incident cases per 1,000"]
inc_ix  <- inc_idx[, .(scenario, year, value = inc_index)][, panel := lab_idx]
inc_comb <- rbind(inc_lvl, inc_ix)
inc_comb[, panel := factor(panel, levels = c("Incident cases per 1,000", lab_idx))]
hl_inc <- data.table(panel = factor(lab_idx, levels = levels(inc_comb$panel)),
                     yy = 100)
p_inc <- ggplot(inc_comb, aes(year, value, colour = scenario)) +
  geom_hline(data = hl_inc, aes(yintercept = yy), linetype = "dashed",
             colour = "grey60", inherit.aes = FALSE) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_colour_manual(values = scenario_cols, labels = scenario_labels_short,
                      breaks = scenario_levels, name = NULL) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = "Year", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
save_fig(p_inc, "fig_incidence.png", 10.5, 4.7)

# ---- FIG: projected deaths (per 1,000 + indexed) ----------------------------
proj_all[, dead_p1000 := dead / pop * 1000]
d_lvl  <- proj_all[year >= 2020, .(scenario, year, value = dead_p1000)][
  , panel := "Deaths per 1,000"]
base_d <- proj_all[year == base_year, .(scenario, bv = dead_p1000)]
d_ix   <- merge(proj_all[year >= base_year, .(scenario, year, dead_p1000)],
                base_d, by = "scenario")
d_ix   <- d_ix[!is.na(bv) & bv > 0]
d_ix[, value := dead_p1000 / bv * 100]
d_ix2  <- d_ix[, .(scenario, year, value)][, panel := lab_idx]
d_comb <- rbind(d_lvl, d_ix2)
d_comb[, panel := factor(panel, levels = c("Deaths per 1,000", lab_idx))]
hl_d <- data.table(panel = factor(lab_idx, levels = levels(d_comb$panel)), yy = 100)
p_d <- ggplot(d_comb, aes(year, value, colour = scenario)) +
  geom_hline(data = hl_d, aes(yintercept = yy), linetype = "dashed",
             colour = "grey60", inherit.aes = FALSE) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_colour_manual(values = scenario_cols, labels = scenario_labels_short,
                      breaks = scenario_levels, name = NULL) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE)) +
  labs(x = "Year", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
save_fig(p_d, "fig_deaths.png", 10.5, 4.7)

# ---- FIG: deaths delayed by scenario (headline result) ----------------------
fig_res <- dd_cum[as.character(scenario) != "baseline"]
fig_res[, scn_lbl := factor(scenario_labels_short[as.character(scenario)],
                            levels = rev(scenario_labels_short[scenario_nb]))]
p_res <- ggplot(fig_res, aes(deaths_delayed / 1e6, scn_lbl, fill = scenario)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.6) +
  scale_fill_manual(values = scenario_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14)), labels = comma) +
  labs(x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
save_fig(p_res, "fig_results_bar.png", 9, 4.8)

# ---- FIG: full package, deaths delayed by cause -----------------------------
p_bc <- ggplot(dd_cause, aes(deaths_delayed / 1e6,
                             reorder(cause_lbl, deaths_delayed), fill = cause)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.6) +
  scale_fill_manual(values = cause_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.14)), labels = comma) +
  labs(x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
save_fig(p_bc, "fig_bycause.png", 8.5, 3.9)

# ---- FIG: full package, deaths delayed by country ---------------------------
dd_ctry <- copy(dd_country)
setorder(dd_ctry, -deaths_delayed)
p_ct <- ggplot(dd_ctry, aes(deaths_delayed / 1e6, reorder(location, deaths_delayed))) +
  geom_col(fill = "#1f78b4", width = 0.72) +
  geom_text(aes(label = fmt_m(deaths_delayed, 2)), hjust = -0.1, size = 3.4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15)), labels = comma) +
  labs(x = "Deaths delayed (millions)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())
save_fig(p_ct, "fig_countries.png", 8.5, 5)

# ---- FIG: benchmark vs 100M-Lives 30% sodium reduction ----------------------
# Always writes fig_benchmark.png so the deck can include it unconditionally;
# falls back to a text placeholder if the benchmark workbook was unavailable.
if (benchmark_available) {
  p_bench_deck <- ggplot(bench_long,
                         aes(dd / 1e6, reorder(location, dd), fill = model)) +
    geom_col(width = 0.7, position = position_dodge(width = 0.78)) +
    geom_text(aes(label = fmt_m(dd, 2)), position = position_dodge(width = 0.78),
              hjust = -0.1, size = 3) +
    scale_fill_manual(values = bench_cols, name = NULL) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.16)), labels = comma) +
    labs(x = "Deaths delayed (millions)", y = NULL) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank())
} else {
  p_bench_deck <- ggplot() +
    annotate("text", x = 0, y = 0,
             label = "Benchmark workbook unavailable", size = 5, colour = "grey40") +
    theme_void()
}
save_fig(p_bench_deck, "fig_benchmark.png", 9.5, 5)

# ---- TABLES (pre-rendered LaTeX) --------------------------------------------
# Effect sizes (Table M2).
tef <- as.data.frame(eff_disp)
names(tef) <- c("Intervention", "Source", "Applied reduction", "Note")
tef$Intervention <- as.character(tef$Intervention)
tex_effect_sizes <- mk_tex(tef, align = c("l", "l", "r", "l"))

# Illustrative salteff by scenario x country (Table M4).
tsl <- as.data.frame(salteff_grid)
tex_salteff <- mk_tex(tsl, align = c("l", rep("r", ncol(tsl) - 1L)))

# Source shares (Table M1).
tss <- as.data.frame(ss_disp)
tex_source_shares <- mk_tex(tss, align = c("l", rep("r", ncol(tss) - 1L)))

# Packaged-food Euromonitor trend (Table M2b).
if (!is.null(pkg_trend_tbl)) {
  tex_packaged_trend <- mk_tex(as.data.frame(pkg_trend_tbl),
                               align = c("l", "r", "r", "r"))
} else {
  tex_packaged_trend <- "\\emph{Packaged-food trend table unavailable.}"
}

# Headline results (compact Table 1).
tr <- dd_cum[as.character(scenario) != "baseline"]
tr[, scenario := factor(as.character(scenario), levels = scenario_levels)]
setorder(tr, scenario)
tres <- data.frame(
  Scenario                    = unname(scenario_labels[as.character(tr$scenario)]),
  `Deaths delayed (millions)` = fmt_m(tr$deaths_delayed, 3),
  `Avg per year (thousands)`  = fmt_k(tr$per_year, 1),
  `% of baseline`             = fmt_pct(tr$pct_vs_base, 2),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
tex_results <- mk_tex(tres, align = c("l", "r", "r", "r"))

# Benchmark vs 100M-Lives 30% sodium reduction (compact; totals row included).
if (benchmark_available) {
  bench_tex_dt <- rbind(
    benchmark[, .(location, dd_current, pct_current, dd_prev, pct_prev, ratio)],
    benchmark[, .(location = "Total (selected)",
                  dd_current = sum(dd_current, na.rm = TRUE), pct_current = NA_real_,
                  dd_prev    = sum(dd_prev,    na.rm = TRUE), pct_prev    = NA_real_,
                  ratio = sum(dd_current, na.rm = TRUE) / sum(dd_prev, na.rm = TRUE))]
  )
  tbn <- data.frame(
    Country            = bench_tex_dt$location,
    `Current (M)`      = fmt_m(bench_tex_dt$dd_current, 2),
    `Prev 30% (M)`     = fmt_m(bench_tex_dt$dd_prev, 2),
    `Cur / prev`       = formatC(bench_tex_dt$ratio, format = "f", digits = 2),
    check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
  )
  tex_benchmark <- mk_tex(tbn, align = c("l", "r", "r", "r"))
} else {
  tex_benchmark <- "\\emph{Benchmark table unavailable.}"
}

# ---- NUMBERS ----------------------------------------------------------------
pkg_row <- dd_cum[as.character(scenario) == "full_package"]
vn_sanity <- as.numeric(compute_total_efficacy("public_procurement", "Viet Nam"))

# Effect sizes come from the PUBLISHED table (RC$intervention_effects); the fiscal
# base + Euromonitor + LSS constants come from run_config -> the deck states no
# raw numbers the pipeline did not publish (Task 4i).
eff_of  <- function(iv, src) { v <- intervention_effects[intervention == iv & source == src, effect]
  if (length(v) == 0) NA_real_ else v[1] }
.fiscal_scn2 <- if ("fiscal_base" %in% scenarios_present) "fiscal_base" else "fiscal"
.he <- as.data.table(htn_eligibility)[location %in% RC$priority_countries]
fa <- as.data.table(fiscal_audit)
fa_base <- fa[variant %in% c("fiscal_base", "fiscal")][1]
fa_low  <- fa[variant == "fiscal_low"][1]
fa_high <- fa[variant == "fiscal_high"][1]

# Representative LSS Na/K→SBP quantities (mean over whole-population s2 rows) for
# the deck's LSS methods slide — read from the audit, not recomputed.
la_dk <- as.data.table(lss_audit)[grepl("^lss_s2", scenario) & method == "na_k_sbp"]

nums <- list(
  n_countries    = n_countries,
  study_years    = study_years,
  impact_window  = impact_window,
  scaleup1       = SCALEUP_YEAR1,
  scaleup2       = SCALEUP_YEAR2,
  horizon_end    = impact_y2,
  # headline deaths delayed (millions, formatted strings)
  pkg_m          = fmt_m(ddc("full_package"), 2),
  pkg_per_year_k = fmt_k(pkg_row$per_year, 0),
  pkg_pct        = fmt_pct(pkg_row$pct_vs_base, 1),
  salt_targets_m = fmt_m(ddc("salt_targets"), 2),
  fopl_m         = fmt_m(ddc("fopl"), 2),
  fiscal_m       = fmt_m(ddc(.fiscal_scn2), 2),
  public_m       = fmt_m(ddc("public_procurement"), 2),
  # LSS primary Na/K-SBP scenarios 2/4/5 + SSaSS benchmark (if present)
  lss_method     = RC$lss$method,
  lss_s2_m       = fmt_m(ddc("lss_s2"), 2),
  lss_s4_m       = fmt_m(ddc("lss_s4"), 2),
  lss_s5_m       = fmt_m(ddc("lss_s5"), 2),
  lss_s4_ssass_m = fmt_m(ddc("lss_s4_ssass"), 2),
  lss_s5_ssass_m = fmt_m(ddc("lss_s5_ssass"), 2),
  # LSS Na/K constants + NCD-RisC eligibility (mean pop share across priority)
  lss_nacl         = fmt_pct(RC$lss$nacl_fraction, 0),
  lss_kcl          = fmt_pct(RC$lss$kcl_fraction, 0),
  lss_adherence    = formatC(RC$lss$adherence, format = "f", digits = 2),
  lss_additivity   = formatC(RC$lss$additivity_factor, format = "f", digits = 2),
  lss_harms        = RC$lss$harms_modelled,
  lss_cov_all      = fmt_pct(RC$lss$coverage_all, 0),
  htn_dx_pop       = fmt_pct(mean(.he$diagnosed_pop, na.rm = TRUE), 0),
  htn_tx_pop       = fmt_pct(mean(.he$treated_pop,   na.rm = TRUE), 0),
  htn_source_year  = RC$htn_source_year,
  # SSaSS benchmark constants (surfaced for the benchmark section only)
  lss_rr_stroke_inc = formatC(LSS_RR_STROKE_INC, format = "f", digits = 2),
  lss_cf_mult       = formatC(LSS_CF_MULT,       format = "f", digits = 3),
  lss_stroke_total  = formatC(LSS_STROKE_TOTAL,  format = "f", digits = 2),
  # applied per-source effect sizes (from the published intervention_effects)
  eff_public       = fmt_pct(eff_of("public_procurement", "public"), 0),
  eff_fopl         = fmt_pct(eff_of("fopl", "packaged"), 1),
  eff_salttargets  = fmt_pct(eff_of("salt_targets", "packaged"), 0),
  eff_fiscal       = fmt_pct(RC$fiscal$base_effect, 0),
  # fiscal pathway (EXPLORATORY) from fiscal_audit base row
  fiscal_status        = "exploratory",
  fiscal_tax_rate      = fmt_pct(RC$fiscal$base$tax_rate, 0),
  fiscal_taxable_share = fmt_pct(RC$fiscal$base$taxable_share, 0),
  fiscal_in_package    = as.character(RC$fiscal$include_in_package),
  # fiscal pathway detail (EXPLORATORY) — base row + low/high, from fiscal_audit
  fiscal_pass_through  = fmt_pct(fa_base$pass_through, 0),
  fiscal_elasticity    = formatC(fa_base$elasticity, format = "f", digits = 2),
  fiscal_price_change  = fmt_pct(fa_base$price_change, 0),
  fiscal_qty_response  = fmt_pct(abs(fa_base$quantity_response), 0),
  fiscal_taxed_pkg_red = fmt_pct(fa_base$taxed_packaged_reduction, 0),
  fiscal_pkg_red       = fmt_pct(fa_base$packaged_reduction, 1),
  fiscal_substitution  = fmt_pct(fa_base$substitution, 0),
  fiscal_reformulation = fmt_pct(fa_base$reformulation, 0),
  fiscal_low_pkg       = fmt_pct(fa_low$packaged_reduction, 1),
  fiscal_high_pkg      = fmt_pct(fa_high$packaged_reduction, 1),
  fiscal_low_m         = fmt_m(ddc("fiscal_low"), 2),
  fiscal_high_m        = fmt_m(ddc("fiscal_high"), 2),
  # LSS Na/K→SBP detail (representative means over whole-population s2 rows)
  lss_uptake           = formatC(RC$lss$uptake, format = "f", digits = 2),
  lss_nonhtn_factor    = formatC(RC$lss$nonhtn_factor, format = "f", digits = 2),
  lss_eff_na_k_disc    = fmt_pct(RC$lss$eff_lss_na_k_discretionary, 0),
  lss_k_mg_per_mmol    = as.character(RC$lss$k_mg_per_mmol),
  lss_k_intake_excr    = as.character(RC$lss$k_intake_to_excretion),
  lss_dsbp_na          = formatC(mean(la_dk$sbp_delta_na_mmHg,       na.rm = TRUE), format = "f", digits = 2),
  lss_dsbp_k           = formatC(mean(la_dk$sbp_delta_k_mmHg,        na.rm = TRUE), format = "f", digits = 2),
  lss_dsbp_tot         = formatC(mean(la_dk$sbp_delta_combined_mmHg, na.rm = TRUE), format = "f", digits = 2),
  lss_na_disp          = formatC(mean(la_dk$sodium_displaced_g,      na.rm = TRUE), format = "f", digits = 2),
  lss_k_add            = formatC(mean(la_dk$potassium_added_g,       na.rm = TRUE), format = "f", digits = 2),
  # published media effect (retained in registry; excluded from reported scenarios)
  eff_media            = fmt_pct(eff_of("media_campaign", "packaged"), 0),
  # external comparators (documented in the report layer, not our pipeline output)
  huang_s2_deaths      = "1.85 million/year",
  saxena_na_red        = "≈100 mg/day (≈10%)",
  saxena_tax           = "20%",
  saxena_horizon       = "20 years",
  eff_lss          = fmt_pct(eff_of("lss", "discretionary"), 0),
  vn_sanity        = formatC(vn_sanity, format = "f", digits = 4),
  # PROVISIONAL: public-procurement share is still the stale school-inclusive 5%.
  public_share     = fmt_pct(source_shares[location == "Viet Nam", public], 1),
  # Euromonitor packaged-food trend (v002 combined) summary stats
  euro_window      = RC$euromonitor$window,
  euro_ceiling     = fmt_pct(RC$euromonitor$ceiling, 0),
  euro_post2030    = RC$euromonitor$post2030,
  pkg_apc_median   = pkg_apc_median,
  pkg_apc_range    = pkg_apc_range,
  # worked composition example: FOPL + salt targets on 'packaged'
  compose_fopl_salt = fmt_pct(1 - (1 - eff_of("fopl","packaged")) * (1 - eff_of("salt_targets","packaged")), 1),
  # SSaSS / LSS composition constants
  ssass_na_mgday   = "-350 mg/day",
  ssass_base_gday  = "4.3 g/day",
  ssass_sbp_mmhg   = "-3.34 mmHg",
  ssass_k_mgday    = "+803 mg/day",
  lss_ratio        = paste0(fmt_pct(RC$lss$nacl_fraction,0), " NaCl / ", fmt_pct(RC$lss$kcl_fraction,0), " KCl"),
  # benchmark vs the 100 Million Lives Saved 30% population sodium reduction
  bench_n          = length(sel_countries),
  bench_current_m  = if (benchmark_available) fmt_m(bench_cur_tot, 2)  else "n/a",
  bench_prev_m     = if (benchmark_available) fmt_m(bench_prev_tot, 2) else "n/a",
  bench_ratio      = if (benchmark_available)
                       formatC(bench_ratio_tot, format = "f", digits = 2) else "n/a"
)

# LaTeX-safe percent signs. The deck inlines these nums both in markdown text
# (where pandoc would escape a bare '%') AND inside raw-LaTeX environments
# (tabular / itemize / block) where it would NOT -- an unescaped '%' there
# becomes a comment and silently eats the rest of the line. Pre-escaping every
# '%' to '\%' is correct in both contexts, so it is done once here.
nums <- lapply(nums, function(v)
  if (is.character(v)) gsub("%", "\\\\%", v, perl = TRUE) else v)

# ---- NEW deck artifacts (Task 4i): year-specific shares, LSS Na/K pathway, ---
# NCD-RisC eligibility, Na/K-vs-SSaSS, fiscal audit. Guarded so a skipped chunk
# never breaks the bundle.
if (exists("fig_pkg_time")) save_fig(fig_pkg_time, "fig_packaged_time.png", 10, 5)
tex_ncd_eligibility <- if (exists("he_tbl")) mk_tex(as.data.frame(he_tbl),
  align = c("l","l", rep("r",6))) else "\\emph{NCD-RisC eligibility unavailable.}"
tex_lss_audit <- if (exists("la_tbl")) mk_tex(as.data.frame(la_tbl),
  align = c("l","l", rep("r",8))) else "\\emph{LSS audit unavailable.}"
tex_fiscal_audit <- if (exists("fa_tbl")) mk_tex(as.data.frame(fa_tbl),
  align = c("l", rep("r",10))) else "\\emph{Fiscal audit unavailable.}"
tex_nak_ssass <- if (exists("cmp_tbl")) mk_tex(as.data.frame(cmp_tbl),
  align = c("l","l","l", rep("r",5))) else "\\emph{Na/K-vs-SSaSS comparison unavailable.}"
tex_fiscal_sensitivity <- if (exists("fr_tbl")) mk_tex(as.data.frame(fr_tbl),
  align = c("l","r","r")) else "\\emph{Fiscal sensitivity results unavailable.}"
tex_huang_ours  <- if (exists("huang_cmp"))  mk_tex(huang_cmp,  align = c("l","l","l")) else "\\emph{Huang comparison unavailable.}"
tex_saxena_ours <- if (exists("saxena_cmp")) mk_tex(saxena_cmp, align = c("l","l","l")) else "\\emph{Saxena comparison unavailable.}"

figs <- list(
  source_shares = "fig_source_shares.png",
  salteff       = "fig_salteff.png",
  salteff_pkg   = "fig_salteff_pkg.png",
  flow          = "fig_flow.png",
  baseline_asmr = "fig_baseline_asmr.png",
  incidence     = "fig_incidence.png",
  deaths        = "fig_deaths.png",
  results_bar   = "fig_results_bar.png",
  bycause       = "fig_bycause.png",
  countries     = "fig_countries.png",
  benchmark      = "fig_benchmark.png",
  packaged_time  = "fig_packaged_time.png",  # NEW: packaged share over time by country
  lss_pathway    = "fig_lss_pathway.png",    # NEW: LSS Na/K causal pathway
  fiscal_pathway = "fig_fiscal_pathway.png"  # NEW: fiscal tax->price->demand->sodium pathway
)

tex <- list(
  effect_sizes    = tex_effect_sizes,
  salteff         = tex_salteff,
  source_shares   = tex_source_shares,
  packaged_trend  = tex_packaged_trend,
  results         = tex_results,
  benchmark       = tex_benchmark,
  ncd_eligibility = tex_ncd_eligibility,   # NEW (Task 4i)
  lss_audit       = tex_lss_audit,         # NEW
  fiscal_audit    = tex_fiscal_audit,      # NEW
  nak_vs_ssass    = tex_nak_ssass,         # NEW
  fiscal_sensitivity = tex_fiscal_sensitivity,  # NEW: deaths delayed low/base/high
  huang_ours      = tex_huang_ours,        # NEW: LSS vs Huang et al. 2026
  saxena_ours     = tex_saxena_ours        # NEW: fiscal vs Saxena et al. 2025
)

deck <- list(nums = nums, figs = figs, tex = tex)
saveRDS(deck, file = file.path(deck_dir, "deck_data.rds"))

message("Deck artifacts written to: ", deck_dir)
message("  figures: ", paste(unlist(figs), collapse = ", "))
message("  bundle : deck_data.rds")
```
