---
title: "Sodium Interventions Model"
subtitle: "Projected cardiovascular health impact of sodium-reduction policies in ten priority countries"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  bookdown::word_document2:
    toc: true
    toc_depth: 2
    number_sections: false
    fig_caption: true
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: false
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
bibliography: references.bib
csl: apa.csl
link-citations: true
---

```{r tr-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 300, out.width = "100%")
options(knitr.kable.NA = "")
suppressMessages({
  library(flextable); library(officer); library(knitr); library(kableExtra)
})

# Locate and load the authoritative artifact bundle (produced by the analysis
# layer). No calculation happens in this document.
.here <- tryCatch(dirname(knitr::current_input(dir = TRUE)), error = function(e) getwd())
assets  <- file.path(.here, "report_assets")
TR <- readRDS(file.path(assets, "tr_bundle.rds"))
N  <- TR$nums
fig_path <- function(key) file.path(assets, TR$fig[[key]]$file)

# ---- one table renderer, two output formats --------------------------------
# `key` looks up the bundle; `id` is the chunk-label / bookmark / cross-ref target
# (must equal the chunk label so \@ref(tab:id) resolves in both Word and PDF).
.is_num_col <- function(x) {
  x <- trimws(as.character(x)); x <- x[nzchar(x)]
  if (!length(x)) return(FALSE)
  mean(grepl("^[-–−(]?[0-9][0-9,.]*%?\\)?$", x)) > 0.6
}
tr_tab <- function(key, id = gsub("_", "-", key)) {
  tb  <- TR$tab[[key]]; df <- tb$display; cap <- paste0(tb$title, ".")
  wide <- ncol(df) >= 6 || identical(tb$orientation, "landscape")
  numcol <- vapply(df, .is_num_col, logical(1)); numcol[1] <- FALSE
  al <- ifelse(numcol, "r", "l")
  if (knitr::is_latex_output()) {
    k <- kbl(df, booktabs = TRUE, caption = cap, linesep = "", align = al) |>
      kable_styling(latex_options = c("striped", "hold_position",
        if (wide) "scale_down" else NULL), font_size = tb$fontsize,
        full_width = FALSE)
    # Wrap a long text first column so narrow tables do not overflow the page.
    if (!wide && max(nchar(df[[1]]), na.rm = TRUE) > 30)
      k <- column_spec(k, 1, width = "16em")
    # Wrap all columns of long multi-column text comparisons (e.g. Huang/Saxena).
    if (ncol(df) == 3 && mean(nchar(unlist(df[-1]))) > 30)
      k <- column_spec(k, 2:3, width = "20em")
    k
  } else {
    ft <- flextable(df)
    ft <- set_caption(ft, caption = cap,
            autonum = run_autonum(seq_id = "tab", pre_label = "Table ", bkm = id))
    ft <- theme_booktabs(ft)
    ft <- bg(ft, part = "header", bg = "#1F4E5F")
    ft <- color(ft, part = "header", color = "white")
    ft <- bold(ft, part = "header")
    ft <- fontsize(ft, size = tb$fontsize, part = "all")
    ft <- padding(ft, padding = 2, part = "all")
    if (any(numcol)) ft <- align(ft, j = which(numcol), align = "right", part = "all")
    ft <- align(ft, j = which(!numcol), align = "left", part = "body")
    ft <- align(ft, part = "header", align = "center")
    ft <- valign(ft, valign = "top", part = "body")
    if (grepl("^Total", tail(df[[1]], 1))) ft <- bold(ft, i = nrow(df), part = "body")
    set_table_properties(ft, layout = "autofit", width = 1)
  }
}
tab_note <- function(key) {
  tb <- TR$tab[[key]]
  bits <- c(if (nzchar(tb$units)) paste0("Units: ", tb$units),
            if (nzchar(tb$period)) paste0("Period: ", tb$period),
            if (nzchar(tb$rounding)) paste0("Rounding: ", tb$rounding),
            if (nzchar(tb$source)) paste0("Source: ", tb$source),
            if (nzchar(tb$notes)) tb$notes)
  paste0("*", paste(bits, collapse = " "), "*")
}
```

`r if (knitr::is_latex_output()) "\\vspace{2em}"`

**Sodium Interventions Model — Technical Report**

*Projected cardiovascular health impact of national sodium-reduction policies in
`r N$n_countries` priority countries, `r N$window`.*

Prepared by the University of Washington in partnership with Resolve to Save Lives.
`r format(Sys.Date(), "%B %Y")`.

This report accompanies a companion results workbook. Health impacts are expressed
as **cardiovascular deaths delayed** relative to a no-intervention baseline. Figures
labelled *exploratory* or *provisional* rest on preliminary inputs and should not be
read as settled estimates.

\newpage

# 1. Executive summary

## 1.1 Purpose and scope

This report estimates how many cardiovascular deaths could be **delayed** by a menu
of population sodium-reduction policies in `r N$n_countries` priority countries —
Bangladesh, Cameroon, China, Ethiopia, India, Malaysia, Nigeria, the Philippines,
Thailand and Viet Nam — over the period **`r N$window`**. It documents the modelling
approach, presents the projected health impacts, and states the assumptions and
limitations plainly so that donors, policy leaders, public-health specialists and
technical reviewers can judge the estimates on their merits.

The policies evaluated are public food procurement, mandatory sodium (salt) targets
for packaged food, front-of-pack warning labelling, a salt tax (treated as
exploratory), and low-sodium salt substitutes. Each is modelled through the same
biological chain: less sodium — and, for salt substitutes, more potassium — lowers
blood pressure, which lowers the incidence of four cardiovascular diseases.

## 1.2 Headline findings

- A coordinated **package** of packaged-food and procurement policies is projected to
  delay about **`r N$pkg_m` million** cardiovascular deaths over `r N$window`
  (`r N$pkg_pct` of baseline cardiovascular deaths in the window). The range across
  salt-tax assumptions is **`r N$pkg_low_m`–`r N$pkg_high_m` million** (Table
  \@ref(tab:main)).
- Among packaged-food policies, **mandatory salt targets** deliver the largest single
  benefit (**`r N$salt_targets_m` million** deaths delayed), followed by front-of-pack
  labelling (**`r N$fopl_m` million**).
- **Low-sodium salt substitutes** show the largest *potential* of any single measure
  where discretionary salt dominates intake: whole-population use delays from
  **`r N$lss_s2_lo_m` million** at `r N$lss_s2_lo_pct` population uptake to
  **`r N$lss_s2_hi_m` million** at `r N$lss_s2_hi_pct` uptake (Table \@ref(tab:lss),
  Figure \@ref(fig:lss-uptake)). These are benefit-only estimates (see limitations).
- Impact is highly concentrated: **China accounts for about `r N$china_share`** of the
  package benefit, reflecting its population size and sodium burden.

## 1.3 Donor-facing takeaways

- Sodium policy is a **high-return, low-cost** cardiovascular investment: a handful of
  packaged-food policies, acting only on the sodium they can realistically reach,
  still delay millions of deaths across ten countries.
- The **best policy mix is country-specific.** Where packaged food dominates sodium
  (for example China and Malaysia), reformulation and labelling matter most; where
  discretionary salt dominates (for example Ethiopia, India, Bangladesh), low-sodium
  salt substitutes carry the largest potential.
- **Combining policies compounds the benefit** without double-counting, because each
  policy is applied only to the sodium sources it can reach.

## 1.4 How to interpret these estimates

These are **deaths delayed**, not permanently averted: the sodium-to-blood-pressure
policies lower disease *incidence* while leaving case fatality unchanged, so they shift
deaths later in life rather than removing them from the life table. Totals therefore
depend on the accounting window (`r N$window`). Estimates are **point projections**
without full uncertainty ranges; the salt-tax results are **exploratory**, and public
food-procurement rests on a **provisional** coverage assumption. They are best read as
the *relative* promise of each policy and the *order of magnitude* of the opportunity,
not as precise forecasts.

# 2. Modeling approach

## 2.1 Overall conceptual framework

The model links each policy to health outcomes through one transparent causal chain
(Figure \@ref(fig:framework)). A policy reduces sodium from the food sources it can
reach; the resulting fall in average sodium intake lowers systolic blood pressure;
lower blood pressure reduces the yearly risk of developing four cardiovascular
diseases; and a population model translates that lower risk into deaths delayed over
time. Low-sodium salt substitutes add a second, blood-pressure-lowering channel
through the potassium they contain.

```{r framework, fig.cap=TR$fig$framework$title, out.width="100%"}
include_graphics(fig_path("framework"))
```

## 2.2 Countries and populations included

The model covers the ten RTSL priority countries and the adult population aged 20 and
older in each (Table \@ref(tab:countries-tab)). Nigeria's sodium-source profile is used as
a stand-in for Cameroon pending country-specific data. Although the client's draft
results tables omitted Viet Nam, it is a modelled priority country and is included
throughout.

```{r countries-tab}
tr_tab("countries", "countries-tab")
```

`r tab_note("countries")`

## 2.3 Baseline year, scale-up, projection horizon and results window

The population model is seeded in **`r TR$meta$baseline_year`** and projected annually
to **`r TR$meta$horizon`**. Policies scale up linearly over **`r N$scaleup`** and then
hold at full implementation. Results are reported cumulatively over the window
**`r N$window`**, with reference years `r N$results_years`.

## 2.4 Sodium-source framework

Baseline sodium intake in each country is divided into **five sources** —
discretionary salt (added at the table or in home cooking), packaged food, food from
restaurants and vendors, food from public and institutional settings, and sodium
inherent to unprocessed foods. These shares sum to 100% (Figure \@ref(fig:shares),
Table \@ref(tab:source-shares)). Each policy is applied **only to the sources it can
plausibly reach** — for example, salt targets and labelling to packaged food, salt
substitutes to discretionary salt — and inherent sodium is never targeted. This
"source-split" design keeps every policy's claimed effect anchored to the sodium it
can actually influence (Figure \@ref(fig:map)).

```{r shares, fig.cap=TR$fig$shares$title, out.width="90%"}
include_graphics(fig_path("shares"))
```

```{r map, fig.cap=TR$fig$map$title, out.width="80%"}
include_graphics(fig_path("map"))
```

## 2.5 Population and demographic assumptions

Country populations, births and deaths follow the United Nations 2024 projections.
Baseline disease and mortality rates come from the Global Burden of Disease study, and
the distribution of blood pressure by age and sex from pooled international data. The
model tracks single years of age from 20 to 95.

## 2.6 Cardiovascular outcomes included

Four blood-pressure-related cardiovascular diseases are modelled: ischaemic heart
disease, ischaemic stroke, intracerebral (haemorrhagic) stroke, and hypertensive heart
disease. All benefits reported here are for these four causes combined unless a
by-cause breakdown is shown.

## 2.7 Intervention effect assumptions

Each policy has a documented **relative reduction** of its target sodium source at full
implementation (Table \@ref(tab:eff-brief)); the evidence behind each value, including
the blood-pressure step, is detailed in Appendix C (Table \@ref(tab:eff-evidence)).
The sodium-to-blood-pressure relationship follows an established dose-response
[@ettehad2016; @Filippini2020]: a reduction in sodium lowers systolic blood pressure by
more in people with already-raised pressure than in those with normal pressure, and the
lower pressure reduces disease incidence by cause.

```{r eff-brief}
tr_tab("eff_brief")
```

`r tab_note("eff_brief")`

## 2.8 Policy-package construction and overlapping interventions

When several policies act on the **same** sodium source (for example labelling and salt
targets both act on packaged food), their effects are combined **multiplicatively**, so
the joint reduction can approach but never exceed 100% of that source. Formally, for a
source acted on by policies with reductions $e_i$, the combined reduction is
$1-\prod_i(1-e_i)$. Each source's combined reduction is then weighted by that source's
share of intake and summed to give a single **total sodium reduction fraction**,
denoted $\eta$, for each country and year. The **full package** combines public food
procurement, salt targets, front-of-pack labelling and the salt tax; low-sodium salt
substitutes are reported separately.

## 2.9 Packaged-food growth and changing sodium-source composition

Packaged food is a growing share of sodium intake in most of these countries. The model
lets the packaged share **grow over time** along a country-specific trend, capped at
80% of sodium sources, with the other sources rescaled so the five always sum to 100%
(Figure \@ref(fig:packaged-time)). This changes the *composition* of sodium intake — and
therefore the reach of packaged-food policies — while holding total baseline intake
fixed [@vandevijvere2019; @marklund2022sources].

```{r packaged-time, fig.cap=TR$fig$packaged_time$title, out.width="90%"}
include_graphics(fig_path("packaged_time"))
```

## 2.10 Low-sodium salt substitutes: the sodium and potassium pathway

Low-sodium salt substitutes replace part of the sodium chloride in salt with potassium
chloride (here `r N$lss_nacl` sodium chloride / `r N$lss_kcl` potassium chloride by
mass). Switching discretionary salt to a substitute therefore does two things: it
**removes sodium** and it **adds potassium** (Figure \@ref(fig:lss-pathway)). Both lower
blood pressure — sodium through the same dose-response as the other policies, potassium
through a separate dose-response calibrated to measured potassium intake
[@Filippini2020; @Reddin2023] — and the two effects are combined. This mechanistic
approach follows the structure of a recent global modelling study [@huang2026]. We
report it as the primary estimate for salt substitutes and use the SSaSS trial
[@Neal2021LSSTrial] only as a benchmark for hypertension-targeted scenarios (Appendix
C). **Importantly, this is a benefit-only estimate**: the model does not include any
potential kidney-related harm of higher potassium, so salt-substitute benefits are not
offset (see limitations).

```{r lss-pathway, fig.cap=TR$fig$lss_pathway$title, out.width="100%"}
include_graphics(fig_path("lss_pathway"))
```

## 2.11 High-level state-transition framework

Health impact is projected with a **well–sick–dead** population model. Each year, some
well people develop disease at the (now lower) incidence rate, some sick people die,
and the surviving population ages by one year. Running the model with and without each
policy, and taking the difference in deaths, yields the **deaths delayed**. The complete
equations are in Appendix A.

# 3. Results

## 3.1 Deaths delayed by intervention and country

Table \@ref(tab:main) reports deaths delayed by country for each packaged-food and
procurement policy, for the salt tax at low, base and high assumptions, and for the full
package at each salt-tax assumption. The country totals reconcile to the "Total" row
within rounding. Figure \@ref(fig:headline) ranks the interventions overall, and Figure
\@ref(fig:countries) shows the full package by country.

```{r main}
tr_tab("main")
```

`r tab_note("main")`

```{r headline, fig.cap=TR$fig$headline$title, out.width="85%"}
include_graphics(fig_path("headline"))
```

## 3.2 Fiscal-policy variants and corresponding package totals

The salt tax is modelled as an **exploratory** pathway from tax rate to consumer price,
demand response and, finally, reduced packaged sodium (Figure \@ref(fig:fiscal-pathway);
Appendix C, Table \@ref(tab:fiscal-path)). It ranges from **`r N$fiscal_low_m` million**
deaths delayed under conservative assumptions to **`r N$fiscal_high_m` million** under
optimistic ones, with a base case of **`r N$fiscal_base_m` million**. Because deaths
delayed do not simply add across policies, the package total is re-computed for each tax
assumption; these are the three "Package total" columns of Table \@ref(tab:main),
ranging from **`r N$pkg_low_m`** to **`r N$pkg_high_m` million**.

```{r fiscal-pathway, fig.cap=TR$fig$fiscal_pathway$title, out.width="100%"}
include_graphics(fig_path("fiscal_pathway"))
```

## 3.3 Low-sodium salt substitutes: uptake and hypertension targeting

Table \@ref(tab:lss) reports salt-substitute results: whole-population use (Scenario 2)
at population uptake levels of 10%, 20%, 30%, 40% and 50%, and two
hypertension-targeted scenarios — adults with **diagnosed** hypertension (Scenario 4)
and adults with **treated** hypertension (Scenario 5). Whole-population benefit scales
almost linearly with uptake (Figure \@ref(fig:lss-uptake)), reaching
**`r N$lss_s2_hi_m` million** deaths delayed at `r N$lss_s2_hi_pct` uptake. The
hypertension-targeted scenarios deliver **`r N$lss_s4_m` million** (diagnosed) and
**`r N$lss_s5_m` million** (treated) — a large benefit from a much smaller target
group, because risk is concentrated in people with high blood pressure. These figures
follow the mechanistic sodium-and-potassium pathway, not the salt-substitute trial's
relative risks.

```{r lss}
tr_tab("lss")
```

`r tab_note("lss")`

```{r lss-uptake, fig.cap=TR$fig$lss_uptake$title, out.width="85%"}
include_graphics(fig_path("lss_uptake"))
```

## 3.4 Which interventions have the greatest impact

Across the ten countries, **mandatory salt targets** are the single most impactful
packaged-food policy, because they act on the whole packaged category at a substantial
reduction. **Front-of-pack labelling** is next. **Public food procurement** is the
smallest of the packaged-and-procurement policies, reflecting the modest and still
**provisional** share of sodium that public institutions reach. The **salt tax** falls
between labelling and procurement but is **exploratory**. **Low-sodium salt substitutes**
stand apart: where discretionary salt dominates intake, their whole-population potential
exceeds the entire packaged-food package — a key strategic finding, subject to the
benefit-only caveat and to achieving real-world uptake and adherence.

## 3.5 Important cross-country differences

The ranking of policies flips with the local sodium profile. In China and Malaysia,
where packaged food is a large share of sodium, reformulation and labelling dominate. In
Ethiopia, India and Bangladesh, where discretionary salt dominates, salt substitutes
carry by far the most potential. Absolute impact tracks population and cardiovascular
burden: China and India account for most deaths delayed under every policy, while the
smallest countries contribute proportionally less (Figure \@ref(fig:countries)).

```{r countries, fig.cap=TR$fig$countries$title, out.width="80%"}
include_graphics(fig_path("countries"))
```

## 3.6 Full-package findings

The full package delays about **`r N$pkg_m` million** cardiovascular deaths
(`r N$pkg_pct` of baseline), the great majority in the largest countries. By cause, the
package's benefit is led by **`r N$top_cause`** (**`r N$top_cause_m` million**,
`r N$top_cause_pct` of the package total), reflecting the strong blood-pressure link of
stroke (Figure \@ref(fig:bycause); Table \@ref(tab:bycause-tab)). The full package excludes
low-sodium salt substitutes, which are reported separately; adding them where
discretionary salt is large would raise the total substantially.

```{r bycause, fig.cap=TR$fig$bycause$title, out.width="75%"}
include_graphics(fig_path("bycause"))
```

```{r bycause-tab}
tr_tab("bycause", "bycause-tab")
```

`r tab_note("bycause")`

## 3.7 Donor-relevant takeaways, without overstating precision

Sodium policy offers a large, credible cardiovascular return across these ten countries.
The estimates here are deliberately conservative in construction — each policy reaches
only the sodium it can — and they still point to millions of deaths delayed. They are
projections, not guarantees: the ranges shown for the salt tax and the salt-substitute
uptake sweep are the honest expression of that uncertainty, and provisional or
exploratory items are labelled as such.

# 4. Assumptions and limitations

The main assumptions and limitations are summarised in Table \@ref(tab:limitations) and
expanded below.

```{r limitations}
tr_tab("limitations")
```

- **Deaths delayed, not averted.** The sodium-to-blood-pressure policies lower incidence
  but not case fatality, so they postpone deaths rather than prevent them outright;
  totals depend on the window.
- **Salt substitutes are benefit-only.** No kidney-related harm of higher potassium is
  modelled, so salt-substitute benefits are not offset and should be read as an upper
  part of the plausible range where such risks matter.
- **The salt tax is exploratory.** Its price, demand and substitution assumptions are
  uncertain, and broad-category demand responses may not transfer to high-sodium foods.
- **Public food procurement is provisional.** It uses a preliminary institutional-reach
  assumption pending adults-only country data.
- **Some inputs are held constant or use proxies.** Hypertension diagnosis and treatment
  shares are age-standardised and held after their source year; potassium intake for two
  countries uses a regional average; several source-share profiles are provisional; and
  future packaged-food growth is an extrapolation.
- **Uncertainty is not fully propagated.** Results are point estimates; no full
  probabilistic sensitivity analysis is run.
- **Mass-media campaigns are excluded** from the reported scenarios and the package.

# 5. Conclusion

Across ten priority countries, a modest set of packaged-food and procurement policies is
projected to delay on the order of **`r N$pkg_m` million** cardiovascular deaths over
`r N$window`, and low-sodium salt substitutes could deliver more still where
discretionary salt dominates intake. The most effective mix differs by country, which
argues for tailoring policy portfolios to each country's sodium profile rather than
applying a single template. With transparent assumptions and clearly-flagged provisional
elements, these estimates make a strong, evidence-based case for sustained investment in
population sodium reduction — while leaving room for the better data and fuller
uncertainty analysis that would sharpen the numbers.

# 6. References

::: {#refs}
:::

# Appendix A: Full mathematical specification

**Source split and total sodium reduction.** Let $w_{c,s,t}$ be the share of country
$c$'s sodium intake from source $s$ in year $t$ (with $\sum_s w_{c,s,t}=1$), and let
$e_{i,s}$ be the fractional reduction that policy $i$ applies to source $s$. For a set of
policies $\mathcal{P}$, the combined reduction of source $s$ is
\begin{equation}
r_{c,s,t}=1-\prod_{i\in\mathcal{P}}\bigl(1-e_{i,s}\bigr),(\#eq:compose)
\end{equation}
and the total sodium reduction fraction is
\begin{equation}
\eta_{c,t}=\sum_{s} w_{c,s,t}\, r_{c,s,t}.(\#eq:eta)
\end{equation}
Equation \@ref(eq:compose) guarantees that overlapping policies on one source never
exceed a 100% reduction.

**Sodium to blood pressure.** With baseline sodium intake $\text{Na}^0$ (g/day) and a
scaled-up reduction $\Delta\text{Na}=\eta\,\text{Na}^0$ (subject to a 2 g/day minimum
intake floor), the systolic blood-pressure reduction is
\begin{equation}
\Delta\text{SBP}=\bigl(2.8\,h+1.0\,(1-h)\bigr)\,\Delta\text{Na},(\#eq:sbp)
\end{equation}
where $h$ is the fraction of the age–sex group with raised blood pressure.

**Blood pressure to incidence.** Within each blood-pressure band $b$, baseline
incidence for cause $k$ is decomposed using relative-risk weights and then reduced in
proportion to $\Delta\text{SBP}$:
\begin{equation}
\text{IR}^{\text{new}}_{k}=\sum_b \pi_b\, \text{IR}_{b,k}\Bigl(1-(1-RR^{10}_{b,k})\,\tfrac{\Delta\text{SBP}}{10}\Bigr),(\#eq:ir)
\end{equation}
where $\pi_b$ is the share of the group in band $b$ and $RR^{10}_{b,k}$ the relative risk
per 10 mmHg. Case fatality is unchanged for these policies.

**Potassium channel (salt substitutes).** Switching discretionary salt to a substitute
adds potassium in proportion to the sodium displaced (about 1.33 g potassium per g
sodium). Baseline and post-switch potassium are converted to measured excretion and
mapped to a blood-pressure change through a dose-response curve; the resulting
$\Delta\text{SBP}_{K}$ is added to the sodium-mediated change and Equation \@ref(eq:ir)
is applied to the combined total (Appendix F).

**Population dynamics.** For each cause, with well ($W$), sick ($S$) and dead ($D$)
counts, incidence $\text{IR}$, case fatality $\text{CF}$ and background mortality $m$:
\begin{equation}
\begin{aligned}
S_t &= S_{t-1}(1-\text{CF}-m)+W_{t-1}\,\text{IR},\\
D_t &= S_{t-1}\,\text{CF},\\
W_t &= P_t - S_t - \textstyle\sum_k D_{t}.
\end{aligned}(\#eq:states)
\end{equation}
**Deaths delayed** for a scenario are baseline deaths minus scenario deaths, summed over
the results window.

# Appendix B: Complete scenario definitions

```{r scn-def}
tr_tab("scn_def")
```

`r tab_note("scn_def")`

# Appendix C: Detailed intervention methods and effect-size evidence

Table \@ref(tab:eff-evidence) lists each effect size with its evidence source. Modelled
values follow the production configuration and may differ from a single published study
(for example, salt targets are modelled at 20% against roughly 15% observed in the
United Kingdom over nine years [@he2013]). Public food procurement draws on institutional
cafeteria studies [@gase2011]; front-of-pack warning labelling on the Chilean evaluation
[@taillie2024]; and the salt tax on Philippine tax modelling [@saxena2025] with support
from Australian work [@cobiac2017].

```{r eff-evidence}
tr_tab("eff_evidence")
```

`r tab_note("eff_evidence")`

The salt tax is an **exploratory** pathway (Table \@ref(tab:fiscal-path)); it does not
assume that a 10% tax yields a 10% sodium reduction, nor that all packaged sodium is
taxable. The salt-substitute benchmark using the SSaSS trial's stroke relative risks
[@Neal2021LSSTrial] is shown alongside the mechanistic estimate for the
hypertension-targeted scenarios in Table \@ref(tab:nak-ssass); the two are distinct
pathways and are never combined.

```{r fiscal-path}
tr_tab("fiscal_path")
```

`r tab_note("fiscal_path")`

```{r nak-ssass}
tr_tab("nak_ssass")
```

`r tab_note("nak_ssass")`

# Appendix D: Sodium-source definitions and country assumptions

The five sodium sources are discretionary salt, packaged food, restaurant and vendor
food, public and institutional food, and inherent sodium. Table
\@ref(tab:source-shares) gives each country's base-year shares; several are provisional,
and any country without its own profile would use a default average.

```{r source-shares}
tr_tab("source_shares")
```

`r tab_note("source_shares")`

# Appendix E: Population, disease, blood-pressure and mortality inputs

Hypertension diagnosis and treatment shares that size the salt-substitute
hypertension-targeted scenarios come from the NCD Risk Factor Collaboration
[@ncdrisc2021] (Table \@ref(tab:ncd-elig)); baseline potassium intake from a global
pooled analysis [@Reddin2023] (Table \@ref(tab:baseline-k)); and the packaged-food growth
trend from commercial sales data [@marklund2022sources] (Table \@ref(tab:pkg-trend)). Key
model inputs and parameters are summarised in Table \@ref(tab:key-inputs).

```{r ncd-elig}
tr_tab("ncd_elig")
```

`r tab_note("ncd_elig")`

```{r baseline-k}
tr_tab("baseline_k")
```

`r tab_note("baseline_k")`

```{r pkg-trend}
tr_tab("pkg_trend")
```

`r tab_note("pkg_trend")`

```{r key-inputs}
tr_tab("key_inputs")
```

`r tab_note("key_inputs")`

# Appendix F: Low-sodium salt substitute parameters and pathway quantities

Table \@ref(tab:lss-audit) reports the sodium displaced, potassium added and
blood-pressure change per country and sex at a representative whole-population coverage;
these quantities scale linearly with coverage. Table \@ref(tab:huang) positions the
mechanistic salt-substitute model against the recent global study whose structure it
follows [@huang2026].

```{r lss-audit}
tr_tab("lss_audit")
```

`r tab_note("lss_audit")`

```{r huang}
tr_tab("huang")
```

`r tab_note("huang")`

# Appendix G: Packaged-food trend method

The packaged share of sodium grows along a country-specific annual rate (Table
\@ref(tab:pkg-trend)), capped at 80% of sodium sources, held constant after 2030, with
the other four sources rescaled each year so the five sum to 100%. Total baseline sodium
intake is held fixed — only the *composition* changes — so the trend changes the reach of
packaged-food policies over time rather than the underlying intake.

# Appendix H: Validation and consistency checks

The following checks were performed and passed for this run: the five sodium-source
shares sum to 100% in every country and year; baseline deaths delayed equal zero; the
whole-population salt-substitute benefit is non-decreasing from 10% to 50% uptake in
every country; hypertension-targeted scenarios use the intended country- and
sex-specific eligibility; salt-substitute results use the mechanistic
sodium-and-potassium pathway rather than the trial relative risks; country totals
reconcile to the reported totals within rounding; and the table and figure values in this
report match the companion workbook exactly. As an external cross-check, the full package
(`r N$bench_cur_m` million deaths delayed across the modelled countries) is compared with
a previous estimate for a uniform 30% sodium cut (`r N$bench_prev_m` million); the current
package is expectedly smaller because it reaches only the sodium its policies can affect
(Table \@ref(tab:benchmark)).

```{r benchmark}
tr_tab("benchmark")
```

`r tab_note("benchmark")`

# Appendix I: Supplemental tables and figures

Table \@ref(tab:scn-summary) gives cumulative deaths delayed by scenario across all
countries and causes, and Table \@ref(tab:saxena) positions the exploratory salt-tax
model against the study that anchors it [@saxena2025].

```{r scn-summary}
tr_tab("scn_summary")
```

`r tab_note("scn_summary")`

```{r saxena}
tr_tab("saxena")
```

`r tab_note("saxena")`
