---
title: "Sodium Interventions Model"
subtitle: "Projected cardiovascular health impact of sodium-reduction policies in ten priority countries"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  bookdown::word_document2:
    toc: true
    toc_depth: 2
    number_sections: false
    fig_caption: true
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: false
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
bibliography: references.bib
csl: apa.csl
link-citations: true
---

```{r tr-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 300, out.width = "100%")
options(knitr.kable.NA = "")
suppressMessages({
  library(flextable); library(officer); library(knitr); library(kableExtra)
})

# Locate and load the authoritative artifact bundle (produced by the analysis
# layer). No calculation happens in this document.
.here <- tryCatch(dirname(knitr::current_input(dir = TRUE)), error = function(e) getwd())
assets  <- file.path(.here, "report_assets")
TR <- readRDS(file.path(assets, "tr_bundle.rds"))
N  <- TR$nums
fig_path <- function(key) file.path(assets, TR$fig[[key]]$file)

# Brand styling comes from the bundle (single palette defined in reports/report.RMD).
STY <- if (!is.null(TR$style)) TR$style else
  list(header_bg = "#382986", header_fg = "#FFFFFF", band = "#F5F6FC")

# ---- identifiers, titles, notes and internal links (all metadata from bundle) ----
# Human-readable identifiers ("Table 3", "Figure 6") are kept SEPARATE from the
# Word-safe bookmark names ("tbl_countries", "fig_map") the bundle stores.
fig_ref   <- function(key) TR$fig[[key]]$id                       # e.g. "Figure 6"
tab_ref   <- function(key) TR$tab[[key]]$id                       # e.g. "Table 3"
# (Comment 2) Clickable internal cross-references: visible identifier -> bookmark
# anchor. Emitted as Markdown links so pandoc makes real Word hyperlinks.
fig_link  <- function(key) paste0("[", TR$fig[[key]]$id, "](#", TR$fig[[key]]$bkm, ")")
tab_link  <- function(key) paste0("[", TR$tab[[key]]$id, "](#", TR$tab[[key]]$bkm, ")")
# The title paragraph (above the figure/table) carries the target bookmark via a Pandoc
# span id, so links land on the title itself, not a nearby heading. Tables use the SAME
# mechanism as figures because flextable's run_autonum(bkm=) does not emit a bookmark that
# a Markdown link can target in this toolchain (verified against the rendered OOXML).
fig_title <- function(key)
  paste0("**[", TR$fig[[key]]$id, ". ", TR$fig[[key]]$title, "]{#", TR$fig[[key]]$bkm, "}**")
tab_title <- function(key)
  paste0("**[", TR$tab[[key]]$id, ". ", TR$tab[[key]]$title, "]{#", TR$tab[[key]]$bkm, "}**")

# Notes (plain paragraph) then, on a separate line, a bold standardized Source line.
# The source is not duplicated inside the notes; the block itself is not italic/bold.
.mk_note <- function(o) {
  g <- function(nm) { v <- o[[nm]]; if (!is.null(v) && nzchar(v)) v else NULL }
  bits <- c(g("notes"),
            if (!is.null(g("population"))) paste0("Population: ", o$population, "."),
            if (!is.null(g("period")))     paste0("Period: ", o$period, "."),
            if (!is.null(g("units")))      paste0("Units: ", o$units, "."),
            if (!is.null(g("scenario_def"))) o$scenario_def,
            if (!is.null(g("rounding")))   paste0("Rounding: ", o$rounding, "."))
  body <- paste(bits, collapse = " ")
  body <- gsub("\\.\\.+", ".", body)            # collapse doubled periods
  body <- gsub("\\.\\s*\\.", ".", body)
  src  <- if (!is.null(g("source"))) paste0("**Source: ", o$source, ".**") else ""
  paste0("Notes: ", body, "\n\n", src)
}
tab_note <- function(key) .mk_note(TR$tab[[key]])
fig_note <- function(key) .mk_note(TR$fig[[key]])

# ---- one table renderer, two output formats --------------------------------
# The table's numbered title + bookmark is rendered ABOVE the table by tab_title()
# (a Pandoc span, like figures), which tab_link() targets, so every in-text table
# citation is a real internal hyperlink. The table itself therefore carries NO caption
# (no duplicate title).
.is_num_col <- function(x) {
  x <- trimws(as.character(x)); x <- x[nzchar(x)]
  if (!length(x)) return(FALSE)
  mean(grepl("^[-–−(]?[0-9][0-9,.]*%?\\)?$", x)) > 0.6
}
tr_tab <- function(key, id = gsub("_", "-", key)) {
  tb  <- TR$tab[[key]]; df <- tb$display
  wide <- ncol(df) >= 6 || identical(tb$orientation, "landscape")
  numcol <- vapply(df, .is_num_col, logical(1)); numcol[1] <- FALSE
  al <- ifelse(numcol, "r", "l")
  if (knitr::is_latex_output()) {
    # scale_down wraps the table in \resizebox, which only shrinks when the table is
    # wider than the line (a no-op otherwise), so it is safe to apply to every table
    # and prevents long-header tables (e.g. the countries table) from overflowing.
    k <- kbl(df, booktabs = TRUE, linesep = "", align = al) |>
      kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
        font_size = tb$fontsize, full_width = FALSE)
    # Wrap a long text first column so narrow tables do not overflow the page.
    if (!wide && max(nchar(df[[1]]), na.rm = TRUE) > 30)
      k <- column_spec(k, 1, width = "16em")
    # Wrap all columns of long multi-column text comparisons (e.g. Huang/Saxena).
    if (ncol(df) == 3 && mean(nchar(unlist(df[-1]))) > 30)
      k <- column_spec(k, 2:3, width = "20em")
    k
  } else {
    ft <- flextable(df)
    ft <- theme_booktabs(ft)
    ft <- bg(ft, part = "header", bg = STY$header_bg)
    ft <- color(ft, part = "header", color = STY$header_fg)
    ft <- bold(ft, part = "header")
    # Restrained zebra banding for readability (does not carry meaning on its own).
    if (nrow(df) > 2) {
      even <- seq(2, nrow(df), by = 2)
      ft <- bg(ft, i = even, bg = STY$band, part = "body")
    }
    ft <- fontsize(ft, size = tb$fontsize, part = "all")
    ft <- padding(ft, padding = 3, part = "all")
    if (any(numcol)) ft <- align(ft, j = which(numcol), align = "right", part = "all")
    ft <- align(ft, j = which(!numcol), align = "left", part = "body")
    ft <- align(ft, part = "header", align = "center")
    ft <- valign(ft, valign = "top", part = "body")
    if (grepl("^Total", tail(df[[1]], 1))) ft <- bold(ft, i = nrow(df), part = "body")
    ft <- set_table_properties(ft, layout = "autofit", width = 1)
    ft <- flextable::fix_border_issues(ft)
    ft
  }
}
```

```{r tr-assertions, include=FALSE}
# Report-level QA. These run at render time and STOP the build if an invariant the
# report depends on is violated (missing element, empty title, broken removal, etc.).
.fail <- function(msg) stop("technical-report QA assertion failed: ", msg, call. = FALSE)
`%||%` <- function(a, b) if (is.null(a)) b else a

# 1. Every ordered table/figure key exists in the bundle.
for (k in TR$table_order)  if (is.null(TR$tab[[k]])) .fail(paste("missing table", k))
for (k in TR$figure_order) if (is.null(TR$fig[[k]])) .fail(paste("missing figure", k))

# 2. Every table/figure has a non-empty title.
for (k in TR$table_order)  if (!nzchar(TR$tab[[k]]$title)) .fail(paste("empty title, table", k))
for (k in TR$figure_order) if (!nzchar(TR$fig[[k]]$title)) .fail(paste("empty title, figure", k))

# 3. Every element has a substantive note and/or source.
.has_note <- function(o) any(nzchar(c(o$notes, o$source, o$units, o$period, o$scenario_def)))
for (k in TR$table_order)  if (!.has_note(TR$tab[[k]])) .fail(paste("no note/source, table", k))
for (k in TR$figure_order) if (!.has_note(TR$fig[[k]])) .fail(paste("no note/source, figure", k))

# 4. Figure/table numbers follow appearance order.
if (!identical(unname(vapply(TR$figure_order, function(k) TR$fig[[k]]$num, integer(1))),
               seq_along(TR$figure_order))) .fail("figure numbering not in appearance order")
if (!identical(unname(vapply(TR$table_order,  function(k) TR$tab[[k]]$num, integer(1))),
               seq_along(TR$table_order)))  .fail("table numbering not in appearance order")

# 5. Comments D/E/G: state figure present; fiscal-pathway and by-cause figures removed.
if (!("state_transition" %in% TR$figure_order)) .fail("comment D: state-transition figure missing")
if ("fiscal_pathway" %in% TR$figure_order)      .fail("comment E: fiscal-pathway figure not removed")
if ("bycause" %in% TR$figure_order)             .fail("comment G: by-cause figure not removed")

# 6. Comments E/G: fiscal-path and by-cause TABLES retained.
if (!("fiscal_path" %in% TR$table_order)) .fail("comment E: fiscal pathway table missing")
if (!("bycause" %in% TR$table_order))     .fail("comment G: by-cause table missing")

# 7. Comment F: countries figure is a Pareto whose cumulative share ends at 100%.
.cty <- TR$fig$countries$data
if (is.null(.cty) || is.null(.cty$cumulative_share)) .fail("comment F: countries figure lacks cumulative share (not a Pareto)")
if (abs(tail(.cty$cumulative_share, 1) - 1) > 0.01)  .fail("comment F: Pareto cumulative share does not end at 100%")

# 8. Comment B: no residual media-campaign text anywhere in the bundle.
if (grepl("\\bmedia\\b", tolower(paste(unlist(TR), collapse = " | ")))) .fail("comment B: residual media text in bundle")

# 9. Country table (comments 3, 4): SBP labelled 2026, NCD-RisC attribution, DISTINCT
#    raised-BP and NCD-RisC 2021 hypertension columns, and 2026 (not 2030) population.
.cn <- TR$tab$countries; .nm <- names(.cn$display)
if (!any(grepl("SBP", .nm) & grepl("2026", .nm))) .fail("comment A: SBP column not labelled 2026")
if (!grepl("NCD-RisC", paste(.cn$source, .cn$notes))) .fail("comment A: SBP not attributed to NCD-RisC")
if (!any(grepl("Raised BP", .nm)))                 .fail("comment 3: raised-BP column missing")
if (!any(grepl("Hypertension", .nm) & grepl("2021", .nm))) .fail("comment 3: NCD-RisC 2021 hypertension column missing")
if (!any(grepl("population, 2026", .nm)))          .fail("comment 4: population column not labelled 2026")
if (any(grepl("2030", .nm)))                       .fail("comment 4: a 2030 label remains in the country table")

# 10. Comment 2: every table/figure has a unique, Word-safe bookmark distinct from its id.
.bk <- c(vapply(TR$table_order, function(k) TR$tab[[k]]$bkm %||% "", character(1)),
         vapply(TR$figure_order, function(k) TR$fig[[k]]$bkm %||% "", character(1)))
if (any(!nzchar(.bk)))                       .fail("comment 2: an element has no bookmark")
if (!all(grepl("^[A-Za-z][A-Za-z0-9_]*$", .bk))) .fail("comment 2: a bookmark is not Word-safe")
if (anyDuplicated(.bk))                      .fail("comment 2: duplicate bookmark names")

# 11. Comment 6: every table/figure source is EXACTLY the standardized string.
.srcs <- c(vapply(TR$table_order, function(k) TR$tab[[k]]$source %||% "", character(1)),
           vapply(TR$figure_order, function(k) TR$fig[[k]]$source %||% "", character(1)))
if (!all(.srcs == "UW/RTSL sodium policy model")) .fail(paste(
  "comment 6: non-standard source:", paste(unique(.srcs[.srcs != "UW/RTSL sodium policy model"]), collapse=" | ")))

# 12. Comment 5: bundle distinguishes 2017 initialization from the 2019 epidemiologic seed.
if (!identical(as.integer(TR$meta$initialization_year), 2017L)) .fail("comment 5: initialization_year is not 2017")
if (!identical(as.integer(TR$meta$epidemiologic_seed_year), 2019L)) .fail("comment 5: epidemiologic_seed_year is not 2019")

# 13. Comment 7: eff_evidence merged into eff_brief and removed; eff_brief has an evidence column.
if ("eff_evidence" %in% TR$table_order) .fail("comment 7: eff_evidence table not removed")
if (!any(grepl("Evidence", names(TR$tab$eff_brief$display)))) .fail("comment 7: eff_brief lacks an evidence column")

# 14. Comment 9: no status words anywhere in the bundle's report-facing content.
.blob <- tolower(paste(unlist(lapply(c(TR$tab, TR$fig), function(o)
  c(o$title, o$notes, o$units, o$period, o$scenario_def, o$source, unlist(o$display)))),
  paste(unlist(TR$dict), collapse = " "), paste(unlist(TR$meta$assumptions), collapse = " "),
  collapse = " "))
for (w in c("provisional", "preliminary", "pending", "exploratory"))
  if (grepl(w, .blob)) .fail(paste0("comment 8/9: status word '", w, "' present in the bundle"))

# 15. Totals reconcile: main table Total row == column sums of the country rows.
.m <- TR$tab$main$raw
if (!is.null(.m)) {
  numc <- vapply(.m, is.numeric, logical(1)); trow <- grepl("^Total", .m[[1]])
  if (any(trow) && any(numc)) {
    got <- as.numeric(.m[trow, numc][1, ]); exp <- colSums(.m[!trow, numc, drop = FALSE])
    if (any(abs(got - exp) > pmax(1, 1e-3 * abs(exp)))) .fail("main table Total row does not reconcile")
  }
}
cat("technical-report QA: validated", length(TR$table_order), "tables /",
    length(TR$figure_order), "figures.\n")
```

`r if (knitr::is_latex_output()) "\\vspace{2em}"`

**Sodium Interventions Model: Technical Report**

*Projected cardiovascular health impact of national sodium-reduction policies in
`r N$n_countries` priority countries, `r N$window`.*

Prepared by the University of Washington in partnership with Resolve to Save Lives.
`r format(Sys.Date(), "%B %Y")`.

This report accompanies a companion results workbook. Health impacts are expressed
as **cardiovascular deaths delayed** relative to a no-intervention baseline, and the
results are point estimates.

\newpage

# 1. Executive summary

High salt intake is one of the largest and most modifiable drivers of cardiovascular
disease. Across the `r N$n_countries` priority countries in this report, average sodium
intake sits well above what the body needs, raising blood pressure and, with it, the
risk of heart disease and stroke. Because the sodium-to-blood-pressure-to-disease chain
is well established, population policies that lower sodium can prevent a large share of
this burden, and most of the sodium people eat is within reach of practical, proven
policy.

## 1.1 The problem and the opportunity

Raised blood pressure from excess dietary sodium is a leading, preventable cause of the
four cardiovascular diseases modelled here: ischaemic heart disease, ischaemic stroke,
intracerebral (haemorrhagic) stroke and hypertensive heart disease. Governments have
several evidence-based levers, including reformulating packaged food, warning labels,
healthier public food procurement, fiscal measures and low-sodium salt substitutes,
each acting on a different part of the food supply. This report estimates how many
cardiovascular deaths each lever, and a coordinated package, could delay in these ten
countries.

## 1.2 What we modelled

We evaluate five policies: public food procurement, mandatory sodium (salt) targets for
packaged food, front-of-pack warning labelling, a salt tax, and low-sodium salt
substitutes, across `r N$n_countries` priority countries (Bangladesh, Cameroon, China,
Ethiopia, India, Malaysia, Nigeria, the Philippines, Thailand and Viet Nam) over
**`r N$window`**. Each policy is modelled through the same biological chain: less sodium
(and, for salt substitutes, more potassium) lowers blood pressure, which lowers the
incidence of the four cardiovascular diseases. Each policy is applied only to the sodium
sources it can realistically reach, so the estimates are conservative by construction.

## 1.3 What the model indicates

- A coordinated **package** of packaged-food and procurement policies is projected to
  delay about **`r N$pkg_m` million** cardiovascular deaths over `r N$window`
  (`r N$pkg_pct` of baseline cardiovascular deaths in the window); across salt-tax
  assumptions the range is **`r N$pkg_low_m`–`r N$pkg_high_m` million**
  (`r tab_link("main")`).
- Among packaged-food policies, **mandatory salt targets** deliver the largest single
  benefit (**`r N$salt_targets_m` million** deaths delayed), followed by front-of-pack
  labelling (**`r N$fopl_m` million**).
- **Low-sodium salt substitutes** show the largest *potential* of any single measure
  where discretionary salt dominates intake: whole-population use delays from
  **`r N$lss_s2_lo_m` million** at `r N$lss_s2_lo_pct` population uptake to
  **`r N$lss_s2_hi_m` million** at `r N$lss_s2_hi_pct` uptake (`r tab_link("lss")`,
  `r fig_link("lss_uptake")`).

## 1.4 How impact differs by country and sodium source

Impact is highly concentrated: **China accounts for about `r N$china_share`** of the
package benefit, and a small number of the largest countries account for roughly 80% of
the ten-country total (`r fig_link("countries")`), reflecting population size and
cardiovascular burden. The *best policy mix* also differs by country. Where packaged
food is a large share of sodium (for example China and Malaysia), reformulation and
labelling matter most; where discretionary salt dominates (for example Ethiopia, India
and Bangladesh), low-sodium salt substitutes carry the largest potential. Combining
policies compounds the benefit without double-counting, because each acts only on the
sodium sources it can reach.

## 1.5 What decision-makers should take from this

A small set of sodium policies, applied only to the sodium they can realistically reach,
is still projected to delay millions of cardiovascular deaths across these ten countries.
The evidence supports tailoring each country's policy portfolio to its own sodium
profile rather than applying a single template, and pursuing packaged-food reformulation
and low-sodium salt substitutes as complementary strategies. This analysis models health
impact only; it does not estimate programme costs or cost-effectiveness.

## 1.6 How to interpret these estimates, and main uncertainties

These are **deaths delayed**, not permanently averted: the sodium-to-blood-pressure
policies lower disease *incidence* while leaving case fatality unchanged, so they shift
deaths later in life rather than removing them from the life table, and totals depend on
the accounting window (`r N$window`). Estimates are **point projections** without full
uncertainty ranges. The salt-tax (fiscal) results are **exploratory**: the tax
pass-through, price elasticity, taxable share and substitution are uncertain, and
broad-category demand responses may not transfer to high-sodium foods. Public food
procurement is modelled using an assumed 5% institutional share of sodium, and
salt-substitute figures are benefit-only because potential kidney-related harms of higher
potassium are not modelled. The results are best read as the *relative* promise of each
policy and the *order of magnitude* of the opportunity, not as precise forecasts.

# 2. Modeling approach

## 2.1 Overall conceptual framework

The model links each policy to health outcomes through one transparent causal chain
(`r fig_link("framework")`). A policy reduces sodium from the food sources it can
reach; the resulting fall in average sodium intake lowers systolic blood pressure;
lower blood pressure reduces the yearly risk of developing four cardiovascular
diseases; and a population model translates that lower risk into deaths delayed over
time. Low-sodium salt substitutes add a second, blood-pressure-lowering channel
through the potassium they contain.

`r fig_title("framework")`

```{r framework, out.width="100%"}
include_graphics(fig_path("framework"))
```

`r fig_note("framework")`

## 2.2 Countries and populations included

The model covers the ten RTSL priority countries and the adult population aged 20 and
older in each (`r tab_link("countries")`). Nigeria's sodium-source profile is applied to
Cameroon because a country-specific profile was unavailable. Although the client's draft
results tables omitted Viet Nam, it is a modelled priority country and is included
throughout.

`r tab_title("countries")`

```{r countries-tab}
tr_tab("countries", "countries-tab")
```

`r tab_note("countries")`

## 2.3 Baseline year, scale-up, projection horizon and results window

The population model is **initialised in `r TR$meta$initialization_year`** to establish
the first transition history, with **`r TR$meta$epidemiologic_seed_year`** as the
pre-pandemic epidemiologic seed (reference) year; it is then projected annually to
**`r TR$meta$horizon`**. Policies scale up linearly over **`r N$scaleup`** and then hold
at full implementation. Results are reported cumulatively over the window
**`r N$window`**, with reference years `r N$results_years`.

## 2.4 Sodium-source framework

Baseline sodium intake in each country is divided into **five sources**: discretionary
salt (added at the table or in home cooking), packaged food, food from restaurants and
vendors, food from public and institutional settings, and sodium inherent to unprocessed
foods. These shares sum to 100% (`r fig_link("shares")`, `r tab_link("source_shares")`).
Each policy is applied **only to the sources it can plausibly reach** (for example, salt
targets and labelling to packaged food, salt substitutes to discretionary salt), and
inherent sodium is never targeted. This "source-split" design keeps every policy's
claimed effect anchored to the sodium it can actually influence (`r fig_link("map")`).

`r fig_title("shares")`

```{r shares, out.width="90%"}
include_graphics(fig_path("shares"))
```

`r fig_note("shares")`

`r fig_title("map")`

```{r map, out.width="80%"}
include_graphics(fig_path("map"))
```

`r fig_note("map")`

## 2.5 Population and demographic assumptions

Country populations, births and deaths follow the United Nations 2024 projections.
Baseline disease and mortality rates come from the Global Burden of Disease study, and
the distribution of blood pressure by age and sex from pooled international data. The
model tracks single years of age from 20 to 95.

## 2.6 Cardiovascular outcomes included

Four blood-pressure-related cardiovascular diseases are modelled: ischaemic heart
disease, ischaemic stroke, intracerebral (haemorrhagic) stroke, and hypertensive heart
disease. All benefits reported here are for these four causes combined unless a
by-cause breakdown is shown.

## 2.7 Intervention effect assumptions

Each policy has a documented **relative reduction** of its target sodium source at full
implementation, with the modelled effect, blood-pressure pathway and evidence source
summarised together in `r tab_link("eff_brief")`. The sodium-to-blood-pressure
relationship follows an established dose-response [@ettehad2016; @Filippini2020]: a
reduction in sodium lowers systolic blood pressure by more in people with already-raised
pressure than in those with normal pressure, and the lower pressure reduces disease
incidence by cause.

`r tab_title("eff_brief")`

```{r eff-brief}
tr_tab("eff_brief")
```

`r tab_note("eff_brief")`

## 2.8 Policy-package construction and overlapping interventions

When several policies act on the **same** sodium source (for example labelling and salt
targets both act on packaged food), their effects are combined **multiplicatively**, so
the joint reduction can approach but never exceed 100% of that source. Formally, for a
source acted on by policies with reductions $e_i$, the combined reduction is
$1-\prod_i(1-e_i)$. Each source's combined reduction is then weighted by that source's
share of intake and summed to give a single **total sodium reduction fraction**,
denoted $\eta$, for each country and year. The **full package** combines public food
procurement, salt targets, front-of-pack labelling and the salt tax; low-sodium salt
substitutes are reported separately.

## 2.9 Packaged-food growth and changing sodium-source composition

Packaged food is a growing share of sodium intake in most of these countries. The model
lets the packaged share **grow over time** along a country-specific trend, capped at
80% of sodium sources (`r fig_link("packaged_time")`). Raising the packaged share is a
**compositional reallocation, not an increase in intake**: the other four source shares
are proportionally rescaled so that all five sources continue to sum to 100%. This
guarantees that **total baseline sodium intake does not grow** in this module; only its
source composition changes, which in turn shifts how much sodium packaged-food policies
can reach over time [@vandevijvere2019; @marklund2022sources].

`r fig_title("packaged_time")`

```{r packaged-time, out.width="90%"}
include_graphics(fig_path("packaged_time"))
```

`r fig_note("packaged_time")`

## 2.10 Low-sodium salt substitutes: the sodium and potassium pathway

Low-sodium salt substitutes replace part of the sodium chloride in salt with potassium
chloride (here `r N$lss_nacl` sodium chloride / `r N$lss_kcl` potassium chloride by
mass). Switching discretionary salt to a substitute therefore does two things: it
**removes sodium** and it **adds potassium** (`r fig_link("lss_pathway")`). Both lower
blood pressure: sodium through the same dose-response as the other policies, and potassium
through a separate dose-response calibrated to measured potassium intake
[@Filippini2020; @Reddin2023]; the two effects are combined. This mechanistic
approach follows the structure of a recent global modelling study [@huang2026]. We
report it as the primary estimate for salt substitutes and use the SSaSS trial
[@Neal2021LSSTrial] only as a benchmark for hypertension-targeted scenarios (Appendix
C). **Importantly, this is a benefit-only estimate**: the model does not include any
potential kidney-related harm of higher potassium, so salt-substitute benefits are not
offset (see limitations).

`r fig_title("lss_pathway")`

```{r lss-pathway, out.width="100%"}
include_graphics(fig_path("lss_pathway"))
```

`r fig_note("lss_pathway")`

## 2.11 High-level state-transition framework

Health impact is projected with a **well–sick–dead** population model
(`r fig_link("state_transition")`). Each year, some well people develop disease at the
(now lower) incidence rate; some people living with disease die from it at the
case-fatality rate; and both groups are subject to background mortality from other
causes. The surviving population then ages by one year and a new cohort enters at age
20. The process runs independently for each of the four cardiovascular causes. Running
the model with and without each policy, and taking the difference in deaths, yields the
**deaths delayed**. The complete equations are in Appendix A.

`r fig_title("state_transition")`

```{r state-transition, out.width="100%"}
include_graphics(fig_path("state_transition"))
```

`r fig_note("state_transition")`

# 3. Results

## 3.1 Deaths delayed by intervention and country

`r tab_link("main")` reports deaths delayed by country for each packaged-food and
procurement policy, for the salt tax at low, base and high assumptions, and for the full
package at each salt-tax assumption. The country totals reconcile to the "Total" row
within rounding. `r fig_link("headline")` ranks the interventions overall, and
`r fig_link("countries")` shows how that impact concentrates across countries (Section
3.5).

`r tab_title("main")`

```{r main}
tr_tab("main")
```

`r tab_note("main")`

`r fig_title("headline")`

```{r headline, out.width="85%"}
include_graphics(fig_path("headline"))
```

`r fig_note("headline")`

## 3.2 Fiscal-policy variants and corresponding package totals

The salt tax is modelled as a pathway from tax rate to consumer price, demand response
and, finally, reduced packaged sodium (the pathway and its parameters are detailed in
Appendix C, `r tab_link("fiscal_path")`). It ranges from **`r N$fiscal_low_m` million**
deaths delayed under conservative assumptions to **`r N$fiscal_high_m` million** under
optimistic ones, with a base case of **`r N$fiscal_base_m` million**. Because deaths
delayed do not simply add across policies, the package total is re-computed for each tax
assumption; these are the three "Package total" columns of `r tab_link("main")`, ranging
from **`r N$pkg_low_m`** to **`r N$pkg_high_m` million**.

## 3.3 Low-sodium salt substitutes: uptake and hypertension targeting

`r tab_link("lss")` reports salt-substitute results: whole-population use (Scenario 2)
at population uptake levels of 10%, 20%, 30%, 40% and 50%, and two hypertension-targeted
scenarios, namely adults with **diagnosed** hypertension (Scenario 4) and adults with
**treated** hypertension (Scenario 5). Whole-population benefit scales almost linearly
with uptake (`r fig_link("lss_uptake")`), reaching
**`r N$lss_s2_hi_m` million** deaths delayed at `r N$lss_s2_hi_pct` uptake. The
hypertension-targeted scenarios deliver **`r N$lss_s4_m` million** (diagnosed) and
**`r N$lss_s5_m` million** (treated), a large benefit from a much smaller target group,
because risk is concentrated in people with high blood pressure. These figures
follow the mechanistic sodium-and-potassium pathway, not the salt-substitute trial's
relative risks.

`r tab_title("lss")`

```{r lss}
tr_tab("lss")
```

`r tab_note("lss")`

`r fig_title("lss_uptake")`

```{r lss-uptake, out.width="85%"}
include_graphics(fig_path("lss_uptake"))
```

`r fig_note("lss_uptake")`

## 3.4 Which interventions have the greatest impact

Across the ten countries, **mandatory salt targets** are the single most impactful
packaged-food policy, because they act on the whole packaged category at a substantial
reduction. **Front-of-pack labelling** is next. **Public food procurement** is the
smallest of the packaged-and-procurement policies, reflecting the modest share of sodium
that public institutions reach under the assumed 5% institutional share. The **salt tax**
falls between labelling and procurement. **Low-sodium salt substitutes** stand apart:
where discretionary salt dominates intake, their whole-population potential exceeds the
entire packaged-food package, a key strategic finding, subject to the benefit-only caveat
and to achieving real-world uptake and adherence.

## 3.5 Important cross-country differences

The ranking of policies flips with the local sodium profile. In China and Malaysia,
where packaged food is a large share of sodium, reformulation and labelling dominate. In
Ethiopia, India and Bangladesh, where discretionary salt dominates, salt substitutes
carry by far the most potential. Absolute impact, however, tracks population and
cardiovascular burden and is highly concentrated: ranked from largest to smallest, a
small number of the biggest countries, led by China and India, account for roughly 80%
of the ten-country full-package total, as the cumulative line of the Pareto chart shows
(`r fig_link("countries")`). This concentration is why a few large countries drive the
headline totals, even though the *best policy mix* is country-specific.

`r fig_title("countries")`

```{r countries, out.width="95%"}
include_graphics(fig_path("countries"))
```

`r fig_note("countries")`

## 3.6 Full-package findings

The full package delays about **`r N$pkg_m` million** cardiovascular deaths
(`r N$pkg_pct` of baseline), the great majority in the largest countries. As
`r tab_link("bycause")` shows, the package's benefit is led by **`r N$top_cause`**
(**`r N$top_cause_m` million**, `r N$top_cause_pct` of the package total), reflecting the
strong blood-pressure link of stroke. The full package excludes low-sodium salt
substitutes, which are reported separately; adding them where discretionary salt is large
would raise the total substantially.

`r tab_title("bycause")`

```{r bycause-tab}
tr_tab("bycause", "bycause-tab")
```

`r tab_note("bycause")`

## 3.7 Donor-relevant takeaways, without overstating precision

Sodium policy offers a large, credible cardiovascular return across these ten countries.
The estimates here are deliberately conservative in construction (each policy reaches
only the sodium it can) and they still point to millions of deaths delayed. They are
projections, not guarantees: the ranges shown for the salt tax and the salt-substitute
uptake sweep are the honest expression of that uncertainty.

# 4. Assumptions and limitations

The main assumptions and limitations are summarised in `r tab_link("limitations")` and
expanded below.

`r tab_title("limitations")`

```{r limitations}
tr_tab("limitations")
```

`r tab_note("limitations")`

- **Deaths delayed, not averted.** The sodium-to-blood-pressure policies lower incidence
  but not case fatality, so they postpone deaths rather than prevent them outright;
  totals depend on the window.
- **Salt substitutes are benefit-only.** No kidney-related harm of higher potassium is
  modelled, so salt-substitute benefits are not offset and should be read as an upper
  part of the plausible range where such risks matter.
- **The salt tax.** Its price, demand and substitution assumptions are uncertain, and
  broad-category demand responses may not transfer to high-sodium foods.
- **Public food procurement.** It is modelled using an assumed 5% institutional share of
  sodium.
- **Some inputs are held constant or use proxies.** Hypertension diagnosis and treatment
  shares are age-standardised and held after their source year; potassium intake for two
  countries uses a regional average; several source-share profiles use proxy or default
  values; and future packaged-food growth is an extrapolation.
- **Uncertainty is not fully propagated.** Results are point estimates; no full
  probabilistic sensitivity analysis is run.

# 5. Conclusion

Across ten priority countries, a modest set of packaged-food and procurement policies is
projected to delay on the order of **`r N$pkg_m` million** cardiovascular deaths over
`r N$window`, and low-sodium salt substitutes could deliver more still where
discretionary salt dominates intake. The most effective mix differs by country, which
argues for tailoring policy portfolios to each country's sodium profile rather than
applying a single template. With transparent assumptions clearly stated, these estimates
make a strong, evidence-based case for sustained investment in population sodium
reduction, while leaving room for the better data and fuller uncertainty analysis that
would sharpen the numbers.

# 6. References

::: {#refs}
:::

# Appendix A: Full mathematical specification

**Source split and total sodium reduction.** Let $w_{c,s,t}$ be the share of country
$c$'s sodium intake from source $s$ in year $t$ (with $\sum_s w_{c,s,t}=1$), and let
$e_{i,s}$ be the fractional reduction that policy $i$ applies to source $s$. For a set of
policies $\mathcal{P}$, the combined reduction of source $s$ is
\begin{equation}
r_{c,s,t}=1-\prod_{i\in\mathcal{P}}\left(1-e_{i,s}\right),(\#eq:compose)
\end{equation}
and the total sodium reduction fraction is
\begin{equation}
\eta_{c,t}=\sum_{s} w_{c,s,t}\, r_{c,s,t}.(\#eq:eta)
\end{equation}
Equation \@ref(eq:compose) guarantees that overlapping policies on one source never
exceed a 100% reduction.

**Sodium to blood pressure.** With baseline sodium intake $\text{Na}^0$ (g/day) and a
scaled-up reduction $\Delta\text{Na}=\eta\,\text{Na}^0$ (subject to a 2 g/day minimum
intake floor), the systolic blood-pressure reduction is
\begin{equation}
\Delta\text{SBP}=\left(2.8\,h+1.0\,(1-h)\right)\,\Delta\text{Na},(\#eq:sbp)
\end{equation}
where $h$ is the fraction of the age–sex group with raised blood pressure.

**Blood pressure to incidence.** Within each blood-pressure band $b$, baseline
incidence for cause $k$ is decomposed using relative-risk weights and then reduced in
proportion to $\Delta\text{SBP}$:
\begin{equation}
\text{IR}^{\text{new}}_{k}=\sum_b \pi_b\, \text{IR}_{b,k}\left(1-(1-RR^{10}_{b,k})\,\frac{\Delta\text{SBP}}{10}\right),(\#eq:ir)
\end{equation}
where $\pi_b$ is the share of the group in band $b$ and $RR^{10}_{b,k}$ the relative risk
per 10 mmHg. Case fatality is unchanged for these policies.

**Potassium channel (salt substitutes).** Switching discretionary salt to a substitute
adds potassium in proportion to the sodium displaced (about 1.33 g potassium per g
sodium). Baseline and post-switch potassium are converted to measured excretion and
mapped to a blood-pressure change through a dose-response curve; the resulting
$\Delta\text{SBP}_{K}$ is added to the sodium-mediated change and Equation \@ref(eq:ir)
is applied to the combined total (Appendix F).

**Population dynamics.** For each cause, with well ($W$), sick ($S$) and dead ($D$)
counts, incidence $\text{IR}$, case fatality $\text{CF}$ and background mortality $m$:
\begin{equation}
\begin{aligned}
S_t &= S_{t-1}(1-\text{CF}-m)+W_{t-1}\,\text{IR},\\
D_t &= S_{t-1}\,\text{CF},\\
W_t &= P_t - S_t - \sum_k D_{t}.
\end{aligned}(\#eq:states)
\end{equation}
**Deaths delayed** for a scenario are baseline deaths minus scenario deaths, summed over
the results window.

# Appendix B: Complete scenario definitions

`r tab_title("scn_def")`

```{r scn-def}
tr_tab("scn_def")
```

`r tab_note("scn_def")`

# Appendix C: Detailed intervention methods and effect-size evidence

Each effect size and its evidence source are summarised alongside the intervention in the
body table (`r tab_link("eff_brief")`). Modelled values follow the production
configuration and may differ from a single published study (for example, salt targets are
modelled at 20% against roughly 15% observed in the United Kingdom over nine years
[@he2013]). Public food procurement draws on institutional cafeteria studies [@gase2011];
front-of-pack warning labelling on the Chilean evaluation [@taillie2024]; and the salt tax
on Philippine tax modelling [@saxena2025] with support from Australian work [@cobiac2017].

The salt-tax pathway (`r tab_link("fiscal_path")`) does not assume that a 10% tax yields a
10% sodium reduction, nor that all packaged sodium is taxable. The salt-substitute
benchmark using the SSaSS trial's stroke relative risks [@Neal2021LSSTrial] is shown
alongside the mechanistic estimate for the hypertension-targeted scenarios in
`r tab_link("nak_ssass")`; the two are distinct pathways and are never combined.

`r tab_title("fiscal_path")`

```{r fiscal-path}
tr_tab("fiscal_path")
```

`r tab_note("fiscal_path")`

`r tab_title("nak_ssass")`

```{r nak-ssass}
tr_tab("nak_ssass")
```

`r tab_note("nak_ssass")`

# Appendix D: Sodium-source definitions and country assumptions

The five sodium sources are discretionary salt, packaged food, restaurant and vendor
food, public and institutional food, and inherent sodium. `r tab_link("source_shares")`
gives each country's base-year shares; where a country lacks its own profile, a proxy or
default average is used.

`r tab_title("source_shares")`

```{r source-shares}
tr_tab("source_shares")
```

`r tab_note("source_shares")`

# Appendix E: Population, disease, blood-pressure and mortality inputs

Hypertension diagnosis and treatment shares that size the salt-substitute
hypertension-targeted scenarios come from the NCD Risk Factor Collaboration
[@ncdrisc2021] (`r tab_link("ncd_elig")`); baseline potassium intake from a global
pooled analysis [@Reddin2023] (`r tab_link("baseline_k")`); and the packaged-food growth
trend from commercial sales data [@marklund2022sources] (`r tab_link("pkg_trend")`). Key
model inputs and parameters are summarised in `r tab_link("key_inputs")`.

`r tab_title("ncd_elig")`

```{r ncd-elig}
tr_tab("ncd_elig")
```

`r tab_note("ncd_elig")`

`r tab_title("baseline_k")`

```{r baseline-k}
tr_tab("baseline_k")
```

`r tab_note("baseline_k")`

`r tab_title("pkg_trend")`

```{r pkg-trend}
tr_tab("pkg_trend")
```

`r tab_note("pkg_trend")`

`r tab_title("key_inputs")`

```{r key-inputs}
tr_tab("key_inputs")
```

`r tab_note("key_inputs")`

# Appendix F: Low-sodium salt substitute parameters and pathway quantities

`r tab_link("lss_audit")` reports the sodium displaced, potassium added and
blood-pressure change per country and sex at a representative whole-population coverage;
these quantities scale linearly with coverage. `r tab_link("huang")` positions the
mechanistic salt-substitute model against the recent global study whose structure it
follows [@huang2026].

`r tab_title("lss_audit")`

```{r lss-audit}
tr_tab("lss_audit")
```

`r tab_note("lss_audit")`

`r tab_title("huang")`

```{r huang}
tr_tab("huang")
```

`r tab_note("huang")`

# Appendix G: Packaged-food trend method

The packaged share of sodium grows along a country-specific annual rate
(`r tab_link("pkg_trend")`), capped at 80% of sodium sources and held constant after 2030.
This is a **compositional reallocation**: each year the other four source shares are
rescaled proportionally so the five continue to sum to 100%. This guarantees that **total
baseline sodium intake does not grow** (only its source composition changes), so the trend
alters the reach of packaged-food policies over time rather than the underlying intake.

# Appendix H: Validation and consistency checks

The following checks were performed and passed for this run: the five sodium-source
shares sum to 100% in every country and year; baseline deaths delayed equal zero; the
whole-population salt-substitute benefit is non-decreasing from 10% to 50% uptake in
every country; hypertension-targeted scenarios use the intended country- and
sex-specific eligibility; salt-substitute results use the mechanistic
sodium-and-potassium pathway rather than the trial relative risks; country totals
reconcile to the reported totals within rounding; and the table and figure values in this
report match the companion workbook exactly. As an external cross-check, the full package
(`r N$bench_cur_m` million deaths delayed across the modelled countries) is compared with
a previous estimate for a uniform 30% sodium cut (`r N$bench_prev_m` million); the current
package is expectedly smaller because it reaches only the sodium its policies can affect
(`r tab_link("benchmark")`).

`r tab_title("benchmark")`

```{r benchmark}
tr_tab("benchmark")
```

`r tab_note("benchmark")`

# Appendix I: Supplemental tables and figures

`r tab_link("scn_summary")` gives cumulative deaths delayed by scenario across all
countries and causes, and `r tab_link("saxena")` positions the salt-tax model against the
study that anchors it [@saxena2025].

`r tab_title("scn_summary")`

```{r scn-summary}
tr_tab("scn_summary")
```

`r tab_note("scn_summary")`

`r tab_title("saxena")`

```{r saxena}
tr_tab("saxena")
```

`r tab_note("saxena")`
