---
title: "Health and Economic Returns to Dietary Sodium Reduction"
subtitle: "Technical review of the source-split CVD model and preliminary results for prioritized countries"
author: "Resolve to Save Lives (RTSL) and University of Washington"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  beamer_presentation:
    latex_engine: xelatex
    keep_tex: false
    slide_level: 2
    toc: true
    citation_package: biblatex
bibliography: references.bib
biblio-style: authoryear
classoption: "aspectratio=169"
header-includes:
  - \input{beamer_preamble.tex}
fontsize: 11pt
---

# Recap and today's plan

```{r setup, include=FALSE}
# SETUP — load deck artifacts only (zero recomputation). Every figure/table/
# number is produced in reports/report.RMD and handed over via docs/deck_assets/.
# This deck performs NO calculations of its own.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "100%")

deck_assets <- "deck_assets"
bundle      <- file.path(deck_assets, "deck_data.rds")
if (!file.exists(bundle))
  stop("deck_data.rds not found in docs/deck_assets/. Render reports/report.RMD ",
       "first — its 'deck-export' chunk generates every figure/table/number ",
       "this deck consumes.")

deck <- readRDS(bundle)
N    <- deck$nums                                  # pre-formatted numbers
fig  <- function(name) file.path(deck_assets, deck$figs[[name]])  # figure path
tbl  <- function(name) cat(deck$tex[[name]])       # emit pre-rendered LaTeX table
```

## Recap: where we left off (May 28 call)

\justifying
\footnotesize

From the previous presentation and discussion:

<!-- - **Modeling framework** --- multi-state CVD life-table (Well $\to$ Sick $\to$ Dead), simulated annually for ages 20--95, by age $\times$ sex $\times$ country, for four causes: IHD, ischemic stroke, hemorrhagic stroke, hypertensive heart disease. -->
<!-- - **Outcome metrics** --- deaths and cases delayed, DALYs averted, plus a full benefit--cost analysis (NPV, BCR). -->
<!-- - **Causal pathway** --- sodium $\to$ systolic BP (Filippini dose--response) $\to$ CVD incidence (Ettehad per-10-mmHg relative risks) $\to$ life-table transitions \autocite{Filippini2021,ettehad2016}. -->
<!-- - **Interventions** --- six policy levers scored in the Sodium Lives Saved Calculator (public procurement, media, low-sodium salt substitutes, front-of-pack labelling, salt targets, fiscal). -->
<!-- - **Deliverables** --- RTSL funders report (**Prioritized**), peer-reviewed paper, and interactive tool. -->

\justifying
\scriptsize

\begin{columns}[T]
\begin{column}{0.52\textwidth}
\textbf{Scope \& data}
\begin{itemize}
  \item Fixed to (64) \textbf{`r N$n_countries` priority countries}.
  \item Sodium sources: \textbf{combined approach} --- use country data where available, else the Bhat/Marklund GDP regression \autocite{marklund2022sources}. Drivers extend beyond GDP (condiments/bouillon in Viet Nam, Nigeria).
  \item \textbf{Do \emph{not} map interventions to SHAKE} --- no longer how the revised SHAKE frames them.
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Policies \& framing}
\begin{itemize}
  \item \textbf{Fiscal policy}: feasibility \& generalizability doubts; RTSL interested in salt taxes??
  \item \textbf{LSSS}: focus scenarios 4/5; include potassium.
  \item \textbf{Media}: revised to a UK basis; may be dropped.
  \item Goal is not only to \emph{maximize} lives saved (e.g.\ prevention framing for Ethiopia).
  \item Primary deliverable: \textbf{BP reinvestment proposal} (memo, tables + figures), \textbf{Sept 2026}.
\end{itemize}
\end{column}
\end{columns}

\medskip
\footnotesize Today we present the **current build** of the model and **preliminary results** for the prioritized countries.

## Today's agenda

\justifying
\footnotesize

\begin{enumerate}
  \item \textbf{Model inputs} --- updated sodium source shares by country (now incl.\ Nigeria/Cameroon) and the applied salt-effect $\eta$ for each intervention and the full package.
  \item \textbf{Methods} --- the source-split logic (sources $\to$ $\eta$), the state-transition engine, and the current low-sodium salt substitute (LSS) model.
  \item \textbf{Preliminary results} --- baseline burden, projected trajectories by scenario, and deaths delayed across the `r N$n_countries` prioritized countries.
  \item \textbf{Limitations.}
  \item \textbf{Discussion} --- pending decisions (fiscal in/out, LSS scenarios, source-share strategy, harmonizing observation periods) and the timeline to the September reinvestment deliverable.
\end{enumerate}

<!-- <!-- ====================================================================== -->
<!--      2. BACKGROUND -->
<!--      ====================================================================== --> -->

<!-- # Background -->

<!-- ## Project objective and policy question -->

<!-- \justifying -->
<!-- \footnotesize -->

<!-- - Estimate the **national-scale health and economic impact** of dietary sodium-reduction policies using a population CVD model. -->
<!-- - Quantify **deaths and cases delayed** (IHD, ischemic stroke, hemorrhagic stroke, hypertensive heart disease) and **DALYs averted** over `r N$scaleup1`--`r N$horizon_end`, with a full **benefit--cost analysis**. -->
<!-- - Focus on **RTSL partner countries** and on policies governments can realistically implement. -->

<!-- \medskip -->
<!-- \begin{block}{The question the model answers} -->
<!-- \footnotesize -->
<!-- How many cardiovascular deaths are \emph{delayed} by each sodium-reduction policy --- and by a coordinated \textbf{package} --- when each policy's documented effect is applied \textbf{only to the sodium sources it can actually reach}? -->
<!-- \end{block} -->

<!-- ## Causal pathway and scope -->

<!-- \justifying -->
<!-- \footnotesize -->

<!-- \textbf{Exposure $\to$ disease chain (unchanged):} -->

<!-- $$ -->
<!-- \text{Sodium} \;\longrightarrow\; \text{SBP} \;\longrightarrow\; \text{CVD incidence} \;\longrightarrow\; \text{Well}\to\text{Sick}\to\text{Dead} -->
<!-- $$ -->

<!-- - **SBP response** to sodium from the Filippini dose--response meta-analysis \autocite{Filippini2021}. -->
<!-- - **Incidence response** to SBP from the Ettehad per-10-mmHg relative risks \autocite{ettehad2016}, applied by BP bin and cause. -->
<!-- - **Four causes** modelled jointly (IHD, ischemic stroke, hemorrhagic stroke, HHD); **case fatality is held at baseline**, so the model shifts deaths later --- we report **deaths delayed**, not "averted". -->
<!-- - **Scope:** `r N$n_countries` prioritized countries; annual projection over **`r N$study_years`**; scale-up `r N$scaleup1`--`r N$scaleup2`. -->

<!-- ======================================================================
     3. MODEL INPUTS
     ====================================================================== -->

# Model inputs

## Model inputs at a glance

\justifying
\scriptsize

\begin{tabular}{p{3.4cm} p{8.6cm}}
\toprule
\textbf{Input} & \textbf{Role} \\
\midrule
Baseline transition rates (GBD) & Age $\times$ sex $\times$ year $\times$ cause incidence (IR), case fatality (CF), background \& COVID mortality, population ($N_x$). \\
BP distribution & Mean/SD systolic BP by age $\times$ sex, raised-BP fraction, diabetes prevalence. \\
Baseline sodium intake & 2024 mean intake, split into five sources. \\
\textbf{Sodium source shares} & Five-source split of baseline sodium per country (this section). \\
GBD relative risks & Log-linear RR per 10 mmHg SBP; sets bin-specific baseline incidence. \\
Ettehad relative risks & RR per 10 mmHg SBP reduction by BP bin $\times$ cause. \\
Scale-up schedule & Linear ramp `r N$scaleup1`$\to$`r N$scaleup2`, then held. \\
\bottomrule
\end{tabular}

\medskip
\footnotesize The source-split logic is the only change from the earlier single-pool build; everything downstream of $\eta$ is unchanged.

## Sodium source shares by country

\justifying
\footnotesize

Baseline sodium is partitioned into **five mutually exclusive sources** (shares sum to 1). `inherent` sodium is **never targeted**; each policy acts only on the source(s) it can reach.

```{r fig-source-shares, out.width="86%"}
knitr::include_graphics(fig("source_shares"))
```

\scriptsize Country data where available; otherwise predicted from GDP per capita \autocite{marklund2022sources}. **Nigeria/Cameroon** use provisional RTSL/Matti shares (bouillon + condiments counted as packaged; Nigeria a proxy for Cameroon).

## Applied salt-effect ($\eta$) by intervention

\justifying
\footnotesize

Each intervention has a documented **relative reduction** applied to its **target source**. 

<!-- Values match `code/07_run_interventions.R` exactly. -->

```{r tbl-effect-sizes, results='asis'}
tbl("effect_sizes")
```

\scriptsize
Media acts on discretionary **and** packaged (`r N$eff_media` each). FOPL, salt targets and fiscal all act on **packaged** and compose multiplicatively when combined: e.g.\ FOPL (`r N$eff_fopl`) + salt targets (`r N$eff_salttargets`) $\Rightarrow 1-(1-0.138)(1-0.20)=$ **`r N$compose_fopl_salt`**. **Fiscal is currently `r N$eff_fiscal` --- excluded pending the team's decision.**

## Illustrative $\eta$ by intervention and full package

\justifying
\footnotesize

`salteff` ($\eta$) is the **fraction of total intake** each scenario reduces at full implementation --- **country-specific**, because it depends on local source shares:
$$
\eta \;=\; \sum_{s}\, \text{share}_s \cdot \Bigl[\,1-\textstyle\prod_i (1-\text{eff}_{i,s})\Bigr].
$$

```{r tbl-salteff, results='asis'}
tbl("salteff")
```

\scriptsize $\eta$ is largest where the target source dominates local intake --- packaged policies in packaged-heavy Viet Nam; LSS in discretionary-heavy Bangladesh/China. The full-package row composes all active policies.

## Full-package $\eta$: composition by country

\justifying
\footnotesize

Bar height = each country's full-package `salteff`; segments show the per-source contributions ($\text{share}_s \times \text{composed}_s$).

```{r fig-salteff-pkg, out.width="82%"}
knitr::include_graphics(fig("salteff_pkg"))
```

<!-- ======================================================================
     4. METHODS
     ====================================================================== -->

# Methods

## The source-split approach: from sources to $\eta$

\justifying
\footnotesize

```{r fig-flow, out.width="100%"}
knitr::include_graphics(fig("flow"))
```

\medskip
- Split baseline sodium into five sources; apply each policy's reduction **only** to the source(s) it reaches.
- Effects on the **same** source compose multiplicatively --- $\text{composed}_s = 1-\prod_i(1-\text{eff}_{i,s})$ --- so overlapping packaged policies never double-count or exceed 100%.
- Share-weight and sum to a single total-intake fraction $\eta$ (`salteff`), which feeds the unchanged epidemiological pathway.

\scriptsize \textbf{Worked check (asserted in code):} public procurement in Viet Nam touches only the public source (share 0.05, effect `r N$eff_public`) $\Rightarrow \eta = 0.20\times0.05 =$ \textbf{`r N$vn_sanity`}.

<!-- ## From $\eta$ to health impact -->

<!-- \justifying -->
<!-- \scriptsize -->

<!-- Given a country's $\eta$, sodium falls and --- through blood pressure --- so does CVD **incidence** (case fatality is unchanged): -->

<!-- 1. **Target reduction:** $\text{salt\_target} = \text{salt}\cdot\eta$, subject to a **2 g/day minimum-intake floor**. -->
<!-- 2. **Linear scale-up** from `r N$scaleup1` to `r N$scaleup2`, then held. -->
<!-- 3. **Sodium $\to$ SBP (Filippini)** \autocite{Filippini2021}: $\;\Delta SBP = \bigl(2.8\cdot\text{raisedBP} + 1.0\cdot(1-\text{raisedBP})\bigr)\cdot \text{salt\_reduction}$. -->
<!-- 4. **Baseline bin-specific incidence** scaled by the GBD log-linear RR per 10 mmHg, normalised to recover the observed $IR$. -->
<!-- 5. **SBP $\to$ incidence (Ettehad)** \autocite{ettehad2016}: bin-specific RR per 10 mmHg drives the proportional incidence change; averaged over BP bins. -->
<!-- 6. **Case fatality unchanged** $\Rightarrow$ **deaths delayed**, not averted. -->

<!-- ## The Lives Saved model: state-transition engine -->

<!-- \justifying -->
<!-- \scriptsize -->

<!-- Each country/sex/cause/scenario tracks **Well / Sick / Dead**, advanced one year at a time (ages 20--95, `r N$study_years`). With last year's cohort values: -->

<!-- $$ -->
<!-- \begin{aligned} -->
<!-- \text{newcases}_t &= \text{well}_{t-1}\cdot IR \\[1pt] -->
<!-- \text{sick}_t     &= \text{sick}_{t-1}\bigl(1-(CF+\text{BG.mx}+\text{covid.mx})\bigr) + \text{well}_{t-1}\cdot IR \\[1pt] -->
<!-- \text{dead}_t     &= \text{sick}_{t-1}\cdot CF \\[1pt] -->
<!-- \text{pop}_t      &= \text{pop}_{t-1} - \text{all.mx}_{t-1}, \qquad -->
<!-- \text{well}_t = \text{pop}_t - \text{all.mx}_t - \text{sick}_t -->
<!-- \end{aligned} -->
<!-- $$ -->

<!-- - Seeded (2017) from GBD prevalence and mortality; every quantity floored at zero. -->
<!-- - The **only** intervention lever is $IR$: sodium reduction lowers incidence via steps 1--6; $CF$, background and COVID mortality are untouched. -->
<!-- - **Aggregation:** the four causes share one population, so `pop`/`all.mx` are identical across cause rows (denominator taken from a single cause); `sick`/`well`/`dead`/`newcases` are cause-specific and summed over causes. -->

## Low-sodium salt substitute (LSS) model --- current version

\justifying
\footnotesize

LSS is no longer a placeholder: it is derived from the **Salt Substitute and Stroke Study** (SSaSS) \autocite{Neal2021LSSTrial}.

- **Effect size:** a `r N$lss_ratio` substitute delivered an ITT net **`r N$ssass_na_mgday`** sodium reduction against a **`r N$ssass_base_gday`** baseline $\Rightarrow$ **`r N$eff_lss` of discretionary sodium** (sensitivity `r N$lss_sens`). Applied to the discretionary source.

\medskip
\begin{uncertaintyblock}{Two deliberately conservative choices}
\footnotesize
\begin{itemize}
  \item \textbf{Hypertensive-only reach.} As its own scenario, the reduction applies only to raised-BP bins (SBP $\geq$ 140); in the \emph{full package} it composes across all bins.
  \item \textbf{No potassium channel.} SSaSS also raised urinary potassium (`r N$ssass_k_mgday`), which drove much of the observed `r N$ssass_sbp_mmhg` SBP effect. With no K\textsuperscript{+} pathway, the model captures only the sodium-mediated benefit --- a \textbf{lower bound}.
\end{itemize}
\end{uncertaintyblock}

\scriptsize Per the May call: include potassium in future work; focus on discretionary scenarios 4/5 \autocite{huang2026}.

<!-- ======================================================================
     5. BASELINE AND PROJECTIONS
     ====================================================================== -->

# Baseline and projections

## Baseline CVD burden, `r N$study_years`

\justifying
\footnotesize

```{r fig-baseline, out.width="72%"}
knitr::include_graphics(fig("baseline_asmr"))
```

\scriptsize Age-standardised mortality (WHO standard, ages 20--95+) declines with forecasted background-mortality and case-fatality trends; absolute CVD deaths still rise with population growth and ageing. Solid = historical, dashed = projected.

## Projected incidence by scenario

\justifying
\footnotesize

```{r fig-incidence, out.width="100%"}
knitr::include_graphics(fig("incidence"))
```

\scriptsize Incident CVD cases per 1,000 (left) and indexed to `r N$scaleup1` = 100 (right). Scenarios coincide until scale-up begins, then fan out; the **full package** falls fastest. **Fiscal overlaps baseline** because its applied $\eta$ is currently `r N$eff_fiscal`.

## Projected deaths by scenario

\justifying
\footnotesize

```{r fig-deaths, out.width="100%"}
knitr::include_graphics(fig("deaths"))
```

\scriptsize CVD deaths per 1,000 (left) and indexed to `r N$scaleup1` = 100 (right). Because mortality responds to the **prior** year's incidence, the scenarios' divergence lags the incidence divergence by about a year.

<!-- ======================================================================
     6. MAIN RESULTS
     ====================================================================== -->

# Main results

## Deaths delayed by scenario, `r N$impact_window`

\justifying
\footnotesize

```{r fig-results, out.width="70%"}
knitr::include_graphics(fig("results_bar"))
```

\vspace{-0.3em}
\begin{exampleblock}{Headline}
\scriptsize
The \textbf{full package} delays `r N$pkg_m` million CVD deaths across `r N$n_countries` countries.
`r N$pkg_pct` of baseline CVD deaths in the window; $\sim$`r N$pkg_per_year_k` thousand/year.
\end{exampleblock}

## Deaths delayed by scenario --- detail

\justifying
\footnotesize

```{r tbl-results, results='asis'}
tbl("results")
```

\scriptsize
Among single policies, **LSS** delays the most (`r N$lss_m` M) --- it targets discretionary salt, which dominates intake in the most populous countries --- followed by **salt targets** (`r N$salt_targets_m` M), **FOPL** (`r N$fopl_m` M), **public procurement** (`r N$public_m` M) and **media** (`r N$media_m` M). **Fiscal** delays `r N$fiscal_m` M (applied $\eta$ = `r N$eff_fiscal`).

## Full package: deaths delayed by cause

\justifying
\footnotesize

```{r fig-bycause, out.width="82%"}
knitr::include_graphics(fig("bycause"))
```

\scriptsize Ischemic heart disease and stroke dominate the delayed deaths, tracking their share of the baseline CVD burden and their SBP-mediated relative risks.

## Full package: deaths delayed by country

\justifying
\footnotesize

```{r fig-countries, out.width="70%"}
knitr::include_graphics(fig("countries"))
```

\scriptsize Absolute counts are dominated by the most populous countries (China, India); per-capita impact depends on baseline burden and local source structure.

<!-- ======================================================================
     7. LIMITATIONS
     ====================================================================== -->

# Limitations

## Limitations and interpretation

\justifying
\footnotesize

\begin{itemize}
  \item \textbf{Deaths delayed, not averted.} Case fatality is held at baseline, so the intervention lowers incidence and shifts mortality later; totals depend on the accounting window.
  \item \textbf{Point estimates only.} The per-source reductions are deterministic --- no probabilistic sensitivity analysis yet.
  \item \textbf{LSS is a conservative lower bound} --- sodium arm only, no potassium channel; hypertensive-only reach as an individual scenario.
  \item \textbf{Fiscal policy currently excluded} pending the team's decision. (applied $\eta$ = `r N$eff_fiscal`) 
  \item \textbf{Provisional source shares.} Nigeria/Cameroon values are working estimates; several others rest on single-study or older survey data.
  \item \textbf{Heterogeneous observation periods.} Documented effects span 1--20 year windows and are not yet harmonized onto a common footing.
  \item \textbf{Independent-cause structure.} The four causes share one population but are modelled independently.
\end{itemize}

<!-- ======================================================================
     8. DISCUSSION AND NEXT STEPS
     ====================================================================== -->

# Discussion and next steps

## Pending decisions --- for discussion

\justifying
\scriptsize

\begin{alertblock}{Open items}
\begin{enumerate}
  \item \textbf{Fiscal policy} --- in or out? Currently $\eta=$ `r N$eff_fiscal`. Feasibility and the generalizability of the Saxena estimate \autocite{saxena2025} were both questioned.
  \item \textbf{LSS scenarios} --- confirm discretionary scenarios 4/5 (prevalent or treated??); add a potassium channel to capture the full SSaSS effect?
  \item \textbf{Observation-period harmonization} --- effects measured over 1--20 yr. Apply as end-state at full scale-up, linearly over the observed window, or via a country-tier heuristic?
  \item \textbf{Source-of-sodium strategy} --- how hard to push for country-specific shares vs.\ the Bhat/Marklund GDP regression \autocite{marklund2022sources}?.
  \item \textbf{Policy framing} --- lives-saved maximization vs.\ a prevention/counterfactual framing (e.g.\ Ethiopia).
  \item \textbf{Media / behavior change} --- keep or drop?
\end{enumerate}
\end{alertblock}

## Next steps and timeline

\justifying
\footnotesize

- **Lock the inputs** --- finalize source shares (incl.\ Nigeria/Cameroon), confirm effect sizes, and resolve the fiscal decision.
- **Add the economic layer** --- policy costing plus benefit--cost outputs (NPV, BCR) and DALYs.
- **Primary deliverable** --- the **BP reinvestment proposal** memo (tables + figures), target **September 2026**.
- **Paper** is secondary; coordinate with Matti on the Bhat/Marklund inputs to avoid scooping the forthcoming paper.

\medskip
\begin{block}{Discussion}
\footnotesize
Where should we push hardest before the reinvestment deliverable --- e.g data (source shares)?
\end{block}

<!-- ======================================================================
     APPENDIX — detailed per-intervention methods (verified vs. code)
     ====================================================================== -->

# Appendix

## Appendix: per-intervention methods

\justifying
\footnotesize

The following slides document each intervention's **evidence base** and the **applied $\eta$** used in the model. Applied values are produced by `reports/report.RMD` and mirror `code/07_run_interventions.R`; study effect sizes are cited from source.

\medskip
\scriptsize Also in this appendix: documented source-share values (Table M1) and the per-scenario $\eta$ decomposition by country.

## A. Public food procurement

\footnotesize

**Target source:** public (institutional meals). **Applied $\eta$: `r N$eff_public` of the public source.**

- \textbf{Evidence:} \textcite{gase2011} --- modeling study, LA County. Weighted **--233 mg/day** ($\approx$7% of a 3,329 mg baseline), i.e.\ $\approx$20.6% of public-meal sodium.
- \textbf{Source adjustment:} for a country with share $x$ of sodium from institutional meals, population impact $= 0.20 \times x$. Worked: Viet Nam $= 0.20\times0.05 =$ `r N$vn_sanity`.
- \textbf{Caveats:} modeled (not observed); US-specific setting; small affected population share.

## B. Behavior change / mass media

\footnotesize

**Target source:** discretionary + packaged. **Applied $\eta$: `r N$eff_media` to each.**

- \textbf{Evidence:} \textcite{do2016} --- COMBI Vietnam, **5%** fall in mean sodium after a 1-year multi-component campaign. \textcite{collins2014} --- UK economic evaluation, **2%** [1--5%].
- The estimate was revised toward a UK basis; a conservative `r N$eff_media` is applied to each of discretionary and packaged.

\medskip
\begin{alertblock}{Open question}
\footnotesize
RTSL funds \textbf{policy change, not mass media for behavior change} (Laura). Keep this lever in the model, or drop it?
\end{alertblock}

## C. Low-sodium salt substitutes (LSS)

\footnotesize

**Target source:** discretionary salt. **Applied $\eta$: `r N$eff_lss` of discretionary** (sensitivity `r N$lss_sens`).

- \textbf{Evidence:} \textcite{Neal2021LSSTrial} (SSaSS) --- `r N$ssass_na_mgday` ITT sodium reduction vs.\ a `r N$ssass_base_gday` baseline; observed `r N$ssass_sbp_mmhg` SBP (much of it potassium-driven, `r N$ssass_k_mgday`).
- \textbf{Modelling:} sodium arm only (no K\textsuperscript{+} channel) $\Rightarrow$ conservative lower bound; hypertensive-only reach as an individual scenario.
- \textbf{Scenarios} \autocite{huang2026}: focus on discretionary-only 2 (all), 4 (diagnosed HTN), 5 (treated HTN).

## D. Front-of-pack labeling

\footnotesize

**Target source:** packaged (estimate already from packaged-food purchases). **Applied $\eta$: `r N$eff_fopl` of packaged.**

- \textbf{Evidence:} \textcite{taillie2024} --- Chile's mandatory warning-label law (observed). **--73.3 mg/capita/day (--13.8%)** in sodium purchases (2,844 households, 2013--2019).
- \textbf{Caveat:} purchases-to-intake translation is uncertain.
- \textbf{Alternative:} Mexico found $\approx$ --7% in packaged sales ($\approx$ --38 mg/capita/day).

## E. Mandatory salt targets (reformulation)

\footnotesize

**Target source:** packaged. **Applied $\eta$: `r N$eff_salttargets` of packaged** (conservative; 15% floor for sensitivity).

- \textbf{Policy:} South African mandatory maximum sodium limits (2013; targets implemented 2016 and 2019).
- \textbf{Evidence:} \textcite{strausskruger2022} --- African-PREDICT, **--15.2%** sodium; baseline/follow-up straddle the target dates, so likely an **underestimate**.
- \textbf{Alternatives:} \autocite{gaziano2025} --8.4% (rural, spot urine); \autocite{charlton2021} --18.4%; UK \autocite{he2013,song2023}; Mexico modeling \autocite{mucinosandoval2025}.

## F. Fiscal policy / taxation

\footnotesize

**Target source:** packaged. **Applied $\eta$: `r N$eff_fiscal` --- currently excluded (see below).**

- \textbf{Evidence:} \textcite{saxena2025} --- Philippines, modeled 20% tax on foods high in salt/sugar; **--100 mg/day (--10%)** at year 20.
- \textbf{Why excluded (for now):} political feasibility of food-category taxes; generalizability of a country-specific modeled estimate; RTSL not interested in salt taxes (May call).
- \textbf{Alternatives:} India \autocite{roche2026}; Australia \autocite{davies2026,cobiac2017}; systematic review \autocite{dodd2020} (1--11% reduction).

## Appendix: sodium source shares --- documented values

\footnotesize

```{r tbl-source-shares, results='asis'}
tbl("source_shares")
```

\scriptsize Fractions of total intake, renormalised to 1.0. The `default` row (mean of the ten) applies to any country without its own row.

## Appendix: $\eta$ decomposition by scenario and country

\footnotesize

```{r fig-salteff, out.width="88%"}
knitr::include_graphics(fig("salteff"))
```

\scriptsize Per-scenario `salteff` for every country, stacked by source contribution (note free y-axes). Fiscal is empty because its applied $\eta$ is currently `r N$eff_fiscal`.

<!-- ======================================================================
     REFERENCES  (biblatex auto-prints the bibliography after this section)
     ====================================================================== -->

# References

## References

\tiny
