---
title: "Health and Economic Returns to Dietary Sodium Reduction"
subtitle: "A national-scale cardiovascular disease modeling and benefit--cost analysis"
author: "Resolve to Save Lives (RTSL) -- UW"
date: ""
output:
  beamer_presentation:
    latex_engine: xelatex
    keep_tex: false
    slide_level: 2
    toc: false
classoption: "aspectratio=169"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{ragged2e}
  - \usetheme{default}
  - \usecolortheme{default}
  - \setlength{\emergencystretch}{2em}
  - \AtBeginSection[]{}
  # --- Add slide number / total slides (no external package needed) ---
  - \setbeamertemplate{footline}[frame number]
  - \setbeamertemplate{navigation symbols}{}
fontsize: 11pt
---

<!-- ======================================================================
     SETUP — LOAD ARTEFACTS ONLY (zero recomputation)
     ====================================================================== -->
     
```{r setup, include=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

<!-- ======================================================================
     Overview
     ====================================================================== -->

# Study Overview

## Project Objective and Policy Impact

\justifying

- Estimate national-scale health impacts of dietary sodium reduction policies using a population CVD model.
- Quantify deaths and cases of ischemic heart disease (IHD), stroke, and hypertensive heart disease, as well as DALYs averted over 2026--2050.
- Conduct a full economic benefit--cost analysis to support government decision-making.
- Focus on RTSL partner countries and on policies that governments can realistically implement.


## RTSL Sodium Reduction Policy Support

\justifying

- Technical assistance to governments to design and implement mandatory sodium targets for packaged foods.
- Support for front-of-pack sodium warning labels to inform consumers and incentivize reformulation.
- Development of nutrition standards for public food procurement in schools, hospitals, and government settings.
- Assistance with policy drafting, legal design, monitoring systems, and enforcement capacity.


<!-- ======================================================================
     SCENARIOS
     ====================================================================== -->

# Scenarios

## WHO SHAKE Strategy: Policy Framework

\justifying

- WHO SHAKE provides a comprehensive framework for population-wide sodium reduction.
- Components include:
  - **S**urveillance
  - **H**arness Industry (food reformulation)
  - **A**dopt Standards for labeling and marketing
  - **K**nowledge (consumer information)
  - **E**nvironments (healthier food settings)
- The largest and most sustained sodium reductions occur when SHAKE components are implemented as a package.
- This project explicitly models and costs realistic SHAKE policy bundles.

## Sodium and Potassium: Complementary Strategies for BP Control

- Sodium reduction policies are intrinsically linked to increasing potassium intake, as both nutrients play complementary roles in lowering BP and reducing CVD risk.

- While excess sodium raises BP, potassium promotes vasodilation and enhances urinary sodium excretion.

**Key Aspects of the Relationship**

- **Synergistic impact**: Lower sodium combined with higher potassium intake produces larger reductions in BP and stroke risk than sodium reduction alone.
- **Public health strategy**: Many national policies incorporate potassium-enriched salt substitutes (e.g., 75% NaCl / 25% KCl) to simultaneously reduce sodium and address widespread potassium deficiencies.
- **Physiological balancing**: Potassium counteracts sodium’s adverse effects, lowering hypertension risk.
- **Targeted goals**: guidelines emphasize a dual approach: reducing sodium intake and increasing daily potassium consumption.

**Common Policy Approaches**

- Promotion of whole, minimally processed foods (fruits, vegetables, legumes) that are naturally high in potassium and low in sodium.

# Data Sources

## Data Sources and Key Variables

```{r data-sources-table}
ds <- data.table(
  Source    = c("GBD 2023 (IHME)","GBD 2019 (IHME)","Ettehad et al.\\ 2016",
                "UNWPP 2024 (UN)","NCD-RisC","Country-specific"),
  Variables = c(
    "IR, CF, prevalence, background mortality (age/sex/country)",
    "Relative risks per 10 mmHg SBP increase (IHD, stroke, HHD, by age)",
    "Trial effect sizes per BP category; diabetes-stratified",
    "Population projections 2019--2050 (single-year age, sex, country)",
    "Baseline HTN control rates ($C_0$); BP distribution by country",
    "HTN control targets 2030 (Progress / Ambitious / Aspirational)"
  ),
  Role = c(
    "Baseline transition rates; initial states",
    "Anchor BP-bin incidence; normalisation factor $\\alpha$",
    "Treatment effect sizes by BP bin $\\times$ cause",
    "Cohort sizes; rate denominators",
    "$C_0$; $P(\\text{BP}_{\\text{cat}})$",
    "$C_{\\text{target}}$ per scenario"
  )
)
kable(ds, align="lll", escape=FALSE, booktabs=TRUE, format="latex",
      col.names=c("Source","Variables extracted","Role in model")) |>
  kable_styling(latex_options=c("striped","hold_position"),
                font_size=8, full_width=TRUE) |>
  column_spec(1, bold=TRUE, width="2.8cm") |>
  column_spec(2, width="6.8cm") |>
  column_spec(3, width="3.4cm")
```


# Mathematical Model

## Model Architecture: Three Health States

```{r state-diagram, fig.height=2.7}
nodes <- data.table(
  x=c(1,3.2,5.4), y=c(1,1,1),
  label=c("Well\n(W)","Sick\n(K)","Dead\n(D)"),
  fill=c("#d4edda","#fff3cd","#f8d7da")
)
arr_main <- data.table(
  x1=c(1.40,3.60), y1=c(1,1), x2=c(2.80,5.00), y2=c(1,1),
  lbl=c("IR \u00d7 \u03b5_IR","CF \u00d7 \u03b5_CF")
)
arr_bg <- data.table(
  x1=c(1.0,3.2,3.0), y1=c(1.25,1.25,1.0),
  x2=c(1.0,3.2,5.0), y2=c(0.62,0.62,0.62),
  lbl=c("BG.mx","BG.mx","BG.mx.all")
)
ggplot() +
  geom_tile(data=nodes, aes(x=x,y=y,fill=fill),
            width=0.75,height=0.42,color="grey30",linewidth=0.8) +
  scale_fill_identity() +
  geom_text(data=nodes, aes(x=x,y=y,label=label), size=4, fontface="bold") +
  geom_segment(data=arr_main,
    aes(x=x1,y=y1,xend=x2,yend=y2),
    arrow=arrow(length=unit(0.22,"cm"),type="closed"),
    linewidth=1.2, color="#2171b5") +
  geom_label(data=arr_main,
    aes(x=(x1+x2)/2, y=y1+0.22, label=lbl),
    size=3.2, fill="white", label.padding=unit(0.12,"cm"), color="#2171b5") +
  geom_segment(data=arr_bg,
    aes(x=x1,y=y1,xend=x2,yend=y2),
    arrow=arrow(length=unit(0.16,"cm"),type="closed"),
    linewidth=0.7, color="grey55", linetype="dashed") +
  geom_label(data=arr_bg,
    aes(x=(x1+x2)/2+0.2, y=(y1+y2)/2, label=lbl),
    size=2.6, fill="white", color="grey50", label.padding=unit(0.08,"cm")) +
  scale_x_continuous(limits=c(0.5,6.0)) +
  scale_y_continuous(limits=c(0.28,1.58)) +
  theme_void(base_size=11) +
  labs(title="Multi-State Transition Model  |  Annual time step  |  Ages 20-95  |  4 CVD causes")
```

\small \textbf{Key:} $\varepsilon_{IR}$, $\varepsilon_{CF}$ = multiplicative effect ratios from the intervention module $\cdot$ BG.mx = background mortality $\cdot$ COVID-19 excess mortality applied 2020--2021 only

## CVD Modeling Engine

\justifying

- Population cohort model linking sodium intake to blood pressure and cardiovascular outcomes.
- Uses country-specific baseline CVD incidence, mortality, and population age structure.
- Policy scenarios translate sodium reductions into blood pressure changes using established dose--response relationships.
- Resulting changes in heart attacks, strokes, and CVD deaths are projected over time.


## Modeling Assumptions

\begin{columns}[T]
\begin{column}{0.47\textwidth}
\textbf{Model structure}
\begin{itemize}\footnotesize
  \item Discrete-time annual step (2019--2050); ages 20--95
  \item Four causes modeled independently
  \item COVID-19 excess mortality applied 2020--2021 only; no long-term effects
\end{itemize}

\medskip
\textbf{Intervention mechanics}
\begin{itemize}\footnotesize
  \item Linear scale-up 2026 $\to$ 2030; held constant 2030--2050
\end{itemize}
\end{column}


# Economic Evaluation

## Health and Economic Benefits Estimation

\justifying

- Reduced sodium intake leads to fewer cardiovascular events and premature deaths.
- Health outcomes are summarized as deaths and DALYs averted.
- Health gains are monetized using the standard economic valuation of mortality risk reduction.
- Averted health-care expenditures are estimated and included as additional economic benefits.

## Costing the SHAKE Policy Package

\justifying

- Costs are estimated using an ingredients-based regulatory costing approach.
- Government costs include policy design, administration, monitoring, and enforcement.
- Industry costs include reformulation, testing, and labeling compliance, counted as real resource costs.
- Both start-up and ongoing recurrent costs are explicitly modeled over time.

## Costing the SHAKE Policy Package (Example)

\scriptsize

\begin{tabular}{p{2.2cm} p{3.1cm} p{1.7cm} p{1.6cm} p{1.8cm} p{2.3cm}}
\toprule
\textbf{SHAKE Component} & \textbf{Activity} & \textbf{Cost Category} & \textbf{Frequency} & \textbf{Coverage} & \textbf{Cost Bearer} \\
\midrule
S -- Surveillance & Population sodium survey & Program & Every 3--5 yrs & National & Government \\
S -- Surveillance & Food sodium monitoring & Program & Annual & Processed foods & Government \\
H -- Harness Industry & Set reformulation targets & Program & One-off & Key food categories & Government \\
H -- Harness Industry & Product reformulation & Compliance & One-off + transition & Affected SKUs & Industry \\
A -- Labeling & Front-of-pack labeling policy & Program & One-off & National & Government \\
A -- Labeling & Label changes & Compliance & One-off & Affected SKUs & Industry \\
K -- Knowledge & Mass media campaign & Program & Annual & Population reached & Government \\
E -- Environment & School meal standards & Program & Annual & Schools & Government \\
E -- Environment & Institutional food reformulation & Compliance & One-off + annual & Institutions & Vendors \\
\bottomrule
\end{tabular}

## Full Benefit--Cost Analysis Outputs

\justifying

- Health benefits and implementation costs are combined to estimate net present value.
- Benefit--cost ratios and return on investment are calculated for each policy scenario.
- Results allow comparison across individual SHAKE components and bundled strategies.
- Sensitivity analyses assess robustness to key assumptions and uncertainties.


# Decision Tool

## Integrated Web-Based Decision Tool

\justifying

- All modeling and economic components are integrated into an interactive \texttt{R Shiny} web application.
- Users can customize country-specific inputs, policy coverage, and implementation timelines.
- The tool enables real-time comparison of alternative policy scenarios and assumptions.
- Designed to support evidence-based policymaking and stakeholder engagement.


# Discussion

## Why This Analysis Matters for RTSL

\justifying

- Provides finance-ministry--ready evidence on the economic value of sodium reduction policies.
- Strengthens regulatory impact assessments and national investment cases.
- Supports prioritization of high-return noncommunicable disease prevention strategies.
- Scalable framework applicable across multiple countries and policy contexts.

## Next Steps

\justifying

- Finalize country-specific data inputs and validate assumptions with national stakeholders.
- Complete deployment and testing of the interactive web tool.
- Use results to support policy adoption, financing, and sustained implementation.
- Position sodium reduction as a high-return national investment in partner countries.



<!-- ======================================================================
     APPENDIX
     ====================================================================== -->

# Appendix

## Notation: Core Model

\footnotesize

Indices: age $a$, sex $s$, cause $c$, location $\ell$, year $t$, BP category $b$.

\medskip
\begin{tabular}{ll}
\toprule
\textbf{Symbol} & \textbf{Meaning} \\
\midrule
$IR_{a,s,c,\ell,t}$ & Baseline cause-specific incidence rate \\
$CF_{a,s,c,\ell,t}$ & Baseline case fatality rate \\
$BG.mx_{a,s,c,\ell,t}$ & Background mortality (disease state) \\
$BG.mx.all_{a,s,\ell,t}$ & All-cause background mortality \\
$PREVt0 / DIS.mx.t0$ & Baseline prevalence / disease mortality \\
$N_{a,s,\ell,t}$ & Population size \\
$P(b \mid a,s,\ell,t)$ & Probability of BP category $b$ \\
$S_{a,s,\ell,t}$ & Baseline mean sodium intake (g/day) \\
$RBP_{a,s,\ell,t}$ & Share of population with raised BP \\
$D_{a,s,b,\ell,t}$ & Diabetes prevalence \\
\bottomrule
\end{tabular}

## Policy Package: Efficacy

\footnotesize

Five policy components; active set $\mathcal{P}$ selected per scenario.

\medskip
\begin{tabular}{lll}
\toprule
\textbf{Component} & \textbf{Efficacy $e_p$} & \textbf{Note} \\
\midrule
Reformulation      & $r \times q_{\text{pkgd}}$          & $r=0.20$ (mandatory), $0.15$ (voluntary) \\
Package labelling  & $0.10 \times q_{\text{pkgd}}$       & \\
Media campaigns    & $0.10$                              & whole population \\
Supportive env.    & $0.20 \times q_{\text{outside}}$    & \\
Other              & $q_{\text{other}}$                  & optional \\
\bottomrule
\end{tabular}

\medskip
Total package efficacy (passed as \texttt{salteff}):
$$
\eta = \sum_{p \in \mathcal{P}} e_p
$$

## Policy Package: Programme Costs

\footnotesize

Per-capita annual cost: $u = \sum_{p \in \mathcal{P}} u_p$.

Scale-up from implementation start $t_0$ to full coverage $t_1$:

$$
Cost_{\ell,t} =
\begin{cases}
0 & t < t_0 \\[4pt]
\dfrac{Pop_{\ell,t}}{4} \times u \times Ex \times \dfrac{t - t_0 + 1}{t_1 - t_0 + 1} & t_0 \le t < t_1 \\[8pt]
\dfrac{Pop_{\ell,t}}{4} \times u \times Ex & t \ge t_1
\end{cases}
$$

$Ex$ = optional exchange-rate multiplier. Population divided by 4 = quarterly scaling convention.

## Sodium Intake Reduction

\footnotesize

Three reduction methods; all subject to a 2 g/day minimum intake floor.

\medskip
\textbf{Percent reduction} (main scenario runner):
$$\Delta S^{*}_{a,s,\ell,t} = S_{a,s,\ell,t} \times \eta$$

\textbf{Target-based} (absolute gram target):
$$\Delta S^{*}_{a,s,\ell,t} = \min(S_{a,s,\ell,t},\; \eta)$$

\textbf{Fixed absolute} (app-based):
$$\Delta S^{*}_{a,s,\ell,t} = \max(0,\; S_{a,s,\ell,t} - \eta)$$

\textbf{Floor adjustment}:
$$
\Delta S_{a,s,\ell,t} =
\begin{cases}
S_{a,s,\ell,t} - 2 & \text{if } S - \Delta S^{*} < 2 \\
\Delta S^{*}_{a,s,\ell,t} & \text{otherwise}
\end{cases}
$$

## Scale-Up Time Path

\footnotesize

Linear interpolation between implementation start ($T_{\text{start}}$) and full implementation ($T_{\text{full}}$):

$$
\Delta S^{\text{scale}}_{a,s,\ell,t} =
\begin{cases}
0 & t < T_{\text{start}} \\[6pt]
\Delta S_{a,s,\ell,t} \times \dfrac{t - T_{\text{start}} + 1}{T_{\text{full}} - T_{\text{start}} + 1} & T_{\text{start}} \le t \le T_{\text{full}} \\[8pt]
\Delta S_{a,s,\ell,t} & t > T_{\text{full}}
\end{cases}
$$

\medskip
Default scenario parameters: $T_{\text{start}} = 2026$, $T_{\text{full}} = 2030$.

## Sodium $\to$ Systolic Blood Pressure Reduction

\footnotesize

Filippini-style dose--response, differentiating raised- and normal-BP subgroups:

$$
\Delta SBP_{a,s,\ell,t}
= \Bigl(2.8 \times RBP_{a,s,\ell,t} \;+\; 1.0 \times (1-RBP_{a,s,\ell,t})\Bigr)
  \times \Delta S^{\text{scale}}_{a,s,\ell,t}
$$

\medskip
\begin{tabular}{ll}
\toprule
\textbf{Subgroup} & \textbf{SBP reduction per g/day Na removed} \\
\midrule
Raised BP ($RBP=1$) & 2.8 mmHg \\
Normal BP ($RBP=0$) & 1.0 mmHg \\
\bottomrule
\end{tabular}

## Baseline Incidence by BP Category

\footnotesize

\textbf{Step 1 -- GBD relative risk by bin.}
Let $m_b$ be the midpoint SBP of bin $b$ (reference: $<$120 mmHg $\to$ midpoint 110):
$$
RR^{GBD}_{b,a,c} = \bigl(RR^{10}_{a,c}\bigr)^{(m_b - 120)/10}, \quad m_b > 120; \quad 1 \text{ otherwise}
$$

\textbf{Step 2 -- Normalisation factor} (ensures consistency with aggregate $IR$):
$$
\alpha_{a,s,c,\ell,t} = \sum_b P(b \mid a,s,\ell,t) \times RR^{GBD}_{b,a,c}
$$

\textbf{Step 3 -- Bin-specific baseline incidence:}
$$
IR^{bin}_{b,a,s,c,\ell,t} = \frac{RR^{GBD}_{b,a,c} \times IR_{a,s,c,\ell,t}}{\alpha_{a,s,c,\ell,t}}
$$

## ETIHAD Effect Sizes

\footnotesize

ETIHAD (Ettehad et al.\ 2016) provides per-10-mmHg RR reductions by BP category and cause from randomised trial meta-analysis.

\medskip
\textbf{Diabetes-weighted cumulative effect} for bin $b$, cause $c$:

$$
E_{b,c,a,s,\ell,t}
= (1 - D_{a,s,b,\ell,t})\,E^{ND}_{b,c}
  \;+\; D_{a,s,b,\ell,t}\,E^{D}_{b,c}
$$

where $E^{ND}_{b,c}$ and $E^{D}_{b,c}$ are the no-diabetes and diabetes-specific cumulative effect sizes tabulated from ETIHAD.

\medskip
$\Rightarrow$ implemented in \texttt{calculate\_etihad\_cumulative\_rr()}.

## Effect of Sodium Reduction on Incidence

\footnotesize

\textbf{Step 1 -- Sodium-induced effect per bin.}
Scale the per-10-mmHg ETIHAD effect proportionally to achieved SBP reduction:

$$
E^{\text{sodium}}_{b,a,s,c,\ell,t}
= \bigl(1 - RR^{ETIHAD}_{b,c}\bigr) \times 0.1 \times \Delta SBP_{a,s,\ell,t}
$$

\textbf{Step 2 -- New bin-specific incidence:}
$$
IR^{bin,new}_{b,a,s,c,\ell,t} = IR^{bin}_{b,a,s,c,\ell,t} \times \bigl(1 - E^{\text{sodium}}_{b,a,s,c,\ell,t}\bigr)
$$

\textbf{Step 3 -- Population-weighted new incidence:}
$$
IR^{new}_{a,s,c,\ell,t} = \sum_b IR^{bin,new}_{b,a,s,c,\ell,t} \times P(b \mid a,s,\ell,t)
$$

Incidence effect ratio carried into the life table:
$\;\varepsilon^{IR}_{a,s,c,\ell,t} = IR^{new} / IR$

## Case Fatality Assumption

\footnotesize

The sodium policy model does \textbf{not} impose a secondary effect on case fatality (unlike antihypertensive treatment models).

$$
CF^{new}_{a,s,c,\ell,t} = CF_{a,s,c,\ell,t}
\qquad \Rightarrow \qquad
\varepsilon^{CF}_{a,s,c,\ell,t} = 1
$$

\medskip
\textbf{Rationale}: Sodium reduction acts entirely through lower incidence of new events. Trial evidence for a direct sodium-pathway effect on survival of existing disease is insufficient to parameterize a case-fatality reduction.

## State-Transition Equations

\footnotesize

Annual transitions (age $a \to a+1$, year $t \to t+1$):

$$
NewCases_{a+1,t+1} = Well_{a,t} \times IR_{a,t}
$$

$$
Sick_{a+1,t+1} = Sick_{a,t}\bigl(1 - CF_{a,t} - BG.mx_{a,t} - covid.mx_{a,t}\bigr) + Well_{a,t} \times IR_{a,t}
$$

$$
Dead_{a+1,t+1} = Sick_{a,t} \times CF_{a,t}
$$

$$
Pop_{a+1,t+1} = Pop_{a,t} - all.mx_{a,t}
$$

$$
Well_{a+1,t+1} = Pop_{a+1,t+1} - all.mx_{a+1,t+1} - Sick_{a+1,t+1}
$$

\medskip
\textbf{Initial conditions} ($t=t_0$ or $a=20$):
$\;Sick_0 = N \cdot PREVt0;\; Dead_0 = N \cdot DIS.mx.t0;\; Well_0 = N(1-PREVt0-BG.mx)$

## Scenario Definitions and Default Efficacies

\footnotesize

Five scenarios; $q_{\text{pkgd}} = q_{\text{outside}} = 30\%$; $u_p = \$0.05$ per component; scale-up 2026--2030.

\medskip
\begin{tabular}{lcc}
\toprule
\textbf{Scenario} & \textbf{Active policies} & \textbf{$\eta$ (salteff)} \\
\midrule
Baseline                  & none             & 0 \\
Reformulation -- mandatory & reform only      & $0.20 \times 0.30 = 0.060$ \\
Reformulation -- voluntary & reform only      & $0.15 \times 0.30 = 0.045$ \\
Full package -- mandatory  & reform + label + media + support & $0.060+0.030+0.100+0.060 = 0.250$ \\
Full package -- voluntary  & reform + label + media + support & $0.045+0.030+0.100+0.060 = 0.235$ \\
\bottomrule
\end{tabular}

\medskip
Per-capita annual costs: \$0.05 (reformulation-only) to \$0.20 (full 4-component package).

