---
title: "Health and Economic Returns to Dietary Sodium Reduction"
subtitle: "A national-scale cardiovascular disease modeling and benefit--cost analysis"
author: "Resolve to Save Lives (RTSL) -- UW"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  beamer_presentation:
    latex_engine: xelatex
    keep_tex: false
    slide_level: 2
    toc: true
classoption: "aspectratio=169"
header-includes:
  - \input{beamer_preamble.tex}
fontsize: 11pt
---

<!-- ======================================================================
     SETUP — LOAD ARTEFACTS ONLY (zero recomputation)
     ====================================================================== -->

```{r setup, include=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

<!-- ======================================================================
     1. Study Overview
     ====================================================================== -->

# Study Overview

## Project Objective and Policy Impact

\justifying

- Estimate national-scale health impacts of dietary sodium reduction policies using a population CVD model.
- Quantify deaths and cases of ischemic heart disease (IHD), stroke, and hypertensive heart disease, as well as DALYs averted over 2026--2050.
- Conduct a full economic benefit--cost analysis to support government decision-making.
- Focus on RTSL partner countries and on policies that governments can realistically implement.

<!-- ======================================================================
     2. Recap of Prior Presentation
     ====================================================================== -->

# Recap of Prior Presentation

## Where We Left Off

\justifying
\footnotesize

- **Modeling framework**: Multi-state CVD life-table (Well $\to$ Sick $\to$ Dead) simulated annually over 2026--2050 for ages 20--95, four CVD causes (IHD, stroke, HHD, other CVD), by age $\times$ sex $\times$ country.
- **Outcome metrics**: Deaths averted, IHD/stroke/HHD cases averted, DALYs averted, plus full benefit--cost analysis (NPV, BCR).
- **Baseline scenario design**: Five SHAKE-aligned policy bundles with total sodium-reduction efficacy \texttt{salteff} ($\eta$) ranging from 0 (baseline) to 0.250 (full mandatory package).
- **Causal pathway**: Sodium $\to$ SBP (Filippini dose--response) $\to$ CVD incidence (Ettehad per-10-mmHg RRs) $\to$ life-table transitions.
- **Deliverables**: Peer-reviewed paper + RTSL report + interactive Shiny tool, originally targeted for end of June 2026 (per Apr 24 timeline).

## Original Timeline (Apr 24 Email)

\justifying
\footnotesize

Andrew Moran circulated the following milestones on April 24, 2026:

\medskip
\begin{tabular}{ll}
\toprule
\textbf{Date} & \textbf{Milestone} \\
\midrule
May 11  & Zina shares preliminary policy/effect-size table with William \\
May 15  & William circulates inputs to full working group \\
May 28  & Group call: review inputs, manuscript outline, shell tables/figures \\
Late June & Review tables/figures, assign writing, pull in RTSL Comms/Design \\
End of June & Report + paper completed; paper submitted to journal \\
\bottomrule
\end{tabular}

\medskip
\textbf{Status}: Timeline has slipped \textbf{+1--2 months} to accommodate low-sodium salt substitutes (LSSS) integration. New target date to be discussed.

<!-- ======================================================================
     3. Email Thread: What's Changed Since April
     ====================================================================== -->

# Email Thread: What's Changed Since April

## LSS Scope Decision

\justifying
\footnotesize

Chronology of the key scope change:

\medskip
\begin{enumerate}
  \item \textbf{May 19 — Zina} raises the question of how to incorporate the JHU/TGI Huang et al.\ LSSS modeling paper (almost published) without duplicating it. Flags as ``outstanding question.''
  \item \textbf{May 21 — William} proposes keeping the UW paper focused on SHAKE-mapped interventions; leave LSSS to Matti's team and revisit later.
  \item \textbf{May 21 — Laura} responds that RTSL \textbf{leadership wants LSSS in the model} --- it is needed for the \textbf{Bloomberg renewal application}. Notes that LSSS is now included in the latest SHAKE package (released week of May 12).
  \item \textbf{May 21 — David Watkins} accepts; estimates \textbf{+1--2 months} to incorporate LSSS data inputs.
  \item \textbf{May 21 — Matti Marklund} confirms his paper is at \emph{BMJ} and is methodologically different (comparative-risk-assessment vs.\ UW's state-transition framework). Offers to share inputs.
\end{enumerate}

## Coordination With JHU/TGI (Matti Marklund)

\justifying
\footnotesize

Matti (JHU / The George Institute) is offering four inputs for our model:

\medskip
\begin{enumerate}
  \item \textbf{GDP-per-capita regression} for predicting \% discretionary salt by country --- from the Bhat/Marklund et al.\ systematic review of sources of sodium \autocite{marklund2022sources}.
  \item \textbf{Floor/ceiling values} to prevent any country from having 0\% or 100\% discretionary salt.
  \item \textbf{BP effects of potassium and sodium} from LSSS trials.
  \item \textbf{Data behind the global LSSS estimate} (Huang et al.\ preprint \autocite{huang2026}).
\end{enumerate}

\medskip
\textbf{Note}: Matti will present details at the \textbf{Thursday UW modeling call (4 pm ET)}.

\medskip
His paper uses a CRA framework (not state-transition), so methodological overlap with UW is limited. Co-authorship/coordination to be confirmed so we do not scoop the \emph{BMJ} submission.

## Sources of Sodium --- Current State

\justifying
\footnotesize

Zina and Nicole have been compiling literature on sodium shares across three categories:

\medskip
\begin{itemize}
  \item \textbf{Packaged foods}: Processed and ultra-processed sources, including salt-based condiments (fish sauce, soy sauce) where separately reported.
  \item \textbf{Discretionary sources}: Salt or MSG added while cooking at home or at the table.
  \item \textbf{Outside-the-home}: Restaurants, cafeterias/canteens, street food. Approximately $\frac{1}{3}$ of outside-home assumed to be institutional (public food procurement).
\end{itemize}

\medskip
This compilation feeds directly into the \textbf{Sodium Lives Saved Calculator} (next section) and determines how intervention effects are attributed to the correct sodium source.

\medskip
\textbf{Key method}: Where country-specific data are unavailable, discretionary share is predicted from GDP per capita using the Bhat/Marklund regression \autocite{marklund2022sources}. Processed share is predicted from discretionary share. Floors: 10\% discretionary (USA), 2\% packaged (Ethiopia).

<!-- ======================================================================
     4. Sodium Lives Saved Calculator
     ====================================================================== -->

# Sodium Lives Saved Calculator

## Calculator Overview

\justifying
\scriptsize

\textbf{Purpose}: Estimate national-scale lives saved from a portfolio of sodium-reducing interventions.

\textbf{Six intervention areas}: (A) Public food procurement, (B) Behavior change / mass media, (C) Low-sodium salt substitutes, (D) Front-of-pack labeling, (E) Mandatory salt targets, (F) Fiscal policy / taxation.

\textbf{Source disaggregation}: Sodium intake split into discretionary, packaged, and outside-the-home (with $\sim\!\frac{1}{3}$ of outside-home = institutional). Each intervention acts on its target source.

\textbf{Multiplicative combination}: Effects within a source are combined multiplicatively, not additively, to prevent ${>}$100\% reductions. E.g., tax (13.7\%) + FOPL (15\%) on packaged $\Rightarrow$ $1-(1-0.137)(1-0.15)=26.6\%$.

\textbf{Discretionary share}: Predicted from GDP per capita \autocite{marklund2022sources}; floors at 10\% (USA) and 2\% packaged (Ethiopia).

\textbf{Packaged-food growth}: Projected using Vandevijvere et al.\ 2019 regional growth rates \autocite{vandevijvere2019}, capped at Singapore (4.3 g Na/day from packaged).

\textbf{Demo}: \url{https://dcp-uw.shinyapps.io/salt_demo/} \quad \textbf{Code}: \url{https://github.com/Disease-Control-Priorities/sodium}

## Calculator Overview: Summary of Effect Sizes (Table 1)

\scriptsize

\begin{tabular}{p{2.8cm} p{2.2cm} p{2.0cm} p{1.6cm} p{1.5cm} p{3.0cm}}
\toprule
\textbf{Intervention} & \textbf{Source targeted} & \textbf{Study} & \textbf{$\Delta$ mg/day} & \textbf{$\Delta$(\%)} & \textbf{Time horizon} \\
\midrule
A. Public food procurement & Institutional & Gase 2011 & $-233$ & 7\% & Modeled \\
B. Behavior change / media & Discret.\ + Packaged & Do 2016 & --- & 5\% & 1 year \\
C. LSSS & Discretionary & Huang 2026 & --- & TBD & TBD \\
D. Front-of-pack labeling & Packaged & Taillie 2024 & $-73.3$ & 13.8\% & 3 years \\
E. Mandatory salt targets & Packaged & Strauss-Kruger 2022 & $-480$ & 15.2\% & 4.56 years \\
F. Fiscal policy / taxation & Packaged & Saxena 2025 & $-100$ & 10\% & 20 years \\
\bottomrule
\end{tabular}

## A. Public Food Procurement

\footnotesize

\textbf{Source targeted}: Institutional share of outside-home consumption.

\textbf{Selected study}: \textcite{gase2011} --- modeling/simulation study, LA County.

\textbf{Policy simulated}: Two approaches --- (1) mandatory reformulation (reduce sodium content of all foods served); (2) choice architecture + incentives (labeling, promotion, subsidies, lower-Na alternatives).

\textbf{Effect size}: Weighted average across adult settings = \textbf{--233 mg/day ($\approx$7\%)} of 3,329 mg baseline. Applies only to settings where individuals eat $\geq$1 meal/day, $\geq$5 days/week, $\geq$8 weeks.

\textbf{Population affected}: Only \textbf{0.176\%} of LA adult population (13,183 / 7,501,453).

\textbf{Source-of-sodium adjustment}: 233 mg = 20.6\% reduction in sodium content of public meals. For a country where $x\%$ of sodium comes from institutional meals, population-level impact = $0.206 \times x\%$.

\textbf{Caveats}: Modeled, not observed; US-specific setting; very small affected population share.

## B. Behavior Change / Mass Media

\footnotesize

\textbf{Source targeted}: Discretionary + packaged.

\textbf{Candidate study 1}: \textcite{do2016} --- COMBI Vietnam. Multi-component intervention (mass media + school + community + high-risk targeting). \textbf{5\% decrease} in mean sodium excretion after 1-year intervention.

\textbf{Candidate study 2}: \textcite{collins2014} --- UK economic evaluation. Advertising campaign promoting healthy eating. \textbf{2\% [1--5\%]} effect size used in their model. Provenance of these numbers could not be verified in the source references.

\medskip

\begin{alertblock}{Open question}
Laura's email notes that RTSL funds \textbf{policy change, not mass media for behavior change}. Should this intervention be included in the model? SHAKE framing argues for inclusion; reinvestment case argues for exclusion.
\end{alertblock}

## C. Low-Sodium Salt Substitutes (LSSS)

\footnotesize

\textbf{Source targeted}: Discretionary salt.

\textbf{Study}: \textcite{huang2026} --- preprint (Marklund/TGI/Polly). Five scenarios:

\scriptsize
\begin{enumerate}
  \item Worldwide replacement of \textbf{all salt} (discretionary + non-discretionary)$^{a,b}$
  \item Worldwide replacement of \textbf{discretionary salt only}$^{a,b}$
  \item Worldwide replacement of non-discretionary salt only$^{a,b}$
  \item Replacement of discretionary salt for people with \textbf{diagnosed} hypertension$^{a,c}$
  \item Replacement of discretionary salt for people with \textbf{treated} hypertension$^{a,c}$
\end{enumerate}
\footnotesize
$^a$ Modeled BP-lowering benefits. $^b$ Includes hyperkalemia harms in CKD G3--G5. $^c$ Assumes clinical screening prevents use in 20\% of CKD G3--G5.

\medskip
\textbf{Working decision}: Focus on scenarios \textbf{2, 4, 5} (discretionary-only). Pending confirmation.

## C. LSSS: Deaths Averted by Scenario

\scriptsize

\begin{tabular}{l r r r}
\toprule
\textbf{Country} & \textbf{Scenario 2} & \textbf{Scenario 4} & \textbf{Scenario 5} \\
 & (Discret., all) & (Discret., dx HTN) & (Discret., tx HTN) \\
\midrule
Global      & 1,860,000 & 600,000 & 470,000 \\
China       & 689,246   & 205,936 & 169,409 \\
India       & 277,196   & 54,096  & 46,445  \\
Nigeria     & 13,214    & 3,257   & 2,047   \\
Ethiopia    & 8,232     & 1,406   & 811     \\
Bangladesh  & 36,044    & 7,510   & 5,931   \\
Malaysia    & 8,798     & 3,200   & 2,930   \\
Vietnam     & 36,831    & 10,672  & 7,530   \\
Thailand    & 17,714    & 5,710   & 5,086   \\
Philippines & 32,899    & 9,106   & 6,533   \\
\bottomrule
\end{tabular}

\medskip
\footnotesize
Source: \textcite{huang2026}. Scenarios 2, 4, 5 model discretionary-salt replacement only; scenarios 4 and 5 restrict to diagnosed/treated hypertensives respectively.

## D. Front-of-Pack Labeling

\footnotesize

\textbf{Source targeted}: Packaged (no adjustment needed --- estimate already from packaged-food purchases).

\textbf{Selected study}: \textcite{taillie2024} --- Chile FOPWL, observed.

\textbf{Policy}: Mandatory black octagonal warning labels for Na, sugar, saturated fat, and calories. Includes school sales bans and marketing restrictions on labeled products.

\textbf{Population}: 2,844 urban Chilean households, 138,391 household-months, 13 of 16 regions. July 2013 -- June 2019.

\textbf{Effect size}: \textbf{--73.3 mg/capita/day (--13.8\%)} in sodium purchases.

\textbf{Caveat}: Purchases-to-intake translation is uncertain. Nadia's US/Canada work used purchases directly.

\textbf{Alternative}: Mexico 2026 study found --7\% (95\% CI --13\%, --2\%) in packaged-food/beverage \emph{sales} $\approx$ --38 mg/capita/day.

## E. Mandatory Salt Targets

\footnotesize

\textbf{Source targeted}: Packaged foods.

\textbf{Policy}: South African mandatory national sodium reduction legislation (2013); maximum sodium limits across 13 categories of processed foods; targets implemented 2016 and 2019.

\textbf{Selected study}: \textcite{strausskruger2022} --- African-PREDICT. $N=311$ healthy 20--30-year-olds, 24-hr urine. Baseline 2013--2017 vs.\ follow-up 2018--2020; mean follow-up 4.56 yrs.

\textbf{Effect size}: \textbf{--15.2\% sodium/day} (derived). Baseline/follow-up straddle target dates $\to$ likely \textbf{underestimates} true effect.

\textbf{Source adjustment}: ${\sim}$62\% of SA sodium from packaged food \autocite{charlton2005}. Raw adjustment yields 24.5\% reduction relative to packaged source; conservative estimate = 20\%.

\textbf{Alternative SA studies}: Gaziano 2025 (--8.4\%, rural, ${\geq}$40 yrs, spot urine) \autocite{gaziano2025}; Charlton 2021 (--18.4\%, national, methodological concerns) \autocite{charlton2021}.

\textbf{UK targets}: He 2013: --15\% over 7 yrs, voluntary \autocite{he2013}; Song 2023: --19.2\% over 11 yrs \autocite{song2023}.

\textbf{Mexico modeling}: Muci\~{n}o-Sandoval 2025: --25.2\% phase 1, --28.6\% phase 2 \autocite{mucinosandoval2025} --- justifies not using Gaziano's 8\%.

## F. Fiscal Policy / Taxation

\footnotesize

\textbf{Source targeted}: Packaged (no adjustment needed).

\textbf{Selected study}: \textcite{saxena2025} --- Philippines, modeled.

\textbf{Policy}: 20\% tax on foods high in salt and sugar; WHO WPRO nutrient-profile model; 100\% pass-through; industry reformulation (${\leq}$50\% above threshold).

\textbf{Population}: Adults of the Philippines, aged ${\geq}$25 years.

\textbf{Effect size}: \textbf{--100 mg Na/day (--10\%)} at year 20.

\textbf{Caveat}: High relative to other estimates; tax does not include fat.

\textbf{Alternative studies}:
\begin{itemize}
  \item India \autocite{roche2026}: --1.63\% over 30 yrs (40\% GST, non-standard tax structure).
  \item Australia \autocite{davies2026}: --110 mg/day ($\approx$ --3.22\%) over lifetime of current population.
  \item Australia \autocite{cobiac2017}: --67 mg/day under \$0.30/g Na excise.
  \item \textcite{dodd2020} systematic review: modeling studies predict 1--11\% decrease in salt intake.
\end{itemize}

## Sources of Sodium by Country

\scriptsize

\begin{tabular}{p{2.5cm} p{4.5cm} r r r}
\toprule
\textbf{Country} & \textbf{Source} & \textbf{\% Discret.} & \textbf{\% Packaged} & \textbf{\% Outside} \\
\midrule
South Africa & Charlton et al.\ 2005 & 33--46\% & 54--68\% & --- \\
Vietnam & Ha (PhD thesis, 2014) & 13.5\% & 85.8\% & --- \\
Philippines & Amara et al.\ 2021 & 7.8--8.5\% & 90.4--93.3\% & --- \\
Argentina & Elorriaga et al.\ 2017 & --- & --- & --- \\
\midrule
\multicolumn{5}{l}{\emph{Pending: Panama, Paraguay, Peru, Thailand}} \\
\bottomrule
\end{tabular}

\medskip
\footnotesize
Sources: \autocite{charlton2005,ha2014,amara2021,elorriaga2017}. Where country-specific data are unavailable, discretionary share is predicted from GDP per capita using the Bhat/Marklund regression \autocite{marklund2022sources}; packaged share predicted from discretionary share.

## Projected Growth of Packaged-Food Sales

\footnotesize

\textbf{Source}: \textcite{vandevijvere2019} --- change in ultra-processed food sales volume, 2002--2016.

\medskip
\begin{tabular}{l r}
\toprule
\textbf{Region} & \textbf{Change in UPF sales volume} \\
\midrule
South and Southeast Asia & +67.3\% \\
North Africa and the Middle East & +57.6\% \\
Central and East Asia & +10.0\% \\
Western Europe & $-3.8$\% \\
North America and Australasia & $-2.2$\% \\
\bottomrule
\end{tabular}

\medskip
Used to project rising packaged-food share over the modeling horizon.

\textbf{Cap}: Singapore's 4.3 g Na/day from packaged foods serves as the ceiling.

\textbf{Implication}: Countries in South/SE Asia will see a growing share of sodium from packaged sources, increasing the impact of reformulation, labeling, and taxation interventions over time.

<!-- ======================================================================
     5. Scenarios (REVISED)
     ====================================================================== -->

# Scenarios

## SHAKE 2.0 / Revamped SHAKE Package

\justifying
\footnotesize

The WHO SHAKE package (2026 revision) now incorporates \textbf{low-sodium salt substitutes} alongside the original five pillars:

\medskip
\begin{enumerate}
  \item \textbf{S}urveillance --- population sodium monitoring
  \item \textbf{H}arness Industry --- food reformulation (mandatory/voluntary targets)
  \item \textbf{A}dopt Standards --- front-of-pack labeling, marketing restrictions
  \item \textbf{K}nowledge --- consumer information and education campaigns
  \item \textbf{E}nvironments --- healthier food settings (institutional procurement)
  \item[\textbf{+}] \textbf{Low-Sodium Salt Substitutes (LSSS)} --- newly added to the SHAKE package, reflecting the WHO guideline on LSSS and growing trial evidence
\end{enumerate}

\medskip
The largest and most sustained sodium reductions occur when SHAKE components are implemented as a \textbf{package}. Our modeling evaluates individual and combined policy bundles mapped to these components, now including LSSS as a sixth intervention lever.

## Sodium and Potassium: Complementary Strategies for BP Control

\footnotesize
- Sodium reduction policies are intrinsically linked to increasing potassium intake, as both nutrients play complementary roles in lowering BP and reducing CVD risk.

- While excess sodium raises BP, potassium promotes vasodilation and enhances urinary sodium excretion.

**Key Aspects of the Relationship**

- **Synergistic impact**: Lower sodium combined with higher potassium intake produces larger reductions in BP and stroke risk than sodium reduction alone.
- **Public health strategy**: Many national policies incorporate potassium-enriched salt substitutes (e.g., 75% NaCl / 25% KCl) to simultaneously reduce sodium and address widespread potassium deficiencies.
- **Physiological balancing**: Potassium counteracts sodium's adverse effects, lowering HTN risk.
- **Targeted goals**: guidelines emphasize a dual approach: reducing sodium intake and increasing daily potassium consumption.

**LSSS is the operational instrument that delivers both sodium reduction and potassium increase simultaneously.**

## Scenario Definitions and Default Efficacies

\footnotesize

Six scenarios; $q_{\text{pkgd}} = q_{\text{outside}} = 30\%$; $u_p = \$0.05$ per component; scale-up 2026--2030.

\medskip
\begin{tabular}{lcc}
\toprule
\textbf{Scenario} & \textbf{Active policies} & \textbf{$\eta$ (salteff)} \\
\midrule
Baseline                  & none             & 0 \\
Reformulation -- mandatory & reform only      & $0.20 \times 0.30 = 0.060$ \\
Reformulation -- voluntary & reform only      & $0.15 \times 0.30 = 0.045$ \\
Full package -- mandatory  & reform + label + media + support & $0.060+0.030+0.100+0.060 = 0.250$ \\
Full package -- voluntary  & reform + label + media + support & $0.045+0.030+0.100+0.060 = 0.235$ \\
LSSS (discretionary)       & LSSS (scenarios 2/4/5)  & TBD -- pending Marklund inputs \\
\bottomrule
\end{tabular}

\medskip
Per-capita annual costs: \$0.05 (reformulation-only) to \$0.20 (full package). LSSS costing TBD.

\medskip
\scriptsize \textbf{Note}: LSSS efficacy values to be calibrated from Huang et al.\ discretionary-only scenarios (2, 4, 5) once Matti shares BP-effect inputs at Thursday's call.


# Mathematical Model

## Model Architecture: Three Health States

```{r state-diagram, fig.height=2.7}
nodes <- data.table(
  x=c(1,3.2,5.4), y=c(1,1,1),
  label=c("Well\n(W)","Sick\n(K)","Dead\n(D)"),
  fill=c("#d4edda","#fff3cd","#f8d7da")
)
arr_main <- data.table(
  x1=c(1.40,3.60), y1=c(1,1), x2=c(2.80,5.00), y2=c(1,1),
  lbl=c("IR × ε_IR","CF × ε_CF")
)
arr_bg <- data.table(
  x1=c(1.0,3.2,3.0), y1=c(1.25,1.25,1.0),
  x2=c(1.0,3.2,5.0), y2=c(0.62,0.62,0.62),
  lbl=c("BG.mx","BG.mx","BG.mx.all")
)
ggplot() +
  geom_tile(data=nodes, aes(x=x,y=y,fill=fill),
            width=0.75,height=0.42,color="grey30",linewidth=0.8) +
  scale_fill_identity() +
  geom_text(data=nodes, aes(x=x,y=y,label=label), size=4, fontface="bold") +
  geom_segment(data=arr_main,
    aes(x=x1,y=y1,xend=x2,yend=y2),
    arrow=arrow(length=unit(0.22,"cm"),type="closed"),
    linewidth=1.2, color="#2171b5") +
  geom_label(data=arr_main,
    aes(x=(x1+x2)/2, y=y1+0.22, label=lbl),
    size=3.2, fill="white", label.padding=unit(0.12,"cm"), color="#2171b5") +
  geom_segment(data=arr_bg,
    aes(x=x1,y=y1,xend=x2,yend=y2),
    arrow=arrow(length=unit(0.16,"cm"),type="closed"),
    linewidth=0.7, color="grey55", linetype="dashed") +
  geom_label(data=arr_bg,
    aes(x=(x1+x2)/2+0.2, y=(y1+y2)/2, label=lbl),
    size=2.6, fill="white", color="grey50", label.padding=unit(0.08,"cm")) +
  scale_x_continuous(limits=c(0.5,6.0)) +
  scale_y_continuous(limits=c(0.28,1.58)) +
  theme_void(base_size=11) +
  labs(title="Multi-State Transition Model  |  Annual time step  |  Ages 20-95  |  4 CVD causes")
```

\small \textbf{Key:} $\varepsilon_{IR}$, $\varepsilon_{CF}$ = multiplicative effect ratios from the intervention module $\cdot$ BG.mx = background mortality $\cdot$ COVID-19 excess mortality applied 2020--2021 only

## Policy Package: Efficacy

\footnotesize

Five policy components; active set $\mathcal{P}$ selected per scenario.

\medskip
\begin{tabular}{lll}
\toprule
\textbf{Component} & \textbf{Efficacy $e_p$} & \textbf{Note} \\
\midrule
Reformulation      & $r \times q_{\text{pkgd}}$          & $r=0.20$ (mandatory), $0.15$ (voluntary) \\
Package labelling  & $0.10 \times q_{\text{pkgd}}$       & \\
Media campaigns    & $0.10$                              & whole population \\
Supportive env.    & $0.20 \times q_{\text{outside}}$    & \\
Other              & $q_{\text{other}}$                  & optional \\
\bottomrule
\end{tabular}

\medskip
Total package efficacy (passed as \texttt{salteff}):
$$
\eta = \sum_{p \in \mathcal{P}} e_p
$$

## Policy Package: Programme Costs

\footnotesize

Per-capita annual cost: $u = \sum_{p \in \mathcal{P}} u_p$.

Scale-up from implementation start $t_0$ to full coverage $t_1$:

$$
Cost_{\ell,t} =
\begin{cases}
0 & t < t_0 \\[4pt]
\dfrac{Pop_{\ell,t}}{4} \times u \times Ex \times \dfrac{t - t_0 + 1}{t_1 - t_0 + 1} & t_0 \le t < t_1 \\[8pt]
\dfrac{Pop_{\ell,t}}{4} \times u \times Ex & t \ge t_1
\end{cases}
$$

$Ex$ = optional exchange-rate multiplier. Population divided by 4 = quarterly scaling convention.

## Sodium Intake Reduction

\footnotesize

Three reduction methods; all subject to a 2 g/day minimum intake floor.

\medskip
\textbf{Percent reduction} (main scenario runner):
$$\Delta S^{*}_{a,s,\ell,t} = S_{a,s,\ell,t} \times \eta$$

\textbf{Target-based} (absolute gram target):
$$\Delta S^{*}_{a,s,\ell,t} = \min(S_{a,s,\ell,t},\; \eta)$$

\textbf{Fixed absolute} (app-based):
$$\Delta S^{*}_{a,s,\ell,t} = \max(0,\; S_{a,s,\ell,t} - \eta)$$

\textbf{Floor adjustment}:
$$
\Delta S_{a,s,\ell,t} =
\begin{cases}
S_{a,s,\ell,t} - 2 & \text{if } S - \Delta S^{*} < 2 \\
\Delta S^{*}_{a,s,\ell,t} & \text{otherwise}
\end{cases}
$$


# Economic Evaluation

## Costing the SHAKE Policy Package

\justifying

- Costs are estimated using an ingredients-based regulatory costing approach.
- Government costs include policy design, administration, monitoring, and enforcement.
- Industry costs include reformulation, testing, and labeling compliance, counted as real resource costs.
- Both start-up and ongoing recurrent costs are explicitly modeled over time.

## Costing the SHAKE Policy Package (Example)

\scriptsize

\begin{tabular}{p{2.2cm} p{3.1cm} p{1.7cm} p{1.6cm} p{1.8cm} p{2.3cm}}
\toprule
\textbf{SHAKE Component} & \textbf{Activity} & \textbf{Cost Category} & \textbf{Frequency} & \textbf{Coverage} & \textbf{Cost Bearer} \\
\midrule
S -- Surveillance & Population sodium survey & Program & Every 3--5 yrs & National & Government \\
S -- Surveillance & Food sodium monitoring & Program & Annual & Processed foods & Government \\
H -- Harness Industry & Set reformulation targets & Program & One-off & Key food categories & Government \\
H -- Harness Industry & Product reformulation & Compliance & One-off + transition & Affected SKUs & Industry \\
A -- Labeling & Front-of-pack labeling policy & Program & One-off & National & Government \\
A -- Labeling & Label changes & Compliance & One-off & Affected SKUs & Industry \\
K -- Knowledge & Mass media campaign & Program & Annual & Population reached & Government \\
E -- Environment & School meal standards & Program & Annual & Schools & Government \\
E -- Environment & Institutional food reformulation & Compliance & One-off + annual & Institutions & Vendors \\
\bottomrule
\end{tabular}

<!-- ======================================================================
     8. Team Discussion
     ====================================================================== -->

# Team Discussion

## Pending Decisions

\scriptsize

\begin{enumerate}
  \item \textbf{LSSS scenarios} --- Confirm we model scenarios 2, 4, 5 (discretionary-only: all adults / diagnosed HTN / treated HTN). Or do we also try scenario 1 (all salt)?
  \item \textbf{Mass media} --- In or out? Laura: RTSL funds policy change, not media for behavior change. SHAKE framing argues for inclusion in the paper; reinvestment case argues for exclusion.
  \item \textbf{Heterogeneity of observation periods} --- Effect sizes measured over very different time windows (1 yr $\to$ 20 yrs). How do we harmonize? Options: (a) assume effects are end-state, apply at full scale-up; (b) apply linearly over observed window; (c) use a country-tier heuristic (low/middle/high).
  \item \textbf{Source-of-sodium estimates} --- How hard do we push for country-specific shares vs.\ fall back on Marklund's GDP regression with floors/ceilings? Flag condiment-heavy countries where the packaged-food transition is escalating.
  \item \textbf{Integration with Marklund/TGI} --- Confirm we use his discretionary-vs-GDP regression and BP-effect inputs; clarify co-authorship/coordination so we do not scoop the \emph{BMJ} paper.
  \item \textbf{Timeline} --- Original end-of-June paper submission is now +1--2 months for LSSS integration. New target?
  \item \textbf{Bloomberg renewal deliverable} --- What does the renewal application need from us, and by when, that is distinct from the paper?
\end{enumerate}

## Agenda for Thursday's Call

\footnotesize

\begin{enumerate}
  \item \textbf{Matti presents} his discretionary-vs-packaged disaggregation approach ($\sim$15 min) --- GDP regression, floor/ceiling values, BP effects of K and Na.
  \item \textbf{Walk through the 6 interventions} in the Sodium Lives Saved Calculator --- confirm effect sizes and source assignments.
  \item \textbf{Decide pending items 1--4}: LSSS scenarios, mass media inclusion, observation-period harmonization, country-specific vs.\ regression-based sodium shares.
  \item \textbf{Lock revised timeline} for paper + report + Bloomberg renewal deliverable.
\end{enumerate}

<!-- ======================================================================
     APPENDIX
     ====================================================================== -->

# Appendix

## Notation: Core Model

\footnotesize

Indices: age $a$, sex $s$, cause $c$, location $\ell$, year $t$, BP category $b$.

\medskip
\begin{tabular}{ll}
\toprule
\textbf{Symbol} & \textbf{Meaning} \\
\midrule
$IR_{a,s,c,\ell,t}$ & Baseline cause-specific incidence rate \\
$CF_{a,s,c,\ell,t}$ & Baseline case fatality rate \\
$BG.mx_{a,s,c,\ell,t}$ & Background mortality (disease state) \\
$BG.mx.all_{a,s,\ell,t}$ & All-cause background mortality \\
$PREVt0 / DIS.mx.t0$ & Baseline prevalence / disease mortality \\
$N_{a,s,\ell,t}$ & Population size \\
$P(b \mid a,s,\ell,t)$ & Probability of BP category $b$ \\
$S_{a,s,\ell,t}$ & Baseline mean sodium intake (g/day) \\
$RBP_{a,s,\ell,t}$ & Share of population with raised BP \\
$D_{a,s,b,\ell,t}$ & Diabetes prevalence \\
\bottomrule
\end{tabular}

## Scale-Up Time Path

\footnotesize

Linear interpolation between implementation start ($T_{\text{start}}$) and full implementation ($T_{\text{full}}$):

$$
\Delta S^{\text{scale}}_{a,s,\ell,t} =
\begin{cases}
0 & t < T_{\text{start}} \\[6pt]
\Delta S_{a,s,\ell,t} \times \dfrac{t - T_{\text{start}} + 1}{T_{\text{full}} - T_{\text{start}} + 1} & T_{\text{start}} \le t \le T_{\text{full}} \\[8pt]
\Delta S_{a,s,\ell,t} & t > T_{\text{full}}
\end{cases}
$$

\medskip
Default scenario parameters: $T_{\text{start}} = 2026$, $T_{\text{full}} = 2030$.

## Sodium $\to$ Systolic Blood Pressure Reduction

\footnotesize

Filippini-style dose--response \autocite{Filippini2021}, differentiating raised- and normal-BP subgroups:

$$
\Delta SBP_{a,s,\ell,t}
= \Bigl(2.8 \times RBP_{a,s,\ell,t} \;+\; 1.0 \times (1-RBP_{a,s,\ell,t})\Bigr)
  \times \Delta S^{\text{scale}}_{a,s,\ell,t}
$$

\medskip
\begin{tabular}{ll}
\toprule
\textbf{Subgroup} & \textbf{SBP reduction per g/day Na removed} \\
\midrule
Raised BP ($RBP=1$) & 2.8 mmHg \\
Normal BP ($RBP=0$) & 1.0 mmHg \\
\bottomrule
\end{tabular}

## Baseline Incidence by BP Category

\footnotesize

\textbf{Step 1 -- GBD relative risk by bin.}
Let $m_b$ be the midpoint SBP of bin $b$ (reference: $<$120 mmHg $\to$ midpoint 110):
$$
RR^{GBD}_{b,a,c} = \bigl(RR^{10}_{a,c}\bigr)^{(m_b - 120)/10}, \quad m_b > 120; \quad 1 \text{ otherwise}
$$

\textbf{Step 2 -- Normalisation factor} (ensures consistency with aggregate $IR$):
$$
\alpha_{a,s,c,\ell,t} = \sum_b P(b \mid a,s,\ell,t) \times RR^{GBD}_{b,a,c}
$$

\textbf{Step 3 -- Bin-specific baseline incidence:}
$$
IR^{bin}_{b,a,s,c,\ell,t} = \frac{RR^{GBD}_{b,a,c} \times IR_{a,s,c,\ell,t}}{\alpha_{a,s,c,\ell,t}}
$$

## Ettehad Effect Sizes

\footnotesize

\textcite{ettehad2016} provides per-10-mmHg RR reductions by BP category and cause from randomised trial meta-analysis.

\medskip
\textbf{Diabetes-weighted cumulative effect} for bin $b$, cause $c$:

$$
E_{b,c,a,s,\ell,t}
= (1 - D_{a,s,b,\ell,t})\,E^{ND}_{b,c}
  \;+\; D_{a,s,b,\ell,t}\,E^{D}_{b,c}
$$

where $E^{ND}_{b,c}$ and $E^{D}_{b,c}$ are the no-diabetes and diabetes-specific cumulative effect sizes tabulated from Ettehad.

\medskip
$\Rightarrow$ implemented in \texttt{calculate\_etihad\_cumulative\_rr()}.

## Effect of Sodium Reduction on Incidence

\footnotesize

\textbf{Step 1 -- Sodium-induced effect per bin.}
Scale the per-10-mmHg Ettehad effect proportionally to achieved SBP reduction:

$$
E^{\text{sodium}}_{b,a,s,c,\ell,t}
= \bigl(1 - RR^{Ettehad}_{b,c}\bigr) \times 0.1 \times \Delta SBP_{a,s,\ell,t}
$$

\textbf{Step 2 -- New bin-specific incidence:}
$$
IR^{bin,new}_{b,a,s,c,\ell,t} = IR^{bin}_{b,a,s,c,\ell,t} \times \bigl(1 - E^{\text{sodium}}_{b,a,s,c,\ell,t}\bigr)
$$

\textbf{Step 3 -- Population-weighted new incidence:}
$$
IR^{new}_{a,s,c,\ell,t} = \sum_b IR^{bin,new}_{b,a,s,c,\ell,t} \times P(b \mid a,s,\ell,t)
$$

Incidence effect ratio carried into the life table:
$\;\varepsilon^{IR}_{a,s,c,\ell,t} = IR^{new} / IR$

## Case Fatality Assumption

\footnotesize

The sodium policy model does \textbf{not} impose a secondary effect on case fatality (unlike antihypertensive treatment models).

$$
CF^{new}_{a,s,c,\ell,t} = CF_{a,s,c,\ell,t}
\qquad \Rightarrow \qquad
\varepsilon^{CF}_{a,s,c,\ell,t} = 1
$$

\medskip
\textbf{Rationale}: Sodium reduction acts entirely through lower incidence of new events. Trial evidence for a direct sodium-pathway effect on survival of existing disease is insufficient to parameterize a case-fatality reduction.

## State-Transition Equations

\footnotesize

Annual transitions (age $a \to a+1$, year $t \to t+1$):

$$
NewCases_{a+1,t+1} = Well_{a,t} \times IR_{a,t}
$$

$$
Sick_{a+1,t+1} = Sick_{a,t}\bigl(1 - CF_{a,t} - BG.mx_{a,t} - covid.mx_{a,t}\bigr) + Well_{a,t} \times IR_{a,t}
$$

$$
Dead_{a+1,t+1} = Sick_{a,t} \times CF_{a,t}
$$

$$
Pop_{a+1,t+1} = Pop_{a,t} - all.mx_{a,t}
$$

$$
Well_{a+1,t+1} = Pop_{a+1,t+1} - all.mx_{a+1,t+1} - Sick_{a+1,t+1}
$$

\medskip
\textbf{Initial conditions} ($t=t_0$ or $a=20$):
$\;Sick_0 = N \cdot PREVt0;\; Dead_0 = N \cdot DIS.mx.t0;\; Well_0 = N(1-PREVt0-BG.mx)$

## References {.allowframebreaks}

\tiny
\printbibliography[heading=none]
